### 2014-10-24 abstract.tex
Initial commit.

Declarative programming is good for parallel programming but makes it hard for
the programmer to fine tune scheduling and data layout of the program.
some programs require more information in order to execute better.
with our approach we permit the programmer to write declarative programs
but then optimize them later using sensing facts and action facts that
understand and change how the programs are executed.

Linear Meld is a concurrent forward-chaining linear logic programming
language where logical facts can be asserted and retracted in a
structured way. In Linear Meld, a program is seen as a database of
logical facts and a set of derivation rules. The database of facts is
partitioned by the nodes of a graph structure which leads to

### 2014-10-24 coordination.tex
Initial commit.

The original Meld was
implemented as an ensemble programming language, targeting modular robotic systems such as
Claytronics~\cite{ashley-rollman-derosa-iros07wksp}. In such systems, there is a natural matching
between computation and processing units, since each robot is represented by a node. This distribution
of data leaves little choice to be made in the division of computation to the various nodes.

On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
since nodes are a program abstraction and part of the program's logic.
We view the set of nodes as a graph data structure where workers will perform work.
A worker is able to process any node, although a node cannot be computed by more than one worker
at the same time. This disallows the manipulation of a node by multiple workers.

Since LM uses linear logic, the order in which nodes are scheduled can impact the
performance and even the results of the program. When only using
classical logic, no matter how computation is done, the end result will be the same since the program
is strictly monotonic.
LM introduces the concept of \emph{coordination} that allows the programmer
to write code that changes how the runtime system schedules nodes.

Randomized and approximation
algorithms can obtain significant benefits from coordination directives because although the final
program results will not be exact, they follow important statistical properties and can be computed faster.
Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.

\subsection{Scheduling Facts}

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the worker level
so that each worker can make a local decision about which node of the graph to run next.
Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).

The following list presents the action facts available to manipulate the scheduling decisions of the system:

\begin{description}
   \item[\texttt{type linear action set-priority(node, float)}]: This sets the priority of a node. If the node already has some priority, we only change the priority if the new one is higher priority. The programmer can decide if priorities are to be ordered in ascending or descending order.
   \item[\texttt{type linear action add-priority(node, float)}]: This gets the current node's priority and increases or decreases it.
   \item[\texttt{type linear action schedule-next(node)}]: The work will fetch the highest priority node's priority $P$ from its set of nodes and set the action's argument node's priority as $P + 1.0$. If using the priorities
   in ascending order, we pick the lowest priority and subtract $1.0$.
   \item[\texttt{type linear action unset-priority(node)}]: Removes the priority, if any, of a given node.
   \item[\texttt{type linear action stop-program(node)}]: Immediately stops the execution of the whole program.
\end{description}

When the highest priority node is picked up for execution, its priority is reset to 0 (the default priority value). This means that
the programmer must set the node's priority again if he wants to prioritize that node.

We intend to add more action facts in the near future. For example, we want the programmer to be able to place specific nodes in workers. This will permit good use of
memory locality by forcing certain computations to be performed in the same worker.

Sensing facts provide information about node placement and node priority. We can use those facts
to express coordination policies. LM provides the following two
sensing facts:

\begin{description}
   \item[\texttt{type linear cpu-id(node, node, int)}]: The third argument indicates the worker's ID where the node of the second argument is currently running.
   \item[\texttt{type linear priority(node, node, float)}]: The third argument is the current priority of the node in the second argument.
\end{description}

Note that when sensing facts are consumed, they are re-derived automatically, except if the programmer explicitly tells the compiler otherwise.

\subsubsection{Global Directives}

We also provide a few global coordination statements:

\begin{description}
   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$. Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
\end{description}

\subsection{Implementation}

How the virtual machine and coordinaction facts are implemented.

\subsection{Programs}

To better understand how coordination directives work, we next present some programs that
take advantage of them.

\subsubsection{Graph Search}

talk about a-star

\subsubsection{Single Source Shortest Path}

The Single Source Shortest Path (SSSP) is a graph algorithm where we want to compute the
distance of all nodes to a single node. The full code is presented in Fig.~\ref{code:shortest_path_program}.
The graph is represented using the \texttt{edge(A,~B,~W)} predicate (line 1), where
\texttt{A} $\rightarrow$ \texttt{B} is a directed edge from node \texttt{A} to \texttt{B}
with distance \texttt{W}. The second predicate is \texttt{path(A,~D,~F)} (line 2), where
\texttt{D} is the distance of node \texttt{A} to our initial node and \texttt{F}
indicates if this path has been processed.

Every node may have several \texttt{path} facts. From this set of facts, the node will select
the path with the least distance (rules in lines 11-13) and then propagate it using the third rule (lines 15-19).
When a path is used for propagation, we make sure that the path flag is \texttt{notused} (line 15)
because we do not want to propagate the same distance twice. When the rule is fired, the flag
then changes to \texttt{used} (line 19).

We start with the axiom \texttt{path(startnode,~0,~notused)} in line 9 (the distance to the starting
node is \texttt{0}). This will kickstart the computation by propagating the initial distance to the
neighbor nodes using rule 3. We also use 2 coordination directives:

\begin{itemize}
   \item \texttt{priority~@order~asc}: paths are picked in ascending order (line 4).
   \item \texttt{set-priority(B,~D~+~W)}: to change the node's priority upon the computation of
a new distance (line 18).
\end{itemize}

With these directives, we ensure that we pick the node with the smallest distance
first. If we are only using 1 thread, then our algorithm will behave like Dijkstra's shortest
path algorithm~\cite{Dijkstra}. When using more than 1 thread, each thread will pick the smallest
path from a subset of nodes. This is an example where using coordination directives will
improve the speed of execution since we avoid propagating some of the paths.

\begin{figure}[h!]
\small\begin{Verbatim}[numbers=left,commandchars=&\[\]]
type route edge(node, node, int).
type linear path(node, int, int).

&underline[priority @order asc].

relax(@0, 0).
path(A, +00).

relax(A, D1), path(A, D2),
D1 < D2
   -o path(A, D1),
      {B, W | !edge(A, B, W) |
         relax(B, D + W),
         &underline[set-priority(B, D + W))]},

relax(A, D1), path(A, D2),
D1 >= D2
   -o path(A, D2).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program}
\end{figure}
\normalsize

\subsubsection{Heat Transfer}

\subsubsection{MiniMax}

\subsubsection{Splash BP}

Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP is an algorithm that maps very well to the graph based model of LM. In its original form, we need to compute
the belief of all nodes for several iterations and also synchronize after each iteration.
However, it is still possible to apply
some optimizations in order to take advantage of the fact that LBP will converge even when using
an asynchronous approach. So, instead of computing the belief iteratively,
we first keep track of all messages sent/received (and overwrite them when we receive a new one)
and recompute the belief when we want, instead of synchronizing between nodes.
This way, we can prioritize the computation of beliefs when
a node's belief value changes significantly. When that happens, we set the priority of its
neighbors so that they can recompute their beliefs.

The asynchronous approach proves to be a nice improvement over the synchronous version. Still, it
is possible to do even better. Gonzalez et al~\cite{Gonzalez+al:aistats09paraml} developed an optimal
algorithm to compute this algorithm by first building a tree and then updating the beliefs of each node twice, first from the leaves to the root and then from the root to the leaves. The root of this tree
is the node with the highest priority (based on belief) while the other nodes in the tree
must have a non-zero priority. Note that the priorities are updated whenever a neighbor updates
their belief. These splash trees are built iteratively until we reach convergence.

The code in Fig.~\ref{code:sbp} presents the coordination code of the Belief Propagation problem.
Please note that we just appended the code in Fig.~\ref{code:sbp} to a working but
unoptimized version of the algorithm.
In the coordination code we have three sections:
\begin{description}
   \item[Tree building]: Each node has a \texttt{waiting} fact that is used to start the tree building process. When the highest priority is picked we create a token that will navigate through the tree. Note that in the rule located in lines 11-18 we check if the priority of the new node to add to the tree is positive and that both nodes are in the same worker. We want the tree to be kept in the same worker.
   \item[First phase]: In the third rule (lines 8-9), when we reach a certain number of nodes in the tree, we generate \texttt{first-phase} in order to update the beliefs of all nodes in tree starting from the leaves and ending at the root. As we update the nodes, we generate \texttt{update} to update the belief values (line 29).
   \item[Second phase]: In the second phase we go from the root to the leaves and update the beliefs a second time (line 39).
\end{description}

When we have several workers, every worker will generate their own trees by taking into account the highest priority node in their own queues.

\begin{figure}[h!]
\small\begin{Verbatim}[numbers=left,commandchars=*\{\}]
// TREE BUILDING
// continue tree
waiting(A), token(A, All, Next) -o token(A, All, Next).
// start tree
waiting(A), *underline{@priority(A, A, P)}, P > 0.0
   -o token(A, [A], [A]).
// end tree building
token(A, All, Next), length(All) > maxnodes
   -o first-phase(A, All, reverse(All)).
// expand tree
token(A, All, [A | Next])
   -o [collect => L | Side | !edge(A, L, Side),
         0 = count(All, L),
         0 = count(Next, L),
         *underline{priority(A, L, P)}, P > 0.0,
         *underline{cpu-id(A, L, Id1)},
         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
         send-token(A, All, Next ++ L)].

send-token(A, All, [])
   -o first-phase(A, All, reverse(All)).
send-token(A, All, [B | Next])
   -o *underline{schedule-next(B)},
      token(B, All ++ [B], [B | Next]).

// FIRST PHASE
first-phase(A, [A], [A]) -o second-phase(A, [], A).
first-phase(A, [A, B | Next], [A])
   -o update(A), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
first-phase(A, All, [A, B | Next])
   -o update(A), *underline{schedule-next(B)},
      first-phase(B, All, [B | Next]).

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
second-phase(A, [A], Back)
   -o update(A), waiting(Back),
      waiting(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), waiting(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program.}
  \label{code:sbp}
\end{figure}
\normalsize

### 2014-10-24 intro.tex
Initial commit.

talk about declarative programming and how hard it is for the programmer to
control how the program is scheduled or how data is laid out.

we took the work made by \cite{cruz-iclp14} and extended it with action and
sensing facts.

### 2014-10-24 language.tex
Initial commit.

give an overview of the language

we already have the ICLP paper that describes the language, so we can be brief
here. \cite{cruz-iclp14}

### 2014-10-30 application.tex
Add implementation.

--- 2014-10-24 coordination.tex
@@
-The original Meld was
-implemented as an ensemble programming language, targeting modular robotic systems such as
-Claytronics~\cite{ashley-rollman-derosa-iros07wksp}. In such systems, there is a natural matching
-between computation and processing units, since each robot is represented by a node. This distribution
-of data leaves little choice to be made in the division of computation to the various nodes.
-
-On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
-since nodes are a program abstraction and part of the program's logic.
-We view the set of nodes as a graph data structure where workers will perform work.
-A worker is able to process any node, although a node cannot be computed by more than one worker
-at the same time. This disallows the manipulation of a node by multiple workers.
-
-Since LM uses linear logic, the order in which nodes are scheduled can impact the
-performance and even the results of the program. When only using
-classical logic, no matter how computation is done, the end result will be the same since the program
-is strictly monotonic.
-LM introduces the concept of \emph{coordination} that allows the programmer
-to write code that changes how the runtime system schedules nodes.
-
-Randomized and approximation
-algorithms can obtain significant benefits from coordination directives because although the final
-program results will not be exact, they follow important statistical properties and can be computed faster.
-Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
-Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.
-
-\subsection{Scheduling Facts}
-
-We can use action facts to change the order in which nodes are evaluated by adding
-\emph{priorities}. Node priority works at the worker level
-so that each worker can make a local decision about which node of the graph to run next.
-Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).
-
-The following list presents the action facts available to manipulate the scheduling decisions of the system:
-
-\begin{description}
-   \item[\texttt{type linear action set-priority(node, float)}]: This sets the priority of a node.
-If the node already has some priority, we only change the priority if the new one is higher priority.
-The programmer can decide if priorities are to be ordered in ascending or descending order.
-   \item[\texttt{type linear action add-priority(node, float)}]: This gets the current node's priority and increases or decreases it.
-   \item[\texttt{type linear action schedule-next(node)}]: The work will fetch the highest priority node's priority $P$ from its set of nodes and set the action's argument node's priority as $P + 1.0$.
-If using the priorities
-   in ascending order, we pick the lowest priority and subtract $1.0$.
-   \item[\texttt{type linear action unset-priority(node)}]: Removes the priority, if any, of a given node.
-   \item[\texttt{type linear action stop-program(node)}]: Immediately stops the execution of the whole program.
-\end{description}
-
-When the highest priority node is picked up for execution, its priority is reset to 0 (the default priority value). This means that
-the programmer must set the node's priority again if he wants to prioritize that node.
-
-We intend to add more action facts in the near future. For example, we want the programmer to be able to place specific nodes in workers. This will permit good use of
-memory locality by forcing certain computations to be performed in the same worker.
-
-Sensing facts provide information about node placement and node priority. We can use those facts
-to express coordination policies. LM provides the following two
-sensing facts:
-
-\begin{description}
-   \item[\texttt{type linear cpu-id(node, node, int)}]: The third argument indicates the worker's ID where the node of the second argument is currently running.
-   \item[\texttt{type linear priority(node, node, float)}]: The third argument is the current priority of the node in the second argument.
-\end{description}
-
-Note that when sensing facts are consumed, they are re-derived automatically, except if the programmer explicitly tells the compiler otherwise.
-
-\subsubsection{Global Directives}
-
-We also provide a few global coordination statements:
-
-\begin{description}
-   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
-   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$.
-Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
-   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
-\end{description}
-
-\subsection{Implementation}
-
-How the virtual machine and coordinaction facts are implemented.
-
-\subsection{Programs}
-
 To better understand how coordination directives work, we next present some programs that
@@
 \begin{figure}[h!]
-\small\begin{Verbatim}[numbers=left,commandchars=&\[\]]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=&\[\]]
 type route edge(node, node, int).
@@
 \begin{figure}[h!]
-\small\begin{Verbatim}[numbers=left,commandchars=*\{\}]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
 // TREE BUILDING

### 2014-10-30 coordination.tex
Add implementation.

@@
 \end{description}
-
-\subsection{Implementation}
-
-How the virtual machine and coordinaction facts are implemented.
-
-\subsection{Programs}
-
-To better understand how coordination directives work, we next present some programs that
-take advantage of them.
-
-\subsubsection{Graph Search}
-
-talk about a-star
-
-\subsubsection{Single Source Shortest Path}
-
-The Single Source Shortest Path (SSSP) is a graph algorithm where we want to compute the
-distance of all nodes to a single node. The full code is presented in Fig.~\ref{code:shortest_path_program}.
-The graph is represented using the \texttt{edge(A,~B,~W)} predicate (line 1), where
-\texttt{A} $\rightarrow$ \texttt{B} is a directed edge from node \texttt{A} to \texttt{B}
-with distance \texttt{W}. The second predicate is \texttt{path(A,~D,~F)} (line 2), where
-\texttt{D} is the distance of node \texttt{A} to our initial node and \texttt{F}
-indicates if this path has been processed.
-
-Every node may have several \texttt{path} facts. From this set of facts, the node will select
-the path with the least distance (rules in lines 11-13) and then propagate it using the third rule (lines 15-19).
-When a path is used for propagation, we make sure that the path flag is \texttt{notused} (line 15)
-because we do not want to propagate the same distance twice. When the rule is fired, the flag
-then changes to \texttt{used} (line 19).
-
-We start with the axiom \texttt{path(startnode,~0,~notused)} in line 9 (the distance to the starting
-node is \texttt{0}). This will kickstart the computation by propagating the initial distance to the
-neighbor nodes using rule 3. We also use 2 coordination directives:
-
-\begin{itemize}
-   \item \texttt{priority~@order~asc}: paths are picked in ascending order (line 4).
-   \item \texttt{set-priority(B,~D~+~W)}: to change the node's priority upon the computation of
-a new distance (line 18).
-\end{itemize}
-
-With these directives, we ensure that we pick the node with the smallest distance
-first. If we are only using 1 thread, then our algorithm will behave like Dijkstra's shortest
-path algorithm~\cite{Dijkstra}. When using more than 1 thread, each thread will pick the smallest
-path from a subset of nodes. This is an example where using coordination directives will
-improve the speed of execution since we avoid propagating some of the paths.
-
-\begin{figure}[h!]
-\small\begin{Verbatim}[numbers=left,commandchars=&\[\]]
-type route edge(node, node, int).
-type linear path(node, int, int).
-
-&underline[priority @order asc].
-
-relax(@0, 0).
-path(A, +00).
-
-relax(A, D1), path(A, D2),
-D1 < D2
-   -o path(A, D1),
-      {B, W | !edge(A, B, W) |
-         relax(B, D + W),
-         &underline[set-priority(B, D + W))]},
-
-relax(A, D1), path(A, D2),
-D1 >= D2
-   -o path(A, D2).
-\end{Verbatim}
-  \caption{Shortest Path Program.}
-  \label{code:shortest_path_program}
-\end{figure}
-\normalsize
-
-\subsubsection{Heat Transfer}
-
-\subsubsection{MiniMax}
-
-\subsubsection{Splash BP}
-
-Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
-used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
-where nodes exchange messages with their immediate neighbors and apply some computations to the messages
-received.
-
-LBP is an algorithm that maps very well to the graph based model of LM. In its original form, we need to compute
-the belief of all nodes for several iterations and also synchronize after each iteration.
-However, it is still possible to apply
-some optimizations in order to take advantage of the fact that LBP will converge even when using
-an asynchronous approach. So, instead of computing the belief iteratively,
-we first keep track of all messages sent/received (and overwrite them when we receive a new one)
-and recompute the belief when we want, instead of synchronizing between nodes.
-This way, we can prioritize the computation of beliefs when
-a node's belief value changes significantly. When that happens, we set the priority of its
-neighbors so that they can recompute their beliefs.
-
-The asynchronous approach proves to be a nice improvement over the synchronous version. Still, it
-is possible to do even better. Gonzalez et al~\cite{Gonzalez+al:aistats09paraml} developed an optimal
-algorithm to compute this algorithm by first building a tree and then updating the beliefs of each node twice, first from the leaves to the root and then from the root to the leaves.
-The root of this tree
-is the node with the highest priority (based on belief) while the other nodes in the tree
-must have a non-zero priority. Note that the priorities are updated whenever a neighbor updates
-their belief. These splash trees are built iteratively until we reach convergence.
-
-The code in Fig.~\ref{code:sbp} presents the coordination code of the Belief Propagation problem.
-Please note that we just appended the code in Fig.~\ref{code:sbp} to a working but
-unoptimized version of the algorithm.
-In the coordination code we have three sections:
-\begin{description}
-   \item[Tree building]: Each node has a \texttt{waiting} fact that is used to start the tree building process.
-When the highest priority is picked we create a token that will navigate through the tree.
-Note that in the rule located in lines 11-18 we check if the priority of the new node to add to the tree is positive and that both nodes are in the same worker.
-We want the tree to be kept in the same worker.
-   \item[First phase]: In the third rule (lines 8-9), when we reach a certain number of nodes in the tree, we generate \texttt{first-phase} in order to update the beliefs of all nodes in tree starting from the leaves and ending at the root.
-As we update the nodes, we generate \texttt{update} to update the belief values (line 29).
-   \item[Second phase]: In the second phase we go from the root to the leaves and update the beliefs a second time (line 39).
-\end{description}
-
-When we have several workers, every worker will generate their own trees by taking into account the highest priority node in their own queues.
-
-\begin{figure}[h!]
-\small\begin{Verbatim}[numbers=left,commandchars=*\{\}]
-// TREE BUILDING
-// continue tree
-waiting(A), token(A, All, Next) -o token(A, All, Next).
-// start tree
-waiting(A), *underline{@priority(A, A, P)}, P > 0.0
-   -o token(A, [A], [A]).
-// end tree building
-token(A, All, Next), length(All) > maxnodes
-   -o first-phase(A, All, reverse(All)).
-// expand tree
-token(A, All, [A | Next])
-   -o [collect => L | Side | !edge(A, L, Side),
-         0 = count(All, L),
-         0 = count(Next, L),
-         *underline{priority(A, L, P)}, P > 0.0,
-         *underline{cpu-id(A, L, Id1)},
-         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
-         send-token(A, All, Next ++ L)].
-
-send-token(A, All, [])
-   -o first-phase(A, All, reverse(All)).
-send-token(A, All, [B | Next])
-   -o *underline{schedule-next(B)},
-      token(B, All ++ [B], [B | Next]).
-
-// FIRST PHASE
-first-phase(A, [A], [A]) -o second-phase(A, [], A).
-first-phase(A, [A, B | Next], [A])
-   -o update(A), *underline{schedule-next(B)},
-      second-phase(B, [B | Next], A).
-first-phase(A, All, [A, B | Next])
-   -o update(A), *underline{schedule-next(B)},
-      first-phase(B, All, [B | Next]).
-
-// SECOND PHASE
-second-phase(A, [], _)
-   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
-second-phase(A, [A], Back)
-   -o update(A), waiting(Back),
-      waiting(A), *underline{set-priority(A, 0.0)}.
-second-phase(A, [A, B | Next], Back)
-   -o update(A), waiting(Back), *underline{schedule-next(B)},
-      second-phase(B, [B | Next], A).
-\end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program.}
-  \label{code:sbp}
-\end{figure}
-\normalsize

### 2014-10-30 implementation.tex
Add implementation.

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.

\subsection{Compilation}

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.

Optimizations such as join optimizations (to allow rule filtering invalid
combinations) and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

\subsection{Node partitioning}

The compiler builds a representation of the graph by inspecting axioms of the
program. Next, it orders the nodes of the graph using a breadth-first
topological sort. This node ordering is written to the byte-code file and then
used to initially partition nodes across available threads.

\subsubsection{Coordination directives}

Coordination directives are compiled in two different ways, depending whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of returning something. While these are still facts from the point of
view of the language, they are consumed immediatelly in order to apply the
operation, instead of being derived like any other regular fact. This is more
efficient since we do not need to create a new fact.

\subsection{Execution}

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible
for executing a subset of nodes. Nodes are either inactive, with no new facts,
or active, where they will be placed in the corresponding thread's queue.
Threads do useful work by processing the nodes in their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread into its
own queues. If unsuccessful, the thread synchronizes with the other threads to
become inactive.

\subsubsection{Nodes}

A node is represented as a collection of facts (per predicate) and an indexing structure that
keeps track of the available predicates and potential candidate rules. We need
a separate indexing structure per node since rules run locally.

Facts need to be stored efficiently because the virtual machine instructions
perform searches on the database by fixing arguments of a predicate to concrete
values. Each predicate is stored using one of the following data structures:

\begin{itemize}
\item \emph{Tries} are used exclusively to store 
  persistent facts. Tries are trees where facts are indexed by common
  prefix arguments.
\item \emph{Doubly Linked Lists} are used to store 
  linear facts. We use a double linked list because it is a very
   efficient way to add and remove facts.
\item \emph{Hash Tables} are used to improve lookup when 
  linked lists are too long and when we need to do search filtered by
  a fixed argument. The virtual machine decides which arguments are
  best to be indexed and then uses a hash table
  indexed by the appropriate argument.
\end{itemize}

A node goes through different states during execution. The state machine in
Fig.~\ref{fig:node_states} represents the valid state transitions with the
following states:

\begin{description}
   \item[working]: the node is being executed.
   \item[idle]: the node is not active, has no new facts and is not in any
   queue for processing.
   \item[queue]: the node is active with new facts and is waiting in some queue
   to be processed.
   \item[stealing]: the node has just been stolen is about to move to another
   thread.
   \item[coordinating]: the node is moving from one queue to another (i.e.
         changing priority).
\end{description}

\begin{figure}[h!]
   \centering
   \includegraphics[width=0.4\textwidth]{node_states.pdf}
   \caption{The node state machine. During the lifetime of a program, each node
   goes through different states as specified by the state machine.}
   \label{fig:node_states}
\end{figure}

Each node is protected by a main spinlock that allows threads to change node
attributes: incoming facts, owner thread, node state and locality information.
There is also an internal spinlock that is locked whenever the node enters into
the \textbf{working} state. If a thread sends facts to a node placed in
another thread, it attempts to lock the internal lock in order to update the
indexing structures of the node, otherwise it adds the facts to the list of
incoming facts that are later processed by the owner thread.

\subsubsection{Threads}

Threads pop active nodes from their queues and execute new candidate rules. 
Each thread has 2 queues, a doubly linked list known as the \emph{standard
queue} and a min/max heap known as the \emph{priority queue}. The regular queue
contains nodes without priorities and
implements the operations \texttt{push\_tail(node)} (push to the tail),
\texttt{pop\_head()} (remove node from the head),
\texttt{remove(node)} (remove an arbitrary node) and \texttt{remove\_half()}
(removes first half). The priority queue contains nodes
with priorities and is implemented as a binary heap array. It supports the
operations \texttt{push(node,priority)} (add new node to the heap),
           \texttt{pop\_min()} (remove the best node),
\texttt{remove(node)} (remove an arbitrary node), \texttt{pop\_half\_min()}
(remove the best half) and \texttt{move(node,new\_priority)} (update priority).
Operations such as \texttt{remove\_half()} are supported in order to support
node stealing, while operations \texttt{remove(node)} or
\texttt{move(node,new\_priority)} allows threads to change the priority of
nodes.

The \texttt{next} and \texttt{prev} pointers of the regular queue are part of
the node structure in order to save space. These pointers are also used as the
index in the priority queue and current priority, respectively.

Both the regular and priority queue are actually implemented as 2 queues each.
The \emph{static queue} contains nodes that are local to a thread (and thus cannot be stolen) and
the \emph{dynamic queue} maybe accessed by other threads during node stealing.

\subsubsection{Communication}

Threads synchronize with each other using mutual exclusion. We use a spinlock in
each queue to protect queue operations and one spinlock to protect the thread's
state. Given threads $T_1$ and $T_2$, we enumerate the most important
synchronization intensive places in the virtual machine:

\begin{description}
   \item[New facts:] When a node executes on $T_1$ and derives facts
   to a node in $T_2$, $T_1$ first buffers the facts 
   and then sends them to the target node. Here, it checks if the
   node is currently \textbf{idle} and then synchronizes with $T_2$ to add the
   node to the $T_2$'s queue.
   \item[Thread activation:] If $T_2$ is inactive when adding facts to a node in
   $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
   \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
   nodes from $T_2$ by removing half of the nodes from $T_2$'s queues.
   \item[Coordination:] If $T_1$ needs to perform coordination operations
   to a node in $T_2$, it may need to sychronize with $T_2$ during priority
   updates in order to move the node in $T_2$'s queues.
\end{description}

\subsubsection{Termination}

There is a global atomic counter, a global boolean flag and one boolean flag
(active/idle) per thread for detecting termination. Once a thread goes idle,
it decrements the global counter and changes its flag to idle. If the counter
goes to zero, the global flag is set to idle. Since every thread will be
busy-waiting and checking the global flag, they will detect the change and exit
the program.

### 2014-10-31 intro.tex
Add introduction.

@@
-talk about declarative programming and how hard it is for the programmer to
-control how the program is scheduled or how data is laid out.
+Writing parallel programs in sequential languages is hard because manipulating
+shared state using multiple threads of execution may result in data race
+conditions. Such issues are handled with low level constructs such as locks and
+condition variables, requiring a fair amount of effort to get right.
+Declarative programming has been hailed as a solution to this issue, since the
+problem of implementing the details of parallelism is moved away from the programmer to the
+compiler and runtime of the language. The programmer writes code
+without having to deal with parallel programming constructs and the compiler
+automatically parallelizes the program in order to take advantage of
+multiple threads of execution.
+The programming paradigm has been adopted with huge success in domain specific
+languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
+Although more general declarative languages have yet to be as successful, the
+future looks promising for this particular approach.
 
-we took the work made by \cite{cruz-iclp14} and extended it with action and
-sensing facts.
+We argue, however, that declarative programming leaves little to no programmer control
+over how execution is scheduled or how data is laid out. This introduces some
+performance issues since a runtime system may be able to parallelize the program
+up to a point, using a general algorithm, but lacks some specific information that would make
+execution better in terms of execution time, memory usage and scalability.
+
+In this paper, we extend Linear Meld (LM), a declarative language by
+\cite{cruz-iclp14}, with coordination facts that give programmer control over
+scheduling and data placement. LM is a linear logic programming language
+intended for programs over graph structures and with
+support for structured manipulation of mutable state using 
+linear logic~\cite{girard-87}. In LM, the nodes of the graph perform computation
+through derivation of rules and communicate with other nodes when logical facts
+are derived.
+
+The operational semantics of LM allow for plenty of
+non-determinism since nodes execute independently and can be executed with
+any specific order. Since LM uses mutable state, the order in which nodes are
+executed may be crucial in terms of performance. We extended the language with
+coordination facts that can be used as a regular fact in the program logic to
+improve program execution.
+The first kind of coordination facts are
+called \emph{sensing facts} and are used to sense information about data
+placement (where is the node being executed). This allows the programmer to
+write logical rules that depend on the current state of the program. The second
+kind of coordination facts are \emph{action facts} that, when deriving, will be
+consumed to apply a scheduling operation during execution. This allows the
+programmer to prioritize node computation or place nodes in different threads.
+Both kinds of facts are \emph{linear facts} since they are consumed and/or change.
+
+To the best of our knowledge, this is the first time that a declarative language allows so much
+control over execution without a significant change in the language itself.
+Our main contributions of this paper are then two-fold: a novel way to coordinate
+declarative programs through user control with or without modifications to the
+original algorithm, introducing measurable performance improvements; several
+examples that take simple declarative algorithms and make them more
+efficient through the introduction of coordination facts in the algorithmic
+logic.

### 2014-10-31 application.tex
Added coordination and language sections.

@@
 
+Randomized and approximation
+algorithms can obtain significant benefits from coordination directives because although the final
+program results will not be exact, they follow important statistical properties and can be computed faster.
+Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
+Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.
+
 Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
@@
 \scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
+set-static(A). // all nodes are static.
+
 // TREE BUILDING

### 2014-10-31 coordination.tex
Added coordination and language sections.

@@
-The original Meld was
-implemented as an ensemble programming language, targeting modular robotic systems such as
-Claytronics~\cite{ashley-rollman-derosa-iros07wksp}. In such systems, there is a natural matching
-between computation and processing units, since each robot is represented by a node. This distribution
-of data leaves little choice to be made in the division of computation to the various nodes.
+Since LM uses linear logic and supports deletion of facts, the order in which
+nodes are scheduled can impact the performance and even the results of the
+program.
 
-On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
-since nodes are a program abstraction and part of the program's logic.
-We view the set of nodes as a graph data structure where workers will perform work.
-A worker is able to process any node, although a node cannot be computed by more than one worker
-at the same time. This disallows the manipulation of a node by multiple workers.
+The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
+declarative but its performance may depend in the order in which nodes are
+executed. If nodes with greater distances are prioritized over other nodes, the
+program will generate more \texttt{relax} facts since it will take longer to
+reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
+clear that the best computational ordering is the following: \texttt{@1},
+\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
+facts are generated. If we had decided to process nodes in order
+\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
+\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
+Therefore, the optimal solution for SSSP is to schedule the node with the
+shortest distance, which is essentially the Dijkstra shortest path
+algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
+the algorithm by simply changing the order of node computation, but still
+retain the declarative nature of the program.
 
-Since LM uses linear logic, the order in which nodes are scheduled can impact the
-performance and even the results of the program. When only using
-classical logic, no matter how computation is done, the end result will be the same since the program
-is strictly monotonic.
-LM introduces the concept of \emph{coordination} that allows the programmer
-to write code that changes how the runtime system schedules nodes.
+We introduce the concept of \emph{coordination facts} that allow the programmer
+to change how the run time schedules nodes and partitions the nodes among
+threads of execution. Coordination facts can be used in the body or head of the
+rules to allow the programmer to reason and change how execution is to be done.
+In this fashion, we keep the language declaratively because we reason logically
+about the state of execution, without the need to introduce extra-logical
+operators into the language that would reduce problems when attempting to prove
+properties about programs.
 
-Randomized and approximation
-algorithms can obtain significant benefits from coordination directives because although the final
-program results will not be exact, they follow important statistical properties and can be computed faster.
-Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
-Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.
+There are two kinds of coordination facts. The first kind of coordination facts are
+called \emph{sensing facts} and are used to sense information about data
+placement (where is the node being executed) or scheduled (what is the priority
+of the node). The second kind of coordination facts are \emph{action facts} that,
+when derived, are consumed to apply a coordination operation during execution.
+These logical rules that use coordination are implemented efficiently in the run
+time system and are hidden from the programmer.
 
@@
 We can use action facts to change the order in which nodes are evaluated by adding
-\emph{priorities}. Node priority works at the worker level
-so that each worker can make a local decision about which node of the graph to run next.
+\emph{priorities}. Node priority works at the thread level
+so that each thread can make a local decision about which node of the graph to run next.
 Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).
 
-The following list presents the action facts available to manipulate the scheduling decisions of the system:
+We have two kinds of priorities: a \emph{temporary priority} and a \emph{default
+priority}. A default priority is used indefinitely, while a temporary priority
+is only used once, so that when node with a temporary priority $P$ is picked to
+run, the priority will default back to $D$, the default priority. Initially,
+all nodes have a default priority of $0$.
 
-\begin{description}
-   \item[\texttt{type linear action set-priority(node, float)}]: This sets the priority of a node.
-If the node already has some priority, we only change the priority if the new one is higher priority.
-The programmer can decide if priorities are to be ordered in ascending or descending order.
-   \item[\texttt{type linear action add-priority(node, float)}]: This gets the current node's priority and increases or decreases it.
-   \item[\texttt{type linear action schedule-next(node)}]: The work will fetch the highest priority node's priority $P$ from its set of nodes and set the action's argument node's priority as $P + 1.0$.
-If using the priorities
-   in ascending order, we pick the lowest priority and subtract $1.0$.
-   \item[\texttt{type linear action unset-priority(node)}]: Removes the priority, if any, of a given node.
-   \item[\texttt{type linear action stop-program(node)}]: Immediately stops the execution of the whole program.
-\end{description}
+The following list presents the action facts available to manipulate the scheduling
+decisions of the system:
 
-When the highest priority node is picked up for execution, its priority is reset to 0 (the default priority value). This means that
-the programmer must set the node's priority again if he wants to prioritize that node.
+\begin{itemize}
+   \item \texttt{type linear set-priority(node, float)}: This sets the
+   temporary priority of a node. If the node already has a priority, we only
+   change the priority if the new one is higher priority. The programmer can
+   decide if priorities are to be ordered in ascending or descending order.
+   \item \texttt{type linear action add-priority(node, float)}: Increments,
+   temporarily, the current priority of the node.
+   \item\texttt{type linear schedule-next(node)}: The work will fetch the
+   highest priority node's priority $P$ from its set of nodes and set the
+   action's argument node's priority as $P + 1.0$. If using the priorities in
+   ascending order, we pick the lowest priority and subtract $1.0$.
+   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
+   priority of the node.
+   \item \texttt{type linear action stop-program(node)}: Immediately stops the
+   execution of the whole program.
+\end{itemize}
 
-We intend to add more action facts in the near future. For example, we want the programmer to be able to place specific nodes in workers. This will permit good use of
-memory locality by forcing certain computations to be performed in the same worker.
+LM provides the sensing fact \texttt{type linear priority(node, node, float)},
+where the second argument is the target node and the third argument is the
+node priority. Sensing facts are only used in the body of the rules in order
+to fetch information from the runtime system.
+Note that when sensing facts are consumed, they are re-derived automatically,
+except if the programmer explicitly tells the compiler otherwise. 
+Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
+value of \texttt{priority} facts, but this done seamlessly by the runtime
+system.
 
-Sensing facts provide information about node placement and node priority. We can use those facts
-to express coordination policies. LM provides the following two
-sensing facts:
+\subsubsection{Partitioning facts}
 
-\begin{description}
-   \item[\texttt{type linear cpu-id(node, node, int)}]: The third argument indicates the worker's ID where the node of the second argument is currently running.
-   \item[\texttt{type linear priority(node, node, float)}]: The third argument is the current priority of the node in the second argument.
-\end{description}
+We provide several coordination facts for dealing with node partitioning among
+the available threads executing the program. In terms of action
+facts, we have the following:
 
-Note that when sensing facts are consumed, they are re-derived automatically, except if the programmer explicitly tells the compiler otherwise.
+\begin{itemize}
+   \item \texttt{type linear set-cpu(node, int)}: Moves the node to a
+   thread of execution.
+   \item \texttt{type linear set-affinity(node, node)}: Places the first node in
+   the same thread as the second node.
+   \item \texttt{type linear set-moving(node)}: Allows the node to move freely
+   between threads.
+   \item \texttt{type linear set-static(node)}: Forces the node to stay in the
+   same thread indefinitely.
+\end{itemize}
 
+For sensing facts, we have the following set of coordination facts:
+
+\begin{itemize}
+   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
+   indicates the thread where the node of the second argument is currently running.
+   \item \texttt{type linear moving(node)}: Fact available if the node in the
+   second argument is allowed to move between threads.
+   \item \texttt{type linear static(node, node)}: Fact available if the node in
+   the second argument is not allowed to move between threads.
+\end{itemize}
+
+\iffalse
 \subsubsection{Global Directives}
@@
 \end{description}
+
+\fi

### 2014-10-31 implementation.tex
Added coordination and language sections.

@@
 the program.
+
+\subsubsection{Node collection}
+
+Whenever a node has an empty database and no references from other nodes, it is
+deleted from the graph. The virtual machine detects such cases by keeping a
+count of fact references to every node. Once the count drops to zero and the
+database is empty, the node is immediatelly deleted in order to reduce memory
+usage.

### 2014-10-31 intro.tex
Added coordination and language sections.

@@
 Writing parallel programs in sequential languages is hard because manipulating
-shared state using multiple threads of execution may result in data race
+shared state using multiple threads may result in data race
 conditions. Such issues are handled with low level constructs such as locks and
@@
 over how execution is scheduled or how data is laid out. This introduces some
-performance issues since a runtime system may be able to parallelize the program
-up to a point, using a general algorithm, but lacks some specific information that would make
-execution better in terms of execution time, memory usage and scalability.
+performance issues because even if the runtime system is able to reasonably
+parallelize the program using a general algorithm, there is a lack of specific
+information about the program that a compiler does not easily understand. In
+turn, such information would make execution better in terms of run time time,
+memory usage and scalability.
 
 In this paper, we extend Linear Meld (LM), a declarative language by
-\cite{cruz-iclp14}, with coordination facts that give programmer control over
+\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer control over
 scheduling and data placement. LM is a linear logic programming language
@@
 
-To the best of our knowledge, this is the first time that a declarative language allows so much
-control over execution without a significant change in the language itself.
-Our main contributions of this paper are then two-fold: a novel way to coordinate
-declarative programs through user control with or without modifications to the
-original algorithm, introducing measurable performance improvements; several
-examples that take simple declarative algorithms and make them more
-efficient through the introduction of coordination facts in the algorithmic
-logic.
+To the best of our knowledge, this is the first time that a declarative language allows 
+control over execution without a significant change in the language itself and
+without using non-declarative operations. This proves specially important since
+programs can proven to be correct even in the presence of coordination facts.
+Our other two main contributions are as follows: a novel way to coordinate
+declarative programs through user control that introduce measurable performance
+improvements; several application where we add programmer control to declarative
+programs in order to make them perform better.

### 2014-10-31 language.tex
Added coordination and language sections.

@@
-give an overview of the language
+Linear Meld (LM) is a forward-chaining linear logic programming
+language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
+Programs are written as a set of rules of the form
+\texttt{a(X), b(Y) -o c(Z)} and are read as following: if \texttt{a(X), b(Y)} (the body of
+the rule) is satisfied using some facts \texttt{a(X)} and \texttt{b(Y)} then
+\texttt{c(Z)} (the head of the rule) is true.
+The facts used during rule derivation are placed in a \emph{database of facts}.
+Logical facts are composed of a predicate and a tuple of concrete values,
+representing the arguments.
+Since LM uses linear logic as its foundation, rule derivations
+may delete facts used in the body of the rule. Program execution starts by
+adding the \emph{axioms} (the initial facts) of the program to the database.
+Next, the rules are recursively applied and the database is updated with new and
+deleted facts. When nore more rules are applicable, the program terminates.
 
-we already have the ICLP paper that describes the language, so we can be brief
-here. \cite{cruz-iclp14}
+LM has been designed for writing graph-based programs. LM accomplishes this by
+partitioning the logical facts by the first argument (the \emph{node}) and by
+restricting the body of every rule to only refer to the same first argument.
+This allows rules to be executed using only facts of a single node. Although the
+body of rules is restricted, the head of the rule may refer to any other node as
+long as the node variable is refered somewhere in the body of the rule. This allows
+derivations of facts in other nodes. The whole database is then composed of smaller
+databases (the nodes) and rule derivations allows smaller databases to derive
+facts in other databases, creating edges in the graph.
+
+In order to fully understand how the language works, we present a very simple
+algorithm: the single source shortest path program~(SSSP). Later in the paper, we
+add coordination facts to improve the execution of the program.
+
+The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
+with the declaration of the predicates. Predicates specify the kinds of facts
+used in the program. The first predicate, \texttt{edge}, specifies edges between
+nodes of the graph, where the third argument represents the weight of the edge.
+The \texttt{route} modifier informs the compiler about the structure of the
+program graph. The predicates \texttt{shortest} and \texttt{relax} are specified
+as linear facts and thus are deleted when used in rules.
+The idea of the algorithm is to compute the shortest distance from node
+\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
+fact that is improved with new \texttt{relax} facts.
+Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
+graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
+for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
+initial distance from \texttt{@1} to \texttt{@1}.
+
+\begin{figure}[h!]
+\scriptsize\begin{Verbatim}[numbers=left]
+type route edge(node, node, int).
+type linear shortest(node, int, list int).
+type linear relax(node, int, list int).
+
+!edge(@1, @2, 3). !edge(@1, @3, 1).
+!edge(@3, @2, 1). !edge(@3, @4, 5).
+!edge(@2, @4, 1).
+shortest(A, +00, []).
+relax(@1, 0, [@1]).
+
+relax(A, D1, P1), shortest(A, D2, P2),
+D1 < D2
+   -o shortest(A, D1, P1),
+      {B, W | !edge(A, B, W) |
+         relax(B, D1 + W, P1 ++ [B])}.
+
+relax(A, D1, P1), shortest(A, D2, P2),
+D1 >= D2
+   -o path(A, D2, P2).
+\end{Verbatim}
+  \caption{Single Source Shortest Path program code.}
+  \label{code:shortest_path_program}
+\end{figure}
+\normalsize
+
+The first rule of the program (lines 11-15) replaces the current path in
+\texttt{shortest} with a shorter one in \texttt{relax}. The rule deletes both
+\texttt{relax} and \texttt{shortest} facts and derives a new \texttt{shortest}
+fact. In lines 14-15, we have a \emph{comprehension} where the program iterates
+over the edges of node \texttt{A} and derives a new \texttt{relax} fact at
+\texttt{B} with the new shorter distance plus the weight of the edge. For
+instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
+\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
+and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
+applying the same rule but at node \texttt{2}.
+
+\begin{figure*}[ht]
+\begin{center}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
+\end{center}
+\caption{Graphical representation of the SSSP program. Figure (a) represents the
+   program after propagating initial distance at node \texttt{@1}, followed by
+   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
+   represents the state of the final program, where all the shortest paths
+   have been computed.}
+\label{fig:shortest_path_program}
+\end{figure*}
+
+The second rule of the program (lines 17-19) deals with the case where the new
+distance in \texttt{relax} is not better than the current one, so it is thrown
+away and the current distance is kept.
+
+There many opportunities for concurrency in the SSSP program. For instance,
+after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
+possible to either apply rules in either node \texttt{@2} or node
+\texttt{@3}. In our implementation, we partition the graph into subgraphs that
+are processed by multiple threads of execution. Eventually, it is no longer to
+possible to apply rules and the final result present in
+Fig.~\ref{fig:shortest_path_program}~(c) is achieved.
+
+%On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
+%since nodes are a program abstraction and part of the program's logic.
+%We view the set of nodes as a graph data structure where workers will perform work.
+%A worker is able to process any node, although a node cannot be computed by more than one worker
+%at the same time. This disallows the manipulation of a node by multiple workers.

### 2014-11-01 application.tex
More updates.

@@
-To better understand how coordination directives work, we next present some programs that
+To better understand how coordination facts are used, we next present some programs that
 take advantage of them.
 
-\subsubsection{Graph Search}
+\subsection{Single Source Shortest Path}
 
-talk about a-star
+The Single Source Shortest Path (SSSP) program was already presented in
+Section~\ref{sec:language}. We are going to update the code in
+Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
+are selected to run before others nodes. The coordinated version of the same
+algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
+uses one coordination fact, namely \texttt{set-priority} (line 15). Note that we
+also use a global program directive to order priorities in ascending order (line
+5).
 
-\subsubsection{Single Source Shortest Path}
-
-The Single Source Shortest Path (SSSP) is a graph algorithm where we want to compute the
-distance of all nodes to a single node. The full code is presented in Fig.~\ref{code:shortest_path_program}.
-The graph is represented using the \texttt{edge(A,~B,~W)} predicate (line 1), where
-\texttt{A} $\rightarrow$ \texttt{B} is a directed edge from node \texttt{A} to \texttt{B}
-with distance \texttt{W}. The second predicate is \texttt{path(A,~D,~F)} (line 2), where
-\texttt{D} is the distance of node \texttt{A} to our initial node and \texttt{F}
-indicates if this path has been processed.
-
-Every node may have several \texttt{path} facts. From this set of facts, the node will select
-the path with the least distance (rules in lines 11-13) and then propagate it using the third rule (lines 15-19).
-When a path is used for propagation, we make sure that the path flag is \texttt{notused} (line 15)
-because we do not want to propagate the same distance twice. When the rule is fired, the flag
-then changes to \texttt{used} (line 19).
-
-We start with the axiom \texttt{path(startnode,~0,~notused)} in line 9 (the distance to the starting
-node is \texttt{0}). This will kickstart the computation by propagating the initial distance to the
-neighbor nodes using rule 3. We also use 2 coordination directives:
-
-\begin{itemize}
-   \item \texttt{priority~@order~asc}: paths are picked in ascending order (line 4).
-   \item \texttt{set-priority(B,~D~+~W)}: to change the node's priority upon the computation of
-a new distance (line 18).
-\end{itemize}
-
-With these directives, we ensure that we pick the node with the smallest distance
-first. If we are only using 1 thread, then our algorithm will behave like Dijkstra's shortest
-path algorithm~\cite{Dijkstra}. When using more than 1 thread, each thread will pick the smallest
-path from a subset of nodes. This is an example where using coordination directives will
-improve the speed of execution since we avoid propagating some of the paths.
+When using a single thread, the algorithm behaves like
+Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
+threads, each thread will pick the smallest distance from their subset of nodes.
+While this is not optimal since threads do not share a global view of priorities,
+many rule derivations are going to be avoided.
 
@@
 type route edge(node, node, int).
-type linear path(node, int, int).
+type linear shortest(node, int, list int).
+type linear relax(node, int, list int).
 
@@
 
-relax(@0, 0).
-path(A, +00).
+shortest(A, +00, \[\]).
+relax(@1, 0, [@1]).
 
-relax(A, D1), path(A, D2),
+relax(A, D1, P1), shortest(A, D2, P2),
 D1 < D2
-   -o path(A, D1),
+   -o shortest(A, D1, P1),
       {B, W | !edge(A, B, W) |
-         relax(B, D + W),
-         &underline[set-priority(B, D + W))]},
+         relax(B, D1 + W, P1 ++ \[B\]),
+         &underline[set-priority(B, float(D1 + W))]}.
 
-relax(A, D1), path(A, D2),
+relax(A, D1, P1), shortest(A, D2, P2),
 D1 >= D2
-   -o path(A, D2).
+   -o path(A, D2, P2).
 \end{Verbatim}
   \caption{Shortest Path Program.}
-  \label{code:shortest_path_program}
+  \label{code:shortest_path_program_coord}
 \end{figure}
@@
 
-\subsubsection{Heat Transfer}
+\begin{figure}[h!]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=4cm]{results/shortest-uspowergrid.png}}
+      \subfloat[]{ \scriptsize\begin{tabular}{ | c | c | c | }
+         \hline                       
+         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
+         1 & 333K & 206K \\
+         2 & 300K & 210K \\
+         4 & 316K & 208K \\
+         8 & 328K & 211K \\
+         16 & 343K & 212K \\
+         \hline  
+         \end{tabular}
+         \normalsize
+      }
+   \end{center}
+   \caption{Single Source Shortest Path program code.}
+   \label{results:sssp_uspowergrid}
+\end{figure}
 
-\subsubsection{MiniMax}
+\subsection{Graph Search}
 
-\subsubsection{Splash BP}
+\subsection{MiniMax}
+
+\begin{figure}[h!]
+   \begin{center}
+      \includegraphics[width=5cm]{figures/minimax_tree}
+   \end{center}
+   \caption{Single Source Shortest Path program code.}
+   \label{minimax}
+\end{figure}
+
+\subsection{Heat Transfer}
+
+\subsection{Splash BP}
 
@@
 
+\begin{figure}[h!]
+   \begin{center}
+      \includegraphics[width=6.5cm]{figures/splash_bp}
+   \end{center}
+   \caption{Single Source Shortest Path program code.}
+   \label{splash_bp}
+\end{figure}
+
 The code in Fig.~\ref{code:sbp} presents the coordination code of the Belief Propagation problem.
@@
 When the highest priority is picked we create a token that will navigate through the tree.
-Note that in the rule located in lines 11-18 we check if the priority of the new node to add to the tree is positive and that both nodes are in the same worker.
-We want the tree to be kept in the same worker.
+Note that in the rule located in lines 11-18 we check if the priority of the new node to add to the tree is positive and that both nodes are in the same thread.
+We want the tree to be kept in the same thread.
    \item[First phase]: In the third rule (lines 8-9), when we reach a certain number of nodes in the tree, we generate \texttt{first-phase} in order to update the beliefs of all nodes in tree starting from the leaves and ending at the root.
@@
 
-When we have several workers, every worker will generate their own trees by taking into account the highest priority node in their own queues.
+When we have several threads, every thread will generate their own trees by taking into account the highest priority node in their own queues.
 

### 2014-11-02 application.tex
Add more text in MiniMax.

@@
 \begin{figure}[h!]
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=&\[\]]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
 type route edge(node, node, int).
@@
 
-&underline[priority @order asc].
+\underline{priority @order asc}.
 
-shortest(A, +00, \[\]).
+shortest(A, +00, []).
 relax(@1, 0, [@1]).
@@
    -o shortest(A, D1, P1),
-      {B, W | !edge(A, B, W) |
-         relax(B, D1 + W, P1 ++ \[B\]),
-         &underline[set-priority(B, float(D1 + W))]}.
+      \{B, W | !edge(A, B, W) |
+         relax(B, D1 + W, P1 ++ [B]),
+         \underline{set-priority(B, float(D1 + W))}\}.
 
@@
 D1 >= D2
-   -o path(A, D2, P2).
+   -o shortest(A, D2, P2).
 \end{Verbatim}
@@
 
+The most interesting property of the SSSP program presented in
+Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
+although it applies rules using smarter ordering. The derivation of
+\texttt{set-priority} does not change the behavior of the logical rules and the
+code remains declarative.
+
+Figure~\ref{results:sssp_uspowergrid} shows experimental results for a variant of
+the SSSP program that computes the SSSP for many of the nodes of the graph.
+There are some situations where unecessary facts are propagated
+because although the shortest distance is selected, sub-optimal distances may be
+propagated because many SSSP distances are computed at the same time.
+However, we see a reduction of over 50\% in the number of
+derived facts when using the coordinated version \textbf{C} over the regular
+version \textbf{R}. The results also show that the cordinated
+version using 16 threads is 1.3 times faster than the regular version using 16
+threads. In turn, this makes the coordinated version 20 times faster
+than the regular version using 1 thread.
+
 \begin{figure}[h!]
    \begin{center}
-      \subfloat[]{\includegraphics[width=4cm]{results/shortest-uspowergrid.png}}
-      \subfloat[]{ \scriptsize\begin{tabular}{ | c | c | c | }
+      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
          \hline                       
@@
       }
+      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
    \end{center}
-   \caption{Single Source Shortest Path program code.}
+   \caption{Experimental results for the SSSP program using US powergrid network.
+   On the left we show the number of facts derived for the regular \textbf{R}
+   and coordinated version \textbf{C} using a variable number of threads
+   \textbf{\# T}. On the right, we have the scalability of the regular version
+   and coordinated version. The speedup values are computed using the execution
+   time of the regular version for 1 thread.}
    \label{results:sssp_uspowergrid}
@@
 
-\subsection{Graph Search}
+\subsection{MiniMax}
 
-\subsection{MiniMax}
+The MiniMax algorithm is a decision rule algorithm for minimizing the
+possible loss for a worst case (maximum loss) scenario in a zero sum
+game for 2 (or more) players that play in
+turns~\cite{Edwards54}.
+
+The algorithm proceeds by building a game tree, where each tree node represents a
+game state and the children represent the possible game moves that can
+be made by either player 1 or player 2. The
+tree levels are built in a way that each player plays in turns and the
+first level represents the moves made by player 1.  An
+evaluation function \texttt{f(State)} is used to compute the score of
+the board for each leaf of the tree. A node is a leaf when it can no longer be
+expanded since the state represents a possible end of the game.  Finally, the
+algorithm recursively minimizes or maximizes the scores of each node.
+To select the best move for player 1, the
+algorithm picks the move maximized at the root node.
+
+In LM, the program starts with a root node (with the initial game state)
+that is expanded with the available moves at each level. The graph of the
+program is dynamic since nodes are created and eventually deleted once they
+have nore more facts or no more references to them. The latter happens when the
+leaves scores are computed or when a node fully minimizes or maximizes the
+children scores. When the program ends, only the root node is have facts in its
+database.
+
+The LM code in Fig.~\ref{minimax:check-end} implements
+the initial expansion of the tree. The first three rules (lines 1-10) deal
+with the cases where no children nodes are created and the last three rules
+(12-29) deal with cases where new nodes may be created.
+Thes rule in lines 12-25 generate new nodes using the
+\texttt{exists} language construct, which creates a new node address
+\texttt{B}, where a new \texttt{play} fact is created with the new
+board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
+communicate with its parent.
+
+Since our implementation
+uses a FIFO ordering for nodes, meaning that when children are expanded, we
+first expand each children and then the children of the children. The default
+ordering of the system works well enough for most practical purposes. We note
+however, that this is not optimal for this program since the complete tree needs
+to be expanded before computing the scores at the leaves. This introduces memory
+issues since, in the worst case, the memory needs of the program are
+$\mathcal{O}(n)$, where $n$ is the number of nodes in the tree.
+
+\begin{figure}[h!]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
+expand(A, Board, [], 0, P, \underline{Depth})
+  -o leaf(A, Board).
+
+expand(A, Board, [], N, P, \underline{Depth}),
+N > 0
+  -o maximize(A, N, -00, 0).
+
+expand(A, Board, [], N, P, \underline{Depth}),
+N > 0, NextPlayer <> RootPlayer
+  -o minimize(A, N, +00, 0).
+
+expand(A, Board, [0 | Xs], N, P, Depth),
+Depth >= 5
+  -o exists B. (\underline{set-static(B)},
+       \underline{set-default-priority(B, float(Depth + 1))},
+       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
+       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
+       parent(B, A)).
+
+expand(A, Board, [0 | Xs], N, P, Depth),
+Depth < 5
+  -o exists B. (\underline{set-default-priority(B, float(Depth + 1))},
+       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
+       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
+       parent(B, A)).
+
+expand(A, Board, [C | Xs], N, P, \underline{Depth})
+C <> 0
+  -o expand(A, Board ++ [C], Xs, N, P, \underline{Depth}).
+\end{Verbatim}
+\caption{MiniMax: checking if the game has ended and expanding the tree}
+\label{minimax:check-end}
+\end{figure}
+\normalsize
+
+With coordination, we set the priority nodes to be the same as the
+depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
+The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
+is the depth of the tree and $t$ is the number of threads. In
+Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
+expanded the tree on the left side. When a second thread comes up, it steals the
+first unexpanded node and expands the tree from that position. Since threads
+prioritize deeper nodes, the scores of the first leaves are immediatelly
+computed and then sent to the parent node. At this point, the leaves are deleted
+and reused for other nodes in the tree, resulting in minimal memory usage.
 
@@
    \begin{center}
-      \includegraphics[width=5cm]{figures/minimax_tree}
+      \includegraphics[width=4.5cm]{figures/minimax_tree}
    \end{center}
-   \caption{Single Source Shortest Path program code.}
-   \label{minimax}
+   \caption{Expanding the MiniMax tree using coordination. By prioritizing
+      deeper nodes, threads are forced to expand the tree using a depth-first
+      approach, which is superior since there is no need to expand the whole
+      tree before computing the node scores.}
+   \label{fig:minimax}
 \end{figure}
+
+We also take advantage of memory locality by using \texttt{set-static} (line
+14), so that nodes after a certain level are not stolen by other threads. While
+this is not critical for performance in shared memory systems where node
+stealing is fairly efficient, we expect that such coordination to be critical in
+distributed systems.
 

### 2014-11-02 application.tex
Splash BP updates.

--- 2014-11-01 application.tex
@@
 \subsection{Single Source Shortest Path}
-
-The Single Source Shortest Path (SSSP) program was already presented in
-Section~\ref{sec:language}. We are going to update the code in
-Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
-are selected to run before others nodes. The coordinated version of the same
-algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
-uses one coordination fact, namely \texttt{set-priority} (line 15). Note that we
-also use a global program directive to order priorities in ascending order (line
-5).
-
-When using a single thread, the algorithm behaves like
-Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
-threads, each thread will pick the smallest distance from their subset of nodes.
-While this is not optimal since threads do not share a global view of priorities,
-many rule derivations are going to be avoided.
-
-\begin{figure}[h!]
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=&\[\]]
-type route edge(node, node, int).
-type linear shortest(node, int, list int).
-type linear relax(node, int, list int).
-
-&underline[priority @order asc].
-
-shortest(A, +00, \[\]).
-relax(@1, 0, [@1]).
-
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 < D2
-   -o shortest(A, D1, P1),
-      {B, W | !edge(A, B, W) |
-         relax(B, D1 + W, P1 ++ \[B\]),
-         &underline[set-priority(B, float(D1 + W))]}.
-
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 >= D2
-   -o path(A, D2, P2).
-\end{Verbatim}
-  \caption{Shortest Path Program.}
-  \label{code:shortest_path_program_coord}
-\end{figure}
-\normalsize
-
-\begin{figure}[h!]
-   \begin{center}
-      \subfloat[]{\includegraphics[width=4cm]{results/shortest-uspowergrid.png}}
-      \subfloat[]{ \scriptsize\begin{tabular}{ | c | c | c | }
-         \hline                       
-         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
-         1 & 333K & 206K \\
-         2 & 300K & 210K \\
-         4 & 316K & 208K \\
-         8 & 328K & 211K \\
-         16 & 343K & 212K \\
-         \hline  
-         \end{tabular}
-         \normalsize
-      }
-   \end{center}
-   \caption{Single Source Shortest Path program code.}
-   \label{results:sssp_uspowergrid}
-\end{figure}
-
-\subsection{Graph Search}
+\input{sssp}
 
@@
 
-\begin{figure}[h!]
-   \begin{center}
-      \includegraphics[width=5cm]{figures/minimax_tree}
-   \end{center}
-   \caption{Single Source Shortest Path program code.}
-   \label{minimax}
-\end{figure}
+\input{minimax}
 
@@
 
-\subsection{Splash BP}
+\input{heat-transfer}
 
-Randomized and approximation
-algorithms can obtain significant benefits from coordination directives because although the final
-program results will not be exact, they follow important statistical properties and can be computed faster.
-Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
-Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.
+\subsection{Splash Belief Propagation}
 
-Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
-used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
-where nodes exchange messages with their immediate neighbors and apply some computations to the messages
-received.
+\input{splash-bp}
 
-LBP is an algorithm that maps very well to the graph based model of LM. In its original form, we need to compute
-the belief of all nodes for several iterations and also synchronize after each iteration.
-However, it is still possible to apply
-some optimizations in order to take advantage of the fact that LBP will converge even when using
-an asynchronous approach. So, instead of computing the belief iteratively,
-we first keep track of all messages sent/received (and overwrite them when we receive a new one)
-and recompute the belief when we want, instead of synchronizing between nodes.
-This way, we can prioritize the computation of beliefs when
-a node's belief value changes significantly. When that happens, we set the priority of its
-neighbors so that they can recompute their beliefs.
-
-The asynchronous approach proves to be a nice improvement over the synchronous version. Still, it
-is possible to do even better. Gonzalez et al~\cite{Gonzalez+al:aistats09paraml} developed an optimal
-algorithm to compute this algorithm by first building a tree and then updating the beliefs of each node twice, first from the leaves to the root and then from the root to the leaves.
-The root of this tree
-is the node with the highest priority (based on belief) while the other nodes in the tree
-must have a non-zero priority. Note that the priorities are updated whenever a neighbor updates
-their belief. These splash trees are built iteratively until we reach convergence.
-
-\begin{figure}[h!]
-   \begin{center}
-      \includegraphics[width=6.5cm]{figures/splash_bp}
-   \end{center}
-   \caption{Single Source Shortest Path program code.}
-   \label{splash_bp}
-\end{figure}
-
-The code in Fig.~\ref{code:sbp} presents the coordination code of the Belief Propagation problem.
-Please note that we just appended the code in Fig.~\ref{code:sbp} to a working but
-unoptimized version of the algorithm.
-In the coordination code we have three sections:
-\begin{description}
-   \item[Tree building]: Each node has a \texttt{waiting} fact that is used to start the tree building process.
-When the highest priority is picked we create a token that will navigate through the tree.
-Note that in the rule located in lines 11-18 we check if the priority of the new node to add to the tree is positive and that both nodes are in the same thread.
-We want the tree to be kept in the same thread.
-   \item[First phase]: In the third rule (lines 8-9), when we reach a certain number of nodes in the tree, we generate \texttt{first-phase} in order to update the beliefs of all nodes in tree starting from the leaves and ending at the root.
-As we update the nodes, we generate \texttt{update} to update the belief values (line 29).
-   \item[Second phase]: In the second phase we go from the root to the leaves and update the beliefs a second time (line 39).
-\end{description}
-
-When we have several threads, every thread will generate their own trees by taking into account the highest priority node in their own queues.
-
-\begin{figure}[h!]
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
-set-static(A). // all nodes are static.
-
-// TREE BUILDING
-// continue tree
-waiting(A), token(A, All, Next) -o token(A, All, Next).
-// start tree
-waiting(A), *underline{@priority(A, A, P)}, P > 0.0
-   -o token(A, [A], [A]).
-// end tree building
-token(A, All, Next), length(All) > maxnodes
-   -o first-phase(A, All, reverse(All)).
-// expand tree
-token(A, All, [A | Next])
-   -o [collect => L | Side | !edge(A, L, Side),
-         0 = count(All, L),
-         0 = count(Next, L),
-         *underline{priority(A, L, P)}, P > 0.0,
-         *underline{cpu-id(A, L, Id1)},
-         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
-         send-token(A, All, Next ++ L)].
-
-send-token(A, All, [])
-   -o first-phase(A, All, reverse(All)).
-send-token(A, All, [B | Next])
-   -o *underline{schedule-next(B)},
-      token(B, All ++ [B], [B | Next]).
-
-// FIRST PHASE
-first-phase(A, [A], [A]) -o second-phase(A, [], A).
-first-phase(A, [A, B | Next], [A])
-   -o update(A), *underline{schedule-next(B)},
-      second-phase(B, [B | Next], A).
-first-phase(A, All, [A, B | Next])
-   -o update(A), *underline{schedule-next(B)},
-      first-phase(B, All, [B | Next]).
-
-// SECOND PHASE
-second-phase(A, [], _)
-   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
-second-phase(A, [A], Back)
-   -o update(A), waiting(Back),
-      waiting(A), *underline{set-priority(A, 0.0)}.
-second-phase(A, [A, B | Next], Back)
-   -o update(A), waiting(Back), *underline{schedule-next(B)},
-      second-phase(B, [B | Next], A).
-\end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program.}
-  \label{code:sbp}
-\end{figure}
-\normalsize
+\subsection{N Queens}

### 2014-11-02 heat-transfer.tex
Splash BP updates.

Randomized and approximation
algorithms can obtain significant benefits from coordination directives because although the final
program results will not be exact, they follow important statistical properties and can be computed faster.
Examples of such programs are Loopy Belief
Propagation~\cite{Gonzalez+al:aistats09paraml} (shown in the next section) and
Heat Transfer.

### 2014-11-02 minimax.tex
Splash BP updates.

The MiniMax algorithm is a decision rule algorithm for minimizing the
possible loss for a worst case (maximum loss) scenario in a zero sum
game for 2 (or more) players that play in
turns~\cite{Edwards54}.

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2. The
tree levels are built in a way that each player plays in turns and the
first level represents the moves made by player 1.  An
evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when it can no longer be
expanded since the state represents a possible end of the game.  Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In LM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and eventually deleted once they
have nore more facts or no more references to them. The latter happens when the
leaves scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node is have facts in its
database.

The LM code in Fig.~\ref{minimax:check-end} implements
the initial expansion of the tree. The first three rules (lines 1-10) deal
with the cases where no children nodes are created and the last three rules
(12-29) deal with cases where new nodes may be created.
Thes rule in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the new
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

Since our implementation
uses a FIFO ordering for nodes, meaning that when children are expanded, we
first expand each children and then the children of the children. The default
ordering of the system works well enough for most practical purposes. We note
however, that this is not optimal for this program since the complete tree needs
to be expanded before computing the scores at the leaves. This introduces memory
issues since, in the worst case, the space complexity of the program are
$\mathcal{O}(n)$, where $n$ is the number of nodes in the tree.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
expand(A, Board, [], 0, P, \underline{Depth})
  -o leaf(A, Board).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0
  -o maximize(A, N, -00, 0).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0, NextPlayer <> RootPlayer
  -o minimize(A, N, +00, 0).

expand(A, Board, [0 | Xs], N, P, Depth),
Depth >= 5
  -o exists B. (\underline{set-static(B)},
       \underline{set-default-priority(B, float(Depth + 1))},
       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
       parent(B, A)).

expand(A, Board, [0 | Xs], N, P, Depth),
Depth < 5
  -o exists B. (\underline{set-default-priority(B, float(Depth + 1))},
       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
       parent(B, A)).

expand(A, Board, [C | Xs], N, P, \underline{Depth})
C <> 0
  -o expand(A, Board ++ [C], Xs, N, P, \underline{Depth}).
\end{Verbatim}
\caption{MiniMax: checking if the game has ended and expanding the tree}
\label{minimax:check-end}
\end{figure}
\normalsize

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. In
Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
expanded the tree on the left side. When a second thread comes up, it steals the
first unexpanded node and expands the tree from that position. Since threads
prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=4.5cm]{figures/minimax_tree}
   \end{center}
   \caption{Expanding the MiniMax tree using coordination. By prioritizing
      deeper nodes, threads are forced to expand the tree using a depth-first
      approach, which is superior since there is no need to expand the whole
      tree before computing the node scores.}
   \label{fig:minimax}
\end{figure}

We also take advantage of memory locality by using \texttt{set-static} (line
14), so that nodes after a certain level are not stolen by other threads. While
this is not critical for performance in shared memory systems where node
stealing is fairly efficient, we expect that such coordination to be critical in
distributed systems.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c |}
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 11.80GB & 0.50MB \\ \hline
         2 & 12.19GB & 1.45MB \\ \hline
         4 & 13.82GB & 2.35MB \\ \hline
         8 & 14.87GB & 4.36MB \\ \hline
         16 & 13.79GB & 8.1MB \\ \hline
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=5cm]{results/min-max-tictactoe.png}}
   \end{center}
   \caption{Memory usage and scalability of the regular and coordinated versions
      of MiniMax.}
   \label{results:memory_minmax}
\end{figure}

### 2014-11-02 splash-bp.tex
Splash BP updates.

Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP is an algorithm that maps very well to the graph based model of LM. In its
original form, the belief values of nodes are computed by synchronous iterations.
LBP offers more concurrency when belief values are computed asynchronously
leading to faster convergence. For this, every node keeps track of all messages
sent/received and recomputes the belief using partial information from neighbor
nodes. It is then possible to prioritize the computation of beliefs when a
neighbor's belief changes significantly.

The asynchronous approach proves to be a nice improvement over the synchronous
version. Still, it is possible to do even better. Gonzalez et
al~\cite{Gonzalez+al:aistats09paraml} developed an optimal algorithm to compute
this algorithm by first building a tree and then updating the beliefs of each
node twice, first from the leaves to the root and then from the root to the
leaves. The root of this tree is the node with the highest priority (based on
belief) while the other nodes in the tree must have a non-zero priority.
Note that the priorities are updated whenever a neighbor updates their belief.
These \emph{splash trees} are built iteratively until we reach convergence.

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=6.5cm]{figures/splash_bp}
   \end{center}
   \caption{Single Source Shortest Path program code.}
   \label{splash_bp}
\end{figure}

The code for Splash Belief Propagation~(SBP) in Fig.~\ref{code:sbp} presents the
coordination code for LBP.  Please note that we just appended the code in
Fig.~\ref{code:sbp} to a working but unoptimized version of the algorithm, every
other logical rule remains the same. We add new logical ruels that coordinate
the creation and execution of the splash trees:

\begin{description}
   \item[Tree building]: Each node has a \texttt{waiting} fact that is used to
   start the tree building process. When the highest priority is picked, a token
   is created that will navigate through the tree. Note that in the rule located
   in lines 13-20, newly added nodes to the tree must have a positive priority
   (due to new belief updates) and must be residing in the same thread.
   We want the tree to be kept in the same thread in order to maximize
   parallelism with 1 tree per thread.
   \item[First phase]: In the third rule (lines 10-11), when the number of nodes
   of the tree reaches a certain limit, a 
   \texttt{first-phase} is generated in order to update the beliefs of all nodes
   in the tree, starting from the leaves and ending at the root
   As the nodes are updated, an \texttt{update} fact is derived to update
   the belief values (line 34).
   \item[Second phase]: In the second phase, the computation of beliefs is
   performed from the root to the leaves and the belief values are updated a
   second time (line 41).
\end{description}

When we have several threads, every thread will generate their own trees by
taking into account the highest priority node in their own queues. This is the
reason we use \texttt{set-static} in line 1 to force all nodes to remain in the
original threads.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
*underline{set-static(A)}. // all nodes are static.

// TREE BUILDING
// continue tree
waiting(A), token(A, All, Next) -o token(A, All, Next).
// start tree
waiting(A), *underline{@priority(A, A, P)}, P > 0.0
   -o token(A, [A], [A]).
// end tree building
token(A, All, Next), length(All) > maxnodes
   -o first-phase(A, All, reverse(All)).
// expand tree
token(A, All, [A | Next])
   -o [collect => L | Side | !edge(A, L, Side),
         0 = count(All, L),
         0 = count(Next, L),
         *underline{priority(A, L, P)}, P > 0.0,
         *underline{cpu-id(A, L, Id1)},
         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
         send-token(A, All, Next ++ L)].

send-token(A, All, [])
   -o first-phase(A, All, reverse(All)).
send-token(A, All, [B | Next])
   -o *underline{schedule-next(B)},
      token(B, All ++ [B], [B | Next]).

// FIRST PHASE
first-phase(A, [A], [A]) -o second-phase(A, [], A).
first-phase(A, [A, B | Next], [A])
   -o update(A), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
first-phase(A, All, [A, B | Next])
   -o update(A), *underline{schedule-next(B)},
      first-phase(B, All, [B | Next]).

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
second-phase(A, [A], Back)
   -o update(A), waiting(Back),
      waiting(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), waiting(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program.}
  \label{code:sbp}
\end{figure}
\normalsize

In this program, coordination assumes a far more important role than we have
seen before. Coordination rules fully drive the behavior of the algorithm and
although the final result of the algorithm is identical to the original
algorithm (minus probabilitistic errors), SBP works in a very different way.
A system that also implements SBP is
GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
GraphLab provides the splash scheduler for these types of inferences as part of
the framework. This particular scheduler is around 350 lines of complicated C++ code.
With our coordination facts, it is possible to create fairly complicated coordination
patterns with only 12 simple logical rules.

### 2014-11-02 sssp.tex
Splash BP updates.

The Single Source Shortest Path (SSSP) program was already presented in
Section~\ref{sec:language}. We are going to update the code in
Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
are selected to run before others nodes. The coordinated version of the same
algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 15). Note that we
also use a global program directive to order priorities in ascending order (line
5).

When using a single thread, the algorithm behaves like
Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
threads, each thread will pick the smallest distance from their subset of nodes.
While this is not optimal since threads do not share a global view of priorities,
many rule derivations are going to be avoided.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

\underline{priority @order asc}.

shortest(A, +00, []).
relax(@1, 0, [@1]).

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      \{B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ [B]),
         \underline{set-priority(B, float(D1 + W))}\}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o shortest(A, D2, P2).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program_coord}
\end{figure}
\normalsize

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering. The derivation of
\texttt{set-priority} does not change the behavior of the logical rules and the
code remains declarative.

Figure~\ref{results:sssp_uspowergrid} shows experimental results for a variant of
the SSSP program that computes the SSSP for many of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads. In turn, this makes the coordinated version 20 times faster
than the regular version using 1 thread.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 333K & 206K \\
         2 & 300K & 210K \\
         4 & 316K & 208K \\
         8 & 328K & 211K \\
         16 & 343K & 212K \\
         \hline  
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
   \end{center}
   \caption{Experimental results for the SSSP program using US powergrid network.
   On the left we show the number of facts derived for the regular \textbf{R}
   and coordinated version \textbf{C} using a variable number of threads
   \textbf{\# T}. On the right, we have the scalability of the regular version
   and coordinated version. The speedup values are computed using the execution
   time of the regular version for 1 thread.}
   \label{results:sssp_uspowergrid}
\end{figure}

### 2014-11-03 splash-bp.tex
Splash BP results.

@@
 These \emph{splash trees} are built iteratively until we reach convergence.
+In Fig.~\ref{splash_bp} we represent two threads creating two different splash
+trees. It is encouraged to create several splash trees concurrently as long as
+we have threads to create them.
 
@@
    \end{center}
-   \caption{Single Source Shortest Path program code.}
+   \caption{Creating splash trees for belief propagation. Each threads picks the
+   highest priority node and creates a tree from that node. The belief values
+   are updated in two phases: first from the leaves to the root and then from
+   the root to the leaves.}
    \label{splash_bp}
@@
 
-When we have several threads, every thread will generate their own trees by
-taking into account the highest priority node in their own queues. This is the
-reason we use \texttt{set-static} in line 1 to force all nodes to remain in the
-original threads.
+Note that the \texttt{set-static} action fact is used in line 1 in order to
+force all nodes to remain in the original threads, so that each thread has a
+well defined area of nodes to build splash trees on.
 
@@
          0 = count(Next, L),
-         *underline{priority(A, L, P)}, P > 0.0,
-         *underline{cpu-id(A, L, Id1)},
-         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
+         *underline{@priority(A, L, P)}, P > 0.0,
+         *underline{@cpu-id(A, L, Id1)},
+         *underline{@cpu-id(A, A, Id2)}, Id1 = Id2 |
          send-token(A, All, Next ++ L)].
@@
 \end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program.}
+  \caption{Coordination code for the Splash Belief Propagation program. When a
+     linear fact is prefixed by \texttt{@}, it indicates that the fact is going
+  to be re-derived.}
   \label{code:sbp}
@@
 patterns with only 12 simple logical rules.
+
+\begin{figure*}[]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
+   \end{center}
+   \caption{Memory usage and scalability of the regular and coordinated versions
+      of MiniMax.}
+   \label{results:splash_bp}
+\end{figure*}

### 2014-11-04 application.tex
More updates.

@@
 To better understand how coordination facts are used, we next present some programs that
-take advantage of them.
+take advantage of them. In our experimental setup, we we used a machine with
+four AMD Six-Core Opteron TM 8425 HE (2100 MHz) chips (24 cores) and 64 GB of
+DDR-2 667MHz (16x4GB) RAM, running GNU/Linux (kernel 3.15.10-201 64 bits).
+We compiled our virtual machine using GCC 4.8.3 (g++) with the flags
+\texttt{-O3 -std=c++11 -fno-rtti -march=x86-64}~\footnote{Implementation
+   available in \url{http://github.com/.../meld}}.
 

### 2014-11-04 experiments.tex
More updates.

\subsection{Absolute runtime}

The measures of absolute run time performance for the VM were done
~\cite{cruz-ppdp14}.
To put our VM in perspective, we first compare it in terms of absolute

execution time with other competing systems using a single thread.

In Table~\ref{tbl:comp_nqueens}, we compare LM's version of the
classic N-Queens puzzle against 3 other versions: a straightforward
sequential program implemented in C using backtracking; a sequential
Python implementation~\cite{vanRossum:1995:PRM}; and a Prolog
implementation executed in YAP
Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient
implementation of Prolog. Numbers less than 1 mean that LM is faster
and larger than 1 mean that LM is slower. We can observe that LM
easily beats Python, but is 5 to 10 times slower than YAP Prolog and
around 15 times slower than C. Note however that, as we will see next,
if we use at least 16 threads in LM, we can beat the sequential
implementation written in C.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{YAP Prolog} \\
\hline\hline
\textbf{10x10} & 16.92 & 0.62 & 5.42 \\
\textbf{11x11} & 21.59 & 0.64 & 6.47 \\
\textbf{12x12} & 10.32 & 0.73 & 7.61 \\
\textbf{13x13} & 14.35 & 0.88 & 10.38 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  N-Queens program}
\label{tbl:comp_nqueens}
\end{table}

In Table~\ref{tbl:comp_bp}, we compare LM's Belief Propagation (BP)
program, a machine learning algorithm to denoise images, against a
sequential C, Python and GraphLab~\cite{GraphLab2010} version of the
algorithm. GraphLab is a parallel C++ library used to solve
graph-based problems in machine learning. C and GraphLab perform about
the same since they are both compiled to machine code, although
the GraphLab version is highly optimized to run on multicore architectures.
Python runs very slowly since it is a
dynamic programming language and BP has many mathematical
computations. We should note, however, that LM's version uses some
external functions written in C++ in order to improve execution time,
therefore the comparison is not totally fair.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{GraphLab} \\
\hline\hline
\textbf{10}  & 1.00 & 0.03 & 1.00 \\
\textbf{50}  & 1.77 & 0.04 & 1.73 \\
\textbf{200} & 1.99 & 0.05 & 1.79 \\
\textbf{400} & 2.00 & 0.04 & 1.80 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  Belief Propagation program}
\label{tbl:comp_bp}
\end{table}

We also compared a LM's version of the PageRank program against a
similar GraphLab version and LM showed to be around 4 to 6 times
slower. Our worse results were obtained for the all-pairs shortest
distance algorithm where a LM's version of the problem was around 50
times slower than a C sequential implementation of the Dijkstra
algorithm, but almost twice as fast when compared to the same
implementation in Python.

### 2014-11-04 heat-transfer.tex
More updates.

@@
 Propagation~\cite{Gonzalez+al:aistats09paraml} (shown in the next section) and
-Heat Transfer.
+Heat Transfer~(HT).
+
+In HT, we have a graph of nodes that exchange heat values between them. The goal
+of the program is to exchange heat to a point where new heat value $\delta = |H_i -
+H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
+asynchronously since heat values can be updated by using new partial information
+coming from neighbor nodes. This increases parallelism since nodes do not need
+to synchronize if we had to compute heat values using iterations.
+
+Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
+neighbor nodes. We added the first rule to increase the priority of the neighbor
+nodes if the current node has a large $\delta$. The idea is to prioritize the
+computation of heat values of nodes (using \texttt{update}) that have a neighbor
+node that changed significantly. With \texttt{set-priority} the node is
+temporarily prioritized and selected to execute before others.
+
+\begin{figure}[h!]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\[\]]
+new-heat(A, New, Old),
+Delta = fabs(New - Old),
+Delta > epsilon * 2.0
+   -o {B, W | !edge(A, B, W) |
+         new-neighbor-heat(B, A, New),
+         update(B), *underline[set-priority(B, Delta)]}.
+
+new-heat(A, New, Old),
+Delta = fabs(New - Old)
+Delta < epsilon * 2.0,
+Delta > epsilon
+   -o {B, W | !edge(A, B, W) |
+         new-neighbor-heat(B, A, New), update(B)}.
+
+new-heat(A, New, Old)
+Delta = fabs(New - Old)
+Delta < epsilon
+   -o {B, W | !edge(A, B, W) |
+         new-neighbor-heat(B, A, New)}.
+\end{Verbatim}
+  \caption{Coordination code for the Heat Transfer program. We added one rule
+     that covers cases where $\Delta$ is at least $2 \epsilon$ in order to
+     increase the priority of neighbors nodes. The code logic remains exactly
+     the same as before, however bigger changes in heat values are now
+     propagated faster.}
+  \label{code:ht}
+\end{figure}
+\normalsize

### 2014-11-04 minimax.tex
More updates.

@@
 \end{figure}
+
+In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
+of the coordinated MiniMax against the regular MiniMax. The coordinated version
+sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
+the regular versio nneeds around 14GB. There is also some
+correlation between memory usage and number of threads. This is explained by the
+need for more data structures for handling multiple threads, specially
+concerning the memory allocator used by the virtual machine. In terms of run
+time, our experimental results show linear scalability for the coordinated
+version and a 20\% run time reduction when using 16 threads.

### 2014-11-04 splash-bp.tex
More updates.

@@
    is created that will navigate through the tree. Note that in the rule located
-   in lines 13-20, newly added nodes to the tree must have a positive priority
+   in lines 14-21, newly added nodes to the tree must have a positive priority
    (due to new belief updates) and must be residing in the same thread.
@@
    parallelism with 1 tree per thread.
-   \item[First phase]: In the third rule (lines 10-11), when the number of nodes
+   \item[First phase]: In the third rule (lines 11-12), when the number of nodes
    of the tree reaches a certain limit, a 
@@
    As the nodes are updated, an \texttt{update} fact is derived to update
-   the belief values (line 34).
+   the belief values (line 35).
    \item[Second phase]: In the second phase, the computation of beliefs is
    performed from the root to the leaves and the belief values are updated a
-   second time (line 41).
+   second time (line 42).
 \end{description}
 
-Note that the \texttt{set-static} action fact is used in line 1 in order to
-force all nodes to remain in the original threads, so that each thread has a
-well defined area of nodes to build splash trees on.
+Note that the \texttt{set-static} and \texttt{set-cpu} action fact are used in
+line 2 in order to (1) force nodes to stay in the thread and (2) to partition
+nodes as a grid of threads. This setups a well defined area of nodes
+for threads to build splash trees on.
 
@@
 \scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
-*underline{set-static(A)}. // all nodes are static.
+!coord(A, X, Y), start(A)
+   -o *underline{set-static(A), set-cpu(A, grid(X, Y))}.
 
@@
 
+\begin{figure*}[ht!]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
+      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
+   \end{center}
+   \caption{Experimental results for LBP and SBP. Figure (a) shows the
+      scalability of LBP for both LM and GraphLab and Figure (b) shows the
+      scalability of SBP. Figure (c) presents the
+      improvements seen in SBP against LBP, where SBP runs, on average, 1.5 to 2.5
+      times faster than LBP.}
+   \label{results:splash_bp}
+\end{figure*}
+
 In this program, coordination assumes a far more important role than we have
@@
 
-\begin{figure*}[]
-   \begin{center}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
-   \end{center}
-   \caption{Memory usage and scalability of the regular and coordinated versions
-      of MiniMax.}
-   \label{results:splash_bp}
-\end{figure*}
+We measured the behavior of LBP and SBP for both LM and GraphLab.
+Fig.~\ref{results:splash_bp} shows that both systems have very similar behavior
+when using a variable number of threads.
+Note that GraphLab avoids message passing between nodes since new neighbor
+belief updates are changed immediatelly, resulting in slightly better
+scalability for GraphLab.

### 2014-11-04 experiments.tex
Add heat transfer. Add more results.

@@
 
-The measures of absolute run time performance for the VM were done
-~\cite{cruz-ppdp14}.
-To put our VM in perspective, we first compare it in terms of absolute
+In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
+compared against implementations of similar programs but written in other
+languages. For instance, the LBP program was found to be around 2 times slower
+than GraphLab, while N Queens program was found to be around 10 to 15 slower
+than a sequential C implementation. Some of those programs were compared against
+a Python implementations and LM faired fairly well. In this section, we measure
+the impact of the coordination mechanisms we have implemented for the new
+virtual machine.
 
-execution time with other competing systems using a single thread.
+Fig.~\ref{results:comparison1}
 
-In Table~\ref{tbl:comp_nqueens}, we compare LM's version of the
-classic N-Queens puzzle against 3 other versions: a straightforward
-sequential program implemented in C using backtracking; a sequential
-Python implementation~\cite{vanRossum:1995:PRM}; and a Prolog
-implementation executed in YAP
-Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient
-implementation of Prolog. Numbers less than 1 mean that LM is faster
-and larger than 1 mean that LM is slower. We can observe that LM
-easily beats Python, but is 5 to 10 times slower than YAP Prolog and
-around 15 times slower than C. Note however that, as we will see next,
-if we use at least 16 threads in LM, we can beat the sequential
-implementation written in C.
-
-\begin{table}[ht]
-\centering
-{\begin{tabular}{c|c|c|c}
-\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
-\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{YAP Prolog} \\
-\hline\hline
-\textbf{10x10} & 16.92 & 0.62 & 5.42 \\
-\textbf{11x11} & 21.59 & 0.64 & 6.47 \\
-\textbf{12x12} & 10.32 & 0.73 & 7.61 \\
-\textbf{13x13} & 14.35 & 0.88 & 10.38 \\
-\end{tabular}}
-\caption{Comparing the absolute execution times (LM/System) for the
-  N-Queens program}
-\label{tbl:comp_nqueens}
-\end{table}
-
-In Table~\ref{tbl:comp_bp}, we compare LM's Belief Propagation (BP)
-program, a machine learning algorithm to denoise images, against a
-sequential C, Python and GraphLab~\cite{GraphLab2010} version of the
-algorithm. GraphLab is a parallel C++ library used to solve
-graph-based problems in machine learning. C and GraphLab perform about
-the same since they are both compiled to machine code, although
-the GraphLab version is highly optimized to run on multicore architectures.
-Python runs very slowly since it is a
-dynamic programming language and BP has many mathematical
-computations. We should note, however, that LM's version uses some
-external functions written in C++ in order to improve execution time,
-therefore the comparison is not totally fair.
-
-\begin{table}[ht]
-\centering
-{\begin{tabular}{c|c|c|c}
-\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
-\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{GraphLab} \\
-\hline\hline
-\textbf{10}  & 1.00 & 0.03 & 1.00 \\
-\textbf{50}  & 1.77 & 0.04 & 1.73 \\
-\textbf{200} & 1.99 & 0.05 & 1.79 \\
-\textbf{400} & 2.00 & 0.04 & 1.80 \\
-\end{tabular}}
-\caption{Comparing the absolute execution times (LM/System) for the
-  Belief Propagation program}
-\label{tbl:comp_bp}
-\end{table}
-
-We also compared a LM's version of the PageRank program against a
-similar GraphLab version and LM showed to be around 4 to 6 times
-slower. Our worse results were obtained for the all-pairs shortest
-distance algorithm where a LM's version of the problem was around 50
-times slower than a C sequential implementation of the Dijkstra
-algorithm, but almost twice as fast when compared to the same
-implementation in Python.
+\begin{figure}[h!]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
+   \end{center}
+   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
+   bars represent the performance of the original virtual machine, while
+   \textbf{New} is the relative performance of the new version that includes
+   coordination mechanisms. Most programs show little to no degradation in
+   performance.}
+   \label{results:comparison1}
+\end{figure}

### 2014-11-04 heat-transfer.tex
Add heat transfer. Add more results.

@@
 computation of heat values of nodes (using \texttt{update}) that have a neighbor
-node that changed significantly. With \texttt{set-priority} the node is
-temporarily prioritized and selected to execute before others.
+node that changed significantly. Multiple \texttt{add-priority} facts will
+increase the priority of a node so that nodes with multiple large deltas will
+have more priority. Note that the LM compiler will merge the three rules in
+Fig.~\ref{code:ht} into a single one that handles the 3 branches.
 
 \begin{figure}[h!]
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\[\]]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
 new-heat(A, New, Old),
@@
          new-neighbor-heat(B, A, New),
-         update(B), *underline[set-priority(B, Delta)]}.
+         update(B), \underline[add-priority(B, Delta)]}.
 
@@
 \normalsize
+
+Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
+and coordinated version. The dataset used is a square grid with an inner square
+with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
+for 16 threads, there's a 18\% reduction.
+
+\begin{figure}[ht!]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=5cm]{results/new-heat-transfer-80}}
+   \end{center}
+   \caption{Experimental results for the HT program. Speedup values for the
+      coordinated and uncoordinated version are computed using the run time of
+      the uncoordinated version using 1 thread.}
+   \label{results:ht}
+\end{figure}

### 2014-11-04 implementation.tex
More updates.
Add heat transfer. Add more results.

@@
 operations and executes procedures to iterate over the database in order to match facts.
+The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
+Here, we resume the most important details of that paper, while focusing on the
+changes for efficiently implementing coordination.
 
@@
 database, performing join operations and then consuming and deriving facts.
-
 Optimizations such as join optimizations (to allow rule filtering invalid
@@
 
-A node goes through different states during execution. The state machine in
-Fig.~\ref{fig:node_states} represents the valid state transitions with the
-following states:
+In order to facilitate the implementation of coordination, we added a state
+variable for each node. The state machine in
+Fig.~\ref{fig:node_states} represents the valid state transitions of a node:
 
@@
 Threads pop active nodes from their queues and execute new candidate rules. 
-Each thread has 2 queues, a doubly linked list known as the \emph{standard
+In order to allow node priorities, we use 2 queues per thread: a doubly linked
+list known as the \emph{standard
 queue} and a min/max heap known as the \emph{priority queue}. The regular queue

### 2014-11-05 abstract.tex
More corrections.

@@
-Declarative programming is good for parallel programming but makes it hard for
-the programmer to fine tune scheduling and data layout of the program.
-some programs require more information in order to execute better.
-with our approach we permit the programmer to write declarative programs
-but then optimize them later using sensing facts and action facts that
-understand and change how the programs are executed.
-
-Linear Meld is a concurrent forward-chaining linear logic programming
-language where logical facts can be asserted and retracted in a
-structured way. In Linear Meld, a program is seen as a database of
-logical facts and a set of derivation rules. The database of facts is
-partitioned by the nodes of a graph structure which leads to
+Declarative programming has been hailed as a promising approach to parallel
+programming since it hides the implementation details of parallelism away from the
+programmer. However, some control is lost when code is
+automatically parallelized, leaving the programmer with very little
+opportunities for fine tuning programs.
+We extended Linear Meld, a concurrent forward-chaining linear logic programming
+language, with coordination facts that allow the programmer to change
+computation scheduling and data layout. Our approach allows the programmer to
+write declarative parallel programs and then optionally use coordination facts
+to fine tune programs, while keeping the program fully declarative and amenable
+to correctness proofs. We show several application examples, experimental
+results and some advantages of our novel approach.

### 2014-11-05 coordination.tex
Add more text in MiniMax.
More corrections.

@@
 rules to allow the programmer to reason and change how execution is to be done.
-In this fashion, we keep the language declaratively because we reason logically
+In this fashion, we keep the language declarative because we reason logically
 about the state of execution, without the need to introduce extra-logical
-operators into the language that would reduce problems when attempting to prove
+operators into the language that would introduce problems when attempting to prove
 properties about programs.
 
-There are two kinds of coordination facts. The first kind of coordination facts are
-called \emph{sensing facts} and are used to sense information about data
-placement (where is the node being executed) or scheduled (what is the priority
-of the node). The second kind of coordination facts are \emph{action facts} that,
-when derived, are consumed to apply a coordination operation during execution.
-These logical rules that use coordination are implemented efficiently in the run
-time system and are hidden from the programmer.
+There are two classes of coordination facts. The first class of coordination facts are
+called \emph{sensing facts} and are used to sense information about the
+underlying runtime system, including node placement and node scheduling.
+The second kind of coordination facts are \emph{action facts} that are deleted
+to apply a coordination operation in the runtime system.
 
-\subsection{Scheduling Facts}
+\subsection{Scheduling Facts}\label{sec:fifo}
 
@@
 \emph{priorities}. Node priority works at the thread level
-so that each thread can make a local decision about which node of the graph to run next.
-Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).
+so that each thread can make a local decision about which node to execute next.
+Note that, by default, nodes are picked using a FIFO approach, because that
+tends to work better for most programs since older nodes tend to have more facts
+to be used.
 
 We have two kinds of priorities: a \emph{temporary priority} and a \emph{default
-priority}. A default priority is used indefinitely, while a temporary priority
-is only used once, so that when node with a temporary priority $P$ is picked to
-run, the priority will default back to $D$, the default priority. Initially,
-all nodes have a default priority of $0$.
+priority}. A temporary priority momentarily changes the default priority $D$ of a
+node, so that once the node is done, the priority will default back to $D$.
+Initially, all nodes have a default priority of $0$.
 
@@
    temporary priority of a node. If the node already has a priority, we only
-   change the priority if the new one is higher priority. The programmer can
+   change the priority if the new one is higher. The programmer can
    decide if priorities are to be ordered in ascending or descending order.
-   \item \texttt{type linear action add-priority(node, float)}: Increments,
-   temporarily, the current priority of the node.
+   \item \texttt{type linear action add-priority(node, float)}: Increases,
+   temporarily, the priority of the node.
    \item\texttt{type linear schedule-next(node)}: The work will fetch the
@@
 
-LM provides the sensing fact \texttt{type linear priority(node, node, float)},
-where the second argument is the target node and the third argument is the
-node priority. Sensing facts are only used in the body of the rules in order
+LM provides the sensing fact \texttt{priority(A, B, P)} in order to sense the
+priority \texttt{P} of node \texttt{B} from node \texttt{A}.
+Sensing facts are only used in the body of the rules in order
 to fetch information from the runtime system.
-Note that when sensing facts are consumed, they are re-derived automatically,
-except if the programmer explicitly tells the compiler otherwise. 
+Note that when sensing facts are deleted, they are re-derived automatically.
 Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
-value of \texttt{priority} facts, but this done seamlessly by the runtime
+value of \texttt{priority} facts, but this done automatically by the runtime
 system.
@@
    indicates the thread where the node of the second argument is currently running.
-   \item \texttt{type linear moving(node)}: Fact available if the node in the
+   \item \texttt{type linear moving(node, node)}: Fact available if the node in the
    second argument is allowed to move between threads.

### 2014-11-05 experiments.tex
More corrections.

@@
-\subsection{Absolute runtime}
-
 In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
@@
 
-Fig.~\ref{results:comparison1}
+Fig.~\ref{results:comparison1} shows the comparison between the original virtual
+machine and the new virtual machine with coordination mechanisms. All the
+programs benchmarked do not use any kind of coordination. As expected, there
+is no degration of performance by adding coordination mechanisms.
 
@@
    \textbf{New} is the relative performance of the new version that includes
-   coordination mechanisms. Most programs show little to no degradation in
-   performance.}
+   coordination mechanisms.}
    \label{results:comparison1}

### 2014-11-05 heat-transfer.tex
More corrections.

@@
-Randomized and approximation
-algorithms can obtain significant benefits from coordination directives because although the final
-program results will not be exact, they follow important statistical properties and can be computed faster.
-Examples of such programs are Loopy Belief
-Propagation~\cite{Gonzalez+al:aistats09paraml} (shown in the next section) and
-Heat Transfer~(HT).
-
-In HT, we have a graph of nodes that exchange heat values between them. The goal
-of the program is to exchange heat to a point where new heat value $\delta = |H_i -
+In the Heat Transfer algorithm~(HT), we have a graph where heat values are
+exchanged between nodes. The goal of the program is to exchange heat to a point
+where new heat value $\delta = |H_i -
 H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works

### 2014-11-05 implementation.tex
More corrections.

@@
 The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
-Here, we resume the most important details of that paper, while focusing on the
+Here, we review the most important details of that paper, while focusing on the
 changes for efficiently implementing coordination.
@@
 database, performing join operations and then consuming and deriving facts.
-Optimizations such as join optimizations (to allow rule filtering invalid
-combinations) and fact updates are implemented by the compiler. The latter
+Optimizations such as join optimizations (to support efficient rule filtering)
+and fact updates are implemented by the compiler. The latter
 transforms fact derivations of the same predicate to a simple update operation,
@@
 
-\subsection{Node partitioning}
-
-The compiler builds a representation of the graph by inspecting axioms of the
-program. Next, it orders the nodes of the graph using a breadth-first
-topological sort. This node ordering is written to the byte-code file and then
-used to initially partition nodes across available threads.
-
 \subsubsection{Coordination directives}
 
-Coordination directives are compiled in two different ways, depending whether they
+Coordination directives are compiled in two different ways, depending on whether they
 appear in the body or in the head of the rule. Coordination facts in the body
@@
 instead of returning something. While these are still facts from the point of
-view of the language, they are consumed immediatelly in order to apply the
+view of the language, they are deleted immediatelly in order to apply the
 operation, instead of being derived like any other regular fact. This is more
@@
 thread becomes idle, it attempts to steal nodes from a random thread into its
-own queues. If unsuccessful, the thread synchronizes with the other threads to
-become inactive.
+own queues. If sucessful, the thread becomes idle and waits for program
+termination while periodically attempting to steal nodes.
 
@@
 There is also an internal spinlock that is locked whenever the node enters into
-the \textbf{working} state. If a thread sends facts to a node placed in
-another thread, it attempts to lock the internal lock in order to update the
-indexing structures of the node, otherwise it adds the facts to the list of
+the \textbf{working} state. If a node sends facts to node is placed in
+another thread, the current thread first attempts to lock the internal lock in order to update the
+indexing structures of the target node, otherwise it adds the facts to the list of
 incoming facts that are later processed by the owner thread.

### 2014-11-05 intro.tex
More corrections.

@@
 languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
-Although more general declarative languages have yet to be as successful, the
+Although general declarative languages have yet to be as successful, the
 future looks promising for this particular approach.
@@
 In this paper, we extend Linear Meld (LM), a declarative language by
-\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer control over
-scheduling and data placement. LM is a linear logic programming language
-intended for programs over graph structures and with
-support for structured manipulation of mutable state using 
-linear logic~\cite{girard-87}. In LM, the nodes of the graph perform computation
-through derivation of rules and communicate with other nodes when logical facts
-are derived.
+\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer
+control over scheduling and data placement. LM is a linear logic programming
+language designed for programs that operate on graphs.  The use of linear
+logic~\cite{girard-87} allows support for structured manipulation of mutable
+state. In LM, computation is divided so that each node of the graph computes
+independently and is allowed to communicate with other nodes.
+Both computation and communication happen through derivation of logical rules.
 
-The operational semantics of LM allow for plenty of
-non-determinism since nodes execute independently and can be executed with
-any specific order. Since LM uses mutable state, the order in which nodes are
-executed may be crucial in terms of performance. We extended the language with
-coordination facts that can be used as a regular fact in the program logic to
-improve program execution.
-The first kind of coordination facts are
-called \emph{sensing facts} and are used to sense information about data
-placement (where is the node being executed). This allows the programmer to
-write logical rules that depend on the current state of the program. The second
-kind of coordination facts are \emph{action facts} that, when deriving, will be
-consumed to apply a scheduling operation during execution. This allows the
-programmer to prioritize node computation or place nodes in different threads.
-Both kinds of facts are \emph{linear facts} since they are consumed and/or change.
+Given that LM uses mutable state, the order in which nodes are executed may be
+crucial in terms of performance. We extended the language with coordination
+facts that can be used as a regular fact in the program logic to improve program
+execution.  The first kind of coordination facts are called \emph{sensing facts}
+and are used to sense information about the system the program is running on.
+This allows the programmer to write logical rules that depend on the current
+state of the program and the system. The second kind of coordination facts are
+\emph{action facts} that when deleted in rules are used to apply a scheduling
+operation during execution. This allows the programmer to prioritize node
+computation or place nodes in different threads.
 
@@
 control over execution without a significant change in the language itself and
-without using non-declarative operations. This proves specially important since
+without using non-declarative operations. This is specially important since
 programs can proven to be correct even in the presence of coordination facts.
-Our other two main contributions are as follows: a novel way to coordinate
-declarative programs through user control that introduce measurable performance
-improvements; several application where we add programmer control to declarative
-programs in order to make them perform better.
+In order to prove our thesis, we first give a brief overview of the
+base language, including a small example. Next, in
+Section~\ref{sec:coordination} we introduce the available coordination facts,
+followed by Section~\ref{sec:implementation}, where we describe the
+implementation of the runtime required to implement the desired coordination
+mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
+programmer control to declarative programs in order to make them perform better.

### 2014-11-05 language.tex
More corrections.

@@
 language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
-Programs are written as a set of rules of the form
-\texttt{a(X), b(Y) -o c(Z)} and are read as following: if \texttt{a(X), b(Y)} (the body of
-the rule) is satisfied using some facts \texttt{a(X)} and \texttt{b(Y)} then
-\texttt{c(Z)} (the head of the rule) is true.
-The facts used during rule derivation are placed in a \emph{database of facts}.
-Logical facts are composed of a predicate and a tuple of concrete values,
-representing the arguments.
-Since LM uses linear logic as its foundation, rule derivations
-may delete facts used in the body of the rule. Program execution starts by
-adding the \emph{axioms} (the initial facts) of the program to the database.
+Programs consist of a set of \emph{rules} and a \emph{database of facts}.
+Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
+and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
+the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
+the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
+A fact is a predicate and its associated tuple of values, representing the
+arguments. Since LM uses linear logic as its foundation, we distinguish between
+\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
+process of deriving rule, while persistent facts are not.
+Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
 Next, the rules are recursively applied and the database is updated with new and
-deleted facts. When nore more rules are applicable, the program terminates.
+deleted facts. When no more rules are applicable, the program terminates.
 
-LM has been designed for writing graph-based programs. LM accomplishes this by
-partitioning the logical facts by the first argument (the \emph{node}) and by
-restricting the body of every rule to only refer to the same first argument.
-This allows rules to be executed using only facts of a single node. Although the
-body of rules is restricted, the head of the rule may refer to any other node as
-long as the node variable is refered somewhere in the body of the rule. This allows
-derivations of facts in other nodes. The whole database is then composed of smaller
-databases (the nodes) and rule derivations allows smaller databases to derive
-facts in other databases, creating edges in the graph.
+LM has been designed for writing programs that operate on graphs. LM partitions
+the database by using the first argument of each fact. The first argument is
+typed as a \emph{node} and represents a node in the graph. For instance, the
+fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
+is stored in node \texttt{@2}. LM restricts
+the body of every rule to only refer to the same node so that nodes can derive
+rules independently. Although the body is restricted, the head of the
+rule may refer to any node as long as the node is refered somewhere in the
+body. This allows communication between nodes during rule derivation, since a
+node must send a fact to another node.
+Rule restrictions in turn make LM implicitly parallel because nodes are able to
+compute independently as long as they have new facts to be applied in rules.
+This makes LM non-deterministic since nodes can be picked to run in any order,
+affecting which rules are applied and which facts are deleted or derived.
 
@@
 with the declaration of the predicates. Predicates specify the kinds of facts
-used in the program. The first predicate, \texttt{edge}, specifies edges between
-nodes of the graph, where the third argument represents the weight of the edge.
-The \texttt{route} modifier informs the compiler about the structure of the
-program graph. The predicates \texttt{shortest} and \texttt{relax} are specified
-as linear facts and thus are deleted when used in rules.
+used in the program. The first predicate, \texttt{edge}, is a persistent
+predicate that describes the
+relationship between the nodes of the graph, where the third argument represents
+the weight of the edge
+(the \texttt{route} modifier informs the compiler about the structure of the
+program graph). The predicates \texttt{shortest} and \texttt{relax} are specified
+as linear facts and thus are deleted when deriving new facts.
 The idea of the algorithm is to compute the shortest distance from node
@@
   \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
+  \hspace{0.4cm}
   \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
+  \hspace{0.4cm}
   \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}

### 2014-11-05 minimax.tex
More corrections.

--- 2014-11-02 minimax.tex
@@
 evaluation function \texttt{f(State)} is used to compute the score of
-the board for each leaf of the tree. A node is a leaf when it can no longer be
-expanded since the state represents a possible end of the game.  Finally, the
+the board for each leaf of the tree. A node is a leaf when the game state can no longer be
+expanded. Finally, the
 algorithm recursively minimizes or maximizes the scores of each node.
@@
 that is expanded with the available moves at each level. The graph of the
-program is dynamic since nodes are created and eventually deleted once they
-have nore more facts or no more references to them. The latter happens when the
-leaves scores are computed or when a node fully minimizes or maximizes the
-children scores. When the program ends, only the root node is have facts in its
+program is dynamic since nodes are created and then deleted once they are no
+longer needed. The latter happens when the
+leaf scores are computed or when a node fully minimizes or maximizes the
+children scores. When the program ends, only the root node has facts in its
 database.
@@
 (12-29) deal with cases where new nodes may be created.
-Thes rule in lines 12-25 generate new nodes using the
+The rule in lines 12-25 generate new nodes using the
 \texttt{exists} language construct, which creates a new node address
@@
 
-Since our implementation
-uses a FIFO ordering for nodes, meaning that when children are expanded, we
-first expand each children and then the children of the children. The default
-ordering of the system works well enough for most practical purposes. We note
+As noted before in Section~\ref{sec:fifo}, we use a FIFO ordering for nodes,
+meaning that when children are expanded, we first expand the children and
+then the children of the children. We note
 however, that this is not optimal for this program since the complete tree needs
-to be expanded before computing the scores at the leaves. This introduces memory
-issues since, in the worst case, the space complexity of the program are
-$\mathcal{O}(n)$, where $n$ is the number of nodes in the tree.
+to be expanded before computing the scores at the leaves. The memory
+complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
+in the tree.
 
@@
 expand(A, Board, [], N, P, \underline{Depth}),
-N > 0
+N > 0, P = player1
   -o maximize(A, N, -00, 0).
@@
 expand(A, Board, [], N, P, \underline{Depth}),
-N > 0, NextPlayer <> RootPlayer
+N > 0, P = player2
   -o minimize(A, N, +00, 0).
@@
 Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
-expanded the tree on the left side. When a second thread comes up, it steals the
+expanded the tree on the left side. When a second thread is idle, it steals the
 first unexpanded node and expands the tree from that position. Since threads
@@
    \begin{center}
-      \subfloat[]{ \begin{tabular}[b]{ | c | c | c |}
+      \subfloat[]{\small\begin{tabular}[b]{ | c | c | c |}
          \hline                       
@@
       }
-      \subfloat[]{\includegraphics[width=5cm]{results/min-max-tictactoe.png}}
+      \subfloat[]{\includegraphics[width=4.5cm]{results/min-max-tictactoe.png}}
    \end{center}
@@
 \end{figure}
+
+In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
+of the coordinated MiniMax against the regular MiniMax. The coordinated version
+sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
+the regular version needs around 14GB. Note that as the number of threads goes
+up, memory usage also goes up. This is an artifact of our parallel memory
+allocator that allocates large chunks of memory before hand.
+In terms of run time, our experimental results show linear scalability for the coordinated
+version and a 20\% run time reduction when using 16 threads.

### 2014-11-05 related.tex
More corrections.

Recently, there has been an increasing interest in declarative and data-centric
languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
popular data-centric programming model that is optimized for large clusters. The
scheduling and data sharing model is very simple: in the \emph{map phase}, data
is transformed at each node and the result reduced to a final result in the
\emph{reduce phase}. Like LM, many programming systems model the program as a
graph where computation will be performed. The Dryad
system~\cite{Isard:2007:DDD:1272996.1273005} combines computational vertices
with communication channels (edges) to form a data-flow graph. The program is
scheduled to run on multiple computers or cores and data is partitioned
automatically during runtime.

Many programming languages follow the so-called \emph{coordination
paradigm}~\cite{Papadopoulos98coordinationmodels}, a form of distributed
programming that divides execution in two parts: \emph{computation}, where the actual
computation is performed, and \emph{coordination}, which deals with
communication and cooperation between processing units. This paradigm attempts
to clearly distinguish between these two parts by providing abstractions for
coordination in an attempt to provide architecture and system-independent forms
of communication.

Linda~\cite{linda} is probably the most famous coordination model. Linda
implements a data-driven coordination model and features a \emph{tuple space}
that can be manipulated using the following coordination directives:
\texttt{out(t)} writes a tuple \texttt{t} into the tuple space; \texttt{in(t)}
reads a tuple using the template \texttt{t}; \texttt{rd(t)} retrieves a copy of
the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} puts a process
\texttt{p} in the tuple space and executes it in parallel. 
Linda is be implemented on top of many
popular languages by simply creating a communication and storage mechanism for
the tuple space and then adding the directives as a language library.

Another coordination language is Delirium~\cite{Delirium}, where instead of being
embbeded in another language like Linda, it actually embeds operators other
languages inside Delirium. The advantages of Delirium are improved abstraction
and easier debugging because sequential operators are isolated from the
coordination language.

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. LM also differs from those languages since it
raises the abstraction level from the processing units to nodes of a graph,
which are part of the program logic.
Furthermore, the language LM is both the coordination and the computation
language and there is no distinction between them.

\iffalse
GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
learning algorithms. GraphLab allows nodes to
have read/write access to different scopes through different concurrent access
models in order to balance performance and data consistency. While some programs
only need to access the local node's data, others may need to update edge
information. Each consistency model will provide different guarantees that are
better adapted to some algorithms. GraphLab provides different schedulers
that dictate the order in which node's are computed, which is a rudimentary form
of coordination. Later in this paper, we will show how certain GraphLab's
schedulers can be easily implemented in LM through the use of coordination
facts.
\fi

### 2014-11-05 splash-bp.tex
More corrections.

@@
+Randomized and approximation algorithms can obtain significant benefits from
+coordination directives because although the final program results will not be
+exact, they follow important statistical properties and can be computed faster.
+Examples of such programs are Loopy Belief Propagation.
 Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
@@
 \end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program. When a
-     linear fact is prefixed by \texttt{@}, it indicates that the fact is going
-  to be re-derived.}
+  \caption{Coordination code for the Splash Belief Propagation program. LM needs
+     50 lines of rules to implement splash trees, while GraphLab needs a
+     total of 350 lines of C++ to implement the same functionality.  Note
+     that when a linear fact is prefixed by \texttt{@}, it indicates that the
+     fact is going to be re-derived.}
   \label{code:sbp}

### 2014-11-05 minimax.tex
Small changes.

@@
 up, memory usage also goes up. This is an artifact of our parallel memory
-allocator that allocates large chunks of memory before hand.
-In terms of run time, our experimental results show linear scalability for the coordinated
-version and a 20\% run time reduction when using 16 threads.
+allocator that allocates large chunks of memory beforehand.
+In terms of run time, our experimental results show a 14-fold speedup for the
+coordinated version against a 12-fold for the regular version when using 16
+threads. When comparing the two versions directly, there is a 20\% run time
+reduction when using 16 threads in the coordinated version.

### 2014-11-05 sssp.tex
More corrections.
Small changes.

@@
 threads. In turn, this makes the coordinated version 20 times faster
-than the regular version using 1 thread.
+than the regular version using 1 thread. Note that the coordinated program is as
+scalable as the regular program.
 
@@
       }
-      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
+      \subfloat[]{\includegraphics[width=4.5cm]{results/shortest-uspowergrid.png}}
    \end{center}
@@
    and coordinated version \textbf{C} using a variable number of threads
-   \textbf{\# T}. On the right, we have the scalability of the regular version
-   and coordinated version. The speedup values are computed using the execution
-   time of the regular version for 1 thread.}
+   \textbf{\# T}. On the right, we have the scalability of the regular and
+   coordinated version. The speedup values are computed using the execution
+   time for 1 thread.}
    \label{results:sssp_uspowergrid}

### 2014-11-05 related.tex
Improve related work.

@@
-Recently, there has been an increasing interest in declarative and data-centric
+There has been an increasing interest in declarative and data-centric
 languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
@@
 is transformed at each node and the result reduced to a final result in the
-\emph{reduce phase}. Like LM, many programming systems model the program as a
-graph where computation will be performed. The Dryad
-system~\cite{Isard:2007:DDD:1272996.1273005} combines computational vertices
+\emph{reduce phase}. In order to facilitate the writing of programs over large
+datasets, SQL-like languages such as
+PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
+PigLatin automatically compiles declarative code into physical plans
+that are executed over Hadoop but provides very little control to the
+programmer.
+
+A declarative approach that is regaining popularity is
+Datalog~\cite{Ullman:1990:PDK:533142}, also a forward-chaining logic programming
+language. Traditionally used in deductive databases, it is now being increasingly used in different fields
+such as distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
+nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
+LM itself is inspired in P2~\cite{Loo-condie-garofalakis-p2} since it also models
+programs as a graph structure. Another programming system that uses the graph
+abstraction is Dryad~\cite{Isard:2007:DDD:1272996.1273005}, which combines computational vertices
 with communication channels (edges) to form a data-flow graph. The program is
-scheduled to run on multiple computers or cores and data is partitioned
+then scheduled to run on multiple computers/cores and data is partitioned
 automatically during runtime.
@@
 
-Linda~\cite{linda} is probably the most famous coordination model. Linda
+Linda~\cite{linda} is arguably the most famous coordination model. Linda
 implements a data-driven coordination model and features a \emph{tuple space}
 that can be manipulated using the following coordination directives:
-\texttt{out(t)} writes a tuple \texttt{t} into the tuple space; \texttt{in(t)}
-reads a tuple using the template \texttt{t}; \texttt{rd(t)} retrieves a copy of
-the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} puts a process
-\texttt{p} in the tuple space and executes it in parallel. 
+\texttt{out(t)} to write a tuple \texttt{t} into the tuple space; \texttt{in(t)}
+to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy of
+the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a process
+\texttt{p} in the tuple space and execute it in parallel. 
 Linda is be implemented on top of many
@@
 
-Another coordination language is Delirium~\cite{Delirium}, where instead of being
-embbeded in another language like Linda, it actually embeds operators other
-languages inside Delirium. The advantages of Delirium are improved abstraction
+Another important coordination language is Delirium~\cite{Delirium}. Unlike
+Linda which is embbeded into another language, Delirium actually embeds
+operators written in other languages inside the Delirium language.
+The advantages of Delirium are improved abstraction
 and easier debugging because sequential operators are isolated from the
@@
 left to the implementation. LM also differs from those languages since it
-raises the abstraction level from the processing units to nodes of a graph,
-which are part of the program logic.
-Furthermore, the language LM is both the coordination and the computation
-language and there is no distinction between them.
+raises the abstraction level from the processing units to nodes of a graph.
+Furthermore, the LM language is both a coordination language and a computation
+language and there is no distinction between the two components.
 

### 2014-11-05 conclusions.tex
New update.

We think that our particular approach is applicable in distributed systems as we
have seen. Explore functional programming and coordination?

### 2014-11-05 nqueens.tex
New update.

The N-Queens puzzle is the problem of placing N chess queens on an NxN
chessboard so that no pair of two queens attack each
other~\cite{8queens}. The challenge of finding all the
distinct solutions to this problem is a good benchmark in designing
parallel algorithms.

The LM solution considers the chess board as a graph where nodes communicate
with each other to build valid configurations. First, we start with the empty
state in the first row of the chess board. Then, each square adds its own
position to the state and sends the state down, to the next row. Once a square
receives new configurations, it attempts to add its position to the
configurations. If valid, that configuration is then sent, recursively, to the
next row, until all rows are traversed. At the end of the program, the nodes at
the bottom row will have all the valid configurations.

Since computation goes from the top row to the bottom row, not all placements of
nodes on threads may be equally performant. This is specially true because the
bottom rows tend to perform most work. The best placement is then to split the
board vertically, so that, in optimal conditions, each thread gets the same
number of columns.  Our experiments show that, on shared memory, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by improving load balancing dynamically. But what if we want run the
program on a distributed system, where moving nodes between workers can be
expensive? We measured the performance of the program without node stealing in
order to assess how important data placement would be for such systems.
We used the default partition of the program as it is, which results in the board being
split by rows and another configuration using the coordination action
\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
the coordinates of the node in the chessboard.

Figure~\ref{results:nqueens} shows the performance results of coordinating the
placing of nodes on threads without work stealing. There is a noticiable
performance improvement when using an optimal data placement strategy. Note
however that in the plot we do not take into account the costs of moving
nodes between workers. In a real system, the performance gap may not be as wide
since in the worst case there may be 144 moves.

\begin{figure}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/13queens.png}}
   \end{center}
   \caption{Experimental results for the 13 Queens program. The coordinated
      version is on average 2 times faster than the version with bad data
      placement.}
   \label{results:nqueens}
\end{figure}

### 2014-11-06 conclusions.tex
More corrections.

@@
+We have presented a way of adding coordination to a declarative language without
+change the nature of the language or introducing non-declarative constructs. We
+took advantage of the fact that LM uses linear logic, which allows us to derive
+coordination facts to delete facts in order to perform actions in the underlying
+runtime system.
+
+To the best of our knowledge, this is the first time that a declarative language allows 
+control over execution without a significant change in the language itself and
+without using non-declarative operations. This is specially important since
+programs can be proven to be correct even in the presence of coordination facts.
+In order to prove our thesis, we first give a brief overview of the
 We think that our particular approach is applicable in distributed systems as we

### 2014-11-06 intro.tex
More updates.
More corrections.

@@
 
-We argue, however, that declarative programming leaves little to no programmer control
-over how execution is scheduled or how data is laid out. This introduces some
+The problem with declarative programming is that it leaves little to no programmer control
+over how execution is scheduled or how data is laid out, making it hard to achieve
+necessary efficiency. This introduces some
 performance issues because even if the runtime system is able to reasonably
@@
 In this paper, we extend Linear Meld (LM), a declarative language by
-\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer
+\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
 control over scheduling and data placement. LM is a linear logic programming
@@
 execution.  The first kind of coordination facts are called \emph{sensing facts}
-and are used to sense information about the system the program is running on.
-This allows the programmer to write logical rules that depend on the current
-state of the program and the system. The second kind of coordination facts are
+and are used to sense information about the system the program is running on,
+e.g., scheduling and node placement on threads. The second kind of coordination facts are
 \emph{action facts} that when deleted in rules are used to apply a scheduling
-operation during execution. This allows the programmer to prioritize node
+operation during execution. Coordination facts allow the programmer to write
+logical rules that depend on the current state of the program and then prioritize node
 computation or place nodes in different threads.
@@
 without using non-declarative operations. This is specially important since
-programs can proven to be correct even in the presence of coordination facts.
+programs can be proven to be correct even in the presence of coordination facts.
 In order to prove our thesis, we first give a brief overview of the

### 2014-11-06 language.tex
Small changes.
More corrections.

@@
 the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
-A fact is a predicate and its associated tuple of values, representing the
+A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
+associated tuple of values, e.g., the values of \texttt{X} and \texttt{Y}, representing the
 arguments. Since LM uses linear logic as its foundation, we distinguish between
@@
 Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
-Next, the rules are recursively applied and the database is updated with new and
-deleted facts. When no more rules are applicable, the program terminates.
+Next, the rules are recursively applied and the database is updated by adding
+new facts or deleting facts used during rule derivation.
+When no more rules are applicable, the program terminates.
+
+\newcommand{\scare}[1]{``#1''}
 
@@
 rule may refer to any node as long as the node is refered somewhere in the
-body. This allows communication between nodes during rule derivation, since a
-node must send a fact to another node.
+body. This allows \scare{communication} between nodes during rule derivation, since a
+node may \scare{send} a fact to another node.
 Rule restrictions in turn make LM implicitly parallel because nodes are able to
-compute independently as long as they have new facts to be applied in rules.
-This makes LM non-deterministic since nodes can be picked to run in any order,
-affecting which rules are applied and which facts are deleted or derived.
+compute independently. This makes LM non-deterministic since nodes can be picked
+to run in any order, affecting which rules are applied and which facts are
+deleted or derived.
 
@@
 used in the program. The first predicate, \texttt{edge}, is a persistent
-predicate that describes the
-relationship between the nodes of the graph, where the third argument represents
-the weight of the edge
-(the \texttt{route} modifier informs the compiler about the structure of the
-program graph). The predicates \texttt{shortest} and \texttt{relax} are specified
-as linear facts and thus are deleted when deriving new facts.
-The idea of the algorithm is to compute the shortest distance from node
-\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
-fact that is improved with new \texttt{relax} facts.
-Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
-graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
-for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
-initial distance from \texttt{@1} to \texttt{@1}.
+predicate that describes the relationship between the nodes of the graph,
+where the third argument represents the weight of the edge (the
+\texttt{route} modifier informs the compiler that the \texttt{edge} predicate
+determines the structure of the graph).
+The predicates \texttt{shortest} and \texttt{relax} are
+specified as linear facts and thus are deleted when deriving new facts.  The
+algorithm computes the shortest distance from node
+\texttt{@1} to all other nodes in the graph. Every node has a
+\texttt{shortest} fact that is improved with new \texttt{relax} facts.  Lines
+5-9 declare the axioms of the program: \texttt{edge} facts describe the
+graph; \texttt{shortest(A, +00, [])} is the initial shortest distance
+(infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
+algorithm with the initial distance from \texttt{@1} to \texttt{@1}.
 
@@
 
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 < D2
-   -o shortest(A, D1, P1),
+shortest(A, D1, P1), D1 > D2, relax(A, D2, P2)
+   -o shortest(A, D2, P2),
       {B, W | !edge(A, B, W) |
-         relax(B, D1 + W, P1 ++ [B])}.
+         relax(B, D2 + W, P2 ++ [B])}.
 
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 >= D2
-   -o path(A, D2, P2).
+shortest(A, D1, P1), D1 <= D2, relax(A, D2, P2)
+   -o shortest(A, D1, P1).
 \end{Verbatim}
@@
 
-The first rule of the program (lines 11-15) replaces the current path in
-\texttt{shortest} with a shorter one in \texttt{relax}. The rule deletes both
-\texttt{relax} and \texttt{shortest} facts and derives a new \texttt{shortest}
-fact. In lines 14-15, we have a \emph{comprehension} where the program iterates
-over the edges of node \texttt{A} and derives a new \texttt{relax} fact at
-\texttt{B} with the new shorter distance plus the weight of the edge. For
+The first rule of the program (lines 11-14) reads as following: if the current
+\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
+than a new path \texttt{relax} with distance \texttt{D2}, then replace the
+current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
+propagate new paths to the neighbors (lines 14-15) using a \emph{comprehension}.
+The comprehension iterates over the edges of node \texttt{A} and derives a new
+\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
+where \texttt{W} is the weight of the edge. For
 instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
@@
 
-The second rule of the program (lines 17-19) deals with the case where the new
-distance in \texttt{relax} is not better than the current one, so it is thrown
-away and the current distance is kept.
+The second rule of the program (lines 16-17) is read as following: if the
+current shortest path \texttt{D1} is shorter than the new path \texttt{D2} then
+delete the new \texttt{relax} fact and keep the old shortest path.
 

### 2014-11-06 nqueens.tex
More corrections.

@@
 
-The LM solution considers the chess board as a graph where nodes communicate
-with each other to build valid configurations. First, we start with the empty
+The LM solution considers the chess board as a graph where nodes exchange valid
+configurations between each other. First, we start with the empty
 state in the first row of the chess board. Then, each square adds its own
@@
 nodes on threads may be equally performant. This is specially true because the
-bottom rows tend to perform most work. The best placement is then to split the
+bottom rows tend to perform the most work. The best placement is then to split the
 board vertically, so that, in optimal conditions, each thread gets the same

### 2014-11-06 related.tex
More corrections.

@@
 PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
-PigLatin automatically compiles declarative code into physical plans
-that are executed over Hadoop but provides very little control to the
-programmer.
-
-A declarative approach that is regaining popularity is
-Datalog~\cite{Ullman:1990:PDK:533142}, also a forward-chaining logic programming
-language. Traditionally used in deductive databases, it is now being increasingly used in different fields
-such as distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
-nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
-LM itself is inspired in P2~\cite{Loo-condie-garofalakis-p2} since it also models
-programs as a graph structure. Another programming system that uses the graph
-abstraction is Dryad~\cite{Isard:2007:DDD:1272996.1273005}, which combines computational vertices
-with communication channels (edges) to form a data-flow graph. The program is
-then scheduled to run on multiple computers/cores and data is partitioned
-automatically during runtime.
+PigLatin builds on top of MapReduce and allows the programmer to write complex
+data-flow graphs, increasing the abstraction of the MapReduce paradigm.
+An alternative to PigLatin/MapReduce is
+Dryad~\cite{Isard:2007:DDD:1272996.1273005}
+that allows programmers to design arbitrary computation patterns using the DAG
+abstraction. It combines computational vertices with communication channels
+(edges). Dryad programs are automatically scheduled to run on multiple
+computers/cores and data is partitioned automatically during runtime.
 
@@
 
-Another important coordination language is Delirium~\cite{Delirium}. Unlike
+Another early coordination language is Delirium~\cite{Delirium}. Unlike
 Linda which is embbeded into another language, Delirium actually embeds
@@
 coordinate the scheduling of processing units, while the placement of data is
-left to the implementation. LM also differs from those languages since it
-raises the abstraction level from the processing units to nodes of a graph.
+left to the implementation. LM differs from those languages since it
+is fully data-driven and coordination acts on data instead of processing units.
+The abstraction is then raised by considering data and algorithmic aspects of
+the program instead of focusing how processing units are used.
 Furthermore, the LM language is both a coordination language and a computation
@@
 
-\iffalse
-GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
-learning algorithms. GraphLab allows nodes to
-have read/write access to different scopes through different concurrent access
-models in order to balance performance and data consistency. While some programs
-only need to access the local node's data, others may need to update edge
-information. Each consistency model will provide different guarantees that are
-better adapted to some algorithms. GraphLab provides different schedulers
-that dictate the order in which node's are computed, which is a rudimentary form
-of coordination. Later in this paper, we will show how certain GraphLab's
-schedulers can be easily implemented in LM through the use of coordination
-facts.
-\fi
+Forward-chaining logic programming was made famous by the Datalog
+language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
+databases, it is now being increasingly used in different fields such as
+distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
+nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
+LM itself is inspired in Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
+programming distributed emsembles of modular robots. Meld introduced the idea of
+sensing and action facts in order to sense and act on the outside world,
+respectively. Coordination facts are inspired in the same principles but are
+used for coordinating data and computation.

### 2014-11-06 sssp.tex
More corrections.

@@
 algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
-uses one coordination fact, namely \texttt{set-priority} (line 15). Note that we
+uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
 also use a global program directive to order priorities in ascending order (line
@@
 
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 < D2
-   -o shortest(A, D1, P1),
+shortest(A, D1, P1), D1 > D2, relax(A, D2, P2)
+   -o shortest(A, D2, P2),
       \{B, W | !edge(A, B, W) |
-         relax(B, D1 + W, P1 ++ [B]),
-         \underline{set-priority(B, float(D1 + W))}\}.
+         relax(B, D2 + W, P2 ++ [B]),
+         \underline{set-priority(B, float(D2 + W))}\}.
 
-relax(A, D1, P1), shortest(A, D2, P2),
-D1 >= D2
-   -o shortest(A, D2, P2).
+shortest(A, D1, P1), D1 <= D2, relax(A, D2, P2)
+   -o shortest(A, D1, P1).
 \end{Verbatim}

### 2014-11-06 conclusions.tex
More improvements.

--- 2014-11-05 conclusions.tex
@@
-We think that our particular approach is applicable in distributed systems as we
-have seen. Explore functional programming and coordination?
+We have presented a novel way of adding coordination to a declarative language without
+changing the nature of the language or introducing non-declarative constructs. We
+took advantage of the fact that LM uses linear logic, which allows us to derive
+coordination facts that can be deleted in order to perform actions in the underlying
+runtime system. In the other hand, sensing facts allow the programmer to reason
+about locality and scheduling of computation in a data-driven fashion.
+We have presented several applications where a judicious use of coordination can lead to
+better performance behavior.
+
+We would like to explore this coordination paradigm in distributed systems,
+where data locality is far more important than in shared memory systems. We also
+think that LM may be useful as a glue distributed/parallel language since it can
+easily model any computation pattern using graphs. Furthermore, we think that
+other declarative paradigms would benefit from using a similar approach. For
+instance, in functional programming, one may consider annotation functions that
+pick the right granularity for each problem.

### 2014-11-06 implementation.tex
More improvements.

@@
 usage.
+
+\subsubsection{Coordination overhead}
+
+In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
+compared against implementations of similar programs but written in other
+languages. For instance, the LBP program was found to be around 2 times slower
+than GraphLab, while N Queens program was found to be around 10 to 15 slower
+than a sequential C implementation. Some of those programs were compared against
+a Python implementations and LM faired fairly well. In this section, we measure
+the impact of the coordination mechanisms we have implemented for the new
+virtual machine.
+
+Fig.~\ref{results:comparison1} shows the comparison between the original virtual
+machine and the new virtual machine with coordination mechanisms. All the
+programs benchmarked do not use any kind of coordination. As expected, there
+is no degration of performance by adding coordination mechanisms.
+
+\begin{figure}[h!]
+   \begin{center}
+      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
+   \end{center}
+   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
+   bars represent the performance of the original virtual machine, while
+   \textbf{New} is the relative performance of the new version that includes
+   coordination mechanisms.}
+   \label{results:comparison1}
+\end{figure}

### 2014-11-06 splash-bp.tex
More improvements.

@@
 \begin{description}
-   \item[Tree building]: Each node has a \texttt{waiting} fact that is used to
-   start the tree building process. When the highest priority is picked, a token
-   is created that will navigate through the tree. Note that in the rule located
-   in lines 14-21, newly added nodes to the tree must have a positive priority
-   (due to new belief updates) and must be residing in the same thread.
-   We want the tree to be kept in the same thread in order to maximize
-   parallelism with 1 tree per thread.
-   \item[First phase]: In the third rule (lines 11-12), when the number of nodes
-   of the tree reaches a certain limit, a 
-   \texttt{first-phase} is generated in order to update the beliefs of all nodes
-   in the tree, starting from the leaves and ending at the root
-   As the nodes are updated, an \texttt{update} fact is derived to update
-   the belief values (line 35).
+   \item[Tree building]: Each node has a \texttt{inactive} fact that is used to
+   start the tree building process. When the highest priority node is picked, a
+   \texttt{tree} is created that will navigate through the tree. In lines 15-21,
+   we use an \emph{aggregate}~\cite{cruz-iclp14} to gather all the neighbor
+   nodes that have a positive priority (due to a new belief update) and are in the
+   same thread. Nodes are collected into list \texttt{L} (line 21) and
+   appended to list \texttt{Next}.
+
+   \item[First phase]: In the
+   third rule (lines 11-12), when the number of nodes of the tree reaches a
+   certain limit, a \texttt{first-phase} is generated to update the beliefs of
+   all nodes in the tree, starting from the leaves and ending at the root As the
+   nodes are updated, an \texttt{update} fact is derived to update the belief
+   values (line 35).
+
    \item[Second phase]: In the second phase, the computation of beliefs is
@@
 // TREE BUILDING
-// continue tree
-waiting(A), token(A, All, Next) -o token(A, All, Next).
-// start tree
-waiting(A), *underline{@priority(A, A, P)}, P > 0.0
-   -o token(A, [A], [A]).
+// expand tree by adding neighbor nodes
+inactive(A), tree(A, All, Next) -o expand-tree(A, All, Next).
+// start tree since we do not have one
+inactive(A), *underline{@priority(A, A, P)}, P > 0.0
+   -o expand-tree(A, [A], [A]).
 // end tree building
-token(A, All, Next), length(All) > maxnodes
+expand-tree(A, All, Next), length(All) >= maxnodes
    -o first-phase(A, All, reverse(All)).
 // expand tree
-token(A, All, [A | Next])
+expand-tree(A, All, [A | Next]), length([A | Next]) < maxnodes
    -o [collect => L | Side | !edge(A, L, Side),
-         0 = count(All, L),
-         0 = count(Next, L),
-         *underline{@priority(A, L, P)}, P > 0.0,
+         0 = count(All, L), // L is not in All
+         0 = count(Next, L), // L is not in Next
+         *underline{@priority(A, L, P), P > 0.0,}
          *underline{@cpu-id(A, L, Id1)},
-         *underline{@cpu-id(A, A, Id2)}, Id1 = Id2 |
-         send-token(A, All, Next ++ L)].
+         *underline{@cpu-id(A, A, Id2), Id1 = Id2} |
+         send-tree(A, All, Next ++ L)].
 
-send-token(A, All, [])
+send-tree(A, All, [])
    -o first-phase(A, All, reverse(All)).
-send-token(A, All, [B | Next])
+send-tree(A, All, [B | Next])
    -o *underline{schedule-next(B)},
-      token(B, All ++ [B], [B | Next]).
+      tree(B, All ++ [B], [B | Next]).
 
@@
 second-phase(A, [], _)
-   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
+   -o *underline{set-priority(A, 0.0)}, inactive(A), update(A).
 second-phase(A, [A], Back)
-   -o update(A), waiting(Back),
-      waiting(A), *underline{set-priority(A, 0.0)}.
+   -o update(A), inactive(Back),
+      inactive(A), *underline{set-priority(A, 0.0)}.
 second-phase(A, [A, B | Next], Back)
-   -o update(A), waiting(Back), *underline{schedule-next(B)},
+   -o update(A), inactive(Back), *underline{schedule-next(B)},
       second-phase(B, [B | Next], A).

### 2014-11-08 abstract.tex
More updates.

@@
 programming since it hides the implementation details of parallelism away from the
-programmer. However, some control is lost when code is
-automatically parallelized, leaving the programmer with very little
-opportunities for fine tuning programs.
-We extended Linear Meld, a concurrent forward-chaining linear logic programming
-language, with coordination facts that allow the programmer to change
+programmer. However, its advantage has also been its downfall
+as it leaves the programmer with no straightforward way
+to optimize programs for performance.
+In this paper, we introduce Coordinated Linear Meld, a concurrent forward-chaining linear logic programming
+language, with a declarative way to coordinate the execution of the program
+allowing the programmer to change
 computation scheduling and data layout. Our approach allows the programmer to
-write declarative parallel programs and then optionally use coordination facts
+write declarative parallel programs and then optionally use coordination 
 to fine tune programs, while keeping the program fully declarative and amenable
-to correctness proofs. We show several application examples, experimental
-results and some advantages of our novel approach.
+to correctness proofs.

### 2014-11-08 application.tex
New update.
More updates.

--- 2014-11-02 application.tex
@@
 To better understand how coordination facts are used, we next present some programs that
-take advantage of them.
+take advantage of them. In our experimental setup, we used a machine with
+four AMD Six-Core Opteron TM 8425 HE (2100 MHz) chips (24 cores) and 64 GB of
+DDR-2 667MHz (16x4GB) RAM, running GNU/Linux (kernel 3.15.10-201 64 bits).
+We compiled our virtual machine using GCC 4.8.3 (g++) with the flags
+\texttt{-O3 -std=c++11 -fno-rtti -march=x86-64}~\footnote{Implementation
+   available in \url{http://github.com/.../meld}}.
 
@@
 \subsection{MiniMax}
-
 \input{minimax}
@@
 \subsection{Heat Transfer}
-
 \input{heat-transfer}
@@
 \subsection{Splash Belief Propagation}
-
 \input{splash-bp}
@@
 \subsection{N Queens}
+\input{nqueens}

### 2014-11-08 coordination.tex
More corrections.
More updates.

@@
-Since LM uses linear logic and supports deletion of facts, the order in which
-nodes are scheduled can impact the performance and even the results of the
-program.
-
 The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
-declarative but its performance may depend in the order in which nodes are
+declarative but its performance depends on the order in which nodes are
 executed. If nodes with greater distances are prioritized over other nodes, the
@@
 reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
-clear that the best computational ordering is the following: \texttt{@1},
+clear that the best scheduling is the following: \texttt{@1},
 \texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
@@
 \texttt{@2}, then 6 \texttt{relax} facts would have been generated.
-Therefore, the optimal solution for SSSP is to schedule the node with the
+The optimal solution for SSSP is to schedule the node with the
 shortest distance, which is essentially the Dijkstra shortest path
@@
 
-We introduce the concept of \emph{coordination facts} that allow the programmer
-to change how the run time schedules nodes and partitions the nodes among
-threads of execution. Coordination facts can be used in the body or head of the
-rules to allow the programmer to reason and change how execution is to be done.
+\emph{Coordination facts} allow the programmer
+to change how the run time schedules nodes and how it partitions the nodes among
+threads of execution. Coordination facts can be used in either the body of the
+rule, the head of the rule or both.
+This allows scheduling and partition decisions to be made based on the state of
+the program and on the state of the underlying machine.
 In this fashion, we keep the language declarative because we reason logically
 about the state of execution, without the need to introduce extra-logical
-operators into the language that would introduce problems when attempting to prove
+operators into the language that would introduce significant issues when proving
 properties about programs.
 
-There are two classes of coordination facts. The first class of coordination facts are
-called \emph{sensing facts} and are used to sense information about the
+There are two kinds of coordination facts: \emph{sensing} and \emph{action}
+facts. Sensing facts are used to sense information about the
 underlying runtime system, including node placement and node scheduling.
-The second kind of coordination facts are \emph{action facts} that are deleted
-to apply a coordination operation in the runtime system.
+Action facts are used to apply a coordination operations on the runtime system.
 
@@
 
-We can use action facts to change the order in which nodes are evaluated by adding
-\emph{priorities}. Node priority works at the thread level
-so that each thread can make a local decision about which node to execute next.
-Note that, by default, nodes are picked using a FIFO approach, because that
-tends to work better for most programs since older nodes tend to have more facts
-to be used.
+In order to allow different scheduling strategies, we introduce the concept of
+\emph{node priority} by assigning a priority value to every node in the program
+and by introducing coordination facts that manipulate such priority values.
+By default, nodes do not have any priority and can be picked in any order. In
+our implementation, we use a FIFO approach because this tends to work better for
+most programs since older nodes tend to a higher number of unexamined facts.
 
@@
 
+\newcommand{\code}[1]{\texttt{\small{#1}}}
+
 \begin{itemize}
-   \item \texttt{type linear set-priority(node, float)}: This sets the
-   temporary priority of a node. If the node already has a priority, we only
-   change the priority if the new one is higher. The programmer can
-   decide if priorities are to be ordered in ascending or descending order.
-   \item \texttt{type linear action add-priority(node, float)}: Increases,
-   temporarily, the priority of the node.
-   \item\texttt{type linear schedule-next(node)}: The work will fetch the
-   highest priority node's priority $P$ from its set of nodes and set the
-   action's argument node's priority as $P + 1.0$. If using the priorities in
-   ascending order, we pick the lowest priority and subtract $1.0$.
-   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
+   \item \code{set-priority(node A, float F)}: This sets the
+   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
+   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
+   can decide if priorities are to be ordered in ascending or descending order.
+   \item \code{add-priority(node A, float F)}: Increases,
+   temporarily, the priority of node \texttt{A} by \texttt{F}.
+   \item \code{schedule-next(node A)}: Changes the temporary priority of node
+   \texttt{A} to be 1 unit higher than the current highest priority.
+   \item \code{set-default-priority(node, float)}: Sets the default
    priority of the node.
-   \item \texttt{type linear action stop-program(node)}: Immediately stops the
+   \item \code{stop-program(node A)}: Immediately stops the
    execution of the whole program.
@@
 
-LM provides the sensing fact \texttt{priority(A, B, P)} in order to sense the
-priority \texttt{P} of node \texttt{B} from node \texttt{A}.
+LM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
+priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
 Sensing facts are only used in the body of the rules in order
 to fetch information from the runtime system.
-Note that when sensing facts are deleted, they are re-derived automatically.
+Note that when sensing facts are used to prove new facts, they are re-derived automatically.
+~\footnote{The observant reader will know that all the coordination facts are
+   linear and XXX
 Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
 value of \texttt{priority} facts, but this done automatically by the runtime
-system.
+system.}
 
@@
 \begin{itemize}
-   \item \texttt{type linear set-cpu(node, int)}: Moves the node to a
-   thread of execution.
-   \item \texttt{type linear set-affinity(node, node)}: Places the first node in
-   the same thread as the second node.
-   \item \texttt{type linear set-moving(node)}: Allows the node to move freely
+   \item \code{set-cpu(node A, int C)}: Moves node \texttt{A} to thread
+   \texttt{C}.
+   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
+   the thread of node \texttt{B}.
+   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
    between threads.
-   \item \texttt{type linear set-static(node)}: Forces the node to stay in the
+   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
    same thread indefinitely.
@@
 \begin{itemize}
-   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
-   indicates the thread where the node of the second argument is currently running.
-   \item \texttt{type linear moving(node, node)}: Fact available if the node in the
-   second argument is allowed to move between threads.
-   \item \texttt{type linear static(node, node)}: Fact available if the node in
-   the second argument is not allowed to move between threads.
+   \item \code{cpu-id(node A, node B, int C)}: Store at node \texttt{A} which
+   thread, \texttt{C}, \texttt{B} is actually running on.
+   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
+   to move between threads.
+   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
+   allowed to move between threads.
 \end{itemize}

### 2014-11-08 heat-transfer.tex
Small changes.
More updates.

@@
 coming from neighbor nodes. This increases parallelism since nodes do not need
-to synchronize if we had to compute heat values using iterations.
+to synchronize between iterations.
 
 Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
-neighbor nodes. We added the first rule to increase the priority of the neighbor
+neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
 nodes if the current node has a large $\delta$. The idea is to prioritize the
 computation of heat values of nodes (using \texttt{update}) that have a neighbor
-node that changed significantly. Multiple \texttt{add-priority} facts will
+that changed significantly. Multiple \texttt{add-priority} facts will
 increase the priority of a node so that nodes with multiple large deltas will
-have more priority. Note that the LM compiler will merge the three rules in
+have more priority. To further improve scalability, we added the third rule to
+ignore to not send small $\delta$ values if the target node is in another
+thread. Note that the LM compiler will merge the three rules in
 Fig.~\ref{code:ht} into a single one that handles the 3 branches.
@@
 Delta = fabs(New - Old),
-Delta > epsilon * 2.0
+Delta > epsilon
    -o {B, W | !edge(A, B, W) |
@@
 
-new-heat(A, New, Old),
-Delta = fabs(New - Old)
-Delta < epsilon * 2.0,
-Delta > epsilon
+new-heat(A, New, Old)
+fabs(New - Old) <= epsilon
+\underline[cpu-id(A, B, C)],
+\underline[cpu-id(A, A, C)]
    -o {B, W | !edge(A, B, W) |
-         new-neighbor-heat(B, A, New), update(B)}.
+         new-neighbor-heat(B, A, New)}.
 
 new-heat(A, New, Old)
-Delta = fabs(New - Old)
-Delta < epsilon
-   -o {B, W | !edge(A, B, W) |
-         new-neighbor-heat(B, A, New)}.
+fabs(New - Old) <= epsilon,
+\underline[cpu-id(A, B, C1)],
+\underline[cpu-id(A, A, C2)],
+\underline[C1 <> C2]
+   -o 1. // nothing is derived
 \end{Verbatim}
-  \caption{Coordination code for the Heat Transfer program. We added one rule
-     that covers cases where $\Delta$ is at least $2 \epsilon$ in order to
-     increase the priority of neighbors nodes. The code logic remains exactly
+  \caption{Coordination code for the Heat Transfer program. We updated the first rule
+     to increase the priority of neighbor nodes and added the third rule to not
+     send small $\delta$ if the neighbor is in another thread. The code logic remains exactly
      the same as before, however bigger changes in heat values are now
@@
    \end{center}
-   \caption{Experimental results for the HT program. Speedup values for the
-      coordinated and uncoordinated version are computed using the run time of
-      the uncoordinated version using 1 thread.}
+   \caption{Experimental results for the HT program. The coordinated version
+      is, on average, 30\% faster than the regular version although it has a
+      slightly worse scalability due to a reduction in work available.}
    \label{results:ht}

### 2014-11-08 implementation.tex
More updates.

@@
 to exist to satisfy the rule. This procedure is executed by the virtual
-machine whenever we have enough facts to satisfy the rule. The procedure
-loops over all possible combinations of the rule, retrieving facts from the
+machine whenever all the facts mentioned in the body exist.
+The procedure loops over all possible combinations of the rule, retrieving facts from the
 database, performing join operations and then consuming and deriving facts.
 Optimizations such as join optimizations (to support efficient rule filtering)
-and fact updates are implemented by the compiler. The latter
-transforms fact derivations of the same predicate to a simple update operation,
-avoiding a deallocation operation followed by an allocation of the new fact.
+and fact updates are implemented by the compiler.
 
@@
 special instructions of the virtual machine, but they perform some action,
-instead of returning something. While these are still facts from the point of
-view of the language, they are deleted immediatelly in order to apply the
-operation, instead of being derived like any other regular fact. This is more
-efficient since we do not need to create a new fact.
+instead of being added to the database as facts.
+Semantically, action facts are like any other. However, since they are
+immediatelly used by the machine, there is no need to store them in the
+database, avoiding unnecessary allocations and deallocations.
 
@@
 The virtual machine is implemented in C++11 and uses the threading system from
-the standard library to implement multithreading. Each thread is responsible
-for executing a subset of nodes. Nodes are either inactive, with no new facts,
-or active, where they will be placed in the corresponding thread's queue.
-Threads do useful work by processing the nodes in their queues. Whenever a
-thread becomes idle, it attempts to steal nodes from a random thread into its
-own queues. If sucessful, the thread becomes idle and waits for program
+the standard library to implement multithreading. Each thread is responsible for
+executing a subset of \emph{active nodes}. A node is active if it has unexamined
+facts. After a node is processed, it becomes inactive until a new fact is
+derived for it. Inactive nodes have no work to do. When a new fact is derived for an
+inactive node, the node is made active and placed on the appropriate queue of
+the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
+thread becomes idle, it attempts to steal nodes from a random thread.
+If unsucessful, the thread becomes idle and waits for program
 termination while periodically attempting to steal nodes.
@@
 A node is represented as a collection of facts (per predicate) and an indexing structure that
-keeps track of the available predicates and potential candidate rules. We need
-a separate indexing structure per node since rules run locally.
+keeps track of the available facts and potential candidate rules. We need
+a separate indexing structure per node since rules run local to each node.
 
@@
 \begin{itemize}
-\item \emph{Tries} are used exclusively to store 
-  persistent facts. Tries are trees where facts are indexed by common
-  prefix arguments.
+\item \emph{Tries} are used exclusively to store persistent facts.
 \item \emph{Doubly Linked Lists} are used to store 
-  linear facts. We use a double linked list because it is a very
-   efficient way to add and remove facts.
+  linear facts. We use a double linked list because we need to efficiently add
+  and remove facts.
 \item \emph{Hash Tables} are used to improve lookup when 
-  linked lists are too long and when we need to do search filtered by
+  linked lists are too long and when we need to do searches filtered by
   a fixed argument. The virtual machine decides which arguments are
@@
 \begin{description}
-   \item[working]: the node is being executed.
-   \item[idle]: the node is not active, has no new facts and is not in any
+   \item[working]: the node is executing.
+   \item[inactive]: the node is inactive, i.e., it has no new facts and is not in any
    queue for processing.
@@
    to be processed.
-   \item[stealing]: the node has just been stolen is about to move to another
-   thread.
-   \item[coordinating]: the node is moving from one queue to another (i.e.
-         changing priority).
+   \item[stealing]: the node has just been stolen and is in the process of being
+   moved to another thread.
+   \item[coordinating]: the node is moving from one queue to another queue.
 \end{description}
@@
    \includegraphics[width=0.4\textwidth]{node_states.pdf}
-   \caption{The node state machine. During the lifetime of a program, each node
-   goes through different states as specified by the state machine.}
+   \caption{The node state machine as represented by the state variable. During
+      the lifetime of a program, each node goes through different states as
+      specified by the state machine.}
    \label{fig:node_states}
@@
 Each node is protected by a main spinlock that allows threads to change node
-attributes: incoming facts, owner thread, node state and locality information.
-There is also an internal spinlock that is locked whenever the node enters into
-the \textbf{working} state. If a node sends facts to node is placed in
-another thread, the current thread first attempts to lock the internal lock in order to update the
+attributes: incoming facts, owner thread, state variable and locality information.
+A database spinlock is locked whenever the node enters into
+the \textbf{working} state. If a node sends facts to another node placed in
+another thread, the current thread first attempts to lock the database lock in order to update the
 indexing structures of the target node, otherwise it adds the facts to the list of
@@
 Threads pop active nodes from their queues and execute new candidate rules. 
-In order to allow node priorities, we use 2 queues per thread: a doubly linked
-list known as the \emph{standard
-queue} and a min/max heap known as the \emph{priority queue}. The regular queue
-contains nodes without priorities and
-implements the operations \texttt{push\_tail(node)} (push to the tail),
-\texttt{pop\_head()} (remove node from the head),
-\texttt{remove(node)} (remove an arbitrary node) and \texttt{remove\_half()}
-(removes first half). The priority queue contains nodes
-with priorities and is implemented as a binary heap array. It supports the
-operations \texttt{push(node,priority)} (add new node to the heap),
-           \texttt{pop\_min()} (remove the best node),
-\texttt{remove(node)} (remove an arbitrary node), \texttt{pop\_half\_min()}
-(remove the best half) and \texttt{move(node,new\_priority)} (update priority).
-Operations such as \texttt{remove\_half()} are supported in order to support
-node stealing, while operations \texttt{remove(node)} or
-\texttt{move(node,new\_priority)} allows threads to change the priority of
-nodes.
+In order to reduce inter-thread communication, node priorities are implemented
+at the thread level. This means that, when a thread picks the highest
+priority node, it is actually the highest priority from the set of nodes owned
+by the thread and not the highest priority node in the whole program.
+To allow node priorities, we use 2 two pairs of queues per thread: a pair of doubly linked
+lists known as the \emph{standard queue} and a pair of min/max heap known as the \emph{priority queue}.
+
+The regular queue contains nodes without priorities and
+supports push into tail, remove node from the head, remove arbitrary node, and
+remove first half of nodes.
+The priority queue contains nodes with priorities and is implemented as a binary
+heap array. It supports the following operations: push into the heap, remove the min node,
+remove an arbitrary node, remove half of the nodes (horizontal split), and
+priority update.
+Operations for removing half of the queue are implemented in order to support
+node stealing, while operations to remove arbitrary nodes or update the
+priority of nodes allows threads to change the priority of nodes.
 
@@
 Threads synchronize with each other using mutual exclusion. We use a spinlock in
-each queue to protect queue operations and one spinlock to protect the thread's
-state. Given threads $T_1$ and $T_2$, we enumerate the most important
+each queue to protect queue operations.
+Given threads $T_1$ and $T_2$, we enumerate the most important
 synchronization intensive places in the virtual machine:
@@
 \begin{description}
-   \item[New facts:] When a node executes on $T_1$ and derives facts
+   \item[New facts:] When a node executes in $T_1$ and derives facts
    to a node in $T_2$, $T_1$ first buffers the facts 
@@
    \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
-   nodes from $T_2$ by removing half of the nodes from $T_2$'s queues.
+   nodes from $T_2$ by removing half of the nodes from one of $T_2$'s queues.
    \item[Coordination:] If $T_1$ needs to perform coordination operations
@@
 
-\subsubsection{Termination}
-
-There is a global atomic counter, a global boolean flag and one boolean flag
-(active/idle) per thread for detecting termination. Once a thread goes idle,
-it decrements the global counter and changes its flag to idle. If the counter
-goes to zero, the global flag is set to idle. Since every thread will be
-busy-waiting and checking the global flag, they will detect the change and exit
-the program.
-
-\subsubsection{Node collection}
+\subsubsection{Garbage collection}
 
 Whenever a node has an empty database and no references from other nodes, it is
-deleted from the graph. The virtual machine detects such cases by keeping a
-count of fact references to every node. Once the count drops to zero and the
-database is empty, the node is immediatelly deleted in order to reduce memory
-usage.
+deleted from the graph. The virtual machine uses reference counting to detect
+such cases.
 
@@
 machine and the new virtual machine with coordination mechanisms. All the
-programs benchmarked do not use any kind of coordination. As expected, there
-is no degration of performance by adding coordination mechanisms.
+programs benchmarked do not use any kind of coordination. As hoped for, there
+is no degration of performance by adding coordination mechanisms. Since we also
+performed general performance improvements, programs such as PageRank and Greedy
+Graph Coloring programs now perform better than before.
 
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
+      \subfloat[]{\includegraphics[width=8cm]{results/comparison1.png}}
    \end{center}

### 2014-11-08 intro.tex
More updates.

@@
 Declarative programming has been hailed as a solution to this issue, since the
-problem of implementing the details of parallelism is moved away from the programmer to the
-compiler and runtime of the language. The programmer writes code
+problem of implementing the details of parallelism is moved from the programmer to the
+compiler and runtime. The programmer writes code
 without having to deal with parallel programming constructs and the compiler
@@
 over how execution is scheduled or how data is laid out, making it hard to achieve
-necessary efficiency. This introduces some
+necessary efficiency. This introduces 
 performance issues because even if the runtime system is able to reasonably
 parallelize the program using a general algorithm, there is a lack of specific
-information about the program that a compiler does not easily understand. In
-turn, such information would make execution better in terms of run time time,
-memory usage and scalability.
+information about the program that a compiler cannot easily deduce. Such
+information could make execution better in terms of run time time,
+memory usage, and scalability.
 
-In this paper, we extend Linear Meld (LM), a declarative language by
-\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
+In this paper, we introduce Coordinated Linear Meld (CLM), a declarative
+language that extends the Linear Meld~(LM) language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
 control over scheduling and data placement. LM is a linear logic programming
 language designed for programs that operate on graphs.  The use of linear
-logic~\cite{girard-87} allows support for structured manipulation of mutable
+logic~\cite{girard-87} supports structured manipulation of mutable
 state. In LM, computation is divided so that each node of the graph computes
-independently and is allowed to communicate with other nodes.
-Both computation and communication happen through derivation of logical rules.
+independently and is allowed to \scare{communicate} with other nodes.
+Both computation and communication happen through the derivation of logical
+rules (which make up the program).
 
-Given that LM uses mutable state, the order in which nodes are executed may be
-crucial in terms of performance. We extended the language with coordination
-facts that can be used as a regular fact in the program logic to improve program
-execution.  The first kind of coordination facts are called \emph{sensing facts}
+The CLM language features coordination
+primitives that can be used in the same way as any other primitive,
+in other words, they are specificied with the same syntax and semantics as the
+rest of the programming language. These coordination primitives can be used to
+improve program execution based on the state of the program and the underlying
+machine. The first kind of coordination facts are called \emph{sensing facts}
 and are used to sense information about the system the program is running on,
@@
 To the best of our knowledge, this is the first time that a declarative language allows 
-control over execution without a significant change in the language itself and
-without using non-declarative operations. This is specially important since
-programs can be proven to be correct even in the presence of coordination facts.
-In order to prove our thesis, we first give a brief overview of the
-base language, including a small example. Next, in
-Section~\ref{sec:coordination} we introduce the available coordination facts,
-followed by Section~\ref{sec:implementation}, where we describe the
-implementation of the runtime required to implement the desired coordination
-mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
-programmer control to declarative programs in order to make them perform better.
+control over execution while staying declarative and without resorting to
+meta-language constructs. This is crucial to our goal of being able to prove
+programs correct since
+proofs can be constructed even in the presence of coordination facts.
+After briefly discussing related work, we present an overview of the base
+language, with an example. In Section~\ref{sec:coordination}, we introduce the
+current set of coordination facts,
+followed by a description of the changes required to implement the desired
+coordination mechanisms. In
+Section~\ref{sec:applications} we present several applications and show
+coordination can improve programs without destroying clarity or provability.

### 2014-11-08 language.tex
More updates.

@@
-Linear Meld (LM) is a forward-chaining linear logic programming
-language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
+Coordinated Linear Meld (CLM) is a forward-chaining linear logic programming
+language based on LM~\cite{cruz-iclp14}.
 Programs consist of a set of \emph{rules} and a \emph{database of facts}.
@@
 A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
-associated tuple of values, e.g., the values of \texttt{X} and \texttt{Y}, representing the
+associated tuple of values, e.g., the concrete values of \texttt{X} and
+\texttt{Y}, substitued for the
 arguments. Since LM uses linear logic as its foundation, we distinguish between
@@
 
-\newcommand{\scare}[1]{``#1''}
+CLM has been designed for writing programs that operate on graphs. CLM partitions
+the database by using the first argument of each fact. The first argument has
+type \emph{node} and represents a node in the graph. To achieve concurrency,
+our implementation partitions the database by the first argument, e.g.,
+the fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while
+fact \texttt{p(@2)} is stored in node \texttt{@2}. CLM restricts the body of
+every rule to facts with the same node so that nodes can derive rules
+independently. Although the body is restricted, the head of the rule may refer
+to any node as long as that node is refered to somewhere in the body. This allows
+\scare{communication} between nodes during rule derivation, since a node may
+\scare{send} a fact to another node.  Rule restrictions in turn make CLM
+implicitly parallel because nodes are able to compute independently. This makes
+CLM non-deterministic since nodes can be picked to run in any order, affecting
+which rules are applied and which facts are deleted or derived.
 
-LM has been designed for writing programs that operate on graphs. LM partitions
-the database by using the first argument of each fact. The first argument is
-typed as a \emph{node} and represents a node in the graph. For instance, the
-fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
-is stored in node \texttt{@2}. LM restricts
-the body of every rule to only refer to the same node so that nodes can derive
-rules independently. Although the body is restricted, the head of the
-rule may refer to any node as long as the node is refered somewhere in the
-body. This allows \scare{communication} between nodes during rule derivation, since a
-node may \scare{send} a fact to another node.
-Rule restrictions in turn make LM implicitly parallel because nodes are able to
-compute independently. This makes LM non-deterministic since nodes can be picked
-to run in any order, affecting which rules are applied and which facts are
-deleted or derived.
-
-In order to fully understand how the language works, we present a very simple
-algorithm: the single source shortest path program~(SSSP). Later in the paper, we
+To make these ideas concrete, we present a simple
+program: the single source shortest path program~(SSSP). Later in the paper, we
 add coordination facts to improve the execution of the program.
@@
 The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
-with the declaration of the predicates. Predicates specify the kinds of facts
+with the declaration of the predicates. Predicates specify the facts
 used in the program. The first predicate, \texttt{edge}, is a persistent
@@
 (infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
-algorithm with the initial distance from \texttt{@1} to \texttt{@1}.
+algorithm by setting the initial distance from \texttt{@1} to \texttt{@1} to be
+0.
 
@@
 where \texttt{W} is the weight of the edge. For
-instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
+example, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
 \texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
@@
 
-There many opportunities for concurrency in the SSSP program. For instance,
+There are many opportunities for concurrency in the SSSP program. For instance,
 after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
 possible to either apply rules in either node \texttt{@2} or node
-\texttt{@3}. In our implementation, we partition the graph into subgraphs that
-are processed by multiple threads of execution. Eventually, it is no longer to
-possible to apply rules and the final result present in
-Fig.~\ref{fig:shortest_path_program}~(c) is achieved.
+\texttt{@3}. This decision depends largely on implementation factors such as node
+partitioning and number of threads in the system.
+Still, it is easy to prove that no matter the scheduling used,
+the final result present in Fig.~\ref{fig:shortest_path_program}~(c) is achieved.
 
-%On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
+%On the other hand, CLM has no natural matching of data and computation to workers (processes, threads),
 %since nodes are a program abstraction and part of the program's logic.

### 2014-11-08 minimax.tex
More updates.

@@
 game state and the children represent the possible game moves that can
-be made by either player 1 or player 2. The
-tree levels are built in a way that each player plays in turns and the
-first level represents the moves made by player 1.  An
-evaluation function \texttt{f(State)} is used to compute the score of
+be made by either player 1 or player 2.
+An evaluation function \texttt{f(State)} is used to compute the score of
 the board for each leaf of the tree. A node is a leaf when the game state can no longer be
@@
 
-The LM code in Fig.~\ref{minimax:check-end} implements
-the initial expansion of the tree. The first three rules (lines 1-10) deal
-with the cases where no children nodes are created and the last three rules
-(12-29) deal with cases where new nodes may be created.
-The rule in lines 12-25 generate new nodes using the
+The code in Fig.~\ref{minimax:check-end} shows part of the program.
+The first three rules (lines 1-10) deal
+with the case where no children nodes are created and the last three rules
+(12-29) deal with the cases for creating new nodes.
+The rule in lines 12-25 generates new nodes using the
 \texttt{exists} language construct, which creates a new node address
-\texttt{B}, where a new \texttt{play} fact is created with the new
+\texttt{B}, where a new \texttt{play} fact is created with the a
 board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
@@
 
-As noted before in Section~\ref{sec:fifo}, we use a FIFO ordering for nodes,
+As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by XXX,
+in a FIFO ordering for nodes,
 meaning that when children are expanded, we first expand the children and
-then the children of the children. We note
-however, that this is not optimal for this program since the complete tree needs
+then the children of the children. In this case, the defined ordering
+is not optimal for since the complete tree needs
 to be expanded before computing the scores at the leaves. The memory
@@
 The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
-is the depth of the tree and $t$ is the number of threads. In
-Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
-expanded the tree on the left side. When a second thread is idle, it steals the
-first unexpanded node and expands the tree from that position. Since threads
-prioritize deeper nodes, the scores of the first leaves are immediatelly
+is the depth of the tree and $t$ is the number of threads. As an example,
+consider a system with 2 threads, $T_1$ and $T_2$, where $T_1$ first expands the root
+node and then the first child. Since $T_2$ is idle, it steals half of the root's
+children nodes and starts expanding one of the nodes in a depth-first fashion.
+Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
 computed and then sent to the parent node. At this point, the leaves are deleted
@@
 
+\iffalse
 \begin{figure}[h!]
@@
 \end{figure}
+\fi
 
@@
 sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
-the regular version needs around 14GB. Note that as the number of threads goes
+the regular version needs almost 14GB. Note that as the number of threads goes
 up, memory usage also goes up. This is an artifact of our parallel memory
 allocator that allocates large chunks of memory beforehand.
-In terms of run time, our experimental results show a 14-fold speedup for the
+In terms of scalability, our experimental results show a 14-fold speedup for the
 coordinated version against a 12-fold for the regular version when using 16

### 2014-11-08 related.tex
More updates.

@@
-There has been an increasing interest in declarative and data-centric
+There has been an increasing interest in declarative 
 languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
-popular data-centric programming model that is optimized for large clusters. The
-scheduling and data sharing model is very simple: in the \emph{map phase}, data
-is transformed at each node and the result reduced to a final result in the
-\emph{reduce phase}. In order to facilitate the writing of programs over large
+popular data-centric programming model that is optimized for large clusters.
+Intrinsic to its popularity is the simplicity of its scheduling and data sharing model.
+In order to facilitate the writing of programs over large
 datasets, SQL-like languages such as
@@
 PigLatin builds on top of MapReduce and allows the programmer to write complex
-data-flow graphs, increasing the abstraction of the MapReduce paradigm.
+data-flow graphs, raising the abstraction and easy of programmability of
+MapReduce programs.
 An alternative to PigLatin/MapReduce is
@@
 coordinate the scheduling of processing units, while the placement of data is
-left to the implementation. LM differs from those languages since it
-is fully data-driven and coordination acts on data instead of processing units.
+left to the implementation. But even more importantly, CLM differs from those
+languages because coordination acts on data instead of processing units.
 The abstraction is then raised by considering data and algorithmic aspects of
@@
 
-Forward-chaining logic programming was made famous by the Datalog
+Forward-chaining logic programming came into its own with the Datalog
 language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
@@
 nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
-LM itself is inspired in Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
+The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
 programming distributed emsembles of modular robots. Meld introduced the idea of
 sensing and action facts in order to sense and act on the outside world,
-respectively. Coordination facts are inspired in the same principles but are
-used for coordinating data and computation.
+respectively. LCM uses sensing facts to sense the underlying system that the program is running on and action facs
+act on that system.

### 2014-11-08 sssp.tex
More updates.

@@
-The Single Source Shortest Path (SSSP) program was already presented in
-Section~\ref{sec:language}. We are going to update the code in
-Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
-are selected to run before others nodes. The coordinated version of the same
-algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
+In Section~\ref{sec:language}, we said that the SSSP program has an optimal
+scheduling if the nodes with the shortest distances are selected to run before
+other nodes.  The coordinated version of the SSSP
+program is presented in Fig.~\ref{code:shortest_path_program_coord} and it
 uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
@@
 threads, each thread will pick the smallest distance from their subset of nodes.
-While this is not optimal since threads do not share a global view of priorities,
-many rule derivations are going to be avoided.
+While this does not yield the optimal program with relation to 1 thread, it
+allows for parallel execution and locally avoid unnecessary work. XXX
 
@@
 
-Figure~\ref{results:sssp_uspowergrid} shows experimental results for a variant of
-the SSSP program that computes the SSSP for many of the nodes of the graph.
+Figure~\ref{results:sssp_uspowergrid} shows experimental results a
+SSSP program that computes the SSSP for 20\% of the nodes of the graph.
 There are some situations where unecessary facts are propagated
@@
       }
-      \subfloat[]{\includegraphics[width=4.5cm]{results/shortest-uspowergrid.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/shortest-uspowergrid.png}}
    \end{center}

### 2014-11-08 coordination.tex
More updates.

@@
 and by introducing coordination facts that manipulate such priority values.
-By default, nodes do not have any priority and can be picked in any order. In
-our implementation, we use a FIFO approach because this tends to work better for
-most programs since older nodes tend to a higher number of unexamined facts.
+By default, nodes have no priority and can be picked in any order. In
+our implementation, we use a FIFO approach because
+older nodes tend to a higher number of unexamined facts, from which to derive
+subsequent new facts.
 
@@
 
-LM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
+CLM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
 priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
@@
 Note that when sensing facts are used to prove new facts, they are re-derived automatically.
-~\footnote{The observant reader will know that all the coordination facts are
-   linear and XXX
-Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
-value of \texttt{priority} facts, but this done automatically by the runtime
-system.}
+~\footnote{The observant reader will notice that all the coordination facts are
+linear and when created in the body of a rule will be consumed. The system
+creates the necessary code to re-derive them without programmer interaction.
+Likewise, \texttt{set-priority} and \texttt{set-default-priority} update the
+value of \texttt{priority} facts by consuming and re-deriving them but this is
+done automatically by the runtime system.}
 

### 2014-11-08 heat-transfer.tex
More updates.

@@
 In the Heat Transfer algorithm~(HT), we have a graph where heat values are
-exchanged between nodes. The goal of the program is to exchange heat to a point
-where new heat value $\delta = |H_i -
-H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
+exchanged between nodes. The goal of the program is to exchange heat between
+nodes. The program stops when the new heat values of every node are
+$\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
 asynchronously since heat values can be updated by using new partial information
@@
 
-Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
+Fig.~\ref{code:ht} shows the HT rules that send new heat values to
 neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
@@
 increase the priority of a node so that nodes with multiple large deltas will
-have more priority. To further improve scalability, we added the third rule to
-ignore to not send small $\delta$ values if the target node is in another
-thread. Note that the LM compiler will merge the three rules in
-Fig.~\ref{code:ht} into a single one that handles the 3 branches.
+have more priority.
 
@@
 
+new-heat(A, New, Old)
+fabs(New - Old) <= epsilon
+   -o {B, W | !edge(A, B, W) |
+         new-neighbor-heat(B, A, New)}.
+\end{Verbatim}
+  \caption{Coordination code for the Heat Transfer program. We updated the first rule
+     to increase the priority of neighbor nodes.  The code logic remains exactly
+     the same as before, however bigger changes in heat values are now
+     propagated faster.}
+  \label{code:ht}
+\end{figure}
+\normalsize
+
+Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
+and coordinated version. The dataset used is a square grid with an inner square
+with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
+for 16 threads, there's a 18\% reduction.
+
+To further improve locality, we split the second rule to
+not send small $\delta$ values if the target node is in another
+thread. XXX talk about results
+
+\begin{figure}[h!]
+\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
 new-heat(A, New, Old)
@@
 \end{Verbatim}
-  \caption{Coordination code for the Heat Transfer program. We updated the first rule
-     to increase the priority of neighbor nodes and added the third rule to not
-     send small $\delta$ if the neighbor is in another thread. The code logic remains exactly
-     the same as before, however bigger changes in heat values are now
-     propagated faster.}
-  \label{code:ht}
+  \caption{To improve locality, we added a third rule to not send small $\delta$
+     values if the neighbor is in another thread.}
+  \label{code:ht_better}
 \end{figure}
 \normalsize
-
-Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
-and coordinated version. The dataset used is a square grid with an inner square
-with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
-for 16 threads, there's a 18\% reduction.
 

### 2014-11-08 implementation.tex
More updates.

@@
 immediatelly used by the machine, there is no need to store them in the
-database, avoiding unnecessary allocations and deallocations.
+database, therefore avoiding unnecessary allocations and deallocations.
 
@@
 facts. After a node is processed, it becomes inactive until a new fact is
-derived for it. Inactive nodes have no work to do. When a new fact is derived for an
+derived for it. When a new fact is derived for an
 inactive node, the node is made active and placed on the appropriate queue of
@@
 
-Threads pop active nodes from their queues and execute new candidate rules. 
+Threads pop active nodes from a thread local queue and execute new candidate rules to
+generate new facts, which in turn can activate other nodes.
 In order to reduce inter-thread communication, node priorities are implemented
@@
 a Python implementations and LM faired fairly well. In this section, we measure
-the impact of the coordination mechanisms we have implemented for the new
+the impact of the coordination mechanisms we have implemented for the CLM
 virtual machine.

### 2014-11-08 minimax.tex
More updates.

@@
 be made by either player 1 or player 2.
-An evaluation function \texttt{f(State)} is used to compute the score of
+An evaluation function is used to compute the score of
 the board for each leaf of the tree. A node is a leaf when the game state can no longer be
@@
 
-In LM, the program starts with a root node (with the initial game state)
+In CLM, the program starts with a root node (with the initial game state)
 that is expanded with the available moves at each level. The graph of the
@@
 
-The code in Fig.~\ref{minimax:check-end} shows part of the program.
+The code in Fig.~\ref{minimax:check-end} shows deals with expanding the tree.
 The first three rules (lines 1-10) deal
@@
 The rule in lines 12-25 generates new nodes using the
-\texttt{exists} language construct, which creates a new node address
-\texttt{B}, where a new \texttt{play} fact is created with the a
-board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
-communicate with its parent.
+\texttt{exists} language construct, which creates a child node 
+\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
+and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
+fact.
 
-As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by XXX,
-in a FIFO ordering for nodes,
-meaning that when children are expanded, we first expand the children and
-then the children of the children. In this case, the defined ordering
-is not optimal for since the complete tree needs
-to be expanded before computing the scores at the leaves. The memory
-complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
-in the tree.
+As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by
+default, which requires a complete expansion of the tree before computing the
+scores at the leaves, leading
+to a memory complexity of $\mathcal{O}(n)$, where $n$ is the number of nodes in
+the tree.
 
@@
 
-With coordination, we set the priority nodes to be the same as the
-depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
-The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
-is the depth of the tree and $t$ is the number of threads. As an example,
+With coordination, we set the priority of a node to be its depth
+so the tree is expanded in a depth-first fashion (lines 15 and 22),
+leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
+is the depth of the tree and $t$ is the number of threads.
+Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
+computed and then sent to the parent node. At this point, the leaves are deleted
+and reused for other nodes in the tree, resulting in minimal memory usage.
+
+\iffalse
+As an example,
 consider a system with 2 threads, $T_1$ and $T_2$, where $T_1$ first expands the root
@@
 children nodes and starts expanding one of the nodes in a depth-first fashion.
-Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
-computed and then sent to the parent node. At this point, the leaves are deleted
-and reused for other nodes in the tree, resulting in minimal memory usage.
+\fi
 
@@
 of the coordinated MiniMax against the regular MiniMax. The coordinated version
-sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
-the regular version needs almost 14GB. Note that as the number of threads goes
+uses significantly less memory (at most 8MB for 16 threads) than the regular
+version (almost 14GB). Note that as the number of threads goes
 up, memory usage also goes up. This is an artifact of our parallel memory

### 2014-11-08 nqueens.tex
More updates.

@@
 
-The LM solution considers the chess board as a graph where nodes exchange valid
+The CLM solution considers the chess board as a graph where nodes exchange valid
 configurations between each other. First, we start with the empty
@@
 Since computation goes from the top row to the bottom row, not all placements of
-nodes on threads may be equally performant. This is specially true because the
+nodes on threads may be equally performant. This is especially true because the
 bottom rows tend to perform the most work. The best placement is then to split the

### 2014-11-08 splash-bp.tex
More updates.

@@
 Randomized and approximation algorithms can obtain significant benefits from
-coordination directives because although the final program results will not be
-exact, they follow important statistical properties and can be computed faster.
-Examples of such programs are Loopy Belief Propagation.
+coordination directives because their inherent non-determinism can be harnessed
+to evaluate rules in different orders.
+An example of such program is Loopy Belief Propagation.
 Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
@@
 
-LBP is an algorithm that maps very well to the graph based model of LM. In its
+LBP is an algorithm that maps very well to the graph based model of CLM. In its
 original form, the belief values of nodes are computed by synchronous iterations.
@@
 These \emph{splash trees} are built iteratively until we reach convergence.
-In Fig.~\ref{splash_bp} we represent two threads creating two different splash
-trees. It is encouraged to create several splash trees concurrently as long as
-we have threads to create them.
+%In Fig.~\ref{splash_bp} we represent two threads creating two different splash trees.
 
+\iffalse
 \begin{figure}[h!]
@@
 Fig.~\ref{code:sbp} to a working but unoptimized version of the algorithm, every
-other logical rule remains the same. We add new logical ruels that coordinate
+other rule remains the same. We add new rules that coordinate
 the creation and execution of the splash trees:
@@
 line 2 in order to (1) force nodes to stay in the thread and (2) to partition
-nodes as a grid of threads. This setups a well defined area of nodes
+nodes as a grid of threads. This sets up a well defined area of nodes
 for threads to build splash trees on.
@@
 \end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program. LM needs
+  \caption{Coordination code for the Splash Belief Propagation program. CLM needs
      50 lines of rules to implement splash trees, while GraphLab needs a
@@
    \caption{Experimental results for LBP and SBP. Figure (a) shows the
-      scalability of LBP for both LM and GraphLab and Figure (b) shows the
+      scalability of LBP for both CLM and GraphLab and Figure (b) shows the
       scalability of SBP. Figure (c) presents the
@@
 algorithm (minus probabilitistic errors), SBP works in a very different way.
-A system that also implements SBP is
-GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
+SBP is implemented in GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
 GraphLab provides the splash scheduler for these types of inferences as part of
-the framework. This particular scheduler is around 350 lines of complicated C++ code.
-With our coordination facts, it is possible to create fairly complicated coordination
-patterns with only 12 simple logical rules.
+the framework. This particular scheduler is around 350 lines of C++ code.
+With our coordination facts, it is possible to create the necessary
+patterns with only 12 simple rules.
 
-We measured the behavior of LBP and SBP for both LM and GraphLab.
+We measured the behavior of LBP and SBP for both CLM and GraphLab.
 Fig.~\ref{results:splash_bp} shows that both systems have very similar behavior
-when using a variable number of threads.
-Note that GraphLab avoids message passing between nodes since new neighbor
-belief updates are changed immediatelly, resulting in slightly better
-scalability for GraphLab.
+when using a variable number of threads.  The differences in performance between
+GraphLab and CLM comes mainly because GraphLab performs in-place updates of
+neighbor belief values, while CLM compiler and runtime system performs naive
+manipulation of facts by deriving rules. The in-place update could be generated
+by the compiler using smarter compilation strategies.

### 2014-11-08 sssp.tex
More updates.

@@
-In Section~\ref{sec:language}, we said that the SSSP program has an optimal
-scheduling if the nodes with the shortest distances are selected to run before
-other nodes.  The coordinated version of the SSSP
-program is presented in Fig.~\ref{code:shortest_path_program_coord} and it
-uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
-also use a global program directive to order priorities in ascending order (line
-5).
+Here we add coordination to the SSSP program described in Section
+The coordinated version of the
+SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
+\texttt{set-priority} (line 14). We also use a global program directive to order
+priorities in ascending order (line 5).
 
-When using a single thread, the algorithm behaves like
-Dijkstra's shortest path algorithm~\cite{Dijkstra}. However, when using multiple
+When run with one thread, the algorithm behaves like
+Dijkstra's shortest path algorithm~\cite{Dijkstra}. When using multiple
 threads, each thread will pick the smallest distance from their subset of nodes.
 While this does not yield the optimal program with relation to 1 thread, it
-allows for parallel execution and locally avoid unnecessary work. XXX
+allows for parallel execution and locally avoids unnecessary work. The result
+scales well and it is close to Dijkstra's algorithm.
 
@@
 
-Figure~\ref{results:sssp_uspowergrid} shows experimental results a
-SSSP program that computes the SSSP for 20\% of the nodes of the graph.
+Figure~\ref{results:sssp_uspowergrid} shows experimental results when the
+program computes the SSSP for 20\% of the nodes of the graph.
 There are some situations where unecessary facts are propagated
-because although the shortest distance is selected, sub-optimal distances may be
+because although the shortest distance is selected locally, sub-optimal distances may be
 propagated because many SSSP distances are computed at the same time.
@@
 version using 16 threads is 1.3 times faster than the regular version using 16
-threads. In turn, this makes the coordinated version 20 times faster
-than the regular version using 1 thread. Note that the coordinated program is as
-scalable as the regular program.
+threads while still showing good scaling.
 

### 2014-11-08 minimax.tex
use new scap and fig macros

--- 2014-11-08 minimax.tex
@@
 
-\begin{figure}[h!]
+\begin{topfig}
 \scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
@@
 \end{Verbatim}
-\caption{MiniMax: checking if the game has ended and expanding the tree}
-\label{minimax:check-end}
-\end{figure}
+\scap{minimax:check-end}{MiniMax: checking if the game has ended and expanding the tree}
+\end{topfig}
 \normalsize
@@
 \iffalse
-\begin{figure}[h!]
+\begin{topfig}
    \begin{center}
@@
    \end{center}
-   \caption{Expanding the MiniMax tree using coordination. By prioritizing
+   \scap{fig:minimax}{Expanding the MiniMax tree using coordination. By prioritizing
       deeper nodes, threads are forced to expand the tree using a depth-first
@@
       tree before computing the node scores.}
-   \label{fig:minimax}
-\end{figure}
+\end{topfig}
 \fi
@@
 
-\begin{figure}[h!]
+\begin{topfig}
    \begin{center}
@@
    \end{center}
-   \caption{Memory usage and scalability of the regular and coordinated versions
+   \scap{results:memory_minmax}{Memory usage and scalability of the regular and coordinated versions
       of MiniMax.}
-   \label{results:memory_minmax}
-\end{figure}
+\end{topfig}
 

### 2014-11-08 splash-bp.tex
use new scap and fig macros

--- 2014-11-06 splash-bp.tex
@@
    \end{center}
-   \caption{Creating splash trees for belief propagation. Each threads picks the
+   \scap{splash_bp}{Creating splash trees for belief propagation. Each threads picks the
    highest priority node and creates a tree from that node. The belief values
@@
    the root to the leaves.}
-   \label{splash_bp}
-\end{figure}
+\end{topfig}
 
@@
 
-\begin{figure}[h!]
+\begin{topfig}
 \scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
@@
 \end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program. LM needs
+  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. LM needs
      50 lines of rules to implement splash trees, while GraphLab needs a
@@
      fact is going to be re-derived.}
-  \label{code:sbp}
-\end{figure}
+\end{topfig}
 \normalsize
 
-\begin{figure*}[ht!]
+\begin{dblfig}
    \begin{center}
@@
    \end{center}
-   \caption{Experimental results for LBP and SBP. Figure (a) shows the
+   \scap{results:splash_bp}{Experimental results for LBP and SBP. Figure (a) shows the
       scalability of LBP for both LM and GraphLab and Figure (b) shows the
@@
       times faster than LBP.}
-   \label{results:splash_bp}
-\end{figure*}
+\end{dblfig}
 

### 2014-11-08 conclusions.tex

@@
 changing the nature of the language or introducing non-declarative constructs. We
-took advantage of the fact that LM uses linear logic, which allows us to derive
+took advantage of the fact that CLM uses linear logic, which allows us to derive
 coordination facts that can be deleted in order to perform actions in the underlying
@@
 where data locality is far more important than in shared memory systems. We also
-think that LM may be useful as a glue distributed/parallel language since it can
+think that CLM may be useful as a glue distributed/parallel language since it can
 easily model any computation pattern using graphs. Furthermore, we think that

### 2014-11-08 heat-transfer.tex
use new scap and fig macros

@@
 
-\begin{figure}[h!]
+\begin{topfig}
 \scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
@@
 \end{Verbatim}
-  \caption{Coordination code for the Heat Transfer program. We updated the first rule
+  \scap{code:ht}{Coordination code for the Heat Transfer program. We updated the first rule
      to increase the priority of neighbor nodes.  The code logic remains exactly
@@
      propagated faster.}
-  \label{code:ht}
-\end{figure}
+\end{topfig}
 \normalsize
@@
 \end{Verbatim}
-  \caption{To improve locality, we added a third rule to not send small $\delta$
+  \scap{code:ht_better}{To improve locality, we added a third rule to not send small $\delta$
      values if the neighbor is in another thread.}
-  \label{code:ht_better}
-\end{figure}
+\end{topfig}
 \normalsize
 
-\begin{figure}[ht!]
+\begin{topfig}
    \begin{center}
@@
    \end{center}
-   \caption{Experimental results for the HT program. The coordinated version
+   \scap{results:ht}{Experimental results for the HT program. The coordinated version
       is, on average, 30\% faster than the regular version although it has a
       slightly worse scalability due to a reduction in work available.}
-   \label{results:ht}
-\end{figure}
+\end{topfig}

### 2014-11-08 language.tex
More updates.
use new scap and fig macros

@@
 \texttt{Y}, substitued for the
-arguments. Since LM uses linear logic as its foundation, we distinguish between
+arguments. Since CLM uses linear logic as its foundation, we distinguish between
 \emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
@@
 
-\begin{figure}[h!]
+\begin{topfig}
 \scriptsize\begin{Verbatim}[numbers=left]
@@
 \end{Verbatim}
-  \caption{Single Source Shortest Path program code.}
-  \label{code:shortest_path_program}
-\end{figure}
+  \scap{code:shortest_path_program}{Single Source Shortest Path program code.}
+\end{topfig}
 \normalsize
@@
 \end{center}
-\caption{Graphical representation of the SSSP program. Figure (a) represents the
+\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. Figure (a) represents the
    program after propagating initial distance at node \texttt{@1}, followed by
@@
    have been computed.}
-\label{fig:shortest_path_program}
 \end{figure*}

### 2014-11-08 splash-bp.tex

--- 2014-11-08 splash-bp.tex
@@
    the root to the leaves.}
-   \label{splash_bp}
-\end{figure}
+\end{topfig}
 \fi
@@
 \end{Verbatim}
-  \caption{Coordination code for the Splash Belief Propagation program. CLM needs
+  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. CLM needs
      50 lines of rules to implement splash trees, while GraphLab needs a
@@
    \end{center}
-   \caption{Experimental results for LBP and SBP. Figure (a) shows the
+   \scap{results:splash_bp}{Experimental results for LBP and SBP. Figure (a) shows the
       scalability of LBP for both CLM and GraphLab and Figure (b) shows the

### 2014-11-08 heat-transfer.tex
Further updates.

@@
 In the Heat Transfer algorithm~(HT), we have a graph where heat values are
-exchanged between nodes. The goal of the program is to exchange heat between
-nodes. The program stops when the new heat values of every node are
+exchanged between nodes. XXX rewrite The program stops when the new heat values of every node are
 $\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
@@
 not send small $\delta$ values if the target node is in another
-thread. XXX talk about results
+thread (Fig.~\ref{code:ht_better}). The \textbf{Local-Only} line in
+Fig.~\ref{results:ht} shows that improving locality by disabling small updates
+improves the execution time of HT. XXX update figure
 
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5cm]{results/new-heat-transfer-80}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/new-heat-transfer-80}}
    \end{center}

### 2014-11-08 implementation.tex
use new scap and fig macros
Further updates.

@@
 
-\begin{figure}[h!]
+\begin{topfig}
    \centering
    \includegraphics[width=0.4\textwidth]{node_states.pdf}
-   \caption{The node state machine as represented by the state variable. During
+   \scap{fig:node_states}{The node state machine as represented by the state variable. During
       the lifetime of a program, each node goes through different states as
       specified by the state machine.}
-   \label{fig:node_states}
-\end{figure}
+\end{topfig}
 
@@
 generate new facts, which in turn can activate other nodes.
+XXX NEEDS REWRITE
 In order to reduce inter-thread communication, node priorities are implemented
@@
    \end{center}
-   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
+   \scap{results:comparison1}{Measuring the overhead of coordination mechanisms. The \textbf{Old}
    bars represent the performance of the original virtual machine, while
@@
    coordination mechanisms.}
-   \label{results:comparison1}
-\end{figure}
+\end{topfig}

### 2014-11-08 splash-bp.tex
Further updates.

@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_belief-propagation-400.png}}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_splash-bp-400.png}}
-      \subfloat[]{\includegraphics[width=5cm]{results/system_improve_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/system_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/system_splash-bp-400.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/system_improve_belief-propagation-400.png}}
    \end{center}

### 2014-11-08 implementation.tex
rewrote threads/queue section.

@@
 
-Each node is protected by a main spinlock that allows threads to change node
-attributes: incoming facts, owner thread, state variable and locality information.
-A database spinlock is locked whenever the node enters into
-the \textbf{working} state. If a node sends facts to another node placed in
-another thread, the current thread first attempts to lock the database lock in order to update the
-indexing structures of the target node, otherwise it adds the facts to the list of
-incoming facts that are later processed by the owner thread.
+Each node is protected by a main spinlock that allows threads to
+change node attributes: incoming facts, owner thread, state variable
+and locality information.  A database spinlock is locked whenever the
+node enters into the \textbf{working} state.
+
+To avoid unneccesary copying, when a node sends facts to a node
+located in another thread, the current thread first attempts to lock
+the database lock of the target node in order to directly update its
+indexing structures, otherwise it adds the facts to the list of
+incoming facts that are later processed by the owner thread of the
+target node.
 
@@
 
-Threads pop active nodes from a thread local queue and execute new candidate rules to
-generate new facts, which in turn can activate other nodes.
-XXX NEEDS REWRITE
-In order to reduce inter-thread communication, node priorities are implemented
-at the thread level. This means that, when a thread picks the highest
-priority node, it is actually the highest priority from the set of nodes owned
-by the thread and not the highest priority node in the whole program.
-To allow node priorities, we use 2 two pairs of queues per thread: a pair of doubly linked
-lists known as the \emph{standard queue} and a pair of min/max heap known as the \emph{priority queue}.
+The main loop for each thread is to get an active node from one of its
+local queues and process all applicable rules for that node generating
+new facts which can result in new facts for the current node or other
+nodes.  The thread will continue processing facts for the current node
+until there are no new facts generated for it.  It will then pick
+another active node.  If its queues are all empty it will try to steal
+work (\ie nodes) from another thread.
 
-The regular queue contains nodes without priorities and
-supports push into tail, remove node from the head, remove arbitrary node, and
-remove first half of nodes.
-The priority queue contains nodes with priorities and is implemented as a binary
-heap array. It supports the following operations: push into the heap, remove the min node,
-remove an arbitrary node, remove half of the nodes (horizontal split), and
-priority update.
-Operations for removing half of the queue are implemented in order to support
-node stealing, while operations to remove arbitrary nodes or update the
-priority of nodes allows threads to change the priority of nodes.
+To support priorities each thread has two pairs of queues: a pair of
+doubly linked lists known as the \emph{standard queue} and a pair of
+min/max heap known as the \emph{priority queue}.  The standard queue
+contains nodes without priorities and supports push into tail, remove
+node from the head, remove arbitrary node, and remove first half of
+nodes.  The priority queue contains nodes with priorities and is
+implemented as a binary heap array. It supports the following
+operations: push into the heap, remove the min node, remove an
+arbitrary node, remove half of the nodes (horizontal split), and
+priority update.  Operations for removing half of the queue are
+implemented in order to support node stealing, while operations to
+remove arbitrary nodes or update the priority of nodes allows threads
+to change the priority of nodes.
 
-The \texttt{next} and \texttt{prev} pointers of the regular queue are part of
-the node structure in order to save space. These pointers are also used as the
-index in the priority queue and current priority, respectively.
+To minimize inter-thread communication, node priorities are
+implemented at the thread level. Thus, when a thread picks the highest
+priority node from the priority queue, it is only the highest priority
+with respect to the set of nodes owned by the thread and not the
+highest priority node in the whole program.
 
-Both the regular and priority queue are actually implemented as 2 queues each.
-The \emph{static queue} contains nodes that are local to a thread (and thus cannot be stolen) and
-the \emph{dynamic queue} maybe accessed by other threads during node stealing.
+The \texttt{next} and \texttt{prev} pointers of the regular queue are
+part of the node structure in order to save space. These pointers are
+also used as the index in the priority queue and current priority,
+respectively.
+
+Both the regular and priority queue are implemented as a pair of
+queues.  This first of the pair is the \emph{static queue} which
+contains nodes that cannot be stolen.  The other queue in each pair is
+the \emph{dynamic queue} which contains nodes which can be stolen by
+other threads.
 

### 2014-11-09 heat-transfer.tex
Update ht plot.

@@
 not send small $\delta$ values if the target node is in another
-thread (Fig.~\ref{code:ht_better}). The \textbf{Local-Only} line in
-Fig.~\ref{results:ht} shows that improving locality by disabling small updates
-improves the execution time of HT. XXX update figure
+thread (Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Coordinated} line in
+Fig.~\ref{results:ht} shows that this new version performs better due to
+improved locality, exhibiting linear scalability over the regular version.
+However, this come at the price of increased errors in the computed heat values.
 

### 2014-11-09 locks.tex
Add locks plots.

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks-local-only.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/minmax-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/sssp-locks.png}}
   \end{center}
\end{topfig}

\begin{topfig}
   \begin{center}
      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/bp.png}}
   \end{center}
\end{topfig}

### 2014-11-09 results/locks/plot.py

@@ -1,0 +1,147 @@
+#!/usr/bin/python
+
+import re
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+
+reg_file = sys.argv[1]
+coord_file = sys.argv[2]
+output_file = sys.argv[3]
+title = sys.argv[4]
+
+class dataobj(object):
+   ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
+
+   def set(self, name, fail, total):
+      self.info_fail[name] = fail
+      self.info_total[name] = total
+
+   def get_total(self, name):
+      return self.info_total[name]
+
+   def get_fail(self, name):
+      return self.info_fail[name]
+
+   def total_locks(self):
+      return sum(total for name, total in self.info_total.iteritems())
+
+   def total_fail(self):
+      return sum(total for name, total in self.info_fail.iteritems())
+
+   def total_coord_locks(self):
+      return sum(self.get_total(name) for name in self.ATTRS)
+
+   def fail_coord_locks(self):
+      return sum(self.get_fail(name) for name in self.ATTRS)
+
+   def frac_coord_locks(self):
+      return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
+
+   def show(self):
+      print "THREADS", self.threads
+      acc = 0
+      for name, total in self.info_total.iteritems():
+         print name, self.get_fail(name), "/", total
+         acc = acc + total
+      print acc
+
+   def __init__(self, th):
+      self.threads = th
+      self.info_total = {}
+      self.info_fail = {}
+
+class experiment(object):
+
+   def add_thread(self, thread, data):
+      self.threads[thread] = data
+
+   def get_thread(self, thread):
+      return self.threads[thread]
+
+   def num_threads(self):
+      return len(self.threads.keys())
+
+   def get_coord_total_locks(self):
+      return [data.total_coord_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_coord_fail_locks(self):
+      return [data.fail_coord_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_fail_locks(self):
[…]

### 2014-11-09 results/locks/plot.py
Change legend.

@@ -143,5 +143,5 @@ pfailreg = ax.bar(ind + width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
 ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.4])
 ax.set_xticks(ind)
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
-ax.legend((ptotal, pcoord, pfail, ptotalreg), ('Total Locks', 'Coordination Locks', 'Coordination Locks Failed', 'Regular Total Locks'), loc=2)
+ax.legend((ptotal, pcoord, pfail, ptotalreg), ('Original Locks', 'Coordination Locks', 'Coordination Locks (Failed)', 'Original Regular Locks'), loc=2)
 plt.savefig(output_file)

### 2014-11-09 results/locks/lib.py
Create plot library.

@@ -1,0 +1,110 @@
+import re
+
+class dataobj(object):
+   ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
+
+   def set(self, name, fail, total):
+      self.info_fail[name] = fail
+      self.info_total[name] = total
+
+   def get_total(self, name):
+      return self.info_total[name]
+
+   def get_fail(self, name):
+      return self.info_fail[name]
+
+   def total_locks(self):
+      return sum(total for name, total in self.info_total.iteritems())
+
+   def total_fail(self):
+      return sum(total for name, total in self.info_fail.iteritems())
+
+   def total_coord_locks(self):
+      return sum(self.get_total(name) for name in self.ATTRS)
+
+   def fail_coord_locks(self):
+      return sum(self.get_fail(name) for name in self.ATTRS)
+
+   def frac_coord_locks(self):
+      return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
+
+   def show(self):
+      print "THREADS", self.threads
+      acc = 0
+      for name, total in self.info_total.iteritems():
+         print name, self.get_fail(name), "/", total
+         acc = acc + total
+      print acc
+
+   def __init__(self, th):
+      self.threads = th
+      self.info_total = {}
+      self.info_fail = {}
+
+class experiment(object):
+
+   def add_thread(self, thread, data):
+      self.threads[thread] = data
+
+   def get_thread(self, thread):
+      return self.threads[thread]
+
+   def num_threads(self):
+      return len(self.threads.keys())
+
+   def get_coord_total_locks(self):
+      return [data.total_coord_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_coord_fail_locks(self):
+      return [data.fail_coord_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_fail_locks(self):
+      return [data.total_fail() for th, data in sorted(self.threads.iteritems())]
+
+   def get_frac_coord_locks(self):
+      return [data.frac_coord_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_total_locks(self):
+      return [data.total_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_threads(self):
+      return [th for th in sorted(self.threads.keys())]
+
+   def get_lock_fraction(self, other):
+      # Compute other.locks / self.locks
[…]

### 2014-11-09 results/locks/plot.py

@@ -8,119 +8,12 @@ import matplotlib as mpl
 import numpy as np
 from matplotlib import rcParams
 from numpy import nanmax
+from lib import read_experiment
 
 reg_file = sys.argv[1]
 coord_file = sys.argv[2]
 output_file = sys.argv[3]
 title = sys.argv[4]
-
-class dataobj(object):
-   ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
-
-   def set(self, name, fail, total):
-      self.info_fail[name] = fail
-      self.info_total[name] = total
-
-   def get_total(self, name):
-      return self.info_total[name]
-
-   def get_fail(self, name):
-      return self.info_fail[name]
-
-   def total_locks(self):
-      return sum(total for name, total in self.info_total.iteritems())
-
-   def total_fail(self):
-      return sum(total for name, total in self.info_fail.iteritems())
-
-   def total_coord_locks(self):
-      return sum(self.get_total(name) for name in self.ATTRS)
-
-   def fail_coord_locks(self):
-      return sum(self.get_fail(name) for name in self.ATTRS)
-
-   def frac_coord_locks(self):
-      return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
-
-   def show(self):
-      print "THREADS", self.threads
-      acc = 0
-      for name, total in self.info_total.iteritems():
-         print name, self.get_fail(name), "/", total
-         acc = acc + total
-      print acc
-
-   def __init__(self, th):
-      self.threads = th
-      self.info_total = {}
-      self.info_fail = {}
-
-class experiment(object):
-
-   def add_thread(self, thread, data):
-      self.threads[thread] = data
-
-   def get_thread(self, thread):
-      return self.threads[thread]
-
-   def num_threads(self):
-      return len(self.threads.keys())
-
-   def get_coord_total_locks(self):
-      return [data.total_coord_locks() for th, data in sorted(self.threads.iteritems())]
-
-   def get_coord_fail_locks(self):
-      return [data.fail_coord_locks() for th, data in sorted(self.threads.iteritems())]
-
-   def get_fail_locks(self):
-      return [data.total_fail() for th, data in sorted(self.threads.iteritems())]
-
-   def get_frac_coord_locks(self):
-      return [data.frac_coord_locks() for th, data in sorted(self.threads.iteritems())]
-
[…]

### 2014-11-09 results/locks/plot_three.py
Merge HT locks plots into one.

@@ -1,0 +1,53 @@
+#!/usr/bin/python
+
+import re
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+from lib import read_experiment
+
+if len(sys.argv) != 7:
+   print "not enough arguments"
+   sys.exit(1)
+
+reg_file = sys.argv[1]
+coord_file = sys.argv[2]
+opt_file = sys.argv[3]
+output_file = sys.argv[4]
+title = sys.argv[5]
+opt_name = sys.argv[6]
+
+reg_exp = read_experiment(reg_file)
+coord_exp = read_experiment(coord_file)
+opt_exp = read_experiment(opt_file)
+
+N = reg_exp.num_threads()
+ind = np.arange(N)
+width = 0.25
+cmap = plt.get_cmap('gray')
+
+fig = plt.figure()
+ax = fig.add_subplot(111)
+ax.set_ylabel('Number of Locks')
+ax.set_xlabel('Threads')
+ax.set_title(title)
+
+ptotalc = ax.bar(ind, coord_exp.get_total_locks(), width, color=cmap(0.9), hatch='/')
+pcoordc = ax.bar(ind, coord_exp.get_coord_total_locks(), width, color=cmap(0.5))
+pfailc = ax.bar(ind, coord_exp.get_fail_locks(), width, color=cmap(0.1))
+
+ptotalo = ax.bar(ind + width, opt_exp.get_total_locks(), width, color=cmap(0.9), hatch='\\')
+pcoordo = ax.bar(ind + width, opt_exp.get_coord_total_locks(), width, color=cmap(0.5))
+pfailo = ax.bar(ind + width, opt_exp.get_fail_locks(), width, color=cmap(0.1))
+
+ptotalreg = ax.bar(ind + 2 * width, reg_exp.get_total_locks(), width, color=cmap(0.9), hatch=".")
+pfailreg = ax.bar(ind + 2 * width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
+ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.4])
+ax.set_xticks(ind)
+ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
[…]

### 2014-11-09 results/locks/lib.py
Update figures with SSSP.

@@ -1,5 +1,7 @@
 import re
 
+YAXIS = 200000000
+
 class dataobj(object):
    ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
 

### 2014-11-09 results/locks/plot.py

@@ -8,7 +8,7 @@ import matplotlib as mpl
 import numpy as np
 from matplotlib import rcParams
 from numpy import nanmax
-from lib import read_experiment
+from lib import read_experiment, YAXIS
 
 reg_file = sys.argv[1]
 coord_file = sys.argv[2]
@@ -34,6 +34,7 @@ pfail = ax.bar(ind, coord_exp.get_fail_locks(), width, color=cmap(0.1))
 ptotalreg = ax.bar(ind + width, reg_exp.get_total_locks(), width, color=cmap(0.9), hatch=".")
 pfailreg = ax.bar(ind + width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
 ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.4])
+#ax.set_ylim([0, YAXIS])
 ax.set_xticks(ind)
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
 ax.legend((ptotal, pcoord, pfail, ptotalreg), ('Original Locks', 'Coordination Locks', 'Coordination Locks (Failed)', 'Original Regular Locks'), loc=2)

### 2014-11-09 results/locks/plot_three.py

@@ -8,7 +8,7 @@ import matplotlib as mpl
 import numpy as np
 from matplotlib import rcParams
 from numpy import nanmax
-from lib import read_experiment
+from lib import read_experiment, YAXIS
 
 if len(sys.argv) != 7:
    print "not enough arguments"
@@ -46,7 +46,8 @@ pfailo = ax.bar(ind + width, opt_exp.get_fail_locks(), width, color=cmap(0.1))
 
 ptotalreg = ax.bar(ind + 2 * width, reg_exp.get_total_locks(), width, color=cmap(0.9), hatch=".")
 pfailreg = ax.bar(ind + 2 * width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
-ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.4])
+ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.5])
+#ax.set_ylim([0, YAXIS])
 ax.set_xticks(ind)
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
 ax.legend((ptotalc, ptotalo, pcoordc, pfailc, ptotalreg), ('Original Locks (Coordinated)', 'Original Locks (' + opt_name + ')', 'Coordination Locks', 'Coordination Locks (Failed)', 'Original Regular Locks'), loc=2)

### 2014-11-09 results/locks/plot_three.py
Update lock plots.

@@ -49,6 +49,8 @@ pfailreg = ax.bar(ind + 2 * width, reg_exp.get_fail_locks(), width, color=cmap(0
 ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.5])
 #ax.set_ylim([0, YAXIS])
 ax.set_xticks(ind)
+ax.get_yaxis().get_major_formatter().set_scientific(True)
+ax.get_yaxis().get_major_formatter().set_powerlimits((-3, 3))
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
 ax.legend((ptotalc, ptotalo, pcoordc, pfailc, ptotalreg), ('Original Locks (Coordinated)', 'Original Locks (' + opt_name + ')', 'Coordination Locks', 'Coordination Locks (Failed)', 'Original Regular Locks'), loc=2)
 plt.savefig(output_file)

### 2014-11-09 results/lib.py
Add scripts.

@@ -1,0 +1,381 @@
+
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from operator import truediv
+from matplotlib import rcParams
+from numpy import nanmax
+
+max_threads = 16
+
+def name2title(name):
+   table = {'greedy-graph-coloring-2000': "Greedy Graph Coloring (2000 nodes)",
+            'greedy-graph-coloring-search_engines': "Greedy Graph Coloring (Search Engines)",
+            'shortest-uspowergrid': "Shortest Distance (US Power Grid)",
+            'shortest-usairports500': "Shortest Distance (US 500 Airports)",
+            'shortest-celegans': "Shortest Distance (Celegans)",
+            'shortest-1000': "Shortest Distance (1000)",
+            'shortest-oclinks': "Shortest Distance (OCLinks)",
+            '8queens-13': "13 Queens",
+            '13queens': "13 Queens",
+            'tree': "Tree",
+            'tree-coord': "Tree (coordinated)",
+            'min-max-tictactoe': "MiniMax",
+            'min-max-tictactoe-coord': "MiniMax (coordinated)",
+            'belief-propagation-200': "Belief Propagation (200x200)",
+            'belief-propagation-400': "Belief Propagation (400x400)",
+            'splash-bp-400': "Belief Propagation (400x400 with Splashes)",
+            'graphlab': "GraphLab Belief Propagation (400x400)",
+            'pagerank-5000': "PageRank (5000 nodes)",
+            'pagerank-search_engines': "PageRank (Search Engines)",
+            'new-heat-transfer-120': "Heat Transfer (120x120)",
+            'new-heat-transfer-80': "Heat Transfer (80x80)"}
+   return table[name]
+
+def coordinated_program(name):
+   if name == 'belief-propagation-400':
+      return 'splash-bp-400'
+   return name + '-coord'
+
+class experiment_set(object):
+   def add_experiment(self, name, threads, time):
+      try:
+         exp = self.experiments[name]
+         exp.add_time(threads, time)
+      except KeyError:
+         exp = experiment(name)
+         self.experiments[name] = exp
+         exp.add_time(threads, time)
+
+   def experiment_names(self):
+      return self.experiments.keys()
+
[…]

### 2014-11-09 results/plot_compare.py

@@ -1,0 +1,44 @@
+#!/usr/bin/python
+
+import sys
+import math
+from lib import name2title, experiment_set, experiment
+
+if len(sys.argv) < 3:
+  print "Usage: plot_compare.py <filename> <output_prefix> [--global-only]"
+  sys.exit(1)
+
+filename = sys.argv[1]
+prefix = sys.argv[2]
+
+# read data.
+new = experiment_set()
+old = experiment_set()
+with open(filename, "r") as fp:
+   for line in fp:
+      line = line.rstrip("\n")
+      if line == "":
+         continue
+      vec = line.split(" ")
+      if len(vec) < 5:
+         continue
+      name = vec[0]
+      spec = vec[1]
+      time = int(vec[len(vec)-1])
+      if spec.startswith("th"):
+         threads = int(spec[2:])
+         new.add_experiment(name, threads, time)
+      else:
+         threads = int(spec[1:])
+         old.add_experiment(name, threads, time)
+
+if len(sys.argv) == 3:
+   # Generate the speedup graphs.
+   for exp_name in new.experiment_names():
+      newexp = new.get_experiment(exp_name)
+      oldexp = old.get_experiment(exp_name)
+      newexp.create_speedup_compare(prefix, oldexp)
+
+# Generate the histogram.
+new.create_histogram_compare(prefix, old, 1)
+new.create_histogram_compare(prefix, old, 16)

### 2014-11-09 results/plot_compare_system.py

@@ -1,0 +1,58 @@
+#!/usr/bin/python
+
+import sys
+import math
+from lib import name2title, experiment_set, experiment, coordinated_program
+
+if len(sys.argv) != 3:
+  print "Usage: plot_compare_system.py <filename> <output_prefix>"
+  sys.exit(1)
+
+filename = sys.argv[1]
+prefix = sys.argv[2]
+
+# read data.
+expset = experiment_set()
+with open(filename, "r") as fp:
+   for line in fp:
+      line = line.rstrip("\n")
+      if line == "":
+         continue
+      vec = line.split(" ")
+      if len(vec) < 3:
+         continue
+      name = vec[0]
+      spec = vec[1]
+      time = int(vec[len(vec)-1])
+      threads = int(spec[2:])
+      expset.add_experiment(name, threads, time)
+
+def name2othersystem(name):
+   if name == 'belief-propagation-400':
+      return 'graphlab'
+   if name == 'splash-bp-400':
+      return 'graphlab-coord'
+   return None
+
+# Generate the speedup graphs.
+for exp_name in expset.experiment_names():
+   lm = expset.get_experiment(exp_name)
+   othersystem = name2othersystem(exp_name)
+   if not othersystem:
+      continue
+   other = expset.get_experiment(othersystem)
+   if not other:
+      continue
+   title = None
+   if exp_name == 'belief-propagation-400':
+      title = 'LBP (400x400)'
+   if exp_name == 'splash-bp-400':
+      title = 'SBP (400x400)'
+   lm.create_speedup_compare_systems(prefix, other, 'GraphLab', title=title)
+   coord_lm = expset.get_experiment(coordinated_program(exp_name))
+   if not coord_lm:
+      continue
+   coord_other = expset.get_experiment(coordinated_program(othersystem))
+   if not coord_other:
+      continue
+   lm.create_comparison_coord_system(prefix, other, coord_lm, coord_other, 'GraphLab', title="SBP/LBP (400x400)")

### 2014-11-09 results/plot_coord.py

@@ -1,0 +1,44 @@
+#!/usr/bin/python
+
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+from lib import name2title, experiment_set, experiment, coordinated_program
+
+if len(sys.argv) != 3:
+  print "Usage: plot_coord.py <filename> <output_prefix>"
+  sys.exit(1)
+
+filename = sys.argv[1]
+prefix = sys.argv[2]
+max_threads = 16
+
+# read data.
+expset = experiment_set()
+with open(filename, "r") as fp:
+   for line in fp:
+      line = line.rstrip("\n")
+      if line == "":
+         continue
+      if line.startswith("#"):
+         continue
+      vec = line.split(" ")
+      if len(vec) < 3:
+         continue
+      name = vec[0]
+      threads = int(vec[1][2:])
+      time = int(vec[len(vec)-1])
+      expset.add_experiment(name, threads, time)
+
+for exp_name in expset.experiment_names():
+   if exp_name.endswith("-coord"):
+      continue
+   coord = expset.get_experiment(coordinated_program(exp_name))
+   if not coord:
+      continue
+   exp = expset.get_experiment(exp_name)
+   exp.create_coord(prefix, coord)

### 2014-11-09 results/plot_ht.py

@@ -1,0 +1,39 @@
+#!/usr/bin/python
+
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+from lib import name2title, experiment_set, experiment, coordinated_program
+
+if len(sys.argv) != 2:
+  print "Usage: plot_ht.py <filename>"
+  sys.exit(1)
+
+filename = sys.argv[1]
+max_threads = 16
+
+# read data.
+expset = experiment_set()
+with open(filename, "r") as fp:
+   for line in fp:
+      line = line.rstrip("\n")
+      if line == "":
+         continue
+      if line.startswith("#"):
+         continue
+      vec = line.split(" ")
+      if len(vec) < 3:
+         continue
+      name = vec[0]
+      threads = int(vec[1][2:])
+      time = int(vec[len(vec)-1])
+      expset.add_experiment(name, threads, time)
+
+reg = expset.get_experiment('new-heat-transfer-80')
+coord = expset.get_experiment('new-heat-transfer-80-coord0')
+localonly = expset.get_experiment('new-heat-transfer-80-coord1')
+reg.create_ht(coord, localonly)

### 2014-11-10 results/locks/lib.py
Update pictures.

@@ -3,7 +3,7 @@ import re
 YAXIS = 200000000
 
 class dataobj(object):
-   ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
+   COORD_ATTRS = ['coord_normal_lock', 'coord_priority_lock', 'schedule_next_lock', 'add_priority_lock', 'set_priority_lock', 'set_moving_lock', 'set_static_lock', 'set_affinity_lock']
 
    def set(self, name, fail, total):
       self.info_fail[name] = fail
@@ -18,14 +18,17 @@ class dataobj(object):
    def total_locks(self):
       return sum(total for name, total in self.info_total.iteritems())
 
+   def total_basic_locks(self):
+      return self.total_locks() - self.total_coord_locks()
+
    def total_fail(self):
       return sum(total for name, total in self.info_fail.iteritems())
 
    def total_coord_locks(self):
-      return sum(self.get_total(name) for name in self.ATTRS)
+      return sum(self.get_total(name) for name in self.COORD_ATTRS)
 
    def fail_coord_locks(self):
-      return sum(self.get_fail(name) for name in self.ATTRS)
+      return sum(self.get_fail(name) for name in self.COORD_ATTRS)
 
    def frac_coord_locks(self):
       return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
@@ -68,6 +71,9 @@ class experiment(object):
 
    def get_total_locks(self):
       return [data.total_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def get_basic_locks(self):
+      return [data.total_basic_locks() for th, data in sorted(self.threads.iteritems())]
 
    def get_threads(self):
       return [th for th in sorted(self.threads.keys())]

### 2014-11-10 results/locks/plot.py

@@ -28,14 +28,15 @@ ax = fig.add_subplot(111)
 ax.set_ylabel('Number of Locks')
 ax.set_xlabel('Threads')
 ax.set_title(title)
-ptotal = ax.bar(ind, coord_exp.get_total_locks(), width, color=cmap(0.9), hatch='/')
-pcoord = ax.bar(ind, coord_exp.get_coord_total_locks(), width, color=cmap(0.5))
-pfail = ax.bar(ind, coord_exp.get_fail_locks(), width, color=cmap(0.1))
-ptotalreg = ax.bar(ind + width, reg_exp.get_total_locks(), width, color=cmap(0.9), hatch=".")
-pfailreg = ax.bar(ind + width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
+
+pbasicreg = ax.bar(ind, reg_exp.get_basic_locks(), width, color=cmap(0.9), hatch=".")
+
+pcoordbasic = ax.bar(ind + width, coord_exp.get_basic_locks(), width, color=cmap(0.9), hatch='/')
+pcoordcoord = ax.bar(ind + width, coord_exp.get_coord_total_locks(), width, color=cmap(0.5), bottom=coord_exp.get_basic_locks())
+
 ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.4])
 #ax.set_ylim([0, YAXIS])
 ax.set_xticks(ind)
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
-ax.legend((ptotal, pcoord, pfail, ptotalreg), ('Original Locks', 'Coordination Locks', 'Coordination Locks (Failed)', 'Original Regular Locks'), loc=2)
+ax.legend((pbasicreg, pcoordbasic, pcoordcoord), ('Regular - Basic', 'Coordinated - Basic', 'Coordinated - Coordination'), loc=2)
 plt.savefig(output_file)

### 2014-11-10 results/locks/plot_three.py

@@ -36,21 +36,20 @@ ax.set_ylabel('Number of Locks')
 ax.set_xlabel('Threads')
 ax.set_title(title)
 
-ptotalc = ax.bar(ind, coord_exp.get_total_locks(), width, color=cmap(0.9), hatch='/')
-pcoordc = ax.bar(ind, coord_exp.get_coord_total_locks(), width, color=cmap(0.5))
-pfailc = ax.bar(ind, coord_exp.get_fail_locks(), width, color=cmap(0.1))
+pbasicreg = ax.bar(ind, reg_exp.get_basic_locks(), width, color=cmap(0.9), hatch=".")
 
-ptotalo = ax.bar(ind + width, opt_exp.get_total_locks(), width, color=cmap(0.9), hatch='\\')
-pcoordo = ax.bar(ind + width, opt_exp.get_coord_total_locks(), width, color=cmap(0.5))
-pfailo = ax.bar(ind + width, opt_exp.get_fail_locks(), width, color=cmap(0.1))
+pbasicc = ax.bar(ind + width, coord_exp.get_basic_locks(), width, color=cmap(0.9), hatch='/')
+pcoordc = ax.bar(ind + width, coord_exp.get_coord_total_locks(), width, color=cmap(0.5), bottom=coord_exp.get_basic_locks())
 
-ptotalreg = ax.bar(ind + 2 * width, reg_exp.get_total_locks(), width, color=cmap(0.9), hatch=".")
-pfailreg = ax.bar(ind + 2 * width, reg_exp.get_fail_locks(), width, color=cmap(0.5))
+pbasico = ax.bar(ind + 2 * width, opt_exp.get_basic_locks(), width, color=cmap(0.9), hatch='\\')
+pcoordo = ax.bar(ind + 2 * width, opt_exp.get_coord_total_locks(), width, color=cmap(0.5), bottom=opt_exp.get_basic_locks())
+
 ax.set_ylim([0, max([max(reg_exp.get_total_locks()), max(coord_exp.get_total_locks())])*1.5])
 #ax.set_ylim([0, YAXIS])
 ax.set_xticks(ind)
 ax.get_yaxis().get_major_formatter().set_scientific(True)
 ax.get_yaxis().get_major_formatter().set_powerlimits((-3, 3))
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
[…]

### 2014-11-10 heat-transfer.tex
Add locking section.

@@
 In the Heat Transfer algorithm~(HT), we have a graph where heat values are
-exchanged between nodes. XXX rewrite The program stops when the new heat values of every node are
+exchanged between nodes. The program stops when the new heat values of every node are
 $\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
@@
 Delta > epsilon
-   -o {B, W | !edge(A, B, W) |
+   -o {B | !edge(A, B) |
          new-neighbor-heat(B, A, New),
@@
 fabs(New - Old) <= epsilon
-   -o {B, W | !edge(A, B, W) |
+   -o {B | !edge(A, B) |
          new-neighbor-heat(B, A, New)}.
@@
 \underline[cpu-id(A, A, C)]
-   -o {B, W | !edge(A, B, W) |
+   -o {B, W | !edge(A, B) |
          new-neighbor-heat(B, A, New)}.

### 2014-11-10 implementation.tex
fix ie
Add locking section.

@@
 
-\subsubsection{Coordination directives}
-
+\paragraph{Coordination directives}
 Coordination directives are compiled in two different ways, depending on whether they
@@
 
-\subsubsection{Nodes}
-
+\paragraph{Nodes} 
 A node is represented as a collection of facts (per predicate) and an indexing structure that
@@
 
-\subsubsection{Threads}
+\paragraph{Threads}
 
@@
 another active node.  If its queues are all empty it will try to steal
-work (\ie nodes) from another thread.
+work (i.e., nodes) from another thread.
 
@@
 
-\subsubsection{Communication}
+\paragraph{Communication}
 
@@
 
-\subsubsection{Garbage collection}
+\paragraph{Garbage collection}
 
@@
 
-\subsubsection{Coordination overhead}
+\paragraph{Coordination overhead}
 

### 2014-11-10 locks.tex
Merge HT locks plots into one.
Update figures with SSSP.
Add locking section.

@@
-\begin{topfig}
-   \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
-   \end{center}
-\end{topfig}
+\paragraph{Locking} The CLM virtual machine uses spinlocks to protect shared data structures between
+threads. As described in Section~\ref{sec:implementation}, coordination
+operations require synchronization since they need to change node information
+and sometimes move nodes between the standard queue and the regular queue.
+In order to measure the locking costs of coordination, we measured the following
+statistics: \emph{basic locks}, the number of locks acquired that were already
+present in the original LM virtual machine and \emph{coordination locks}, the
+number of locks acquired for performing coordination operations. We also
+measured how often a \texttt{lock()} operation fails on the first attempt, but
+such statistics are not represented in the figures below since the values are
+too small.
 
-\begin{topfig}
-   \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks-local-only.png}}
-   \end{center}
-\end{topfig}
-
-\begin{topfig}
-   \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/minmax-locks.png}}
-   \end{center}
-\end{topfig}
+The first program we measured was SSSP, represented in
+Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
+Coordinated versions, we have the Buffered version where we
+optimized the virtual machine to buffer coordination actions before applying
+them. The idea is to avoid applying multiple \texttt{set-priority} operations to
+the same node during a single node execution. We thus gather the higher priority
+values and then perform coordination after the node has executed.
+Although the number of locks is significantly reduced, we did not see a
+similar performance improvement, since the Buffered version is only
+5\% faster than the Coordinated version.
 
@@
    \end{center}
+  \scap{locks:SSSP}{Locking statistics for SSSP.
+     \texttt{Coordinated} refers to the program in
+     Fig.~\ref{results:sssp_uspowergrid}, while \texttt{Buffered} is a more
+     optimized version of the same program where coordination operations are
+     buffered before application.}
 \end{topfig}
+
+In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
+regular version, the coordinated version and the local-only version as in
+Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
+coordination and then a further reduction in the local-only version because less
+facts are exchanged between threads.
+
+\begin{topfig}
+   \begin{center}
+      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
+   \end{center}
+  \scap{locks:HT}{Locking statistics for HT. Coordination reduces
+     the overall number of locks used, while further optimization for locality
+     reduces the locks even further.}
+\end{topfig}
+
+Finally, in Fig.~\ref{locks:LBP} we present lock statistics for LBP and SBP. LBP
+is represented as the Regular bar, while SBP is represented as the Coordinated
+bar. We notice that LBP has quite a lot of locking for 1 thread since LBP
+is an asynchronous algorithm and tends to be a slow sequential algorithm,
+requiring more derivations when using 1 thread. The numbers go down as more
+threads are used. For SBP, there is a consistent amount of locking without
+little variations between the number of threads used. This is because threads
+interactly very little with each other and the amount of work to do is the same
+for all threads.
 
@@
    \end{center}
+  \scap{locks:LBP}{Locking statistics for LBP and SBP. Locking is constant for
+     SBP because threads do not interact with each other.}
 \end{topfig}
+
+These experiments indicate that handling coordination has a small cost that is
+greatly offset by improved performance. In general, coordination reduces the
+amount of locking required in the overall program because the number of facts
+needed to be derived is reduced. Even in cases where the amount of locking
+increases, it is still better to use coordination due to less synchronization
+between threads.

### 2014-11-10 results/locks/average_queue_operations.py
Added to script to compute average number of queue OPS.

@@ -1,0 +1,18 @@
+#!/usr/bin/python
+
+import re
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+from lib import read_experiment, YAXIS
+
+filename = sys.argv[1]
+
+exp = read_experiment(filename)
+threads = exp.get_threads()
+for ops in exp.queue_instruction_count():
+   print ops

### 2014-11-10 results/locks/lib.py

@@ -1,4 +1,5 @@
 import re
+import math
 
 YAXIS = 200000000
 
@@ -8,6 +9,9 @@ class dataobj(object):
    def set(self, name, fail, total):
       self.info_fail[name] = fail
       self.info_total[name] = total
+
+   def set_average_priority_queue(self, avg):
+      self.avg_prio_queue = avg
 
    def get_total(self, name):
       return self.info_total[name]
@@ -30,6 +34,25 @@ class dataobj(object):
    def fail_coord_locks(self):
       return sum(self.get_fail(name) for name in self.COORD_ATTRS)
 
+   def normal_queue_operations(self):
+      return self.get_total('normal_lock')
+
+   def priority_queue_operations(self):
+      return self.get_total('priority_lock')
+
+   def average_priority_queue(self):
+      try:
+         return self.get_total('average_priority')
+      except KeyError:
+         return 0
+
+   def queue_instruction_count(self):
+      normal = self.normal_queue_operations()
+      prio = self.priority_queue_operations()
+      if self.average_priority_queue() == 0:
+         return normal
+      return normal + int(prio * math.log(float(self.average_priority_queue()), 2))
+
    def frac_coord_locks(self):
       return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
 
@@ -45,6 +68,7 @@ class dataobj(object):
       self.threads = th
       self.info_total = {}
       self.info_fail = {}
+      self.avg_prio_queue = 0
 
 class experiment(object):
 
@@ -74,6 +98,15 @@ class experiment(object):
 
    def get_basic_locks(self):
       return [data.total_basic_locks() for th, data in sorted(self.threads.iteritems())]
+
+   def normal_queue_operations(self):
+      return [data.normal_queue_operations() for th, data in sorted(self.threads.iteritems())]
+
+   def priority_queue_operations(self):
+      return [data.priority_queue_operations() for th, data in sorted(self.threads.iteritems())]
+
+   def queue_instruction_count(self):
+      return [data.queue_instruction_count() for th, data in sorted(self.threads.iteritems())]
 
    def get_threads(self):
       return [th for th in sorted(self.threads.keys())]
@@ -109,9 +142,16 @@ def read_experiment(filename):
          name = vec[0]
          res = vec[1].lstrip(" ")
          resvec = re.split('\s+', res)
-         fail = int(resvec[0])
-         total = int(resvec[2])
-         data.set(name, fail, total)
+         if len(resvec) == 3:
[…]

### 2014-11-11 locks.tex
Update paper with new table.

@@
 
-These experiments indicate that handling coordination has a small cost that is
-greatly offset by improved performance. In general, coordination reduces the
-amount of locking required in the overall program because the number of facts
-needed to be derived is reduced. Even in cases where the amount of locking
-increases, it is still better to use coordination due to less synchronization
-between threads.
+We also measured the number of queue instructions required to compute programs.
+While the regular queue operations have constant complexity, priority queue
+operations are usually logarithmic and require more computation. We counted the
+number of instructions required for queue operations during program execution.
+Each normal queue operation costs 1 instruction, while priority queue costs as
+many \emph{percolate-up}/\emph{percolate-down} instructions are required to maintain the heap property.
+
+Table~\ref{queue_instructions} presents the total queue instructions for several
+programs. All the coordinated programs require far more queue instructions than
+the regular version. For instance, the coordinated SSSP program requires 3 to 4
+times more instructions to perform queue operations. Interestingly, the buffered
+version requires even more instructions although it performs less overall
+coordination. It is likely that this is due to increased fact computation and
+subsequent queue manipulation. We argue that the 5\% performance improvement in the
+buffered version is due mostly to reduced locking.
+
+For the HT program, the coordinated version requires twice more queue
+instructions than the regular version. As expected, the local-only version
+requires less queue operations because the number of derived facts is reduced.
+
+\begin{topfig}
+\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
+\hline
+\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
+   Instructions}} \\ \cline{2-6}
+& 1 & 2 & 4 & 8 & 16 \\ \hline
+\hline
+SSSP - Regular & 298K & 305K & 309K & 286K & 325K \\ \hline
+SSSP - Coordinated & 961K & 919K & 903K & 900K & 922K \\ \hline
+SSSP - Buffered & 939K & 961K & 999K & 926K & 1M \\ \hline
+\hline
+HT - Buffered & 38M & 49M & 50M & 52M & 54M \\ \hline
+HT - Coordinated & 57M & 63M & 95M & 89M & 82M \\ \hline
+HT - Local-Only & 57M & 63M & 91M & 80M & 74M \\ \hline
+\hline
+LBP - Regular & 23M & 26M & 24M & 17M & 12M \\ \hline
+SBP - Coordinated & 95M & 93M & 88M & 81M & 75M \\ \hline
+\hline
+MiniMax - Regular & 24M & 26M & 27M & 26M & 27M \\ \hline
+MiniMax - Coordinated & 51M & 51M & 52M & 51M & 52M \\ \hline
+\end{tabular}
+\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
+   operation to a regular queue counts as 1 instruction, while an operation to
+   a priority queue costs the number of
+   \emph{percolate-up}/\emph{percolate-down} steps performed.}
+\end{topfig}
+
+These experiments indicate that handling coordination has a small cost in terms
+of synchronization but some cost in terms of queue operations with
+up to 5 times more queue instructions when using priority queues.
+Since coordination reduces the number of facts derived due to better scheduling
+decisions, the costs of manipulating priority queues become irrelevant.

### 2014-11-11 results/locks/average_queue_operations.py

@@ -1,18 +1,0 @@
-#!/usr/bin/python
-
-import re
-import sys
-import math
-import matplotlib.pyplot as plt
-import matplotlib as mpl
-import numpy as np
-from matplotlib import rcParams
-from numpy import nanmax
-from lib import read_experiment, YAXIS
-
-filename = sys.argv[1]
-
-exp = read_experiment(filename)
-threads = exp.get_threads()
-for ops in exp.queue_instruction_count():
-   print ops

### 2014-11-11 results/locks/lib.py

@@ -10,8 +10,8 @@ class dataobj(object):
       self.info_fail[name] = fail
       self.info_total[name] = total
 
-   def set_average_priority_queue(self, avg):
-      self.avg_prio_queue = avg
+   def set_heap_operations(self, ops):
+      self.heap_operations = ops
 
    def get_total(self, name):
       return self.info_total[name]
@@ -40,18 +40,9 @@ class dataobj(object):
    def priority_queue_operations(self):
       return self.get_total('priority_lock')
 
-   def average_priority_queue(self):
-      try:
-         return self.get_total('average_priority')
-      except KeyError:
-         return 0
-
    def queue_instruction_count(self):
       normal = self.normal_queue_operations()
-      prio = self.priority_queue_operations()
-      if self.average_priority_queue() == 0:
-         return normal
-      return normal + int(prio * math.log(float(self.average_priority_queue()), 2))
+      return normal + self.heap_operations
 
    def frac_coord_locks(self):
       return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
@@ -68,7 +59,7 @@ class dataobj(object):
       self.threads = th
       self.info_total = {}
       self.info_fail = {}
-      self.avg_prio_queue = 0
+      self.heap_operations = 0
 
 class experiment(object):
 
@@ -147,11 +138,8 @@ def read_experiment(filename):
             total = int(resvec[2])
             data.set(name, fail, total)
          else:
-            if resvec[0] == '-nan':
-               total = 0
-            else:
-               total = int(float(resvec[0]))
-            data.set_average_priority_queue(total)
+            total = int(resvec[0])
+            data.set_heap_operations(total)
    if data:
       exp.add_thread(thread, data)
    return exp

### 2014-11-11 results/locks/queue_operations.py

@@ -1,0 +1,53 @@
+#!/usr/bin/python
+
+import re
+import sys
+import math
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import numpy as np
+from matplotlib import rcParams
+from numpy import nanmax
+from lib import read_experiment, YAXIS
+
+sssp_reg = read_experiment("sssp.txt")
+sssp_coord = read_experiment("sssp-coord.txt")
+sssp_buffered = read_experiment("sssp-buffer.txt")
+ht_reg = read_experiment("ht-reg.txt")
+ht_coord = read_experiment("ht-coord.txt")
+ht_local_only = read_experiment("ht-local-only.txt")
+lbp = read_experiment("bp.txt")
+sbp = read_experiment("sbp.txt")
+minmax_reg = read_experiment("minmax-reg.txt")
+minmax_coord = read_experiment("minmax-coord.txt")
+
+def translate(num):
+   if num > 1000000:
+      return str(num/1000000) + "M"
+   elif num > 1000:
+      return str(num/1000) + "K"
+   else:
+      return str(num)
+
+def show(exp, name):
+   count = exp.queue_instruction_count()
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+
+show(sssp_reg, "SSSP - Regular")
+show(sssp_coord, "SSSP - Coordinated")
+show(sssp_buffered, "SSSP - Buffered")
+print "\hline"
+
+show(ht_reg, "HT - Buffered")
+show(ht_coord, "HT - Coordinated")
+show(ht_local_only, "HT - Local-Only")
+
+print "\hline"
+
+show(lbp, "LBP - Regular")
+show(sbp, "SBP - Coordinated")
+
+print "\hline"
+
+show(minmax_reg, "MiniMax - Regular")
+show(minmax_coord, "MiniMax - Coordinated")

### 2014-11-11 results/locks/queue_operations.py
Update benchmarks.

@@ -20,6 +20,8 @@ lbp = read_experiment("bp.txt")
 sbp = read_experiment("sbp.txt")
 minmax_reg = read_experiment("minmax-reg.txt")
 minmax_coord = read_experiment("minmax-coord.txt")
+nqueens_reg = read_experiment("nqueens-reg.txt")
+nqueens_coord = read_experiment("nqueens-coord.txt")
 
 def translate(num):
    if num > 1000000:
@@ -51,3 +53,8 @@ print "\hline"
 
 show(minmax_reg, "MiniMax - Regular")
 show(minmax_coord, "MiniMax - Coordinated")
+
+print "\hline"
+
+show(nqueens_reg, "N Queens - Regular")
+show(nqueens_coord, "N Queens - Coordinated")

### 2014-11-11 locks.tex
Update Related Work.

@@
 \hline
-SSSP - Regular & 298K & 305K & 309K & 286K & 325K \\ \hline
-SSSP - Coordinated & 961K & 919K & 903K & 900K & 922K \\ \hline
-SSSP - Buffered & 939K & 961K & 999K & 926K & 1M \\ \hline
+SSSP - Regular & 298K & 300K & 313K & 284K & 317K \\ \hline
+SSSP - Coordinated & 953K & 960K & 861K & 911K & 987K \\ \hline
+SSSP - Buffered & 951K & 1M & 1M & 1M & 1M \\ \hline
 \hline
-HT - Buffered & 38M & 49M & 50M & 52M & 54M \\ \hline
-HT - Coordinated & 57M & 63M & 95M & 89M & 82M \\ \hline
-HT - Local-Only & 57M & 63M & 91M & 80M & 74M \\ \hline
+HT - Buffered & 38M & 49M & 52M & 74M & 55M \\ \hline
+HT - Coordinated & 57M & 64M & 95M & 87M & 83M \\ \hline
+HT - Local-Only & 57M & 63M & 90M & 85M & 77M \\ \hline
 \hline
 LBP - Regular & 23M & 26M & 24M & 17M & 12M \\ \hline
-SBP - Coordinated & 95M & 93M & 88M & 81M & 75M \\ \hline
+SBP - Coordinated & 96M & 92M & 90M & 78M & 74M \\ \hline
 \hline
-MiniMax - Regular & 24M & 26M & 27M & 26M & 27M \\ \hline
-MiniMax - Coordinated & 51M & 51M & 52M & 51M & 52M \\ \hline
+MiniMax - Regular & 24M & 25M & 26M & 26M & 27M \\ \hline
+MiniMax - Coordinated & 51M & 52M & 52M & 52M & 52M \\ \hline
+\hline
+N Queens - Regular & 12K & 24K & 25K & 365K & 5M \\ \hline
+N Queens - Coordinated & 57M & 11M & 904K & 145K & 44K \\ \hline
 \end{tabular}

### 2014-11-11 related.tex
More updates.
Update Related Work.

@@
-There has been an increasing interest in declarative 
+\paragraph{Declarative Languages}
+Declarative programming paradigms such as \emph{logic} and \emph{functional}
+programming have been extensively exploited for their implicit parallelism.
+In logic languages such as Prolog, researchers took advantage of the
+non-determinism of proof-search to evaluate subgoals in parallel with models
+such as \emph{or-parallelism}~\cite{ali-86} and
+\emph{and-parallelism}~\cite{Shen-92}.  In functional languages, the stateless
+nature of computation allows multiple expressions to evaluate safely in
+parallel. This has been explored in several languages such as
+NESL~\cite{Blelloch:1996:PPA:227234.227246}, Id~\cite{Nikhil93anoverview}, and
+more recently Data Parallel Haskell~\cite{nepal}.
+
+\paragraph{Data Centric Languages}
+Recently, there has been an increasing interest in declarative data-centric
 languages.  MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
 popular data-centric programming model that is optimized for large clusters.
-Intrinsic to its popularity is the simplicity of its scheduling and data sharing model.
-In order to facilitate the writing of programs over large
-datasets, SQL-like languages such as
-PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
-PigLatin builds on top of MapReduce and allows the programmer to write complex
-data-flow graphs, raising the abstraction and easy of programmability of
-MapReduce programs.
-An alternative to PigLatin/MapReduce is
-Dryad~\cite{Isard:2007:DDD:1272996.1273005}
-that allows programmers to design arbitrary computation patterns using the DAG
-abstraction. It combines computational vertices with communication channels
-(edges). Dryad programs are automatically scheduled to run on multiple
-computers/cores and data is partitioned automatically during runtime.
+Intrinsic to its popularity is the simplicity of its scheduling and data sharing
+model.  In order to facilitate the writing of programs over large datasets,
+SQL-like languages such as PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have
+been developed.  PigLatin builds on top of MapReduce and allows the programmer
+to write complex data-flow graphs, raising the abstraction and easy of
+programmability of MapReduce programs.  An alternative to PigLatin/MapReduce is
+Dryad~\cite{Isard:2007:DDD:1272996.1273005} that allows programmers to design
+arbitrary computation patterns using the DAG abstraction. It combines
+computational vertices with communication channels (edges) that are
+automatically scheduled to run on multiple computers/cores.
 
+GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
+learning algorithms. GraphLab allows nodes to have read/write access to
+different scopes through different concurrent access models in order to balance
+performance and data consistency.  GraphLab also provides different schedulers
+that dictate the order in which node's are computed, allowing the programmer to
+optimize the program. Later in this paper, we will show how one GraphLab
+scheduler can be implemented in CLM through the use of coordination facts.
+
+\paragraph{Coordination Languages}
 Many programming languages follow the so-called \emph{coordination
@@
 \texttt{out(t)} to write a tuple \texttt{t} into the tuple space; \texttt{in(t)}
-to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy of
-the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a process
-\texttt{p} in the tuple space and execute it in parallel. 
-Linda is be implemented on top of many
-popular languages by simply creating a communication and storage mechanism for
-the tuple space and then adding the directives as a language library.
+to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy
+of the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a
+process \texttt{p} in the tuple space and execute it in parallel.  Linda is be
+implemented on top of many popular languages by simply creating a communication
+and storage mechanism for the tuple space and then adding the directives as a
+language library.
 
-Another early coordination language is Delirium~\cite{Delirium}. Unlike
-Linda which is embbeded into another language, Delirium actually embeds
-operators written in other languages inside the Delirium language.
-The advantages of Delirium are improved abstraction
-and easier debugging because sequential operators are isolated from the
-coordination language.
+Another early coordination language is Delirium~\cite{Delirium}. Unlike Linda
+which is embbeded into another language, Delirium actually embeds operators
+written in other languages inside the Delirium language.  The advantages of
+Delirium are improved abstraction and easier debugging because sequential
+operators are isolated from the coordination language.
 
@@
 the program instead of focusing how processing units are used.
-Furthermore, the LM language is both a coordination language and a computation
+Furthermore, the CLM language is both a coordination language and a computation
 language and there is no distinction between the two components.
+
+\paragraph{Forward-Chaining Logic Programming}
 
@@
 nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
-The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
-programming distributed emsembles of modular robots. Meld introduced the idea of
-sensing and action facts in order to sense and act on the outside world,
-respectively. LCM uses sensing facts to sense the underlying system that the program is running on and action facs
-act on that system.
+The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a
+Datalog-like language for programming distributed emsembles of modular robots.
+Meld introduced the idea of sensing and action facts in order to sense and act
+on the outside world, respectively. LCM uses sensing facts to sense the
+underlying system that the program is running on and action facs act on that
+system.

### 2014-11-11 locks.tex
Change N Queens section.

--- 2014-11-11 locks.tex
@@
 
+\paragraph{Queue Instructions}
+
 We also measured the number of queue instructions required to compute programs.
@@
 programs. All the coordinated programs require far more queue instructions than
-the regular version. For instance, the coordinated SSSP program requires 3 to 4
+the regular version. For instance, the coordinated SSSP program requires 4 to 5
 times more instructions to perform queue operations. Interestingly, the buffered
@@
 coordination. It is likely that this is due to increased fact computation and
-subsequent queue manipulation. We argue that the 5\% performance improvement in the
+subsequent queue manipulation because coordination operations are buffered.
+We argue that the 5\% performance improvement in the
 buffered version is due mostly to reduced locking.
@@
 requires less queue operations because the number of derived facts is reduced.
+The MiniMax program also sees twices as many queue instructions using
+coordination. This does not happen with the N Queens program, where the number
+of queue instructions is far greater when using coordination, especially when
+using a small number of threads. We argue that this happens because higher
+priority nodes at the bottom are often added to the priority queue, resulting in
+many \emph{percolate-up} operations.
 
@@
 \hline
-SSSP - Regular & 298K & 305K & 309K & 286K & 325K \\ \hline
-SSSP - Coordinated & 961K & 919K & 903K & 900K & 922K \\ \hline
-SSSP - Buffered & 939K & 961K & 999K & 926K & 1M \\ \hline
+SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\ \hline
+SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\ \hline
+SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\ \hline
 \hline
-HT - Buffered & 38M & 49M & 50M & 52M & 54M \\ \hline
-HT - Coordinated & 57M & 63M & 95M & 89M & 82M \\ \hline
-HT - Local-Only & 57M & 63M & 91M & 80M & 74M \\ \hline
+HT - Buffered & 25M & 32M & 34M & 34M & 36M \\ \hline
+HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\ \hline
+HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\ \hline
 \hline
-LBP - Regular & 23M & 26M & 24M & 17M & 12M \\ \hline
-SBP - Coordinated & 95M & 93M & 88M & 81M & 75M \\ \hline
+LBP - Regular & 15M & 17M & 16M & 11M & 8M \\ \hline
+SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\ \hline
 \hline
-MiniMax - Regular & 24M & 26M & 27M & 26M & 27M \\ \hline
-MiniMax - Coordinated & 51M & 51M & 52M & 51M & 52M \\ \hline
+MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\ \hline
+MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\ \hline
+\hline
+N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
+N Queens - Coordinated & 57M & 18M & 676K & 125K & 41K \\ \hline
 \end{tabular}

### 2014-11-11 nqueens.tex
use new scap and fig macros
Further updates.
Change N Queens section.

@@
 matter much if we start with a bad placement since node stealing overcomes those
-issues by improving load balancing dynamically. But what if we want run the
-program on a distributed system, where moving nodes between workers can be
-expensive? We measured the performance of the program without node stealing in
-order to assess how important data placement would be for such systems.
-We used the default partition of the program as it is, which results in the board being
-split by rows and another configuration using the coordination action
-\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
-thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
-the coordinates of the node in the chessboard.
+issues by improving load balancing dynamically. However, we noticed interesting
+cache effects when prioritizing nodes in bottom rows. The N Queens program
+builds and shares lists representing valid board states and needs to iterate
+over those lists often when checking if a new position does not invalidate a state.
+We thus used coordination axioms \texttt{set-default-priority(A, X)} and \texttt{set-cpu(A,
+vertical(X, Y)}, where \texttt{X} is the row number (0 is the top row) and
+\texttt{Y} is the column number.
 
-Figure~\ref{results:nqueens} shows the performance results of coordinating the
-placing of nodes on threads without work stealing. There is a noticiable
-performance improvement when using an optimal data placement strategy. Note
-however that in the plot we do not take into account the costs of moving
-nodes between workers. In a real system, the performance gap may not be as wide
-since in the worst case there may be 144 moves.
+We first ran the coordinated program using dynamic partitioning (with work
+stealing) and saw some performance improvements. However, using dynamic
+partitioning is bad for memory locality since threads will steal nodes often.
+We then added \texttt{set-static(A)} as a coordination axiom and saw
+performance improvement, especially when the board columns are perfectly
+partitioned by the number of threads (in our case 13 threads).
+Figure~\ref{results:nqueens} shows the performance gains saw for the 13 Queens
+program.
 
-\begin{figure}[ht!]
+\begin{topfig}
    \begin{center}
-      \subfloat[]{\includegraphics[width=5cm]{results/13queens.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/13queens.png}}
    \end{center}
-   \caption{Experimental results for the 13 Queens program. The coordinated
-      version is on average 2 times faster than the version with bad data
-      placement.}
-   \label{results:nqueens}
-\end{figure}
+   \scap{results:nqueens}{Experimental results for the 13 Queens program.}
+\end{topfig}

### 2014-11-11 results/locks/lib.py

@@ -13,6 +13,9 @@ class dataobj(object):
    def set_heap_operations(self, ops):
       self.heap_operations = ops
 
+   def set_normal_operations(self, ops):
+      self.normal_operations = ops
+
    def get_total(self, name):
       return self.info_total[name]
 
@@ -41,8 +44,7 @@ class dataobj(object):
       return self.get_total('priority_lock')
 
    def queue_instruction_count(self):
-      normal = self.normal_queue_operations()
-      return normal + self.heap_operations
+      return self.normal_operations + self.heap_operations
 
    def frac_coord_locks(self):
       return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
@@ -60,6 +62,7 @@ class dataobj(object):
       self.info_total = {}
       self.info_fail = {}
       self.heap_operations = 0
+      self.normal_operations = 0
 
 class experiment(object):
 
@@ -139,7 +142,10 @@ def read_experiment(filename):
             data.set(name, fail, total)
          else:
             total = int(resvec[0])
-            data.set_heap_operations(total)
+            if name.startswith("normal_operations"):
+               data.set_normal_operations(total)
+            else:
+               data.set_heap_operations(total)
    if data:
       exp.add_thread(thread, data)
    return exp

### 2014-11-11 nqueens.tex
Update 8 Queens section.

@@
 We thus used coordination axioms \texttt{set-default-priority(A, X)} and \texttt{set-cpu(A,
-vertical(X, Y)}, where \texttt{X} is the row number (0 is the top row) and
+vertical(X, Y))}, where \texttt{X} is the row number (0 is the top row) and
 \texttt{Y} is the column number.
 
+Experimenta results are presented in Fig.~\ref{results:nqueens}.
 We first ran the coordinated program using dynamic partitioning (with work
-stealing) and saw some performance improvements. However, using dynamic
+stealing) and saw some performance improvements (\textbf{Coordinated}). However, using dynamic
 partitioning is bad for memory locality since threads will steal nodes often.
 We then added \texttt{set-static(A)} as a coordination axiom and saw
-performance improvement, especially when the board columns are perfectly
+performance improvement (\textbf{Static}), especially when the board columns are perfectly
 partitioned by the number of threads (in our case 13 threads).
-Figure~\ref{results:nqueens} shows the performance gains saw for the 13 Queens
-program.
 
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/13queens.png}}
+      \subfloat[]{\includegraphics[width=5.5cm]{results/8queens-13.png}}
    \end{center}
-   \scap{results:nqueens}{Experimental results for the 13 Queens program.}
+   \scap{results:nqueens}{Experimental results for the 13 Queens program.
+   Coordination works best when 1 column is assigned to one thread and bottom
+   rows have higher priority.}
 \end{topfig}

### 2014-11-11 results/lib.py

@@ -152,7 +152,7 @@ class experiment(object):
       name = prefix + self.name + ".png"
       plt.savefig(name)
 
-   def create_ht(self, coord, localonly):
+   def create_ht(self, coord, localonly, optimized):
       fig = plt.figure()
       ax = fig.add_subplot(111)
 
@@ -183,7 +183,7 @@ class experiment(object):
       ax.plot(self.x_axis(), self.linear_speedup(),
         label='Linear', linestyle='-', color=cmap(0.2))
       ax.legend([reg, coordcoord, coordreg, loreg], ["Regular/Regular",
-            "Coordinated/Coordinated", "Coordinated/Regular", "Local-Only/Regular"], loc=2, fontsize=18, markerscale=2)
+            "Coordinated/Coordinated", "Coordinated/Regular", optimized+"/Regular"], loc=2, fontsize=18, markerscale=2)
 
       setup_lines(ax, cmap)
 

### 2014-11-11 results/plot_queens.py

@@ -10,7 +10,7 @@ from numpy import nanmax
 from lib import name2title, experiment_set, experiment, coordinated_program
 
 if len(sys.argv) != 2:
-  print "Usage: plot_ht.py <filename>"
+  print "Usage: plot_queens.py <filename>"
   sys.exit(1)
 
 filename = sys.argv[1]
@@ -33,7 +33,7 @@ with open(filename, "r") as fp:
       time = int(vec[len(vec)-1])
       expset.add_experiment(name, threads, time)
 
-reg = expset.get_experiment('new-heat-transfer-80')
-coord = expset.get_experiment('new-heat-transfer-80-coord0')
-localonly = expset.get_experiment('new-heat-transfer-80-coord1')
-reg.create_ht(coord, localonly, "Local-Only")
+reg = expset.get_experiment('8queens-13')
+coord0 = expset.get_experiment('8queens-13-coord0')
+coord1 = expset.get_experiment('8queens-13-coord1')
+reg.create_ht(coord0, coord1, "Static")

### 2014-11-11 results/locks/queue_operations.py
Add N Queens Static to queue table.

@@ -22,6 +22,7 @@ minmax_reg = read_experiment("minmax-reg.txt")
 minmax_coord = read_experiment("minmax-coord.txt")
 nqueens_reg = read_experiment("nqueens-reg.txt")
 nqueens_coord = read_experiment("nqueens-coord.txt")
+nqueens_static = read_experiment("nqueens-static.txt")
 
 def translate(num):
    if num > 1000000:
@@ -58,3 +59,4 @@ print "\hline"
 
 show(nqueens_reg, "N Queens - Regular")
 show(nqueens_coord, "N Queens - Coordinated")
+show(nqueens_static, "N Queens - Static")

### 2014-11-11 conclusions.tex
Rewrite some text.

@@
 about locality and scheduling of computation in a data-driven fashion.
-We have presented several applications where a judicious use of coordination can lead to
-better performance behavior.
+Even though the programs are optimized, the proofs of correctness for CLM
+programs remain simple since coordination does not change the language.
 
-We would like to explore this coordination paradigm in distributed systems,
-where data locality is far more important than in shared memory systems. We also
-think that CLM may be useful as a glue distributed/parallel language since it can
-easily model any computation pattern using graphs. Furthermore, we think that
-other declarative paradigms would benefit from using a similar approach. For
-instance, in functional programming, one may consider annotation functions that
-pick the right granularity for each problem.
+We would like to explore CLM's coordination principles in distributed systems,
+where data locality is far more important than in shared memory systems.
+Furthermore, we think that other declarative paradigms would benefit
+from giving programmers the freedom to fine-tune their programs while keeping
+the program declarative and easy to reason about.

### 2014-11-11 intro.tex
Update Related Work.
Rewrite some text.

@@
 Writing parallel programs in sequential languages is hard because manipulating
-shared state using multiple threads may result in data race
-conditions. Such issues are handled with low level constructs such as locks and
-condition variables, requiring a fair amount of effort to get right.
-Declarative programming has been hailed as a solution to this issue, since the
-problem of implementing the details of parallelism is moved from the programmer to the
-compiler and runtime. The programmer writes code
-without having to deal with parallel programming constructs and the compiler
-automatically parallelizes the program in order to take advantage of
-multiple threads of execution.
-The programming paradigm has been adopted with huge success in domain specific
+shared state using multiple threads may result in data race conditions. Such
+issues are handled with low level constructs such as locks and condition
+variables, requiring a fair amount of effort to get right.  Declarative
+programming has been hailed as a solution to this issue, since the problem of
+implementing the details of parallelism is moved from the programmer to the
+compiler and runtime. The programmer writes code without having to deal with
+parallel programming constructs and the compiler automatically parallelizes the
+program in order to take advantage of multiple threads of execution.  The
+programming paradigm has been adopted with huge success in domain specific
 languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
-Although general declarative languages have yet to be as successful, the
-future looks promising for this particular approach.
+Although general declarative languages have yet to be as successful, the future
+looks promising for this particular approach.
 
-The problem with declarative programming is that it leaves little to no programmer control
-over how execution is scheduled or how data is laid out, making it hard to achieve
-necessary efficiency. This introduces 
-performance issues because even if the runtime system is able to reasonably
-parallelize the program using a general algorithm, there is a lack of specific
-information about the program that a compiler cannot easily deduce. Such
-information could make execution better in terms of run time time,
-memory usage, and scalability.
+The problem with declarative programming is that it leaves little to no
+programmer control over how execution is scheduled or how data is laid out,
+making it hard to achieve necessary efficiency. This introduces
+performance issues because even if the runtime system is able to
+reasonably parallelize the program using a general algorithm, there
+is a lack of specific information about the program that a compiler
+cannot easily deduce. Such information could make execution better in
+terms of run time time, memory usage, and scalability.
 
-In this paper, we introduce Coordinated Linear Meld (CLM), a declarative
-language that extends the Linear Meld~(LM) language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
-control over scheduling and data placement. LM is a linear logic programming
-language designed for programs that operate on graphs.  The use of linear
-logic~\cite{girard-87} supports structured manipulation of mutable
+In this paper, we introduce Coordinated Linear Meld (CLM), a data-centric declarative
+language that extends the Linear Meld~(LM)
+language~\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give
+programmer control over scheduling and data placement. LM is a linear logic
+programming language designed for programs that operate on graphs.  The use
+of linear logic~\cite{girard-87} supports structured manipulation of mutable
 state. In LM, computation is divided so that each node of the graph computes
-independently and is allowed to \scare{communicate} with other nodes.
-Both computation and communication happen through the derivation of logical
-rules (which make up the program).
+independently and is allowed to \scare{communicate} with other nodes.  Both
+computation and communication happen through the derivation of logical rules
+(which make up the program).
 
-The CLM language features coordination
-primitives that can be used in the same way as any other primitive,
-in other words, they are specificied with the same syntax and semantics as the
-rest of the programming language. These coordination primitives can be used to
-improve program execution based on the state of the program and the underlying
-machine. The first kind of coordination facts are called \emph{sensing facts}
-and are used to sense information about the system the program is running on,
-e.g., scheduling and node placement on threads. The second kind of coordination facts are
-\emph{action facts} that when deleted in rules are used to apply a scheduling
-operation during execution. Coordination facts allow the programmer to write
-logical rules that depend on the current state of the program and then prioritize node
-computation or place nodes in different threads.
+The CLM language features coordination primitives that can be used in the same
+way as any other primitive, in other words, they are specificied with the same
+syntax and semantics as the rest of the programming language. These coordination
+primitives can be used to improve program execution based on the state of the
+program and the underlying machine. The first kind of coordination facts are
+called \emph{sensing facts} and are used to sense information about the system
+the program is running on, e.g., scheduling and node placement on threads. The
+second kind of coordination facts are \emph{action facts} that when deleted in
+rules are used to apply a scheduling operation during execution. Coordination
+facts allow the programmer to write logical rules that depend on the current
+state of the program and then prioritize node computation or place nodes in
+different threads.
 
-To the best of our knowledge, this is the first time that a declarative language allows 
-control over execution while staying declarative and without resorting to
+To the best of our knowledge, this is the first time that a declarative language
+allows control over execution while staying declarative and without resorting to
 meta-language constructs. This is crucial to our goal of being able to prove
-programs correct since
-proofs can be constructed even in the presence of coordination facts.
-After briefly discussing related work, we present an overview of the base
-language, with an example. In Section~\ref{sec:coordination}, we introduce the
-current set of coordination facts,
-followed by a description of the changes required to implement the desired
-coordination mechanisms. In
-Section~\ref{sec:applications} we present several applications and show
-coordination can improve programs without destroying clarity or provability.
+programs correct since proofs can be constructed even in the presence of
+coordination facts.  After briefly discussing related work, we present an
+overview of the base language, with an example. In
+Section~\ref{sec:coordination}, we introduce the current set of coordination
+facts, followed by a description of the changes required to implement the
+desired coordination mechanisms. In Section~\ref{sec:applications} we present
+several applications and show coordination can improve programs without
+destroying clarity or provability.

### 2014-11-12 implementation.tex
Fixed typos.

@@
 Semantically, action facts are like any other. However, since they are
-immediatelly used by the machine, there is no need to store them in the
+immediately used by the machine, there is no need to store them in the
 database, therefore avoiding unnecessary allocations and deallocations.
@@
 thread becomes idle, it attempts to steal nodes from a random thread.
-If unsucessful, the thread becomes idle and waits for program
+If unsuccessful, the thread becomes idle and waits for program
 termination while periodically attempting to steal nodes.
@@
 
-Each node is protected by a main spinlock that allows threads to
+Each node is protected by a main spin-lock that allows threads to
 change node attributes: incoming facts, owner thread, state variable
-and locality information.  A database spinlock is locked whenever the
+and locality information.  A database spin-lock is locked whenever the
 node enters into the \textbf{working} state.
 
-To avoid unneccesary copying, when a node sends facts to a node
+To avoid unnecessary copying, when a node sends facts to a node
 located in another thread, the current thread first attempts to lock
@@
 
-Threads synchronize with each other using mutual exclusion. We use a spinlock in
+Threads synchronize with each other using mutual exclusion. We use a spin-lock in
 each queue to protect queue operations.
@@
    \item[Coordination:] If $T_1$ needs to perform coordination operations
-   to a node in $T_2$, it may need to sychronize with $T_2$ during priority
+   to a node in $T_2$, it may need to synchronize with $T_2$ during priority
    updates in order to move the node in $T_2$'s queues.
@@
 programs benchmarked do not use any kind of coordination. As hoped for, there
-is no degration of performance by adding coordination mechanisms. Since we also
+is no degradation of performance by adding coordination mechanisms. Since we also
 performed general performance improvements, programs such as PageRank and Greedy

### 2014-11-12 intro.tex
Fixed typos.

@@
 The CLM language features coordination primitives that can be used in the same
-way as any other primitive, in other words, they are specificied with the same
+way as any other primitive, in other words, they are specified with the same
 syntax and semantics as the rest of the programming language. These coordination

### 2014-11-12 language.tex
Add N Queens Static to queue table.
Fixed typos.

@@
 associated tuple of values, e.g., the concrete values of \texttt{X} and
-\texttt{Y}, substitued for the
+\texttt{Y}, substituted for the
 arguments. Since CLM uses linear logic as its foundation, we distinguish between
@@
 independently. Although the body is restricted, the head of the rule may refer
-to any node as long as that node is refered to somewhere in the body. This allows
+to any node as long as that node is referred to somewhere in the body. This allows
 \scare{communication} between nodes during rule derivation, since a node may
@@
 
+\paragraph{SSSP Example}
 To make these ideas concrete, we present a simple
@@
 \end{topfig}
-\normalsize
+
+\begin{dblfig}[ht]
+\begin{center}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
+  \hspace{0.4cm}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
+  \hspace{0.4cm}
+  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
+\end{center}
+\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. Figure (a) represents the
+   program after propagating initial distance at node \texttt{@1}, followed by
+   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
+   represents the state of the final program, where all the shortest paths
+   have been computed.}
+\end{dblfig}
 
@@
 applying the same rule but at node \texttt{2}.
-
-\begin{figure*}[ht]
-\begin{center}
-  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
-  \hspace{0.4cm}
-  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
-  \hspace{0.4cm}
-  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
-\end{center}
-\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. Figure (a) represents the
-   program after propagating initial distance at node \texttt{@1}, followed by
-   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
-   represents the state of the final program, where all the shortest paths
-   have been computed.}
-\end{figure*}
 

### 2014-11-12 locks.tex
Add N Queens Static to queue table.
Fixed typos.

@@
-\paragraph{Locking} The CLM virtual machine uses spinlocks to protect shared data structures between
+\paragraph{Locking} The CLM virtual machine uses spin-locks to protect shared data structures between
 threads. As described in Section~\ref{sec:implementation}, coordination
@@
 little variations between the number of threads used. This is because threads
-interactly very little with each other and the amount of work to do is the same
+interact very little with each other and the amount of work to do is the same
 for all threads.
@@
 requires less queue operations because the number of derived facts is reduced.
-The MiniMax program also sees twices as many queue instructions using
+The MiniMax program also sees twice as many queue instructions using
 coordination. This does not happen with the N Queens program, where the number
@@
 N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
-N Queens - Coordinated & 57M & 18M & 676K & 125K & 41K \\ \hline
+N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\ \hline
+N Queens - Static & 57M & 44M & 1M & 1M & 151K \\ \hline
+\hline
 \end{tabular}

### 2014-11-12 related.tex
Fixed typos.

@@
 Another early coordination language is Delirium~\cite{Delirium}. Unlike Linda
-which is embbeded into another language, Delirium actually embeds operators
+which is embedded into another language, Delirium actually embeds operators
 written in other languages inside the Delirium language.  The advantages of
@@
 The LM language was inspired by Meld~\cite{ashley-rollman-iclp09}, a
-Datalog-like language for programming distributed emsembles of modular robots.
+Datalog-like language for programming distributed ensembles of modular robots.
 Meld introduced the idea of sensing and action facts in order to sense and act
 on the outside world, respectively. LCM uses sensing facts to sense the
-underlying system that the program is running on and action facs act on that
+underlying system that the program is running on and action facts act on that
 system.

### 2014-11-12 splash-bp.tex
Fixed typos.

@@
 although the final result of the algorithm is identical to the original
-algorithm (minus probabilitistic errors), SBP works in a very different way.
+algorithm (minus probabilistic errors), SBP works in a very different way.
 SBP is implemented in GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.

### 2014-11-12 sssp.tex
use new scap and fig macros
Fixed typos.

@@
 \end{Verbatim}
-  \caption{Shortest Path Program.}
-  \label{code:shortest_path_program_coord}
-\end{figure}
+  \scap{code:shortest_path_program_coord}{Shortest Path Program.}
+\end{topfig}
 \normalsize
@@
 program computes the SSSP for 20\% of the nodes of the graph.
-There are some situations where unecessary facts are propagated
+There are some situations where unnecessary facts are propagated
 because although the shortest distance is selected locally, sub-optimal distances may be
@@
 derived facts when using the coordinated version \textbf{C} over the regular
-version \textbf{R}. The results also show that the cordinated
+version \textbf{R}. The results also show that the coordinated
 version using 16 threads is 1.3 times faster than the regular version using 16
@@
    \end{center}
-   \caption{Experimental results for the SSSP program using US powergrid network.
+   \scap{results:sssp_uspowergrid}{Experimental results for the SSSP program using US power grid network.
    On the left we show the number of facts derived for the regular \textbf{R}
@@
    time for 1 thread.}
-   \label{results:sssp_uspowergrid}
-\end{figure}
+\end{topfig}

### 2014-11-12 nqueens.tex
Fixed typos.
english and some extra info

@@
 
-The CLM solution considers the chess board as a graph where nodes exchange valid
-configurations between each other. First, we start with the empty
-state in the first row of the chess board. Then, each square adds its own
-position to the state and sends the state down, to the next row. Once a square
-receives new configurations, it attempts to add its position to the
-configurations. If valid, that configuration is then sent, recursively, to the
-next row, until all rows are traversed. At the end of the program, the nodes at
-the bottom row will have all the valid configurations.
+The CLM solution considers the squares of the chess board as a graph
+of nodes which exchange valid configurations with each
+other. Initially every square in the first row of the board gets the
+empty state.  Then, each square adds its own position to the state and
+sends the state down, to the next row. Once a square receives new
+configurations, it attempts to add its position to the
+configurations. If valid, that configuration is then sent,
+recursively, to the next row, until all rows are traversed. At the end
+of the program, the squares at the bottom row will have all the valid
+configurations.
 
-Since computation goes from the top row to the bottom row, not all placements of
-nodes on threads may be equally performant. This is especially true because the
-bottom rows tend to perform the most work. The best placement is then to split the
-board vertically, so that, in optimal conditions, each thread gets the same
-number of columns. Our experiments show that, on shared memory, it does not
-matter much if we start with a bad placement since node stealing overcomes those
-issues by improving load balancing dynamically. However, we noticed interesting
-cache effects when prioritizing nodes in bottom rows. The N Queens program
-builds and shares lists representing valid board states and needs to iterate
-over those lists often when checking if a new position does not invalidate a state.
-We thus used coordination axioms \texttt{set-default-priority(A, X)} and \texttt{set-cpu(A,
-vertical(X, Y))}, where \texttt{X} is the row number (0 is the top row) and
-\texttt{Y} is the column number.
+Since computation goes from the top row to the bottom row, not all
+placements of nodes to threads performane equally. This is especially
+true because the bottom rows tend to perform the most work. The best
+placement is then to split the board vertically so each thread gets
+the same number of columns. Our experiments show that, on a shared
+memory system, it does not matter much if we start with a bad
+placement since node stealing overcomes those issues by load balancing
+dynamically. However, we noticed interesting cache effects when
+prioritizing nodes in bottom rows.\snote{Expand on this?  What are
+they?} The N Queens program builds and shares lists representing valid
+board states and needs to iterate over those lists often when checking
+if a new position does not invalidate a state.  We thus used
+coordination axioms \texttt{set-default-priority(A, X)}
+and \texttt{set-cpu(A, vertical(X, Y))}, where \texttt{X} is the row
+number (0 is the top row) \texttt{Y} is the column number to
+prioritize the evaluation of squares at the XXXX of the board and to
+keep board configurations in the same cache. \snote{check the previous}
 
-Experimenta results are presented in Fig.~\ref{results:nqueens}.
-We first ran the coordinated program using dynamic partitioning (with work
-stealing) and saw some performance improvements (\textbf{Coordinated}). However, using dynamic
-partitioning is bad for memory locality since threads will steal nodes often.
-We then added \texttt{set-static(A)} as a coordination axiom and saw
-performance improvement (\textbf{Static}), especially when the board columns are perfectly
-partitioned by the number of threads (in our case 13 threads).
+Experimental results are presented in Fig.~\ref{results:nqueens}.  We
+first ran the coordinated program using dynamic partitioning (with
+work stealing) and saw some performance improvements
+(\textbf{Coordinated}). However, using dynamic partitioning is bad for
+memory locality since threads will often steal nodes.  We then
+added \texttt{set-static(A)} as a coordination axiom and saw
+performance improvement (\textbf{Static}), especially when the board
+columns are perfectly partitioned by the number of threads (in our
+case 13 threads).
 

### 2014-11-12 related.tex
Update related work.

@@
 more recently Data Parallel Haskell~\cite{nepal}.
+NESL often got good performance, but was limited to a particular
+application space. We also limit our application space, but also give the
+programmer a declarative way to coordinate computation to improve the
+performance of the program.
 
@@
 languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
-popular data-centric programming model that is optimized for large clusters.
+popular programming model that is optimized for large clusters.
 Intrinsic to its popularity is the simplicity of its scheduling and data sharing
@@
 
-GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
+GraphLab~\cite{GraphLab2010} is a C++ framework for developing graph-based parallel machine
 learning algorithms. GraphLab allows nodes to have read/write access to
@@
 scheduler can be implemented in CLM through the use of coordination facts.
+Another, more restrictive, graph-based system is
+Pregel~\cite{Malewicz:2010:PSL:1807167.1807184}, where graph algorithms must be
+executed as a sequence of iterations of computation and message passing.
+Both those systems provide excellent performance, however they are not
+declarative and have a steep learning curve.
 

### 2014-11-12 implementation.tex
Remove fig 4.

@@
 than a sequential C implementation. Some of those programs were compared against
-a Python implementations and LM faired fairly well. In this section, we measure
-the impact of the coordination mechanisms we have implemented for the CLM
-virtual machine.
-
-Fig.~\ref{results:comparison1} shows the comparison between the original virtual
-machine and the new virtual machine with coordination mechanisms. All the
-programs benchmarked do not use any kind of coordination. As hoped for, there
-is no degradation of performance by adding coordination mechanisms. Since we also
-performed general performance improvements, programs such as PageRank and Greedy
-Graph Coloring programs now perform better than before.
-
-\begin{topfig}
-   \begin{center}
-      \subfloat[]{\includegraphics[width=8cm]{results/comparison1.png}}
-   \end{center}
-   \scap{results:comparison1}{Measuring the overhead of coordination mechanisms. The \textbf{Old}
-   bars represent the performance of the original virtual machine, while
-   \textbf{New} is the relative performance of the new version that includes
-   coordination mechanisms.}
-\end{topfig}
+a Python implementations and LM faired fairly well. We measured
+the impact of the coordination mechanisms implemented for the CLM
+virtual machine on regular LM programs and saw very little to no degration.
+Since we also performed general performance improvements, programs such as
+PageRank and Greedy Graph Coloring programs now perform better than before.

### 2014-11-12 locks.tex
make it a table
Remove unneeded subfloats.

@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/sssp-locks.png}}
+      \includegraphics[width=6.5cm]{results/locks/sssp-locks.png}
    \end{center}
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/ht-locks.png}}
+      \includegraphics[width=6.5cm]{results/locks/ht-locks.png}
    \end{center}
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=6.5cm]{results/locks/bp.png}}
+      \includegraphics[width=6.5cm]{results/locks/bp.png}
    \end{center}

### 2014-11-12 language.tex
spacing, ref

@@
 
-\paragraph{SSSP Example}
+\paragraph{SSSP Example}\label{sect:ssspex}
 To make these ideas concrete, we present a simple
@@
 
-\begin{dblfig}[ht]
+\begin{dblfig}
 \begin{center}
@@
 \end{center}
-\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. Figure (a) represents the
+\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. (a) represents the
    program after propagating initial distance at node \texttt{@1}, followed by
-   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
+   (b) where the first rule is applied in node \texttt{@3}. (c)
    represents the state of the final program, where all the shortest paths

### 2014-11-12 minimax.tex
Fixed typos.
small edits for space

@@
 
-The code in Fig.~\ref{minimax:check-end} shows deals with expanding the tree.
+The code in Fig.~\ref{minimax:check-end} deals with expanding the tree.
 The first three rules (lines 1-10) deal
@@
 
-As noted before in Section~\ref{sec:fifo}, the runtime schedules nodes by
-default, which requires a complete expansion of the tree before computing the
-scores at the leaves, leading
-to a memory complexity of $\mathcal{O}(n)$, where $n$ is the number of nodes in
-the tree.
+As noted in \sectref{sec:fifo}, the default scheduler is
+breadth-first, which in this case leads to a complete expansion of the
+tree before computing the scores at the leaves.  This results in using
+$\mathcal{O}(n)$ memory, where $n$ is the number of nodes in the tree.
 
@@
 is the depth of the tree and $t$ is the number of threads.
-Since threads prioritize deeper nodes, the scores of the first leaves are immediatelly
+Since threads prioritize deeper nodes, the scores of the first leaves are immediately
 computed and then sent to the parent node. At this point, the leaves are deleted
@@
 \begin{topfig}
+\vspace*{-2ex}
    \begin{center}

### 2014-11-12 splash-bp.tex
spacing

@@
    \end{center}
-   \scap{results:splash_bp}{Experimental results for LBP and SBP. Figure (a) shows the
-      scalability of LBP for both CLM and GraphLab and Figure (b) shows the
-      scalability of SBP. Figure (c) presents the
+   \scap{results:splash_bp}{Experimental results for LBP and SBP. (a) shows the
+      scalability of LBP for both CLM and GraphLab and (b) shows the
+      scalability of SBP. (c) presents the
       improvements seen in SBP against LBP, where SBP runs, on average, 1.5 to 2.5

### 2014-11-12 sssp.tex
spacing

@@
-Here we add coordination to the SSSP program described in Section
+Here we add coordination to the SSSP program described in \sectref{sect:ssspex}.
 The coordinated version of the
@@
 
-Figure~\ref{results:sssp_uspowergrid} shows experimental results when the
+Fig.~\ref{results:sssp_uspowergrid} shows experimental results when the
 program computes the SSSP for 20\% of the nodes of the graph.
@@
 \begin{topfig}
+\vspace*{-5ex}
    \begin{center}
@@
       }
-      \subfloat[]{\includegraphics[width=5.5cm]{results/shortest-uspowergrid.png}}
+      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
    \end{center}
-   \scap{results:sssp_uspowergrid}{Experimental results for the SSSP program using US power grid network.
-   On the left we show the number of facts derived for the regular \textbf{R}
-   and coordinated version \textbf{C} using a variable number of threads
-   \textbf{\# T}. On the right, we have the scalability of the regular and
-   coordinated version. The speedup values are computed using the execution
-   time for 1 thread.}
+   \scap{results:sssp_uspowergrid}{Results for SSSP run on the 
+   US power grid network.  On the left we show the number of facts
+   derived for the regular \textbf{R} and coordinated
+   version \textbf{C} using a variable number of threads \textbf{\#
+   T}. On the right, we show the scalability of the regular and
+   coordinated version. The speedup values are computed using the
+   execution time for 1 thread.}
+\vspace*{-3ex plus 0pt minus 1ex}
 \end{topfig}

### 2014-11-12 locks.tex

@@
 \paragraph{Locking} The CLM virtual machine uses spin-locks to protect shared data structures between
-threads. As described in Section~\ref{sec:implementation}, coordination
+threads. As described in \sectref{sec:implementation}, coordination
 operations require synchronization since they need to change node information
@@
 
-\begin{topfig}
+\begin{toptab}
 \footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
@@
 \end{tabular}
+\vspace*{.5ex}
 \scap{queue_instructions}{Number of queue \scare{instructions} per program. An
@@
    \emph{percolate-up}/\emph{percolate-down} steps performed.}
-\end{topfig}
+\end{toptab}
 

### 2014-11-12 results/locks/queue_operations.py
Update script to generate table.

@@ -36,27 +36,40 @@ def show(exp, name):
    count = exp.queue_instruction_count()
    print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
 
-show(sssp_reg, "SSSP - Regular")
-show(sssp_coord, "SSSP - Coordinated")
-show(sssp_buffered, "SSSP - Buffered")
-print "\hline"
+def show_separated(exp, name):
+   normal = exp.normal_instruction_count()
+   prio = exp.heap_instruction_count()
+   print name + "& " + translate(normal[0]) + "+" + translate(prio[0]) + " & " + translate(normal[1]) + "+" + translate(prio[1]) + " & " + translate(normal[2]) + "+" + translate(prio[2]) + " & " + translate(normal[4]) + "+" + translate(prio[4]) + " & " + translate(normal[8]) + "+" + translate(prio[8])
 
-show(ht_reg, "HT - Buffered")
-show(ht_coord, "HT - Coordinated")
-show(ht_local_only, "HT - Local-Only")
+def display_data(fun):
+   fun(sssp_reg, "SSSP - Regular")
+   fun(sssp_coord, "SSSP - Coordinated")
+   fun(sssp_buffered, "SSSP - Buffered")
+   print "\hline"
 
-print "\hline"
+   fun(ht_reg, "HT - Regular")
+   fun(ht_coord, "HT - Coordinated")
+   fun(ht_local_only, "HT - Local-Only")
 
-show(lbp, "LBP - Regular")
-show(sbp, "SBP - Coordinated")
+   print "\hline"
 
-print "\hline"
+   fun(lbp, "LBP - Regular")
+   fun(sbp, "SBP - Coordinated")
 
-show(minmax_reg, "MiniMax - Regular")
-show(minmax_coord, "MiniMax - Coordinated")
+   print "\hline"
 
-print "\hline"
+   fun(minmax_reg, "MiniMax - Regular")
+   fun(minmax_coord, "MiniMax - Coordinated")
 
-show(nqueens_reg, "N Queens - Regular")
-show(nqueens_coord, "N Queens - Coordinated")
-show(nqueens_static, "N Queens - Static")
+   print "\hline"
+
+   fun(nqueens_reg, "N Queens - Regular")
+   fun(nqueens_coord, "N Queens - Coordinated")
[…]

### 2014-11-12 heat-transfer.tex
Remove unneeded subfloats.
some rewording

@@
-In the Heat Transfer algorithm~(HT), we have a graph where heat values are
-exchanged between nodes. The program stops when the new heat values of every node are
-$\delta = |H_i - H_{i-1}| \le \epsilon$. The algorithm works
-asynchronously since heat values can be updated by using new partial information
-coming from neighbor nodes. This increases parallelism since nodes do not need
-to synchronize between iterations.
+In the Heat Transfer algorithm~(HT), we have a graph where heat values
+are exchanged between nodes. The program stops when the new heat
+values of every node are $\delta = |H_i - H_{i-1}| \le \epsilon$. The
+algorithm works asynchronously, i.e., heat values are updated using
+information as it arrives from neighboring nodes. This increases
+parallelism since nodes do not need to synchronize between iterations.
 
@@
 \end{Verbatim}
-  \scap{code:ht}{Coordination code for the Heat Transfer program. We updated the first rule
-     to increase the priority of neighbor nodes.  The code logic remains exactly
-     the same as before, however bigger changes in heat values are now
-     propagated faster.}
+  \scap{code:ht}{Coordination code for the Heat Transfer program. The
+      first rule now increases the priority of neighbor nodes.  The
+      logic remains the same as before, however bigger
+      changes in heat are now propagated faster.}
 \end{topfig}
@@
 
-To further improve locality, we split the second rule to
-not send small $\delta$ values if the target node is in another
-thread (Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Coordinated} line in
-Fig.~\ref{results:ht} shows that this new version performs better due to
-improved locality, exhibiting linear scalability over the regular version.
-However, this come at the price of increased errors in the computed heat values.
+To further improve locality, we split the second rule to avoid sending
+small $\delta$ values if the target node is in another thread
+(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Coordinated} line
+in Fig.~\ref{results:ht} shows that this performs better due to
+improved locality, exhibiting linear scalability over the regular
+version.  However, this come at the price of increased errors in the
+computed heat values.
 
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/new-heat-transfer-80}}
+      \includegraphics[width=5.5cm]{results/new-heat-transfer-80}
    \end{center}

### 2014-11-12 locks.tex
rewrote

@@
-\paragraph{Locking} The CLM virtual machine uses spin-locks to protect shared data structures between
-threads. As described in \sectref{sec:implementation}, coordination
-operations require synchronization since they need to change node information
-and sometimes move nodes between the standard queue and the regular queue.
-In order to measure the locking costs of coordination, we measured the following
-statistics: \emph{basic locks}, the number of locks acquired that were already
-present in the original LM virtual machine and \emph{coordination locks}, the
-number of locks acquired for performing coordination operations. We also
-measured how often a \texttt{lock()} operation fails on the first attempt, but
-such statistics are not represented in the figures below since the values are
-too small.
+\begin{toptab}
+\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
+\hline
+\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
+   Instructions}} \\ \cline{2-6}
+& 1 & 2 & 4 & 8 & 16 \\ \hline
+\hline
+SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\ \hline
+SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\ \hline
+SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\ \hline
+\hline
+HT - Regular & 25M & 32M & 34M & 34M & 36M \\ \hline
+HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\ \hline
+HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\ \hline
+\hline
+LBP - Regular & 15M & 17M & 16M & 11M & 8M \\ \hline
+SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\ \hline
+\hline
+MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\ \hline
+MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\ \hline
+\hline
+N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
+N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\ \hline
+N Queens - Static & 57M & 44M & 1M & 1M & 151K \\ \hline
+\hline
+\end{tabular}
+\vspace*{.5ex}
+\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
+   operation to a regular queue counts as 1 instruction, while an operation to
+   a priority queue costs the number of
+   \emph{percolate-up}/\emph{percolate-down} steps performed.}
+\end{toptab}
 
-The first program we measured was SSSP, represented in
-Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
-Coordinated versions, we have the Buffered version where we
-optimized the virtual machine to buffer coordination actions before applying
-them. The idea is to avoid applying multiple \texttt{set-priority} operations to
-the same node during a single node execution. We thus gather the higher priority
-values and then perform coordination after the node has executed.
-Although the number of locks is significantly reduced, we did not see a
-similar performance improvement, since the Buffered version is only
-5\% faster than the Coordinated version.
+Coordination introduces overhead in two major ways.  First, in
+manipulating the priority queues which require operations on a min
+heap.  Second, in potentially requiring more lock operations as we
+move nodes between queues and within the priority queue.  As we show
+here, this overhead is minimal and the resulting improved schedules
+and improved locality more than make up for the overhead.
+
+As can be seen in Table~\ref{queue_instructions} all the coordinated
+versions of the programs require significantly more queue operations.
+The data for this table comes from recording, during execution, the
+number of normal queue operations executed (each costs 1), plus the
+number of \emph{percolate-up}/\emph{percolate-down} operations
+executed for each manipuilation of the priority queue.  As expected,
+programs which rely on prioritizing the order of execution to obtain
+better asymptotic complexity (e.g., SSSP) improve in overall running
+time at the cost of more queue operations.  In all cases, coordination
+adds a significant number of total queue operations, but the resulting
+overhead is more than compensated for by an improved schedule.
+
+An interesting artifact of effectively parallelizing the code is that
+the priority queue on each thread is smaller, so the total number of
+priority queue operations is significantly reduced as the depth of the
+min heap on each thread is reduced.  This is particularly evident for
+the N Queens program.
 
@@
 
-In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
-regular version, the coordinated version and the local-only version as in
-Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
-coordination and then a further reduction in the local-only version because less
-facts are exchanged between threads.
-
 \begin{topfig}
@@
 \end{topfig}
+
+We next investigate the effect of coordination on the syncronization
+overhead that comes froms from the use of more locks.  We measured the
+\emph{basic locks}, the number of locks acquired
+in the original LM virtual machine and \emph{coordination locks}, the
+number of locks acquired to perform coordination operations.  We also
+measured how often \texttt{lock()} operations that fail to acquired
+the lock on the first attempt.  Lock failures happen so infrequently
+(less than X\% \snote{please fill this in} of the time) that they
+cannot be seen on the figures.
+
+Fig.~\ref{results:sssp_uspowergrid} shows the number of locks acquired
+for the SSSP program.  The first bar in each group shows the number of
+locks acquired for the uncoordinated version of the program.  The
+second bar shows the regular and coordination locks acquired for SSSP
+when every newly derived fact and priority operation is carried out
+immediately.  This leads to a significant increase in total locks
+acquired often because the priority of a node is change many times
+before it is operated on.  The third bar shows the effect of buffering
+priority operations before applying them.  This significantly reduces
+the lock overhead, but as can be seen in
+Table~\ref{queue_instructions} increase the number of queue operations
+slightly.  This slight increase comes from the fact the buffering
+delays the change in priorities and some longer paths can sometimes be
+investigated unnecessarily.
+
+Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
+program.  The behavior is representative of the other programs.  There
+is drop in the total number of locks acquired and coordination locks
+represent a small fraction of the total locks acquired.  The locality
+coordination further reduces locking by keeping more derivation local
+to a thread.
+
+Overall, the additional overhead introduced by supporting coordination
+is more than offset by improved schedules and increased locality.
+While there are more locks to manage the total cost of lock operations
+does not change significantly (and is often reduced) when coordination
+is added.  And, although the priority coordination significantly
+increases the cost of queueing it is offset by reducing the number of
+total facts derived.
+
+\iffalse
+
+  The
+SSSP program is The first program we measured was SSSP, represented in
+Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
+Coordinated versions, we have the Buffered version where we optimized
+the virtual machine to buffer coordination actions before applying
+them. The idea is to avoid applying multiple \texttt{set-priority}
+operations to the same node during a single node execution. We thus
+gather the higher priority values and then perform coordination after
+the node has executed.  Although the number of locks is significantly
+reduced, we did not see a similar performance improvement, since the
+Buffered version is only 5\% faster than the Coordinated version.
+
+In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
+regular version, the coordinated version and the local-only version as in
+Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
+coordination and then a further reduction in the local-only version because less
+facts are exchanged between threads.
 
@@
 
-We also measured the number of queue instructions required to compute programs.
-While the regular queue operations have constant complexity, priority queue
-operations are usually logarithmic and require more computation. We counted the
-number of instructions required for queue operations during program execution.
-Each normal queue operation costs 1 instruction, while priority queue costs as
-many \emph{percolate-up}/\emph{percolate-down} instructions are required to maintain the heap property.
-
 Table~\ref{queue_instructions} presents the total queue instructions for several
@@
 
-\begin{toptab}
-\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
-\hline
-\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
-   Instructions}} \\ \cline{2-6}
-& 1 & 2 & 4 & 8 & 16 \\ \hline
-\hline
-SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\ \hline
-SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\ \hline
-SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\ \hline
-\hline
-HT - Buffered & 25M & 32M & 34M & 34M & 36M \\ \hline
-HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\ \hline
-HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\ \hline
-\hline
-LBP - Regular & 15M & 17M & 16M & 11M & 8M \\ \hline
-SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\ \hline
-\hline
-MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\ \hline
-MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\ \hline
-\hline
-N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
-N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\ \hline
-N Queens - Static & 57M & 44M & 1M & 1M & 151K \\ \hline
-\hline
-\end{tabular}
-\vspace*{.5ex}
-\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
-   operation to a regular queue counts as 1 instruction, while an operation to
-   a priority queue costs the number of
-   \emph{percolate-up}/\emph{percolate-down} steps performed.}
-\end{toptab}
-
 These experiments indicate that handling coordination has a small cost in terms
@@
 decisions, the costs of manipulating priority queues become irrelevant.
+\fi

### 2014-11-12 minimax.tex
spacing
rewording, spacing

@@
 
-The algorithm proceeds by building a game tree, where each tree node represents a
+The algorithm builds a game tree, where each tree node represents a
 game state and the children represent the possible game moves that can
-be made by either player 1 or player 2.
-An evaluation function is used to compute the score of
-the board for each leaf of the tree. A node is a leaf when the game state can no longer be
-expanded. Finally, the
-algorithm recursively minimizes or maximizes the scores of each node.
-To select the best move for player 1, the
-algorithm picks the move maximized at the root node.
+be made by either player 1 or player 2.  An evaluation function is
+used to compute the score of the board for each leaf of the tree. A
+node is a leaf when the game state can no longer be expanded. Finally,
+the algorithm recursively minimizes or maximizes the scores of each
+node.  To select the best move for player 1, the algorithm picks the
+move maximized at the root node.
 
@@
 \scap{minimax:check-end}{MiniMax: checking if the game has ended and expanding the tree}
+\vspace*{-2ex}
 \end{topfig}
@@
 \begin{topfig}
-\vspace*{-2ex}
+\vspace*{-3ex}
    \begin{center}
@@
       of MiniMax.}
+\vspace*{-2ex}
 \end{topfig}

### 2014-11-12 splash-bp.tex
rewording, use tightdescription

@@
 
-\begin{description}
+\begin{tightdescription}
    \item[Tree building]: Each node has a \texttt{inactive} fact that is used to
@@
    second time (line 42).
-\end{description}
+\end{tightdescription}
 
-Note that the \texttt{set-static} and \texttt{set-cpu} action fact are used in
-line 2 in order to (1) force nodes to stay in the thread and (2) to partition
-nodes as a grid of threads. This sets up a well defined area of nodes
-for threads to build splash trees on.
+The \texttt{set-static} and \texttt{set-cpu} action facts are used in
+line 2 to (1) force nodes to stay in the thread and (2) 
+partition nodes as a grid of threads. This sets up well defined areas
+of nodes for threads to build splash trees on.
 
@@
 
-In this program, coordination assumes a far more important role than we have
-seen before. Coordination rules fully drive the behavior of the algorithm and
-although the final result of the algorithm is identical to the original
-algorithm (minus probabilistic errors), SBP works in a very different way.
-SBP is implemented in GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine algorithms.
-GraphLab provides the splash scheduler for these types of inferences as part of
-the framework. This particular scheduler is around 350 lines of C++ code.
-With our coordination facts, it is possible to create the necessary
-patterns with only 12 simple rules.
+In this program, coordination assumes a far more important role than
+we have seen before. Coordination rules fully drive the behavior of
+the algorithm and although the final result of the algorithm is
+statistically identical to the original algorithm, SBP works very
+differently than LBP.  SBP is also implemented in
+GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine
+algorithms.  GraphLab provides the splash scheduler as part its the
+framework. It is 350 lines of C++ code.  With our coordination facts,
+it is possible to create the necessary scheduling with only 12 rules.
 

### 2014-11-12 nqueens.tex
Remove unneeded subfloats.

@@
 placement is then to split the board vertically so each thread gets
-the same number of columns. Our experiments show that, on a shared
+the same number of columns with axiom \texttt{set-cpu(A, vertical(X, Y)}, where
+\texttt{X} and \texttt{Y} are the coordinates of the board. Our experiments show that, on a shared
 memory system, it does not matter much if we start with a bad
 placement since node stealing overcomes those issues by load balancing
-dynamically. However, we noticed interesting cache effects when
-prioritizing nodes in bottom rows.\snote{Expand on this?  What are
-they?} The N Queens program builds and shares lists representing valid
-board states and needs to iterate over those lists often when checking
-if a new position does not invalidate a state.  We thus used
-coordination axioms \texttt{set-default-priority(A, X)}
-and \texttt{set-cpu(A, vertical(X, Y))}, where \texttt{X} is the row
-number (0 is the top row) \texttt{Y} is the column number to
-prioritize the evaluation of squares at the XXXX of the board and to
-keep board configurations in the same cache. \snote{check the previous}
+dynamically.  However, the N Queens program incrementally builds and shares lists representing
+valid board states that are transmitted from top to bottom.
+In order to improve memory locality we should then manipulate board
+states that share a significant number of elements because each board state
+needs to be iterated before extended with a new position.
+To accomplish this, we used coordination axioms \texttt{set-default-priority(A,
+X)}, where \texttt{X} is the row number (0 is the top row) in order to
+prioritize the evaluation of squares closer to the bottom of the board.
 
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/8queens-13.png}}
+      \includegraphics[width=5.5cm]{results/8queens-13.png}
    \end{center}

### 2014-11-12 intro.tex

@@
 overview of the base language, with an example. In
-Section~\ref{sec:coordination}, we introduce the current set of coordination
+\sectref{sec:coordination}, we introduce the current set of coordination
 facts, followed by a description of the changes required to implement the
-desired coordination mechanisms. In Section~\ref{sec:applications} we present
+desired coordination mechanisms. In \sectref{sec:applications} we present
 several applications and show coordination can improve programs without

### 2014-11-12 results/locks/lib.py
Fill percent fail.

@@ -87,6 +87,9 @@ class experiment(object):
    def get_coord_fail_locks(self):
       return [data.fail_coord_locks() for th, data in sorted(self.threads.iteritems())]
 
+   def percent_fail(self):
+      return [(float(data.total_fail())/float(data.total_locks())) * 100 for th, data in sorted(self.threads.iteritems())]
+
    def get_fail_locks(self):
       return [data.total_fail() for th, data in sorted(self.threads.iteritems())]
 

### 2014-11-12 results/locks/plot.py

@@ -40,3 +40,6 @@ ax.set_xticks(ind)
 ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
 ax.legend((pbasicreg, pcoordbasic, pcoordcoord), ('Regular - Basic', 'Coordinated - Basic', 'Coordinated - Coordination'), loc=2)
 plt.savefig(output_file)
+
+print reg_exp.percent_fail()
+print coord_exp.percent_fail()

### 2014-11-12 results/locks/plot_three.py

@@ -53,3 +53,6 @@ ax.set_xticklabels(('1', '2', '4', '6', '8', '10', '12', '14', '16'))
 ax.legend((pbasicreg, pbasicc, pcoordc, pbasico, pcoordo), ('Regular - Basic', 'Coordinated - Basic',
          'Coordinated - Coordination', opt_name + ' - Basic', opt_name + ' - Coordination'), loc=2)
 plt.savefig(output_file)
+
+print reg_exp.percent_fail()
+print coord_exp.percent_fail()

### 2014-11-12 abstract.tex
Add facts derived

@@
 write declarative parallel programs and then optionally use coordination 
-to fine tune programs, while keeping the program fully declarative and amenable
-to correctness proofs.
+to fine tune programs, while keeping the program fully declarative and easy to
+reason about.

### 2014-11-12 conclusions.tex
Add facts derived

@@
 runtime system. In the other hand, sensing facts allow the programmer to reason
-about locality and scheduling of computation in a data-driven fashion.
-Even though the programs are optimized, the proofs of correctness for CLM
-programs remain simple since coordination does not change the language.
+about data locality and scheduling of computation in a data-driven fashion.
+Even though the programs are optimized, programs remain declarative and proving
+interesting properties of the programs remains simple because it is possible to
+reason about the logical rules that drive coordination.
 
@@
 where data locality is far more important than in shared memory systems.
-Furthermore, we think that other declarative paradigms would benefit
-from giving programmers the freedom to fine-tune their programs while keeping
-the program declarative and easy to reason about.
+Furthermore, we think that other declarative paradigms such as functional
+programming would benefit from giving programmers the freedom to fine-tune their
+programs.

### 2014-11-12 results/locks/lib.py

@@ -16,6 +16,9 @@ class dataobj(object):
    def set_normal_operations(self, ops):
       self.normal_operations = ops
 
+   def set_facts_derived(self, ops):
+      self.facts_derived = ops
+
    def get_total(self, name):
       return self.info_total[name]
 
@@ -51,6 +54,9 @@ class dataobj(object):
 
    def queue_instruction_count(self):
       return self.normal_operations + self.heap_operations
+
+   def facts_derived_count(self):
+      return self.facts_derived
 
    def frac_coord_locks(self):
       return (float(self.total_coord_locks()) / float(self.total_locks())) * 100
@@ -111,6 +117,9 @@ class experiment(object):
    def queue_instruction_count(self):
       return [data.queue_instruction_count() for th, data in sorted(self.threads.iteritems())]
 
+   def facts_derived_count(self):
+      return [data.facts_derived_count() for th, data in sorted(self.threads.iteritems())]
+
    def get_threads(self):
       return [th for th in sorted(self.threads.keys())]
 
@@ -153,8 +162,10 @@ def read_experiment(filename):
             total = int(resvec[0])
             if name.startswith("normal_operations"):
                data.set_normal_operations(total)
-            else:
+            elif name.startswith("heap_operations"):
                data.set_heap_operations(total)
+            else:
+               data.set_facts_derived(total)
    if data:
       exp.add_thread(thread, data)
    return exp

### 2014-11-12 results/locks/queue_operations.py

@@ -36,6 +36,12 @@ def show(exp, name):
    count = exp.queue_instruction_count()
    print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
 
+def show_facts_per_op(exp, name):
+   instrs = exp.queue_instruction_count()
+   facts = exp.facts_derived_count()
+   count = [round(float(ft)/float(it), 2) for it, ft in zip(instrs, facts)]
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+
 def show_separated(exp, name):
    normal = exp.normal_instruction_count()
    prio = exp.heap_instruction_count()
@@ -73,3 +79,7 @@ print
 print
 print
 display_data(show_separated)
+print
+print
+print
+display_data(show_facts_per_op)

### 2014-11-12 results/locks/queue_operations.py
Add more tables.

@@ -26,9 +26,9 @@ nqueens_static = read_experiment("nqueens-static.txt")
 
 def translate(num):
    if num > 1000000:
-      return str(num/1000000) + "M"
+      return str(round(num/1000000, 1)) + "M"
    elif num > 1000:
-      return str(num/1000) + "K"
+      return str(round(num/1000, 1)) + "K"
    else:
       return str(num)
 
@@ -40,6 +40,10 @@ def show_facts_per_op(exp, name):
    instrs = exp.queue_instruction_count()
    facts = exp.facts_derived_count()
    count = [round(float(ft)/float(it), 2) for it, ft in zip(instrs, facts)]
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+
+def show_facts_derived(exp, name):
+   count = exp.facts_derived_count()
    print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
 
 def show_separated(exp, name):
@@ -83,3 +87,7 @@ print
 print
 print
 display_data(show_facts_per_op)
+print
+print
+print
+display_data(show_facts_derived)

### 2014-11-12 implementation.tex
spacing
use tightitemize
Add N Queens cachegrind description. Add Miguel's suggestions.

@@
 database, performing join operations and then consuming and deriving facts.
-Optimizations such as join optimizations (to support efficient rule filtering)
-and fact updates are implemented by the compiler.
+The compiler implements join optimizations (to support efficient rule filtering)
+and fact updates (to avoid allocations).
 
@@
 
-\begin{itemize}
+\begin{tightitemize}
 \item \emph{Tries} are used exclusively to store persistent facts.
@@
   indexed by the appropriate argument.
-\end{itemize}
+\end{tightitemize}
 
@@
 
-\begin{description}
+\begin{tightdescription}
    \item[working]: the node is executing.
@@
    queue for processing.
-   \item[queue]: the node is active with new facts and is waiting in some queue
+   \item[queue]: the node is active with new facts and is in some queue waiting
    to be processed.
@@
    \item[coordinating]: the node is moving from one queue to another queue.
-\end{description}
+\end{tightdescription}
 
@@
    \centering
-   \includegraphics[width=0.4\textwidth]{node_states.pdf}
+   \includegraphics[width=0.3\textwidth]{node_states.pdf}
+\vspace*{.5ex}
    \scap{fig:node_states}{The node state machine as represented by the state variable. During
@@
 The main loop for each thread is to get an active node from one of its
-local queues and process all applicable rules for that node generating
-new facts which can result in new facts for the current node or other
-nodes.  The thread will continue processing facts for the current node
+local queues and process all applicable rules for that node, generating
+new facts, which can result in new facts for the current node or other
+nodes.  The thread continues processing facts for the current node
 until there are no new facts generated for it.  It will then pick
@@
 doubly linked lists known as the \emph{standard queue} and a pair of
-min/max heap known as the \emph{priority queue}.  The standard queue
+\emph{min/max} heap known as the \emph{priority queue}.  The standard queue
 contains nodes without priorities and supports push into tail, remove
@@
 implemented as a binary heap array. It supports the following
-operations: push into the heap, remove the min node, remove an
+operations: push into the heap, remove the \emph{min} node, remove an
 arbitrary node, remove half of the nodes (horizontal split), and
@@
 
-\begin{description}
+\begin{tightdescription}
    \item[New facts:] When a node executes in $T_1$ and derives facts
@@
    updates in order to move the node in $T_2$'s queues.
-\end{description}
+\end{tightdescription}
 

### 2014-11-12 nqueens.tex
Add N Queens cachegrind description. Add Miguel's suggestions.

@@
 columns are perfectly partitioned by the number of threads (in our
-case 13 threads).
+case 13 threads). We also analyzed N Queens execution with Valgrind's
+CacheGrind~\ref{Nethercote:2007:VFH:1250734.1250746} and found that the
+\texttt{Static} version has half the number of data caches misses (last level)
+of the \texttt{Regular} version and a slight more than half of the
+\texttt{Coordinated} version. This validates our hypothesis about the
+improvement seen in the \texttt{Static} version.
 

### 2014-11-12 intro.tex
Fixed several typos.

@@
-Writing parallel programs in sequential languages is hard because manipulating
+Parallel programming in sequential languages is hard because manipulating a
 shared state using multiple threads may result in data race conditions. Such
-issues are handled with low level constructs such as locks and condition
+issues are handled with low level constructs such as locks, semaphores and/or condition
 variables, requiring a fair amount of effort to get right.  Declarative
-programming has been hailed as a solution to this issue, since the problem of
+programming has been hailed as a alternative solution to this issue, since the problem of
 implementing the details of parallelism is moved from the programmer to the
-compiler and runtime. The programmer writes code without having to deal with
+compiler and runtime environment. The programmer writes code without having to deal with
 parallel programming constructs and the compiler automatically parallelizes the
-program in order to take advantage of multiple threads of execution.  The
+program in order to take advantage of multiple threads of execution.  This
 programming paradigm has been adopted with huge success in domain specific
@@
 programmer control over how execution is scheduled or how data is laid out,
-making it hard to achieve necessary efficiency. This introduces
+making it hard to improve efficiency. This introduces
 performance issues because even if the runtime system is able to
@@
 cannot easily deduce. Such information could make execution better in
-terms of run time time, memory usage, and scalability.
+terms of run time time, memory usage, and/or scalability.
 
@@
 state. In LM, computation is divided so that each node of the graph computes
-independently and is allowed to \scare{communicate} with other nodes.  Both
+independently but is allowed to \scare{communicate} with other nodes.  Both
 computation and communication happen through the derivation of logical rules
@@
 
-The CLM language features coordination primitives that can be used in the same
-way as any other primitive, in other words, they are specified with the same
+The CLM language features two kinds of coordination primitives that can be used in the same
+way as any other primitive, i.e., they are specified with the same
 syntax and semantics as the rest of the programming language. These coordination
@@
 the program is running on, e.g., scheduling and node placement on threads. The
-second kind of coordination facts are \emph{action facts} that when deleted in
-rules are used to apply a scheduling operation during execution. Coordination
+second kind of coordination facts are \emph{action facts} that when detected in
+rules, are used to apply a scheduling operation during execution. Coordination
 facts allow the programmer to write logical rules that depend on the current
@@
 desired coordination mechanisms. In \sectref{sec:applications} we present
-several applications and show coordination can improve programs without
+several applications and show how coordination can improve programs without
 destroying clarity or provability.

### 2014-11-12 related.tex
Fixed several typos.

@@
 \paragraph{Declarative Languages}
-Declarative programming paradigms such as \emph{logic} and \emph{functional}
+Programming paradigms such as \emph{logic} and \emph{functional}
 programming have been extensively exploited for their implicit parallelism.
@@
 
-GraphLab~\cite{GraphLab2010} is a C++ framework for developing graph-based parallel machine
+Another interesting system is GraphLab~\cite{GraphLab2010}, a C++ framework for developing graph-based parallel machine
 learning algorithms. GraphLab allows nodes to have read/write access to
@@
 of the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a
-process \texttt{p} in the tuple space and execute it in parallel.  Linda is be
+process \texttt{p} in the tuple space and execute it in parallel.  Linda was
 implemented on top of many popular languages by simply creating a communication
@@
 coordinate the scheduling of processing units, while the placement of data is
-left to the implementation. But even more importantly, CLM differs from those
+left to the implementation. CLM differs from those
 languages because coordination acts on data instead of processing units.
 The abstraction is then raised by considering data and algorithmic aspects of
-the program instead of focusing how processing units are used.
+the program instead of focusing on how processing units are used.
 Furthermore, the CLM language is both a coordination language and a computation
@@
 \paragraph{Forward-Chaining Logic Programming}
-
-Forward-chaining logic programming came into its own with the Datalog
+This paradigm came into its own with the Datalog
 language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
@@
 Datalog-like language for programming distributed ensembles of modular robots.
-Meld introduced the idea of sensing and action facts in order to sense and act
+Meld also introduced the idea of sensing and action facts in order to sense and act
 on the outside world, respectively. LCM uses sensing facts to sense the
-underlying system that the program is running on and action facts act on that
-system.
+underlying system and action facts to act on that system.

### 2014-11-12 language.tex
Fixed more typos.

@@
 Programs consist of a set of \emph{rules} and a \emph{database of facts}.
-Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
+Rules have the form \texttt{a(X), b(Y) -o c(X, Y)} and can be read as follows: if fact \texttt{a(X)}
 and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
@@
 associated tuple of values, e.g., the concrete values of \texttt{X} and
-\texttt{Y}, substituted for the
-arguments. Since CLM uses linear logic as its foundation, we distinguish between
+\texttt{Y}. Since CLM uses linear logic as its foundation, we distinguish between
 \emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
-process of deriving rule, while persistent facts are not.
-Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
-Next, the rules are recursively applied and the database is updated by adding
-new facts or deleting facts used during rule derivation.
+process of deriving a rule, while persistent facts are not.
+Program execution starts by adding the \emph{axioms} (the initial facts of the
+program) to the database. Next, rules are recursively applied and the
+database is updated by adding new facts or deleting facts used during rule derivation.
 When no more rules are applicable, the program terminates.
 
-CLM has been designed for writing programs that operate on graphs. CLM partitions
-the database by using the first argument of each fact. The first argument has
-type \emph{node} and represents a node in the graph. To achieve concurrency,
-our implementation partitions the database by the first argument, e.g.,
+CLM has been designed for writing programs that operate on graphs. To achieve
+concurrency, CLM partitions the database by using the first argument of each fact. The first argument has
+type \emph{node} and represents a node in the graph. For example,
 the fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while
@@
 every rule to facts with the same node so that nodes can derive rules
-independently. Although the body is restricted, the head of the rule may refer
+independently. The head of the rule may refer
 to any node as long as that node is referred to somewhere in the body. This allows
@@
 \paragraph{SSSP Example}\label{sect:ssspex}
-To make these ideas concrete, we present a simple
-program: the single source shortest path program~(SSSP). Later in the paper, we
+To make these ideas concrete, Fig.~\ref{code:shortest_path_program} presents
+a simple program: the single source shortest path program~(SSSP). Later in the paper, we
 add coordination facts to improve the execution of the program.
 
-The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
-with the declaration of the predicates. Predicates specify the facts
-used in the program. The first predicate, \texttt{edge}, is a persistent
-predicate that describes the relationship between the nodes of the graph,
-where the third argument represents the weight of the edge (the
-\texttt{route} modifier informs the compiler that the \texttt{edge} predicate
-determines the structure of the graph).
-The predicates \texttt{shortest} and \texttt{relax} are
-specified as linear facts and thus are deleted when deriving new facts.  The
-algorithm computes the shortest distance from node
-\texttt{@1} to all other nodes in the graph. Every node has a
-\texttt{shortest} fact that is improved with new \texttt{relax} facts.  Lines
-5-9 declare the axioms of the program: \texttt{edge} facts describe the
-graph; \texttt{shortest(A, +00, [])} is the initial shortest distance
-(infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
-algorithm by setting the initial distance from \texttt{@1} to \texttt{@1} to be
-0.
+The SSSP program starts (lines 1-3) with the declaration of the predicates.
+Predicates specify the facts used in the program. The first predicate,
+\texttt{edge}, is a persistent predicate that describes the
+relationship between the nodes of the graph, where the third argument
+represents the weight of the edge (the \texttt{route} modifier
+informs the compiler that the \texttt{edge} predicate
+determines the structure of the graph).  The predicates
+\texttt{shortest} and \texttt{relax} are specified as linear facts
+and thus are deleted when deriving new facts.  The algorithm computes
+the shortest distance from node \texttt{@1} to all other nodes in the
+graph. Every node has a \texttt{shortest} fact that is improved with
+new \texttt{relax} facts.  Lines 5-9 declare the axioms of the
+program: \texttt{edge} facts describe the graph; \texttt{shortest(A,
++00, [])} is the initial shortest distance (infinity) for all
+nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm by
+setting the initial distance from \texttt{@1} to \texttt{@1} to be 0.
 
@@
 \begin{dblfig}
+\vspace*{-2ex}
 \begin{center}
@@
    have been computed.}
+\vspace*{-1ex}
 \end{dblfig}
@@
 current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
-propagate new paths to the neighbors (lines 14-15) using a \emph{comprehension}.
+propagate new paths to the neighbors (lines 13-14) using a \emph{comprehension}.
 The comprehension iterates over the edges of node \texttt{A} and derives a new
@@
 current shortest path \texttt{D1} is shorter than the new path \texttt{D2} then
-delete the new \texttt{relax} fact and keep the old shortest path.
+delete the new \texttt{relax} fact and keep the current shortest path.
 
@@
 possible to either apply rules in either node \texttt{@2} or node
-\texttt{@3}. This decision depends largely on implementation factors such as node
-partitioning and number of threads in the system.
-Still, it is easy to prove that no matter the scheduling used,
-the final result present in Fig.~\ref{fig:shortest_path_program}~(c) is achieved.
+\texttt{@3}. This decision depends largely on implementation factors such
+as node partitioning and number of threads in the system.  Still, it is
+easy to prove that no matter the scheduling used, the final result, as
+presented in Fig.~\ref{fig:shortest_path_program}~(c), is achieved.
 

### 2014-11-13 coordination.tex
Add locking section.
Fix more typos.

@@
-The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
-declarative but its performance depends on the order in which nodes are
+The SSSP program is concise and declarative but its performance depends on the order in which nodes are
 executed. If nodes with greater distances are prioritized over other nodes, the
@@
 
-\begin{itemize}
+\begin{tightitemize}
    \item \code{set-priority(node A, float F)}: This sets the
@@
    temporarily, the priority of node \texttt{A} by \texttt{F}.
+   \item \code{remove-priority(node A)}: Removes the temporary priority from node
+   \texttt{A}.
    \item \code{schedule-next(node A)}: Changes the temporary priority of node
@@
    execution of the whole program.
-\end{itemize}
+\end{tightitemize}
 
@@
 
-\subsubsection{Partitioning facts}
+\subsection{Partitioning facts}
 
 We provide several coordination facts for dealing with node partitioning among
-the available threads executing the program. In terms of action
-facts, we have the following:
+the running threads. In terms of action facts, we have the following:
 
-\begin{itemize}
-   \item \code{set-cpu(node A, int C)}: Moves node \texttt{A} to thread
-   \texttt{C}.
+\begin{tightitemize}
+   \item \code{set-cpu(node A, int T)}: Moves node \texttt{A} to thread
+   \texttt{T}.
    \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
-   the thread of node \texttt{B}.
+   the thread of node \texttt{A}.
    \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
@@
    same thread indefinitely.
-\end{itemize}
+\end{tightitemize}
 
-For sensing facts, we have the following set of coordination facts:
+For partitioning facts, we have the following set of sensing facts:
 
-\begin{itemize}
-   \item \code{cpu-id(node A, node B, int C)}: Store at node \texttt{A} which
-   thread, \texttt{C}, \texttt{B} is actually running on.
+\begin{tightitemize}
+   \item \code{cpu-id(node A, node B, int T)}: Store at node \texttt{A} which
+   thread \texttt{T}, \texttt{B} is actually running on.
    \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
@@
    allowed to move between threads.
-\end{itemize}
+\end{tightitemize}
 
@@
 
-\begin{description}
+\begin{tightdescription}
    \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
@@
    \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
-\end{description}
+\end{tightdescription}
 

### 2014-11-13 heat-transfer.tex
small space savings
Fix more typos.

@@
 
-Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
+Fig.~\ref{results:ht} presents the scalability results for the regular
 and coordinated version. The dataset used is a square grid with an inner square
-with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
-for 16 threads, there's a 18\% reduction.
+with high heat nodes. Comparing the coordinated version with the regular
+version, with 1 thread there is a 33\% reduction in run time, while
+for 16 threads there is a 18\% reduction.
 
-To further improve locality, we split the second rule to avoid sending
+To further improve locality, we have split the second rule to avoid sending
 small $\delta$ values if the target node is in another thread
-(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Coordinated} line
-in Fig.~\ref{results:ht} shows that this performs better due to
-improved locality, exhibiting linear scalability over the regular
-version.  However, this come at the price of increased errors in the
-computed heat values.
+(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Regular} line
+in Fig.~\ref{results:ht} shows that this has the best performance.  However, this
+come at the price of increased errors in the computed heat values.
 

### 2014-11-13 implementation.tex
Fix more typos.

@@
 compared against implementations of similar programs but written in other
-languages. For instance, the LBP program was found to be around 2 times slower
-than GraphLab, while N Queens program was found to be around 10 to 15 slower
-than a sequential C implementation. Some of those programs were compared against
+languages. For instance, the Loopy Belief Propgation program was found to be around 2 times slower
+than GraphLab, while the N Queens program was found to be around 10 to 15 slower
+than a sequential C implementation. Some of those programs were also compared against
 a Python implementations and LM faired fairly well. We measured
@@
 Since we also performed general performance improvements, programs such as
-PageRank and Greedy Graph Coloring programs now perform better than before.
+PageRank and Greedy Graph Coloring now perform slightly better than before.

### 2014-11-13 locks.tex
Fill percent fail.
Fix more typos.

@@
 \hline
-SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\ \hline
-SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\ \hline
-SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\ \hline
+SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\
+SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\
+SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\
 \hline
-HT - Regular & 25M & 32M & 34M & 34M & 36M \\ \hline
-HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\ \hline
-HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\ \hline
+HT - Regular & 25M & 32M & 34M & 34M & 36M \\
+HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\
+HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\
 \hline
-LBP - Regular & 15M & 17M & 16M & 11M & 8M \\ \hline
-SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\ \hline
+LBP - Regular & 15M & 17M & 16M & 11M & 8M \\
+SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\
 \hline
-MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\ \hline
-MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\ \hline
+MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\
+MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\
 \hline
-N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
-N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\ \hline
-N Queens - Static & 57M & 44M & 1M & 1M & 151K \\ \hline
+N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\
+N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\
+N Queens - Static & 57M & 44M & 1M & 1M & 151K \\
 \hline
@@
 move nodes between queues and within the priority queue.  As we show
-here, this overhead is minimal and the resulting improved schedules
-and improved locality more than make up for the overhead.
+next, this overhead is minimal and is compensated by the resulting improved schedules
+and memory locality that more than make up for the overhead.
 
@@
 number of \emph{percolate-up}/\emph{percolate-down} operations
-executed for each manipuilation of the priority queue.  As expected,
+executed for each manipulation of the priority queue.  As expected,
 programs which rely on prioritizing the order of execution to obtain
@@
 We next investigate the effect of coordination on the syncronization
-overhead that comes froms from the use of more locks.  We measured the
+overhead that comes from the use of more locks.  We measured the
 \emph{basic locks}, the number of locks acquired
@@
 the lock on the first attempt.  Lock failures happen so infrequently
-(less than X\% \snote{please fill this in} of the time) that they
-cannot be seen on the figures.
+(less than 1\% of the time) that they cannot be seen on the figures.
 
@@
 for the SSSP program.  The first bar in each group shows the number of
-locks acquired for the uncoordinated version of the program.  The
-second bar shows the regular and coordination locks acquired for SSSP
-when every newly derived fact and priority operation is carried out
+locks acquired for the regular version of the program.  The
+second bar shows the basic and coordination locks acquired for the coordinated
+version of SSSP when every newly derived fact and priority operation is carried out
 immediately.  This leads to a significant increase in total locks
@@
 the lock overhead, but as can be seen in
-Table~\ref{queue_instructions} increase the number of queue operations
-slightly.  This slight increase comes from the fact the buffering
+Table~\ref{queue_instructions}, increases the number of queue operations
+slightly.  This slight increase comes from the fact that buffering
 delays the change in priorities and some longer paths can sometimes be
@@
 Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
-program.  The behavior is representative of the other programs.  There
-is drop in the total number of locks acquired and coordination locks
-represent a small fraction of the total locks acquired.  The locality
-coordination further reduces locking by keeping more derivation local
+program.  The behavior is representative of the other programs. In general, there
+is a drop in the total number of locks acquired and coordination locks
+represent a small fraction of the total number of locks acquired.  The \textbf{Local-Only}
+version further reduces locking by keeping more derivations local
 to a thread.
@@
 is more than offset by improved schedules and increased locality.
-While there are more locks to manage the total cost of lock operations
-does not change significantly (and is often reduced) when coordination
+While there are more locks to manage the total cost of lock operations,
+the numbers do not change significantly (and is often reduced) when coordination
 is added.  And, although the priority coordination significantly

### 2014-11-13 minimax.tex
Fix more typos.

@@
 
-The code in Fig.~\ref{minimax:check-end} deals with expanding the tree.
+The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
 The first three rules (lines 1-10) deal
 with the case where no children nodes are created and the last three rules
-(12-29) deal with the cases for creating new nodes.
-The rule in lines 12-25 generates new nodes using the
+(12-29) deal with the cases for that create new nodes. In particular, the two
+rules in lines 12-25 generate new nodes using the
 \texttt{exists} language construct, which creates a child node 
@@
 
-With coordination, we set the priority of a node to be its depth
-so the tree is expanded in a depth-first fashion (lines 15 and 22),
+With coordination, we set the priority of a node to be its depth (lines 15 and
+22) so the tree is expanded in a depth-first fashion,
 leading to a memory complexity of $\mathcal{O}(d t)$, where $d$

### 2014-11-13 nqueens.tex
Fix more typos.

@@
 other~\cite{8queens}. The challenge of finding all the
-distinct solutions to this problem is a good benchmark in designing
+distinct solutions to this problem is a well-known benchmark in designing
 parallel algorithms.
@@
 Since computation goes from the top row to the bottom row, not all
-placements of nodes to threads performane equally. This is especially
+placements of nodes to threads perform equally. This is especially
 true because the bottom rows tend to perform the most work. The best
@@
 case 13 threads). We also analyzed N Queens execution with Valgrind's
-CacheGrind~\ref{Nethercote:2007:VFH:1250734.1250746} and found that the
+CacheGrind~\cite{Nethercote:2007:VFH:1250734.1250746} and found that the
 \texttt{Static} version has half the number of data caches misses (last level)

### 2014-11-13 splash-bp.tex
removed redundant info about graphlab code
Fix more typos.

@@
 to evaluate rules in different orders.
-An example of such program is Loopy Belief Propagation.
-Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
+An example of such program is Loopy Belief Propagation~(LBP).
+LBP~\cite{Murphy99loopybelief} is an approximate inference algorithm
 used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
@@
 
-LBP is an algorithm that maps very well to the graph based model of CLM. In its
+LBP maps very well to the graph based model of CLM. In its
 original form, the belief values of nodes are computed by synchronous iterations.
@@
    nodes that have a positive priority (due to a new belief update) and are in the
-   same thread. Nodes are collected into list \texttt{L} (line 21) and
-   appended to list \texttt{Next}.
+   same thread. Nodes are collected into list \texttt{L} and
+   appended to list \texttt{Next} (line 21).
 
-   \item[First phase]: In the
-   third rule (lines 11-12), when the number of nodes of the tree reaches a
-   certain limit, a \texttt{first-phase} is generated to update the beliefs of
-   all nodes in the tree, starting from the leaves and ending at the root As the
-   nodes are updated, an \texttt{update} fact is derived to update the belief
+   \item[First phase]: When the number of nodes in the tree reaches a certain
+   limit, a \texttt{first-phase} (lines 11-12) is generated to update the beliefs of
+   all nodes in the tree. As the nodes are updated, starting from the leaves and
+   ending at the root, an \texttt{update} fact is derived to update the belief
    values (line 35).
@@
 second-phase(A, [], _)
-   -o *underline{set-priority(A, 0.0)}, inactive(A), update(A).
+   -o *underline{remove-priority(A)}, inactive(A), update(A).
 second-phase(A, [A], Back)
    -o update(A), inactive(Back),
-      inactive(A), *underline{set-priority(A, 0.0)}.
+      inactive(A), *underline{remove-priority(A)}.
 second-phase(A, [A, B | Next], Back)
@@
 \end{Verbatim}
-  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. CLM needs
-     50 lines of rules to implement splash trees, while GraphLab needs a
-     total of 350 lines of C++ to implement the same functionality.  Note
+  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. 
+%CLM needs 50 lines of rules to implement splash trees, while GraphLab needs a total of 350 lines of C++ to implement the same functionality.
+  Note
      that when a linear fact is prefixed by \texttt{@}, it indicates that the
@@
 GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine
-algorithms.  GraphLab provides the splash scheduler as part its the
-framework. It is 350 lines of C++ code.  With our coordination facts,
+algorithms.  GraphLab provides the splash scheduler as part of its the
+framework. It includes 350 lines of C++ code.  With our coordination facts,
 it is possible to create the necessary scheduling with only 12 rules.

### 2014-11-13 sssp.tex
Fix more typos.

@@
-Here we add coordination to the SSSP program described in \sectref{sect:ssspex}.
+We start by adding coordination to the SSSP program described before.
 The coordinated version of the
 SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
-\texttt{set-priority} (line 14). We also use a global program directive to order
+\texttt{set-priority} (line 14) and a global program directive to order
 priorities in ascending order (line 5).
@@
 Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
-although it applies rules using smarter ordering. The derivation of
-\texttt{set-priority} does not change the behavior of the logical rules and the
-code remains declarative.
+although it applies rules using smarter ordering and the code remains
+declarative. The derivation of \texttt{set-priority} does not change the
+behavior of the logical rules.
 
@@
 propagated because many SSSP distances are computed at the same time.
-However, we see a reduction of over 50\% in the number of
+However, we see a reduction of over 40\% in the number of
 derived facts when using the coordinated version \textbf{C} over the regular

### 2014-11-13 implementation.tex
Fixed a few typos.

@@
 compared against implementations of similar programs but written in other
-languages. For instance, the Loopy Belief Propgation program was found to be around 2 times slower
+languages. For instance, the Loopy Belief Propagation program was found to be around 2 times slower
 than GraphLab, while the N Queens program was found to be around 10 to 15 slower
@@
 the impact of the coordination mechanisms implemented for the CLM
-virtual machine on regular LM programs and saw very little to no degration.
+virtual machine on regular LM programs and saw very little to no degradation.
 Since we also performed general performance improvements, programs such as

### 2014-11-13 abstract.tex
Change the abstract slightly.

@@
 Declarative programming has been hailed as a promising approach to parallel
-programming since it hides the implementation details of parallelism away from the
-programmer. However, its advantage has also been its downfall
-as it leaves the programmer with no straightforward way
-to optimize programs for performance.
-In this paper, we introduce Coordinated Linear Meld, a concurrent forward-chaining linear logic programming
-language, with a declarative way to coordinate the execution of the program
-allowing the programmer to change
+programming since it hides the implementation details of parallelism away from
+the programmer. However, its advantage has also been its downfall as it leaves
+the programmer with no straightforward way to optimize programs for performance.
+In this paper, we introduce Coordinated Linear Meld, a concurrent
+forward-chaining linear logic programming language, with a declarative way to
+coordinate the execution of the program allowing the programmer to change
 computation scheduling and data layout. Our approach allows the programmer to
-write declarative parallel programs and then optionally use coordination 
-to fine tune programs, while keeping the program fully declarative and easy to
-reason about.
+write graph-based declarative programs and then optionally use coordination to fine
+tune performance, while keeping the program fully declarative and easy to reason
+about.

### 2014-11-13 results/locks/queue_operations.py
Generate unclutered tables.

@@ -34,17 +34,17 @@ def translate(num):
 
 def show(exp, name):
    count = exp.queue_instruction_count()
-   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
 
 def show_facts_per_op(exp, name):
    instrs = exp.queue_instruction_count()
    facts = exp.facts_derived_count()
    count = [round(float(ft)/float(it), 2) for it, ft in zip(instrs, facts)]
-   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
 
 def show_facts_derived(exp, name):
    count = exp.facts_derived_count()
-   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\ \hline"
+   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
 
 def show_separated(exp, name):
    normal = exp.normal_instruction_count()

### 2014-11-13 results/locks/queue_operations.py
Improve queue script.

@@ -26,15 +26,18 @@ nqueens_static = read_experiment("nqueens-static.txt")
 
 def translate(num):
    if num > 1000000:
-      return str(round(num/1000000, 1)) + "M"
+      return str(round(num/1000000.0, 1)) + "M"
    elif num > 1000:
-      return str(round(num/1000, 1)) + "K"
+      return str(round(num/1000.0, 1)) + "K"
    else:
       return str(num)
 
-def show(exp, name):
+def show_instructions0(exp):
    count = exp.queue_instruction_count()
-   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
+   return translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8])
+
+def show_instructions(exp, name):
+   print name + " & " + show_instructions0(exp) + " \\\\"
 
 def show_facts_per_op(exp, name):
    instrs = exp.queue_instruction_count()
@@ -42,14 +45,20 @@ def show_facts_per_op(exp, name):
    count = [round(float(ft)/float(it), 2) for it, ft in zip(instrs, facts)]
    print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
 
-def show_facts_derived(exp, name):
+def show_facts_derived0(exp):
    count = exp.facts_derived_count()
-   print name + " & " + translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8]) + " \\\\"
+   return translate(count[0]) + " & " + translate(count[1]) + " & " + translate(count[2]) + " & " + translate(count[4]) + " & " + translate(count[8])
+
+def show_facts_derived(exp, name):
+   print name + " & " + show_facts_derived0(exp) + "\\\\"
 
 def show_separated(exp, name):
    normal = exp.normal_instruction_count()
    prio = exp.heap_instruction_count()
[…]

### 2014-11-13 locks.tex
Fixed a few typos.
Add facts derived to table.

@@
-\begin{toptab}
-\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
+\begin{dbltab}
+   \begin{center}
+\footnotesize\begin{tabular}{ | l || c | c | c | c | c | c | c | c | c | c |}
 \hline
 \multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
-   Instructions}} \\ \cline{2-6}
-& 1 & 2 & 4 & 8 & 16 \\ \hline
+   Instructions}} & \multicolumn{5}{c |}{\textbf{Facts Derived}} \\ \cline{2-11}
+& 1 & 2 & 4 & 8 & 16 & 1 & 2 & 4 & 8 & 16\\ \hline
 \hline
-SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\
-SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\
-SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\
+SSSP - Regular & 199.1K & 203.1K & 202.7K & 203.8K & 206.4K & 3.3M & 3.0M & 3.4M & 3.5M & 3.3M \\
+SSSP - Coordinated & 959.0K & 924.3K & 891.0K & 966.3K & 1.0M & 2.0M & 2.3M & 2.3M & 2.2M & 2.3M \\
+SSSP - Buffered & 1.0M & 1.0M & 1.1M & 1.0M & 1.1M & 2.0M & 2.1M & 2.3M & 2.3M & 2.4M \\
 \hline
-HT - Regular & 25M & 32M & 34M & 34M & 36M \\
-HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\
-HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\
+HT - Regular & 25.6M & 33.3M & 34.8M & 36.1M & 34.8M & 235.7M & 298.3M & 312.1M & 317.5M & 310.3M \\
+HT - Coordinated & 51.6M & 55.2M & 85.9M & 79.9M & 75.6M & 129.2M & 183.4M & 254.1M & 242.0M & 252.4M \\
+HT - Local-Only & 51.6M & 54.9M & 81.0M & 76.7M & 67.3M & 129.2M & 180.8M & 221.5M & 217.6M & 207.4M \\
 \hline
-LBP - Regular & 15M & 17M & 16M & 11M & 8M \\
-SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\
+LBP - Regular & 15.7M & 17.8M & 16.0M & 11.3M & 8.5M & 160.2M & 171.0M & 149.9M & 107.5M & 85.1M \\
+SBP - Coordinated & 97.2M & 97.4M & 90.0M & 84.3M & 79.2M & 76.8M & 80.7M & 79.2M & 79.0M & 79.3M \\
+\hline \hline
+MiniMax - Regular & 16.4M & 18.3M & 20.5M & 19.4M & 20.9M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
+MiniMax - Coordinated & 52.0M & 52.0M & 52.0M & 52.0M & 52.0M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
 \hline
-MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\
-MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\
-\hline
-N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\
-N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\
-N Queens - Static & 57M & 44M & 1M & 1M & 151K \\
+N Queens - Regular & 8.4K & 14.4K & 17.7K & 16.6K & 19.3K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
+N Queens - Coordinated & 57.1M & 24.7M & 1.1M & 141.1K & 41.6K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
+N Queens - Static & 57.1M & 47.2M & 3.4M & 1.5M & 209.2K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
 \hline
@@
 \vspace*{.5ex}
-\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
+\end{center}
+\scap{queue_instructions}{Total number of queue \scare{instructions} and facts derived per program. An
    operation to a regular queue counts as 1 instruction, while an operation to
@@
    \emph{percolate-up}/\emph{percolate-down} steps performed.}
-\end{toptab}
+\end{dbltab}
 
@@
 versions of the programs require significantly more queue operations.
-The data for this table comes from recording, during execution, the
-number of normal queue operations executed (each costs 1), plus the
+The data for this table comes from recording the number of \emph{queue instructions} and the number of facts derived.
+Queue instructions represent the
+number of normal queue operations executed (each costs 1) plus the
 number of \emph{percolate-up}/\emph{percolate-down} operations
@@
 adds a significant number of total queue operations, but the resulting
-overhead is more than compensated for by an improved schedule.
+overhead is more than compensated for by an improved schedule and reduction
+in number of facts produced. For MiniMax and N Queens, the number of facts derived
+is the same for all configurations, as expected. The performance seen for those programs
+arise from reduced memory usage and improved memory locality.
 
@@
 
-We next investigate the effect of coordination on the syncronization
+We next investigate the effect of coordination on the synchronization
 overhead that comes from the use of more locks.  We measured the

### 2014-11-13 sssp.tex
Add facts derived to table.

@@
 program computes the SSSP for 20\% of the nodes of the graph.
-There are some situations where unnecessary facts are propagated
+The scalability results show that the coordinated
+version using 16 threads is 1.3 times faster than the regular version using 16
+threads while still showing good scaling. Overall, the coordinated version produces 40\% less facts
+derived when compared to the regular version.
+Note that there are some situations where unnecessary facts are propagated
 because although the shortest distance is selected locally, sub-optimal distances may be
 propagated because many SSSP distances are computed at the same time.
-However, we see a reduction of over 40\% in the number of
-derived facts when using the coordinated version \textbf{C} over the regular
-version \textbf{R}. The results also show that the coordinated
-version using 16 threads is 1.3 times faster than the regular version using 16
-threads while still showing good scaling.
 
@@
    \begin{center}
-      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
-         \hline                       
-         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
-         1 & 333K & 206K \\
-         2 & 300K & 210K \\
-         4 & 316K & 208K \\
-         8 & 328K & 211K \\
-         16 & 343K & 212K \\
-         \hline  
-         \end{tabular}
-         \normalsize
-      }
-      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
+      \includegraphics[width=5cm]{results/shortest-uspowergrid.png}
    \end{center}
-   \scap{results:sssp_uspowergrid}{Results for SSSP run on the 
-   US power grid network.  On the left we show the number of facts
-   derived for the regular \textbf{R} and coordinated
-   version \textbf{C} using a variable number of threads \textbf{\#
-   T}. On the right, we show the scalability of the regular and
-   coordinated version. The speedup values are computed using the
+   \scap{results:sssp_uspowergrid}{Scalability results for SSSP run on the 
+   US power grid network. The speedup values are computed using the
    execution time for 1 thread.}

### 2014-11-13 coordination.tex
Several corrections. Reduce size of texttt.

@@
 executed. If nodes with greater distances are prioritized over other nodes, the
-program will generate more \texttt{relax} facts since it will take longer to
+program will generate more \mytt{relax} facts since it will take longer to
 reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
-clear that the best scheduling is the following: \texttt{@1},
-\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
+clear that the best scheduling is the following: \mytt{@1},
+\mytt{@3}, \mytt{@2} and then \mytt{@4}, where only 4 \mytt{relax}
 facts are generated. If we had decided to process nodes in order
-\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
-\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
+\mytt{@1}, \mytt{@2}, \mytt{@4}, \mytt{@3}, \mytt{@4},
+\mytt{@2}, then 6 \mytt{relax} facts would have been generated.
 The optimal solution for SSSP is to schedule the node with the
@@
 
-\newcommand{\code}[1]{\texttt{\small{#1}}}
+\newcommand{\code}[1]{\mytt{\small{#1}}}
 
@@
    \item \code{set-priority(node A, float F)}: This sets the
-   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
-   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
+   temporary priority of \mytt{A} to \mytt{F}. If \mytt{A} has a priority
+   \mytt{F'}, we only change the priority if \mytt{F} is higher. The programmer
    can decide if priorities are to be ordered in ascending or descending order.
    \item \code{add-priority(node A, float F)}: Increases,
-   temporarily, the priority of node \texttt{A} by \texttt{F}.
+   temporarily, the priority of node \mytt{A} by \mytt{F}.
    \item \code{remove-priority(node A)}: Removes the temporary priority from node
-   \texttt{A}.
+   \mytt{A}.
    \item \code{schedule-next(node A)}: Changes the temporary priority of node
-   \texttt{A} to be 1 unit higher than the current highest priority.
+   \mytt{A} to be 1 unit higher than the current highest priority.
    \item \code{set-default-priority(node, float)}: Sets the default
@@
 CLM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
-priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
+priority \mytt{P} of node \mytt{B} (at node \mytt{A}).
 Sensing facts are only used in the body of the rules in order
@@
 creates the necessary code to re-derive them without programmer interaction.
-Likewise, \texttt{set-priority} and \texttt{set-default-priority} update the
-value of \texttt{priority} facts by consuming and re-deriving them but this is
+Likewise, \mytt{set-priority} and \mytt{set-default-priority} update the
+value of \mytt{priority} facts by consuming and re-deriving them but this is
 done automatically by the runtime system.}
@@
 \begin{tightitemize}
-   \item \code{set-cpu(node A, int T)}: Moves node \texttt{A} to thread
-   \texttt{T}.
-   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
-   the thread of node \texttt{A}.
-   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
+   \item \code{set-cpu(node A, int T)}: Moves node \mytt{A} to thread
+   \mytt{T}.
+   \item \code{set-affinity(node A, node B)}: Places node \mytt{B} in
+   the thread of node \mytt{A}.
+   \item \code{set-moving(node A)}: Allows node \mytt{A} to move freely
    between threads.
-   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
+   \item \code{set-static(node A)}: Forces node \mytt{A} to stay in the
    same thread indefinitely.
@@
 \begin{tightitemize}
-   \item \code{cpu-id(node A, node B, int T)}: Store at node \texttt{A} which
-   thread \texttt{T}, \texttt{B} is actually running on.
-   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
+   \item \code{cpu-id(node A, node B, int T)}: Store at node \mytt{A} which
+   thread \mytt{T}, \mytt{B} is actually running on.
+   \item \code{moving(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is allowed
    to move between threads.
-   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
+   \item \code{static(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is not
    allowed to move between threads.
@@
 \begin{tightdescription}
-   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
-   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$.
-Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
-   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
+   \item[\mytt{priority @order ORDER.}] \mytt{ORDER} can be either \mytt{asc} or \mytt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
+   \item[\mytt{priority @initial P.}] The \mytt{initial} statement informs the runtime system that all nodes must start with priority $P$.
+Alternatively, the programmer can define an \mytt{set-priority(A, P)} axiom.
+   \item[\mytt{priority @static.}] The \mytt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
 \end{tightdescription}

### 2014-11-13 heat-transfer.tex
Several corrections. Reduce size of texttt.

@@
 Fig.~\ref{code:ht} shows the HT rules that send new heat values to
-neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
+neighbor nodes. In the first rule we added \mytt{add-priority} to increase the priority of the neighbor
 nodes if the current node has a large $\delta$. The idea is to prioritize the
-computation of heat values of nodes (using \texttt{update}) that have a neighbor
-that changed significantly. Multiple \texttt{add-priority} facts will
+computation of heat values of nodes (using \mytt{update}) that have a neighbor
+that changed significantly. Multiple \mytt{add-priority} facts will
 increase the priority of a node so that nodes with multiple large deltas will
@@
    \begin{center}
-      \includegraphics[width=5.5cm]{results/new-heat-transfer-80}
+      \includegraphics[width=\figsize]{results/new-heat-transfer-80}
    \end{center}

### 2014-11-13 implementation.tex
Several corrections. Reduce size of texttt.

@@
 are compiled into special instructions that inspect the state of the virtual
-machine. For example, the directive \texttt{node-priority} will inspect the
+machine. For example, the fact \mytt{priority} will inspect the
 target node, retrieve the current priority and assign the priority to a
@@
 
-The \texttt{next} and \texttt{prev} pointers of the regular queue are
+The \mytt{next} and \mytt{prev} pointers of the regular queue are
 part of the node structure in order to save space. These pointers are

### 2014-11-13 language.tex
Several corrections. Reduce size of texttt.

@@
 Programs consist of a set of \emph{rules} and a \emph{database of facts}.
-Rules have the form \texttt{a(X), b(Y) -o c(X, Y)} and can be read as follows: if fact \texttt{a(X)}
-and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
-the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
-the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
-A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
-associated tuple of values, e.g., the concrete values of \texttt{X} and
-\texttt{Y}. Since CLM uses linear logic as its foundation, we distinguish between
+Rules have the form \mytt{a(X), b(Y) -o c(X, Y)} and can be read as follows: if fact \mytt{a(X)}
+and fact \mytt{b(Y)} exist in the database then \mytt{c(X, Y)} is added to
+the database. The expression \mytt{a(X), b(Y)} is called the \emph{body} of
+the rule and \mytt{c(X, Y)} is the \emph{head} of the rule.
+A fact is a predicate, e.g., \mytt{a}, \mytt{b} or \mytt{c}, and its
+associated tuple of values, e.g., the concrete values of \mytt{X} and
+\mytt{Y}. Since CLM uses linear logic as its foundation, we distinguish between
 \emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
@@
 type \emph{node} and represents a node in the graph. For example,
-the fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while
-fact \texttt{p(@2)} is stored in node \texttt{@2}. CLM restricts the body of
+the fact \mytt{f(@1, 2)} is stored in node \mytt{@1}, while
+fact \mytt{p(@2)} is stored in node \mytt{@2}. CLM restricts the body of
 every rule to facts with the same node so that nodes can derive rules
@@
 Predicates specify the facts used in the program. The first predicate,
-\texttt{edge}, is a persistent predicate that describes the
+\mytt{edge}, is a persistent predicate that describes the
 relationship between the nodes of the graph, where the third argument
-represents the weight of the edge (the \texttt{route} modifier
-informs the compiler that the \texttt{edge} predicate
+represents the weight of the edge (the \mytt{route} modifier
+informs the compiler that the \mytt{edge} predicate
 determines the structure of the graph).  The predicates
-\texttt{shortest} and \texttt{relax} are specified as linear facts
+\mytt{shortest} and \mytt{relax} are specified as linear facts
 and thus are deleted when deriving new facts.  The algorithm computes
-the shortest distance from node \texttt{@1} to all other nodes in the
-graph. Every node has a \texttt{shortest} fact that is improved with
-new \texttt{relax} facts.  Lines 5-9 declare the axioms of the
-program: \texttt{edge} facts describe the graph; \texttt{shortest(A,
+the shortest distance from node \mytt{@1} to all other nodes in the
+graph. Every node has a \mytt{shortest} fact that is improved with
+new \mytt{relax} facts.  Lines 5-9 declare the axioms of the
+program: \mytt{edge} facts describe the graph; \mytt{shortest(A,
 +00, [])} is the initial shortest distance (infinity) for all
-nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm by
-setting the initial distance from \texttt{@1} to \texttt{@1} to be 0.
+nodes; and \mytt{relax(@1, 0, [@1])} starts the algorithm by
+setting the initial distance from \mytt{@1} to \mytt{@1} to be 0.
 
@@
 \scap{fig:shortest_path_program}{Graphical representation of the SSSP program. (a) represents the
-   program after propagating initial distance at node \texttt{@1}, followed by
-   (b) where the first rule is applied in node \texttt{@3}. (c)
+   program after propagating initial distance at node \mytt{@1}, followed by
+   (b) where the first rule is applied in node \mytt{@2}. (c)
    represents the state of the final program, where all the shortest paths
@@
 The first rule of the program (lines 11-14) reads as following: if the current
-\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
-than a new path \texttt{relax} with distance \texttt{D2}, then replace the
-current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
+\mytt{shortest} path \mytt{P1} with distance \mytt{D1} is larger
+than a new path \mytt{relax} with distance \mytt{D2}, then replace the
+current shortest path with \mytt{D2}, delete the new \mytt{relax} path and
 propagate new paths to the neighbors (lines 13-14) using a \emph{comprehension}.
-The comprehension iterates over the edges of node \texttt{A} and derives a new
-\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
-where \texttt{W} is the weight of the edge. For
+The comprehension iterates over the edges of node \mytt{A} and derives a new
+\mytt{relax} fact for each node \mytt{B} with the distance \mytt{D2 + W},
+where \mytt{W} is the weight of the edge. For
 example, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
-\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
-and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
-applying the same rule but at node \texttt{2}.
+\mytt{@1} where two new \mytt{relax} facts are derived at node \mytt{@2}
+and \mytt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
+applying the same rule but at node \mytt{2}.
 
 The second rule of the program (lines 16-17) is read as following: if the
-current shortest path \texttt{D1} is shorter than the new path \texttt{D2} then
-delete the new \texttt{relax} fact and keep the current shortest path.
+current shortest path \mytt{D1} is shorter than the new path \mytt{D2} then
+delete the new \mytt{relax} fact and keep the current shortest path.
 
@@
 after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
-possible to either apply rules in either node \texttt{@2} or node
-\texttt{@3}. This decision depends largely on implementation factors such
+possible to either apply rules in either node \mytt{@2} or node
+\mytt{@3}. This decision depends largely on implementation factors such
 as node partitioning and number of threads in the system.  Still, it is

### 2014-11-13 locks.tex
Several corrections. Reduce size of texttt.

@@
 \multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
-   Instructions}} & \multicolumn{5}{c |}{\textbf{Facts Derived}} \\ \cline{2-11}
+   Operations}} & \multicolumn{5}{c |}{\textbf{Facts Derived}} \\ \cline{2-11}
 & 1 & 2 & 4 & 8 & 16 & 1 & 2 & 4 & 8 & 16\\ \hline
@@
 \end{center}
-\scap{queue_instructions}{Total number of queue \scare{instructions} and facts derived per program. An
+\scap{queue_instructions}{Total number of queue operations and facts derived per program. An
    operation to a regular queue counts as 1 instruction, while an operation to
@@
 versions of the programs require significantly more queue operations.
-The data for this table comes from recording the number of \emph{queue instructions} and the number of facts derived.
+The data for this table comes from recording the number of \emph{queue
+operations} and facts derived.
 Queue instructions represent the
@@
   \scap{locks:SSSP}{Locking statistics for SSSP.
-     \texttt{Coordinated} refers to the program in
-     Fig.~\ref{results:sssp_uspowergrid}, while \texttt{Buffered} is a more
+     \mytt{Coordinated} refers to the program in
+     Fig.~\ref{results:sssp_uspowergrid}, while \mytt{Buffered} is a more
      optimized version of the same program where coordination operations are
@@
 number of locks acquired to perform coordination operations.  We also
-measured how often \texttt{lock()} operations that fail to acquired
+measured how often \mytt{lock()} operations that fail to acquired
 the lock on the first attempt.  Lock failures happen so infrequently
@@
 is a drop in the total number of locks acquired and coordination locks
-represent a small fraction of the total number of locks acquired.  The \textbf{Local-Only}
-version further reduces locking by keeping more derivations local
-to a thread.
+represent a small fraction of the total number of locks acquired. \textbf{Local-Only}
+further reduces locking by keeping more derivations local to a thread.
 
@@
 total facts derived.
-
-\iffalse
-
-  The
-SSSP program is The first program we measured was SSSP, represented in
-Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
-Coordinated versions, we have the Buffered version where we optimized
-the virtual machine to buffer coordination actions before applying
-them. The idea is to avoid applying multiple \texttt{set-priority}
-operations to the same node during a single node execution. We thus
-gather the higher priority values and then perform coordination after
-the node has executed.  Although the number of locks is significantly
-reduced, we did not see a similar performance improvement, since the
-Buffered version is only 5\% faster than the Coordinated version.
-
-In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
-regular version, the coordinated version and the local-only version as in
-Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
-coordination and then a further reduction in the local-only version because less
-facts are exchanged between threads.
-
-Finally, in Fig.~\ref{locks:LBP} we present lock statistics for LBP and SBP. LBP
-is represented as the Regular bar, while SBP is represented as the Coordinated
-bar. We notice that LBP has quite a lot of locking for 1 thread since LBP
-is an asynchronous algorithm and tends to be a slow sequential algorithm,
-requiring more derivations when using 1 thread. The numbers go down as more
-threads are used. For SBP, there is a consistent amount of locking without
-little variations between the number of threads used. This is because threads
-interact very little with each other and the amount of work to do is the same
-for all threads.
-
-\begin{topfig}
-   \begin{center}
-      \includegraphics[width=6.5cm]{results/locks/bp.png}
-   \end{center}
-  \scap{locks:LBP}{Locking statistics for LBP and SBP. Locking is constant for
-     SBP because threads do not interact with each other.}
-\end{topfig}
-
-\paragraph{Queue Instructions}
-
-Table~\ref{queue_instructions} presents the total queue instructions for several
-programs. All the coordinated programs require far more queue instructions than
-the regular version. For instance, the coordinated SSSP program requires 4 to 5
-times more instructions to perform queue operations. Interestingly, the buffered
-version requires even more instructions although it performs less overall
-coordination. It is likely that this is due to increased fact computation and
-subsequent queue manipulation because coordination operations are buffered.
-We argue that the 5\% performance improvement in the
-buffered version is due mostly to reduced locking.
-
-For the HT program, the coordinated version requires twice more queue
-instructions than the regular version. As expected, the local-only version
-requires less queue operations because the number of derived facts is reduced.
-The MiniMax program also sees twice as many queue instructions using
-coordination. This does not happen with the N Queens program, where the number
-of queue instructions is far greater when using coordination, especially when
-using a small number of threads. We argue that this happens because higher
-priority nodes at the bottom are often added to the priority queue, resulting in
-many \emph{percolate-up} operations.
-
-These experiments indicate that handling coordination has a small cost in terms
-of synchronization but some cost in terms of queue operations with
-up to 5 times more queue instructions when using priority queues.
-Since coordination reduces the number of facts derived due to better scheduling
-decisions, the costs of manipulating priority queues become irrelevant.
-\fi

### 2014-11-13 minimax.tex
Several corrections. Reduce size of texttt.

@@
 rules in lines 12-25 generate new nodes using the
-\texttt{exists} language construct, which creates a child node 
-\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
-and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
+\mytt{exists} language construct, which creates a child node 
+\mytt{B}. We link \mytt{B} with its parent (\mytt{parent(B, A)})
+and kick start the expansion of that node \mytt{B} by adding a \mytt{play}
 fact.
@@
 
-We also take advantage of memory locality by using \texttt{set-static} (line
+We also take advantage of memory locality by using \mytt{set-static} (line
 14), so that nodes after a certain level are not stolen by other threads. While

### 2014-11-13 nqueens.tex
Several corrections. Reduce size of texttt.

@@
 placement is then to split the board vertically so each thread gets
-the same number of columns with axiom \texttt{set-cpu(A, vertical(X, Y)}, where
-\texttt{X} and \texttt{Y} are the coordinates of the board. Our experiments show that, on a shared
+the same number of columns with axiom \mytt{set-cpu(A, vertical(X, Y)}, where
+\mytt{X} and \mytt{Y} are the coordinates of the board. Our experiments show that, on a shared
 memory system, it does not matter much if we start with a bad
@@
 needs to be iterated before extended with a new position.
-To accomplish this, we used coordination axioms \texttt{set-default-priority(A,
-X)}, where \texttt{X} is the row number (0 is the top row) in order to
+To accomplish this, we used coordination axioms \mytt{set-default-priority(A,
+X)}, where \mytt{X} is the row number (0 is the top row) in order to
 prioritize the evaluation of squares closer to the bottom of the board.
@@
 memory locality since threads will often steal nodes.  We then
-added \texttt{set-static(A)} as a coordination axiom and saw
-performance improvement (\textbf{Static}), especially when the board
-columns are perfectly partitioned by the number of threads (in our
-case 13 threads). We also analyzed N Queens execution with Valgrind's
-CacheGrind~\cite{Nethercote:2007:VFH:1250734.1250746} and found that the
-\texttt{Static} version has half the number of data caches misses (last level)
-of the \texttt{Regular} version and a slight more than half of the
-\texttt{Coordinated} version. This validates our hypothesis about the
-improvement seen in the \texttt{Static} version.
+added \mytt{set-static(A)} as a coordination axiom (line
+\textbf{Static/Regular}) and saw a significant drop in data cache misses and an
+improvement in performance, specially when using 13 threads (the number of
+columns).
 
@@
    \begin{center}
-      \includegraphics[width=5.5cm]{results/8queens-13.png}
+      \includegraphics[width=\figsize]{results/8queens-13.png}
    \end{center}

### 2014-11-13 related.tex
Several corrections. Reduce size of texttt.

@@
 that can be manipulated using the following coordination directives:
-\texttt{out(t)} to write a tuple \texttt{t} into the tuple space; \texttt{in(t)}
-to read a tuple using the template \texttt{t}; \texttt{rd(t)} to retrieve a copy
-of the tuple \texttt{t} from the tuple space; and \texttt{eval(p)} to add a
-process \texttt{p} in the tuple space and execute it in parallel.  Linda was
+\mytt{out(t)} to write a tuple \mytt{t} into the tuple space; \mytt{in(t)}
+to read a tuple using the template \mytt{t}; \mytt{rd(t)} to retrieve a copy
+of the tuple \mytt{t} from the tuple space; and \mytt{eval(p)} to add a
+process \mytt{p} in the tuple space and execute it in parallel.  Linda was
 implemented on top of many popular languages by simply creating a communication

### 2014-11-13 splash-bp.tex
Several corrections. Reduce size of texttt.

@@
 \begin{tightdescription}
-   \item[Tree building]: Each node has a \texttt{inactive} fact that is used to
+   \item[Tree building]: Each node has a \mytt{inactive} fact that is used to
    start the tree building process. When the highest priority node is picked, a
-   \texttt{tree} is created that will navigate through the tree. In lines 15-21,
+   \mytt{tree} is created that will navigate through the tree. In lines 15-21,
    we use an \emph{aggregate}~\cite{cruz-iclp14} to gather all the neighbor
    nodes that have a positive priority (due to a new belief update) and are in the
-   same thread. Nodes are collected into list \texttt{L} and
-   appended to list \texttt{Next} (line 21).
+   same thread. Nodes are collected into list \mytt{L} and
+   appended to list \mytt{Next} (line 21).
 
    \item[First phase]: When the number of nodes in the tree reaches a certain
-   limit, a \texttt{first-phase} (lines 11-12) is generated to update the beliefs of
+   limit, a \mytt{first-phase} (lines 11-12) is generated to update the beliefs of
    all nodes in the tree. As the nodes are updated, starting from the leaves and
-   ending at the root, an \texttt{update} fact is derived to update the belief
+   ending at the root, an \mytt{update} fact is derived to update the belief
    values (line 35).
@@
 
-The \texttt{set-static} and \texttt{set-cpu} action facts are used in
+The \mytt{set-static} and \mytt{set-cpu} action facts are used in
 line 2 to (1) force nodes to stay in the thread and (2) 
@@
   Note
-     that when a linear fact is prefixed by \texttt{@}, it indicates that the
+     that when a linear fact is prefixed by \mytt{@}, it indicates that the
      fact is going to be re-derived.}
@@
    \begin{center}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/system_belief-propagation-400.png}}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/system_splash-bp-400.png}}
-      \subfloat[]{\includegraphics[width=5.5cm]{results/system_improve_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=\figsize]{results/system_belief-propagation-400.png}}
+      \subfloat[]{\includegraphics[width=\figsize]{results/system_splash-bp-400.png}}
+      \subfloat[]{\includegraphics[width=\figsize]{results/system_improve_belief-propagation-400.png}}
    \end{center}

### 2014-11-13 sssp.tex
Several corrections. Reduce size of texttt.

@@
 SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
-\texttt{set-priority} (line 14) and a global program directive to order
+\mytt{set-priority} (line 14) and a global program directive to order
 priorities in ascending order (line 5).
@@
 although it applies rules using smarter ordering and the code remains
-declarative. The derivation of \texttt{set-priority} does not change the
-behavior of the logical rules.
+declarative. The proof of correctness considers that, eventually, the shortest
+path is computed at all nodes of the graph until all shortest paths are found.
+The derivation of \mytt{set-priority} does not change this
+proof.
 
@@
    \begin{center}
-      \includegraphics[width=5cm]{results/shortest-uspowergrid.png}
+      \includegraphics[width=\figsize]{results/shortest-uspowergrid.png}
    \end{center}

### 2014-11-13 abstract.tex
small edit

@@
-Declarative programming has been hailed as a promising approach to parallel
-programming since it hides the implementation details of parallelism away from
-the programmer. However, its advantage has also been its downfall as it leaves
-the programmer with no straightforward way to optimize programs for performance.
-In this paper, we introduce Coordinated Linear Meld, a concurrent
-forward-chaining linear logic programming language, with a declarative way to
-coordinate the execution of the program allowing the programmer to change
-computation scheduling and data layout. Our approach allows the programmer to
-write graph-based declarative programs and then optionally use coordination to fine
-tune performance, while keeping the program fully declarative and easy to reason
-about.
+Declarative programming has been hailed as a promising approach to
+parallel programming since it hides the implementation details of
+parallelism from the programmer. However, its advantage has also been
+its downfall as it leaves the programmer with no straightforward way
+to optimize programs for performance.  In this paper, we introduce
+Coordinated Linear Meld, a concurrent forward-chaining linear logic
+programming language, with a declarative way to coordinate the
+execution of the program allowing the programmer to change computation
+scheduling and data layout. Our approach allows the programmer to
+write graph-based declarative programs and then optionally use
+coordination to fine tune performance, while keeping the program fully
+declarative and easy to reason about.

### 2014-11-13 heat-transfer.tex
space saving edits

@@
 
-To further improve locality, we have split the second rule to avoid sending
+To further improve locality, we split the second rule to avoid sending
 small $\delta$ values if the target node is in another thread
-(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Regular} line
-in Fig.~\ref{results:ht} shows that this has the best performance.  However, this
-come at the price of increased errors in the computed heat values.
+(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Regular} line in
+Fig.~\ref{results:ht} shows that this performs best.
+However, this come at the price of increased errors in the computed
+heat values.
 
@@
       slightly worse scalability due to a reduction in work available.}
+\vspace*{-2ex}
 \end{topfig}

### 2014-11-13 implementation.tex
space saving edits

@@
 
-The compiler translates each rule to a procedure and a list of facts that need
-to exist to satisfy the rule. This procedure is executed by the virtual
+The compiler translates each rule to a procedure and a list of facts needed
+to satisfy the rule. This procedure is executed by the virtual
 machine whenever all the facts mentioned in the body exist.
@@
    moved to another thread.
-   \item[coordinating]: the node is moving from one queue to another queue.
+   \item[coordinating]: the node is moving from one queue to another.
 \end{tightdescription}
@@
 
-In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
-compared against implementations of similar programs but written in other
-languages. For instance, the Loopy Belief Propagation program was found to be around 2 times slower
-than GraphLab, while the N Queens program was found to be around 10 to 15 slower
-than a sequential C implementation. Some of those programs were also compared against
-a Python implementations and LM faired fairly well. We measured
-the impact of the coordination mechanisms implemented for the CLM
-virtual machine on regular LM programs and saw very little to no degradation.
-Since we also performed general performance improvements, programs such as
-PageRank and Greedy Graph Coloring now perform slightly better than before.
+In the paper~\cite{cruz-ppdp14}, LM programs were compared to similar
+programs written in other languages. The Loopy Belief Propagation
+program was around 2 times slower than GraphLab, while the N Queens
+program was 10 to 15x slower than a sequential C implementation. The
+LM programs had similar performance to programs written in Python.
+The introduction of coordination in CLM has little or no impact on the
+overall performance of these programs.  In fact, due to some
+improvements in the virtual machine, programs such as PageRank and
+Greedy Graph Coloring now perform slightly better than before.

### 2014-11-13 splash-bp.tex
space saving edits

@@
 \begin{dblfig}
+\vspace*{-1ex}
    \begin{center}
@@
       times faster than LBP.}
+\vspace*{-1ex}
 \end{dblfig}
@@
 In this program, coordination assumes a far more important role than
-we have seen before. Coordination rules fully drive the behavior of
-the algorithm and although the final result of the algorithm is
-statistically identical to the original algorithm, SBP works very
-differently than LBP.  SBP is also implemented in
-GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine
-algorithms.  GraphLab provides the splash scheduler as part of its the
-framework. It includes 350 lines of C++ code.  With our coordination facts,
-it is possible to create the necessary scheduling with only 12 rules.
+we have seen before. Coordination rules drive the behavior of the
+algorithm and while the result of the algorithm is statistically
+identical to the original algorithm, SBP works very differently than
+LBP.  SBP is also implemented in GraphLab~\cite{GraphLab2010}, a C++
+framework for writing machine algorithms.  GraphLab provides the
+splash scheduler as part of its the framework. It includes 350 lines
+of C++ code.  With our coordination facts, it is possible to create
+the necessary scheduling with only 12 rules.
 

### 2014-11-13 sssp.tex
space saving edits

@@
 
-Fig.~\ref{results:sssp_uspowergrid} shows experimental results when the
-program computes the SSSP for 20\% of the nodes of the graph.
-The scalability results show that the coordinated
-version using 16 threads is 1.3 times faster than the regular version using 16
-threads while still showing good scaling. Overall, the coordinated version produces 40\% less facts
-derived when compared to the regular version.
-Note that there are some situations where unnecessary facts are propagated
-because although the shortest distance is selected locally, sub-optimal distances may be
-propagated because many SSSP distances are computed at the same time.
+Fig.~\ref{results:sssp_uspowergrid} shows experimental results when
+the program computes the SSSP for 20\% of the nodes of the graph.  The
+coordinated version scales well and compared to the regular version
+using 16 threads is 1.3 times faster.  the coordinated version
+produces 40\% fewer facts when compared to the regular version.
+Note that there are some situations where unnecessary facts are
+propagated because although the shortest distance is selected locally,
+sub-optimal distances may be propagated because many SSSP distances
+are computed at the same time.
 

### 2014-11-13 locks.tex
small edits

@@
 versions of the programs require significantly more queue operations.
-The data for this table comes from recording the number of \emph{queue
-operations} and facts derived.
-Queue instructions represent the
-number of normal queue operations executed (each costs 1) plus the
-number of \emph{percolate-up}/\emph{percolate-down} operations
-executed for each manipulation of the priority queue.  As expected,
-programs which rely on prioritizing the order of execution to obtain
-better asymptotic complexity (e.g., SSSP) improve in overall running
-time at the cost of more queue operations.  In all cases, coordination
-adds a significant number of total queue operations, but the resulting
-overhead is more than compensated for by an improved schedule and reduction
-in number of facts produced. For MiniMax and N Queens, the number of facts derived
-is the same for all configurations, as expected. The performance seen for those programs
-arise from reduced memory usage and improved memory locality.
+The data for this table comes from recording the number of queue
+operations and facts derived.  Queue operations represent the number
+of normal queue operations executed (each costs 1) plus the number of
+\emph{percolate-up}/\emph{percolate-down} operations executed for each
+manipulation of the priority queue.  As expected, programs which rely
+on prioritizing the order of execution to obtain better asymptotic
+complexity (e.g., SSSP) improve in overall running time at the cost of
+more queue operations.  In all cases, coordination adds a significant
+number of total queue operations, but the resulting overhead is more
+than compensated for by an improved schedule and reduction in number
+of facts produced. For MiniMax and N Queens, the number of facts
+derived is the same for all configurations, as expected. The
+performance seen for those programs arise from reduced memory usage
+and improved memory locality.
 
@@
 number of locks acquired to perform coordination operations.  We also
-measured how often \mytt{lock()} operations that fail to acquired
+measured how often \mytt{lock()} operations fail to acquire
 the lock on the first attempt.  Lock failures happen so infrequently

### 2014-11-13 minimax.tex
Fix verbatim code.

@@
 (12-29) deal with the cases for that create new nodes. In particular, the two
-rules in lines 12-25 generate new nodes using the
+rules in lines 12-26 generate new nodes using the
 \mytt{exists} language construct, which creates a child node 
@@
 
-expand(A, Board, [0 | Xs], N, P, Depth),
+expand(A, Board, [0 | Xs], N, P, \underline{Depth}),
 Depth >= 5
@@
 
-expand(A, Board, [0 | Xs], N, P, Depth),
+expand(A, Board, [0 | Xs], N, P, \underline{Depth}),
 Depth < 5
-  -o exists B. (\underline{set-default-priority(B, float(Depth + 1))},
+  -o exists B. (
+       \underline{set-default-priority(B, float(Depth + 1))},
        play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
@@
 
-expand(A, Board, [C | Xs], N, P, \underline{Depth})
-C <> 0
+expand(A, Board, [C | Xs], N, P, \underline{Depth}) C <> 0
   -o expand(A, Board ++ [C], Xs, N, P, \underline{Depth}).
@@
 With coordination, we set the priority of a node to be its depth (lines 15 and
-22) so the tree is expanded in a depth-first fashion,
+23) so the tree is expanded in a depth-first fashion,
 leading to a memory complexity of $\mathcal{O}(d t)$, where $d$

### 2014-11-13 splash-bp.tex
Fix verbatim code.

@@
 
-// TREE BUILDING
-// expand tree by adding neighbor nodes
-inactive(A), tree(A, All, Next) -o expand-tree(A, All, Next).
+// TREE BUILDING: expand tree by adding nodes
+inactive(A), tree(A, All, Next)
+   -o expand-tree(A, All, Next).
 // start tree since we do not have one
@@
 // end tree building
-expand-tree(A, All, Next), length(All) >= maxnodes
+expand-tree(A, All, Next), len(All) >= maxnodes
    -o first-phase(A, All, reverse(All)).
 // expand tree
-expand-tree(A, All, [A | Next]), length([A | Next]) < maxnodes
+expand-tree(A, All, [A | Next]), len(Next) < maxnodes-1
    -o [collect => L | Side | !edge(A, L, Side),
@@
 GraphLab and CLM comes mainly because GraphLab performs in-place updates of
-neighbor belief values, while CLM compiler and runtime system performs naive
+neighbor belief values, while CLM compiler and runtime system perform na\"{i}ve
 manipulation of facts by deriving rules. The in-place update could be generated

### 2014-11-13 language.tex
Remove dangling word.

@@
 nodes; and \mytt{relax(@1, 0, [@1])} starts the algorithm by
-setting the initial distance from \mytt{@1} to \mytt{@1} to be 0.
+setting the distance from \mytt{@1} to \mytt{@1} to be 0.
 

### 2014-11-13 heat-transfer.tex
Fix verbatim issues.

@@
 \begin{topfig}
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
+\scriptsize\begin{Verbatim}[numbers=left,xleftmargin=7mm,commandchars=\\\[\]]
 new-heat(A, New, Old),
@@
 \begin{topfig}
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
+\scriptsize\begin{Verbatim}[numbers=left,xleftmargin=7mm,commandchars=\\\[\]]
 new-heat(A, New, Old)

### 2014-11-13 conclusions.tex
Fix more typos.
small changes. eliminated last paragraph.

@@
-We have presented a novel way of adding coordination to a declarative language without
-changing the nature of the language or introducing non-declarative constructs. We
-took advantage of the fact that CLM uses linear logic, which allows us to derive
-coordination facts that can be deleted in order to perform actions in the underlying
-runtime system. In the other hand, sensing facts allow the programmer to reason
-about data locality and scheduling of computation in a data-driven fashion.
-Even though the programs are optimized, programs remain declarative and proving
-interesting properties of the programs remains simple because it is possible to
-reason about the logical rules that drive coordination.
+We have presented a novel way of adding coordination to a declarative
+language without changing the nature of the language or introducing
+non-declarative constructs. We took advantage of the fact that CLM
+uses linear logic, which allows us to derive coordination facts that
+can be deleted in order to perform actions in the underlying runtime
+system. Sensing facts allow the programmer to reason about data
+locality and scheduling of computation in a data-driven fashion.
+Action can then be used to affect how the program runs on the
+underlying system.  The resulting programs perform well and yet the
+programs remain declarative and proving interesting properties of the
+programs remains simple because it is possible to reason about the
+logical rules that drive both the computation and coordination.
 
-We would like to explore CLM's coordination principles in distributed systems,
-where data locality is far more important than in shared memory systems.
-Furthermore, we think that other declarative paradigms such as functional
-programming would benefit from giving programmers the freedom to fine-tune their
-programs.
+\iffalse
+We would like to explore CLM's coordination principles in distributed
+systems, where data locality is far more important than in shared
+memory systems.  Furthermore, we think that other declarative
+paradigms such as functional programming would benefit from giving
+programmers the freedom to fine-tune their programs.
+\fi

### 2014-11-13 intro.tex

@@
 primitives can be used to improve program execution based on the state of the
-program and the underlying machine. The first kind of coordination facts are
+program and the underlying machine. The first kind of coordination primitives are
 called \emph{sensing facts} and are used to sense information about the system
 the program is running on, e.g., scheduling and node placement on threads. The
-second kind of coordination facts are \emph{action facts} that when detected in
+second kind of coordination primitives are \emph{action facts} that when detected in
 rules, are used to apply a scheduling operation during execution. Coordination

### 2014-11-13 related.tex

@@
 \emph{and-parallelism}~\cite{Shen-92}.  In functional languages, the stateless
-nature of computation allows multiple expressions to evaluate safely in
+computation allows multiple expressions to evaluate safely in
 parallel. This has been explored in several languages such as
@@
 application space. We also limit our application space, but also give the
-programmer a declarative way to coordinate computation to improve the
-performance of the program.
+programmer a declarative way to coordinate computation.
 

### 2014-11-13 coordination.tex
Fixed more typos.

@@
 our implementation, we use a FIFO approach because
-older nodes tend to a higher number of unexamined facts, from which to derive
+older nodes tend to have a higher number of unexamined facts, from which to derive
 subsequent new facts.
@@
 node, so that once the node is done, the priority will default back to $D$.
-Initially, all nodes have a default priority of $0$.
+Initially, all nodes have a default priority of $-\infty$.
 
@@
    \item \code{set-priority(node A, float F)}: This sets the
-   temporary priority of \mytt{A} to \mytt{F}. If \mytt{A} has a priority
+   temporary priority of \mytt{A} to \mytt{F}. If \mytt{A} has priority
    \mytt{F'}, we only change the priority if \mytt{F} is higher. The programmer
@@
    \item \code{schedule-next(node A)}: Changes the temporary priority of node
-   \mytt{A} to be 1 unit higher than the current highest priority.
+   \mytt{A} to be $+\infty$.
    \item \code{set-default-priority(node, float)}: Sets the default
@@
 ~\footnote{The observant reader will notice that all the coordination facts are
-linear and when created in the body of a rule will be consumed. The system
+linear and when used in the body of a rule will be consumed. The system
 creates the necessary code to re-derive them without programmer interaction.
@@
    thread \mytt{T}, \mytt{B} is actually running on.
-   \item \code{moving(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is allowed
+   \item \code{is-moving(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is allowed
    to move between threads.
-   \item \code{static(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is not
+   \item \code{is-static(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is not
    allowed to move between threads.

### 2014-11-13 language.tex
Fix verbatim issues.
Small improvements.
Fixed more typos.

@@
 process of deriving a rule, while persistent facts are not.
-Program execution starts by adding the \emph{axioms} (the initial facts of the
-program) to the database. Next, rules are recursively applied and the
+Program execution starts by adding the \emph{axioms} (the initial facts)
+to the database. Next, rules are recursively applied and the
 database is updated by adding new facts or deleting facts used during rule derivation.
@@
 a simple program: the single source shortest path program~(SSSP). Later in the paper, we
-add coordination facts to improve the execution of the program.
+add coordination facts to improve the execution of this program.
 

### 2014-11-13 related.tex
Fixed more typos.

@@
 language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
-databases, it is now being increasingly used in different fields such as
+databases, it is now being used for
 distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor

### 2014-11-13 implementation.tex
Fix verbatim issues.
Fixed more typos.
Fixed several typos.

@@
-The implementation of LM includes a compiler and a virtual machine that runs
+The implementation of CLM includes a compiler and a virtual machine that runs
 byte-code. The virtual machine uses 32 registers for
@@
 the appropriate thread.  Threads do useful work by processing active nodes from their queues. Whenever a
-thread becomes idle, it attempts to steal nodes from a random thread.
+thread does not have nodes to process, it attempts to steal nodes from a random thread.
 If unsuccessful, the thread becomes idle and waits for program
@@
 keeps track of the available facts and potential candidate rules. We need
-a separate indexing structure per node since rules run local to each node.
+a separate indexing structure per node since rules are locally run.
 
@@
 
-\begin{tightdescription}
-   \item[working]: the node is executing.
-   \item[inactive]: the node is inactive, i.e., it has no new facts and is not in any
+\begin{tightitemize}
+   \item \textbf{working}: the node is executing.
+   \item \textbf{inactive}: the node is inactive, i.e., it has no new facts and is not in any
    queue for processing.
-   \item[queue]: the node is active with new facts and is in some queue waiting
+   \item \textbf{queue}: the node is active with new facts and is in some queue waiting
    to be processed.
-   \item[stealing]: the node has just been stolen and is in the process of being
+   \item \textbf{stealing}: the node has just been stolen and is in the process of being
    moved to another thread.
-   \item[coordinating]: the node is moving from one queue to another.
-\end{tightdescription}
+   \item \textbf{coordinating}: the node is moving from one queue to another.
+\end{tightitemize}
 
@@
 change node attributes: incoming facts, owner thread, state variable
-and locality information.  A database spin-lock is locked whenever the
-node enters into the \textbf{working} state.
+and locality information.  A database spin-lock is locked when the
+node is in the \textbf{working} state.
 
@@
 doubly linked lists known as the \emph{standard queue} and a pair of
-\emph{min/max} heap known as the \emph{priority queue}.  The standard queue
+\emph{min/max} heaps known as the \emph{priority queue}.  The standard queue
 contains nodes without priorities and supports push into tail, remove
@@
 implemented in order to support node stealing, while operations to
-remove arbitrary nodes or update the priority of nodes allows threads
+remove arbitrary nodes or update priority allows threads
 to change the priority of nodes.
@@
 also used as the index in the priority queue and current priority,
-respectively.
-
-Both the regular and priority queue are implemented as a pair of
-queues.  This first of the pair is the \emph{static queue} which
-contains nodes that cannot be stolen.  The other queue in each pair is
+respectively. Both the regular and priority queue are implemented as a pair of
+queues.  This first queue is the \emph{static queue} which
+contains nodes that cannot be stolen.  The other queue is
 the \emph{dynamic queue} which contains nodes which can be stolen by
@@
 each queue to protect queue operations.
-Given threads $T_1$ and $T_2$, we enumerate the most important
-synchronization intensive places in the virtual machine:
+Given threads $T_1$ and $T_2$, we enumerate the synchronization hotspots
+in the virtual machine:
 
-\begin{tightdescription}
-   \item[New facts:] When a node executes in $T_1$ and derives facts
+\begin{tightitemize}
+   \item \textbf{New facts}: When a node executes in $T_1$ and derives facts
    to a node in $T_2$, $T_1$ first buffers the facts 
@@
    node to the $T_2$'s queue.
-   \item[Thread activation:] If $T_2$ is inactive when adding facts to a node in
+   \item \textbf{Thread activation}: If $T_2$ is inactive when adding facts to a node in
    $T_2$, $T_1$ also synchronizes with $T_2$ to change $T_2$'s state to \emph{active}.
-   \item[Node stealing:] $T_1$ synchronizes with $T_2$ when it attempts to steal
+   \item \textbf{Node stealing}: $T_1$ synchronizes with $T_2$ when it attempts to steal
    nodes from $T_2$ by removing half of the nodes from one of $T_2$'s queues.
-   \item[Coordination:] If $T_1$ needs to perform coordination operations
+   \item \textbf{Coordination}: If $T_1$ needs to perform coordination operations
    to a node in $T_2$, it may need to synchronize with $T_2$ during priority
    updates in order to move the node in $T_2$'s queues.
-\end{tightdescription}
+\end{tightitemize}
 
@@
 program was around 2 times slower than GraphLab, while the N Queens
-program was 10 to 15x slower than a sequential C implementation. The
-LM programs had similar performance to programs written in Python.
-The introduction of coordination in CLM has little or no impact on the
-overall performance of these programs.  In fact, due to some
+program was 10 to 15x slower than a sequential C implementation.
+Furthermore, LM programs had better performance than programs written in Python.
+We measured the impact of our implementation changes on unmodified LM programs
+and our experiments indicate that there is little to no impact on the overall
+performance of these programs. In fact, due to some
 improvements in the virtual machine, programs such as PageRank and

### 2014-11-13 locks.tex
Fixed several typos.

@@
 on prioritizing the order of execution to obtain better asymptotic
-complexity (e.g., SSSP) improve in overall running time at the cost of
+complexity (e.g., SSSP), improve in overall running time at the cost of
 more queue operations.  In all cases, coordination adds a significant
@@
 immediately.  This leads to a significant increase in total locks
-acquired often because the priority of a node is change many times
+acquired often because the priority of a node is changed many times
 before it is operated on.  The third bar shows the effect of buffering

### 2014-11-13 minimax.tex
Fix verbatim issues.
Fixed several typos.

@@
 with the case where no children nodes are created and the last three rules
-(12-29) deal with the cases for that create new nodes. In particular, the two
+(12-29) deal with the cases that create new nodes. In particular, the two
 rules in lines 12-26 generate new nodes using the
@@
 With coordination, we set the priority of a node to be its depth (lines 15 and
-23) so the tree is expanded in a depth-first fashion,
+23) so that the tree is expanded in a depth-first fashion,
 leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
@@
    \begin{center}
-      \subfloat[]{\small\begin{tabular}[b]{ | c | c | c |}
+      \subfloat[]{\footnotesize\begin{tabular}[b]{ | c | c | c |}
          \hline                       

### 2014-11-13 nqueens.tex
space saving edits
Fixed several typos.

@@
 
-Since computation goes from the top row to the bottom row, not all
-placements of nodes to threads perform equally. This is especially
-true because the bottom rows tend to perform the most work. The best
-placement is then to split the board vertically so each thread gets
-the same number of columns with axiom \mytt{set-cpu(A, vertical(X, Y)}, where
-\mytt{X} and \mytt{Y} are the coordinates of the board. Our experiments show that, on a shared
-memory system, it does not matter much if we start with a bad
-placement since node stealing overcomes those issues by load balancing
-dynamically.  However, the N Queens program incrementally builds and shares lists representing
-valid board states that are transmitted from top to bottom.
-In order to improve memory locality we should then manipulate board
-states that share a significant number of elements because each board state
-needs to be iterated before extended with a new position.
-To accomplish this, we used coordination axioms \mytt{set-default-priority(A,
-X)}, where \mytt{X} is the row number (0 is the top row) in order to
-prioritize the evaluation of squares closer to the bottom of the board.
+Since computation goes from the top row to the bottom row, not all placements of
+nodes to threads perform equally. This is especially true because the bottom
+rows tend to perform the most work. The best placement is then to split the
+board vertically with axiom \mytt{set-cpu(A, vertical(X, Y))} so each thread
+gets the same number of columns, where \mytt{X} and \mytt{Y} are the coordinates
+of the board. Our experiments show that, on a shared memory system, it does not
+matter much if we start with a bad placement since node stealing overcomes those
+issues by load balancing dynamically.  However, the N Queens program
+incrementally builds and shares lists representing valid board states that are
+transmitted from top to bottom.  In order to improve memory locality we should
+then manipulate board states that share a significant number of elements because
+each board state needs to be iterated before extended with a new position.  To
+accomplish this, we used coordination axioms \mytt{set-default-priority(A, X)},
+where \mytt{X} is the row number (0 is the top row) in order to
+prioritize the evaluation of squares closer to the bottom of the
+board.
 
@@
 \begin{topfig}
+\vspace*{-1ex}
    \begin{center}

### 2014-11-13 sssp.tex
Fix verbatim issues.
Fixed several typos.

@@
 Dijkstra's shortest path algorithm~\cite{Dijkstra}. When using multiple
-threads, each thread will pick the smallest distance from their subset of nodes.
+threads, each thread will pick the shortest distance from their subset of nodes.
 While this does not yield the optimal program with relation to 1 thread, it
@@
 \begin{topfig}
-\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
+\scriptsize\begin{Verbatim}[numbers=left,xleftmargin=7mm,commandchars=\\\{\}]
 type route edge(node, node, int).
@@
 although it applies rules using smarter ordering and the code remains
-declarative. The proof of correctness considers that, eventually, the shortest
-path is computed at all nodes of the graph until all shortest paths are found.
-The derivation of \mytt{set-priority} does not change this
-proof.
+declarative. Since the proof of correctness considers that, eventually, the shortest
+path is computed at all nodes of the graph, the use of \mytt{set-priority}
+does not change this at all.
 
@@
 the program computes the SSSP for 20\% of the nodes of the graph.  The
-coordinated version scales well and compared to the regular version
-using 16 threads is 1.3 times faster.  the coordinated version
-produces 40\% fewer facts when compared to the regular version.
+coordinated version scales well and when compared to the regular version
+using 16 threads, it is 1.3 times faster. The coordinated version
+also produces 40\% fewer facts with any number of threads.
 Note that there are some situations where unnecessary facts are

### 2014-12-17 rebuttal.txt
Add rebuttal.

We greatly appreciate the reviews of our paper.  We clearly sparked
interest, but need to improve the related work and experimental
results sections.  We address both these issues below.

Each reviewer asked for more (but different) information which would
require us to remove something else in the paper.  We will happily
take advice from the committee about what we should remove to make
room for something else.  However, our primary goal with this paper
was to show the expressive power of making coordination a first class
programming contruct and chose to include five examples to give
readers a flavor of the power of CLM.  This pedagogic approach
highlights the conciseness and expressiveness of CLM as well as the
readability of a high-level declarative approach to fine-tuning the
performance of a program.

Related Work:

The reviews pointed out numerous systems that we did not cite.  Please
note that CLM stands alone in
- making coordination (both scheduling and partitioning) a first-class
   programming contruct.
- supporting data-driven dynamic (at run-time) coordination,
   particularly for irregular data structures.

We do not claim novelty in the mechanisms, but rather in how they are
integrated into a declarative language - as a first class concept.  We
will do a better job of showing where the basic ideas come from.

In reference to works cited by the reviewers:

- Halide supports distinct coordination, but only as "pragmas" in the
  domain of image processing, i.e., regular data-structures.

- Galois supports a rich set of schedulers through customizable
  containers (see: Nguyen&Pingali), but it requires extra-language
  mechanisms to do things such as determining the location of data.

- Elixir's uses a pre-defined set of schedulers and limits
  coordination to scheduling.

- Grace supports scheduling through priorities.  (It is very much like GraphLab.)

- Ligra, also oriented towards graph processing, is limited to shared
  memory and only handles partitioning.

- Cilk, OpenMP and TBB all provide low-level imperative approaches for
  coordination.

We will expand the related work section, but stress the unique nature
of CLM which results in clean concise coordination, e.g., we can write
the Splash scheduler used for Splash BP less than 60 lines (versus 500
in GraphLab).  It is hard (or impossible) to write in all the above
languages.

Experiments:

Mea Culpa on not explaining the superlinear speed-up.  We did not
implement the randomized scheduler needed for the best sequential
implementation.  We will address this better in the final version.

We did compare CLM to an optimized C++ implementation of 2 programs
(HT and SBP in GraphLab): The CLM implementations are 1.8X-3X slower,
but scale equally well.  Our focus was on expressiveness and
scalability, not raw performance. (Our implementation of CLM is
interpreted by an unoptimized virtual machine. It could even be
compiled.)  From an expressivness point of view, the GraphLab
scheduler is 500 lines of customized library code.  CLM's version is
less than 1 page!  If requested, we can also include a third-party
solution to SSSP.  As well as including several different synthetic
graphs to test the scalability over different data sets.

We can include comparisons to sequential solutions:
- interpreted solutions (Python which is 2x-10x slower than CLM)
- compiled solutions in C (which is 2-10x faster depending on the
  program)

Other Stuff (And addressing some of the smaller, but interesting
points taken up by the reviewers):

CLM's busy-wait locks are ticket locks that use atomic
instructions. We have scalable locks but they need more testing.

CLM can do synchronous or asynchronous algorithms, depending on the
needs of the programmer.  We have implemented map-reduce style
algorithms, asynchronous algorithms, and algorithms with mutable
graphs.

Program proofs: we did not include them due to lack of space.
However, we have proven interesting properties of our programs, e.g.,
correctness, termination, dead-lock freedom, liveness, etc.  We can
remove the references to program proofs in the final version. (thats
      another paper)

Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
fact, we have found that the larger the data set the better the
scaling.  We can increase the sizes by 10x if the reviews prefer.

CLM is also important since it is not restricted to shared memory
systems and can be easily extended to distributed computers. That
simply cannot be said for most of these systems.  The use of linear
logic, a sound logical framework, has allowed us to easily prove
properties about programs using coordination.

With regards to the poorly constructed sentence at the end of section
6.3, there are no additional errors.  The values computed all meet the
error bound, but are different for different schedulers.

The difference between Coordinated/Coordinated and Coordinated/Regular:
- Coordinated/Regular: speedup of the coordinated program on
  n-processors over the unannotated (e.g., regular program) on 1
  processor.
- Coordinated/Coordinated: speedup of the coordinated program on
  n-processors over the coordinated program on 1 processor.

We agree that a measure of rule compelexity would be a great thing to
have.  Any suggestions?  Obviously lines of code is not enough.  The
"gut" intution here is that CLM is easy to read and doesn't require
many lines of code.  A better metric would be great.

The scalability results were not presented upto 24 threads due to an
error in our presentation, that info will be in the final paper.
(Just so you know, the graphs continue to look the same up through 24
threads.)

reviewer B:
- get-priority and set-priority are not sufficient.  The temporary
  priority is reset each time the node is processed, which comes in
  quite handy in the splash scheduler, for example.
- The arguments to be indexed and such as described in another paper.
  Happy to forward to committee (or provide the citation if needed.)
- There was a typo in the paper.  We actually do a vertical split, not
  a horizontal split when we do work stealing.
- We do reference counting for garbage counting, but no locks are used
  and the threads do it individually and do not need to stop and
  collect data.
- The implementation of SSSP devolves to Dijkstra's algorithm on 1
  thread, on multiple threads there are times when a non-optimal node
  in the graph is chosen, but overall it approaches the lower bounds
  of Dijkstra's algorithm while allowing for significantly more
  parallelism.

We thank the reviewers for the time and expertise they have invested
in these reviews.  We will of course avail ourselves of all the
comments on presentation, grammer, etc.  Again, many thanks.

### 2014-12-17 rebuttal.txt
organized rebuttal.

@@
 
+=================================================================
 Other Stuff (And addressing some of the smaller, but interesting
@@
 
-CLM's busy-wait locks are ticket locks that use atomic
+-> reviewer A:
+- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
+   3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
+   fact, we have found that the larger the data set the better the
+   scaling.  We can increase the sizes by 10x if the reviews prefer.
+- CLM's busy-wait locks are ticket locks that use atomic
    instructions. We have scalable locks but they need more testing.
-
-CLM can do synchronous or asynchronous algorithms, depending on the
+- CLM can do synchronous or asynchronous algorithms, depending on the
    needs of the programmer.  We have implemented map-reduce style
@@
    graphs.
-
-Program proofs: we did not include them due to lack of space.
-However, we have proven interesting properties of our programs, e.g.,
-correctness, termination, dead-lock freedom, liveness, etc.  We can
-remove the references to program proofs in the final version. (thats
-      another paper)
-
-Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
-3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
-fact, we have found that the larger the data set the better the
-scaling.  We can increase the sizes by 10x if the reviews prefer.
-
-CLM is also important since it is not restricted to shared memory
-systems and can be easily extended to distributed computers. That
-simply cannot be said for most of these systems.  The use of linear
-logic, a sound logical framework, has allowed us to easily prove
-properties about programs using coordination.
-
-With regards to the poorly constructed sentence at the end of section
-6.3, there are no additional errors.  The values computed all meet the
-error bound, but are different for different schedulers.
-
-The difference between Coordinated/Coordinated and Coordinated/Regular:
-- Coordinated/Regular: speedup of the coordinated program on
-  n-processors over the unannotated (e.g., regular program) on 1
-  processor.
-- Coordinated/Coordinated: speedup of the coordinated program on
-  n-processors over the coordinated program on 1 processor.
-
-We agree that a measure of rule compelexity would be a great thing to
-have.  Any suggestions?  Obviously lines of code is not enough.  The
-"gut" intution here is that CLM is easy to read and doesn't require
-many lines of code.  A better metric would be great.
-
-The scalability results were not presented upto 24 threads due to an
-error in our presentation, that info will be in the final paper.
-(Just so you know, the graphs continue to look the same up through 24
-threads.)
-
-reviewer B:
 - get-priority and set-priority are not sufficient.  The temporary
@@
 
+-> reviewer B:
+- Program proofs: we did not include them due to lack of space.
+   However, we have proven interesting properties of our programs, e.g.,
+   correctness, termination, dead-lock freedom, liveness, etc.  We can
+   remove the references to program proofs in the final version. (thats
+         another paper)
+- With regards to the poorly constructed sentence at the end of section
+   6.3, there are no additional errors.  The values computed all meet the
+   error bound, but are different for different schedulers.
+- We agree that a measure of rule compelexity would be a great thing to
+   have.  Any suggestions?  Obviously lines of code is not enough.  The
+   "gut" intution here is that CLM is easy to read and doesn't require
+   many lines of code.  A better metric would be great.
+- The difference between Coordinated/Coordinated and Coordinated/Regular:
+   - Coordinated/Regular: speedup of the coordinated program on
+      n-processors over the unannotated (e.g., regular program) on 1
+      processor.
+   - Coordinated/Coordinated: speedup of the coordinated program on
+      n-processors over the coordinated program on 1 processor.
+
+-> reviewer A and C:
+- The scalability results were not presented upto 24 threads due to an
+error in our presentation, that info will be in the final paper.
+(Just so you know, the graphs continue to look the same up through 24
+threads.)
+
+Finally, we note that CLM is noteworthy since it is not restricted to shared memory
+systems and can be easily extended to distributed systems. That
+simply cannot be said for most of the other systems.  The use of linear
+logic, a sound logical framework, has allowed us to easily prove
+properties about programs using coordination.
+
 We thank the reviewers for the time and expertise they have invested

### 2014-12-17 rebuttal.txt
final edits

@@
 was to show the expressive power of making coordination a first class
-programming contruct and chose to include five examples to give
+programming construct and chose to include five examples to give
 readers a flavor of the power of CLM.  This pedagogic approach
@@
 - making coordination (both scheduling and partitioning) a first-class
-   programming contruct.
+   programming construct.
 - supporting data-driven dynamic (at run-time) coordination,
@@
 interpreted by an unoptimized virtual machine. It could even be
-compiled.)  From an expressivness point of view, the GraphLab
+compiled.)  From an expressiveness point of view, the GraphLab
 scheduler is 500 lines of customized library code.  CLM's version is
@@
 
-=================================================================
 Other Stuff (And addressing some of the smaller, but interesting
@@
    correctness, termination, dead-lock freedom, liveness, etc.  We can
-   remove the references to program proofs in the final version. (thats
+   remove the references to program proofs in the final version. (that's
          another paper)
@@
    error bound, but are different for different schedulers.
-- We agree that a measure of rule compelexity would be a great thing to
+- We agree that a measure of rule complexity would be a great thing to
    have.  Any suggestions?  Obviously lines of code is not enough.  The
-   "gut" intution here is that CLM is easy to read and doesn't require
+   "gut" intuition here is that CLM is easy to read and doesn't require
    many lines of code.  A better metric would be great.
@@
 -> reviewer A and C:
-- The scalability results were not presented upto 24 threads due to an
+- The scalability results were not presented up to 24 threads due to an
 error in our presentation, that info will be in the final paper.
@@
 
-Finally, we note that CLM is noteworthy since it is not restricted to shared memory
-systems and can be easily extended to distributed systems. That
-simply cannot be said for most of the other systems.  The use of linear
-logic, a sound logical framework, has allowed us to easily prove
-properties about programs using coordination.
+Finally, we note that CLM is noteworthy since it is not restricted to
+shared memory systems and can be easily extended to distributed
+systems. That simply cannot be said for most of the other systems.
+The use of linear logic, a sound logical framework, has allowed us to
+easily prove properties about programs using coordination.
 
@@
 in these reviews.  We will of course avail ourselves of all the
-comments on presentation, grammer, etc.  Again, many thanks.
+comments on presentation, grammar, etc.  Again, many thanks.
+
+LocalWords:  Halide pragmas customizable Pingali pre GraphLab Ligra
+LocalWords:  Cilk OpenMP TBB Mea Culpa superlinear SBP scalability
+LocalWords:  unoptimized CLM's SSSP MiniMax scalable liveness
+LocalWords:  unannotated

### 2014-12-17 rebuttal.txt
final formatting

@@
 Each reviewer asked for more (but different) information which would
-require us to remove something else in the paper.  We will happily
+require us to remove something in the paper.  We will happily
 take advice from the committee about what we should remove to make
-room for something else.  However, our primary goal with this paper
-was to show the expressive power of making coordination a first class
-programming construct and chose to include five examples to give
-readers a flavor of the power of CLM.  This pedagogic approach
-highlights the conciseness and expressiveness of CLM as well as the
-readability of a high-level declarative approach to fine-tuning the
-performance of a program.
+room for something else.
+
+Our primary goal with this paper was to show the expressive power of
+making coordination a first class programming construct and chose to
+include five examples to give readers a flavor of the power of CLM.
+This approach highlights the conciseness and expressiveness of CLM as
+well as the readability of a high-level declarative approach to
+fine-tuning the performance of a program.
 
@@
 We do not claim novelty in the mechanisms, but rather in how they are
-integrated into a declarative language - as a first class concept.  We
-will do a better job of showing where the basic ideas come from.
+integrated into a declarative language - as a first class entity.  We
+will do a better job of showing where the ideas come from.
 
@@
 
-Mea Culpa on not explaining the superlinear speed-up.  We did not
+Mea Culpa on not explaining the superlinear speed-up.  We didn't
 implement the randomized scheduler needed for the best sequential
-implementation.  We will address this better in the final version.
+implementation.  We will address this in the final version.
 
 We did compare CLM to an optimized C++ implementation of 2 programs
-(HT and SBP in GraphLab): The CLM implementations are 1.8X-3X slower,
-but scale equally well.  Our focus was on expressiveness and
-scalability, not raw performance. (Our implementation of CLM is
-interpreted by an unoptimized virtual machine. It could even be
-compiled.)  From an expressiveness point of view, the GraphLab
-scheduler is 500 lines of customized library code.  CLM's version is
-less than 1 page!  If requested, we can also include a third-party
-solution to SSSP.  As well as including several different synthetic
-graphs to test the scalability over different data sets.
+(HT and SBP): The CLM implementations are 1.8X-3X slower, but scale
+equally well.  Our focus was on expressiveness and scalability, not
+raw performance. (CLM is currently interpreted by an
+unoptimized virtual machine. It could even be compiled.)  From an
+expressiveness point of view, the GraphLab scheduler is 500 lines of
+customized library code.  CLM's version is less than 1 page!  If
+requested, we can also include a third-party solution to SSSP.  As
+well as including several different synthetic graphs to test the
+scalability over different data sets.
 

### 2014-12-17 rebuttal-forweb.txt
intial add

--- 2014-12-17 rebuttal.txt
@@
-We greatly appreciate the reviews of our paper.  We clearly sparked
-interest, but need to improve the related work and experimental
-results sections.  We address both these issues below.
+We greatly appreciate the reviews of our paper.  We clearly sparked interest, but need to improve the related work and experimental results sections.  We address both these issues below.
 
-Each reviewer asked for more (but different) information which would
-require us to remove something in the paper.  We will happily
-take advice from the committee about what we should remove to make
-room for something else.
+Each reviewer asked for more (but different) information which would require us to remove something in the paper.
+We will happily take advice from the committee about what we should remove to make room for something else.
 
-Our primary goal with this paper was to show the expressive power of
-making coordination a first class programming construct and chose to
-include five examples to give readers a flavor of the power of CLM.
-This approach highlights the conciseness and expressiveness of CLM as
-well as the readability of a high-level declarative approach to
-fine-tuning the performance of a program.
+Our primary goal with this paper was to show the expressive power of making coordination a first class programming construct and chose to include five examples to give readers a flavor of the power of CLM.
+This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tuning the performance of a program.
 
@@
 
-The reviews pointed out numerous systems that we did not cite.  Please
-note that CLM stands alone in
-- making coordination (both scheduling and partitioning) a first-class
-  programming construct.
-- supporting data-driven dynamic (at run-time) coordination,
-  particularly for irregular data structures.
+The reviews pointed out numerous systems that we did not cite.  Please note that CLM stands alone in
+- making coordination (both scheduling and partitioning) a first-class programming construct.
+- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
 
-We do not claim novelty in the mechanisms, but rather in how they are
-integrated into a declarative language - as a first class entity.  We
-will do a better job of showing where the ideas come from.
+We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity.  We will do a better job of showing where the ideas come from.
 
 In reference to works cited by the reviewers:
+- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
+- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
+- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
+- Grace supports scheduling through priorities.  (It is very much like GraphLab.)
+- Ligra, also oriented towards graph processing, is limited to shared memory and only handles partitioning.
+- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.
 
-- Halide supports distinct coordination, but only as "pragmas" in the
-  domain of image processing, i.e., regular data-structures.
-
-- Galois supports a rich set of schedulers through customizable
-  containers (see: Nguyen&Pingali), but it requires extra-language
-  mechanisms to do things such as determining the location of data.
-
-- Elixir's uses a pre-defined set of schedulers and limits
-  coordination to scheduling.
-
-- Grace supports scheduling through priorities.  (It is very much like GraphLab.)
-
-- Ligra, also oriented towards graph processing, is limited to shared
-  memory and only handles partitioning.
-
-- Cilk, OpenMP and TBB all provide low-level imperative approaches for
-  coordination.
-
-We will expand the related work section, but stress the unique nature
-of CLM which results in clean concise coordination, e.g., we can write
-the Splash scheduler used for Splash BP less than 60 lines (versus 500
-in GraphLab).  It is hard (or impossible) to write in all the above
-languages.
+We will expand the related work section, but stress the unique nature of CLM which results in clean concise coordination, e.g., we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).
+It is hard (or impossible) to write in all the above languages.
 
@@
 
-Mea Culpa on not explaining the superlinear speed-up.  We didn't
-implement the randomized scheduler needed for the best sequential
-implementation.  We will address this in the final version.
+Mea Culpa on not explaining the superlinear speed-up.  We didn't implement the randomized scheduler needed for the best sequential implementation.  We will address this in the final version.
 
-We did compare CLM to an optimized C++ implementation of 2 programs
-(HT and SBP): The CLM implementations are 1.8X-3X slower, but scale
-equally well.  Our focus was on expressiveness and scalability, not
-raw performance. (CLM is currently interpreted by an
-unoptimized virtual machine. It could even be compiled.)  From an
-expressiveness point of view, the GraphLab scheduler is 500 lines of
-customized library code.  CLM's version is less than 1 page!  If
-requested, we can also include a third-party solution to SSSP.  As
-well as including several different synthetic graphs to test the
-scalability over different data sets.
+We did compare CLM to an optimized C++ implementation of 2 programs (HT and SBP): The CLM implementations are 1.8X-3X slower, but scale equally well.
+Our focus was on expressiveness and scalability, not raw performance.
+(CLM is currently interpreted by an unoptimized virtual machine.
+It could even be compiled.)  From an expressiveness point of view, the GraphLab scheduler is 500 lines of customized library code.
+CLM's version is less than 1 page!
+If requested, we can also include a third-party solution to SSSP.
+As well as including several different synthetic graphs to test the scalability over different data sets.
 
@@
 - interpreted solutions (Python which is 2x-10x slower than CLM)
-- compiled solutions in C (which is 2-10x faster depending on the
-  program)
+- compiled solutions in C (which is 2-10x faster depending on the program)
 
-Other Stuff (And addressing some of the smaller, but interesting
-points taken up by the reviewers):
+Other Stuff (And addressing some of the smaller, but interesting points taken up by the reviewers):
 
 -> reviewer A:
-- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses
-   3,600,000 nodes. SSSP computes around 5,000,000 shortest paths.  In
-   fact, we have found that the larger the data set the better the
-   scaling.  We can increase the sizes by 10x if the reviews prefer.
-- CLM's busy-wait locks are ticket locks that use atomic
-   instructions. We have scalable locks but they need more testing.
-- CLM can do synchronous or asynchronous algorithms, depending on the
-   needs of the programmer.  We have implemented map-reduce style
-   algorithms, asynchronous algorithms, and algorithms with mutable
-   graphs.
-- get-priority and set-priority are not sufficient.  The temporary
-  priority is reset each time the node is processed, which comes in
-  quite handy in the splash scheduler, for example.
-- The arguments to be indexed and such as described in another paper.
-  Happy to forward to committee (or provide the citation if needed.)
-- There was a typo in the paper.  We actually do a vertical split, not
-  a horizontal split when we do work stealing.
-- We do reference counting for garbage counting, but no locks are used
-  and the threads do it individually and do not need to stop and
-  collect data.
-- The implementation of SSSP devolves to Dijkstra's algorithm on 1
-  thread, on multiple threads there are times when a non-optimal node
-  in the graph is chosen, but overall it approaches the lower bounds
-  of Dijkstra's algorithm while allowing for significantly more
-  parallelism.
+- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses    3,600,000 nodes.
+SSSP computes around 5,000,000 shortest paths.
+In    fact, we have found that the larger the data set the better the    scaling.
+We can increase the sizes by 10x if the reviews prefer.
+- CLM's busy-wait locks are ticket locks that use atomic    instructions. We have scalable locks but they need more testing.
+- CLM can do synchronous or asynchronous algorithms, depending on the    needs of the programmer.
+We have implemented map-reduce style    algorithms, asynchronous algorithms, and algorithms with mutable    graphs.
+- get-priority and set-priority are not sufficient.  The temporary   priority is reset each time the node is processed, which comes in   quite handy in the splash scheduler, for example.
+- The arguments to be indexed and such as described in another paper.   Happy to forward to committee (or provide the citation if needed.)
+- There was a typo in the paper.  We actually do a vertical split, not   a horizontal split when we do work stealing.
+- We do reference counting for garbage counting, but no locks are used   and the threads do it individually and do not need to stop and   collect data.
+- The implementation of SSSP devolves to Dijkstra's algorithm on 1   thread, on multiple threads there are times when a non-optimal node   in the graph is chosen, but overall it approaches the lower bounds   of Dijkstra's algorithm while allowing for significantly more   parallelism.
 
 -> reviewer B:
+
 - Program proofs: we did not include them due to lack of space.
-   However, we have proven interesting properties of our programs, e.g.,
-   correctness, termination, dead-lock freedom, liveness, etc.  We can
-   remove the references to program proofs in the final version. (that's
-         another paper)
-- With regards to the poorly constructed sentence at the end of section
-   6.3, there are no additional errors.  The values computed all meet the
-   error bound, but are different for different schedulers.
-- We agree that a measure of rule complexity would be a great thing to
-   have.  Any suggestions?  Obviously lines of code is not enough.  The
-   "gut" intuition here is that CLM is easy to read and doesn't require
-   many lines of code.  A better metric would be great.
+However, we have proven interesting properties of our programs, e.g.,    correctness, termination, dead-lock freedom, liveness, etc.
+We can    remove the references to program proofs in the final version.
+(that's another paper)
+- With regards to the poorly constructed sentence at the end of section    6.3, there are no additional errors.
+The values computed all meet the    error bound, but are different for different schedulers.
+- We agree that a measure of rule complexity would be a great thing to    have.
+Any suggestions?
+Obviously lines of code is not enough.
+The    "gut" intuition here is that CLM is easy to read and doesn't require    many lines of code.
+A better metric would be great.
 - The difference between Coordinated/Coordinated and Coordinated/Regular:
-   - Coordinated/Regular: speedup of the coordinated program on
-      n-processors over the unannotated (e.g., regular program) on 1
-      processor.
-   - Coordinated/Coordinated: speedup of the coordinated program on
-      n-processors over the coordinated program on 1 processor.
+   - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
+   - Coordinated/Coordinated: speedup of the coordinated program on       n-processors over the coordinated program on 1 processor.
 
 -> reviewer A and C:
-- The scalability results were not presented up to 24 threads due to an
-error in our presentation, that info will be in the final paper.
-(Just so you know, the graphs continue to look the same up through 24
-threads.)
 
-Finally, we note that CLM is noteworthy since it is not restricted to
-shared memory systems and can be easily extended to distributed
-systems. That simply cannot be said for most of the other systems.
-The use of linear logic, a sound logical framework, has allowed us to
-easily prove properties about programs using coordination.
-
-We thank the reviewers for the time and expertise they have invested
-in these reviews.  We will of course avail ourselves of all the
-comments on presentation, grammar, etc.  Again, many thanks.
+- The scalability results were not presented up to 24 threads due to an error in our presentation, that info will be in the final paper.
+(Just so you know, the graphs continue to look the same up through 24 threads.)
+Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems.
+That simply cannot be said for most of the other systems.
+The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
+We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.
 

### 2015-01-20 rebuttal2.txt
Add reviews and rebuttal.

--- 2014-12-17 rebuttal-forweb.txt
@@
-We greatly appreciate the reviews of our paper.  We clearly sparked interest, but need to improve the related work and experimental results sections.  We address both these issues below.
+Coordination is a programming language design issue that requires novel approaches based on sound formal foundations.
+The implementation challenges are important but not as important from a PL perspective.
+While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue.
+Our solution to this PL design challenge fits the scope of PLDI fairly well.
+We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.
 
+CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
+Because the paper was written from this PL perspective, we unfairly neglected important related work.
+
+Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM.
+This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program.
 Each reviewer asked for more (but different) information which would require us to remove something in the paper.
 We will happily take advice from the committee about what we should remove to make room for something else.
-
-Our primary goal with this paper was to show the expressive power of making coordination a first class programming construct and chose to include five examples to give readers a flavor of the power of CLM.
-This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tuning the performance of a program.
 
@@
 - supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
+- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
 
@@
 - Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
-- Grace supports scheduling through priorities.  (It is very much like GraphLab.)
-- Ligra, also oriented towards graph processing, is limited to shared memory and only handles partitioning.
+- Grace supports scheduling through priorities.  (It is very much like GraphLab)
+   - Ligra, also oriented towards graph processing, only handles partitioning.
    - Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.
@@
 
--> reviewer A:
-- Regarding the benchmark sizes, BP uses 160,000 nodes and MiniMax uses    3,600,000 nodes.
-SSSP computes around 5,000,000 shortest paths.
-In    fact, we have found that the larger the data set the better the    scaling.
-We can increase the sizes by 10x if the reviews prefer.
-- CLM's busy-wait locks are ticket locks that use atomic    instructions. We have scalable locks but they need more testing.
-- CLM can do synchronous or asynchronous algorithms, depending on the    needs of the programmer.
-We have implemented map-reduce style    algorithms, asynchronous algorithms, and algorithms with mutable    graphs.
-- get-priority and set-priority are not sufficient.  The temporary   priority is reset each time the node is processed, which comes in   quite handy in the splash scheduler, for example.
-- The arguments to be indexed and such as described in another paper.   Happy to forward to committee (or provide the citation if needed.)
-- There was a typo in the paper.  We actually do a vertical split, not   a horizontal split when we do work stealing.
-- We do reference counting for garbage counting, but no locks are used   and the threads do it individually and do not need to stop and   collect data.
-- The implementation of SSSP devolves to Dijkstra's algorithm on 1   thread, on multiple threads there are times when a non-optimal node   in the graph is chosen, but overall it approaches the lower bounds   of Dijkstra's algorithm while allowing for significantly more   parallelism.
+Please refer to 1st response for the other points.
 
--> reviewer B:
+-> reviewer D:
 
-- Program proofs: we did not include them due to lack of space.
-However, we have proven interesting properties of our programs, e.g.,    correctness, termination, dead-lock freedom, liveness, etc.
-We can    remove the references to program proofs in the final version.
-(that's another paper)
-- With regards to the poorly constructed sentence at the end of section    6.3, there are no additional errors.
-The values computed all meet the    error bound, but are different for different schedulers.
-- We agree that a measure of rule complexity would be a great thing to    have.
-Any suggestions?
-Obviously lines of code is not enough.
-The    "gut" intuition here is that CLM is easy to read and doesn't require    many lines of code.
-A better metric would be great.
-- The difference between Coordinated/Coordinated and Coordinated/Regular:
+- Small increment over Meld: we agree that this is true on the surface since the language is actually the same, but we do not agree that this is a point against since the language is able to describe normal computation and coordination using the same constructs.
+However, the runtime system actually looks quite different because now we have to deal with programmable scheduling and data partitioning.
+These two coordination mechanisms required significant changes in the implementation and compilation with the introduction of different queues for nodes and the ability to move and pin nodes to threads.
+- CLM target: our current target is shared memory multi-cores.
+However the CLM programming model fits and is easily ported to clusters of computers.
+It’s just a matter of implementing a suitable runtime system for those architectures.
+- Contribution of CLM: Galois and many others are not declarative programming languages.
+CLM solves a problem of programming language design by seamlessly combining computation and coordination into a single declarative programming model.
+Galois is a runtime system and does not solve this particular design problem.
+- Technical challenges: there is a PL design language: how to design a declarative language that allows the programmer to dynamically coordinate execution while keeping the language declarative; we show that linear logic allows us to do this easily.
+There is an implementation challenge where we had to add support for all the coordination directives.
+- Proofs: for most programs they do not complicate the proofs. However, for programs such as SBP, you have to take coordination into account because it changes how the program is computed.
+
+Reviewer E:
+
+- Programming efficiency: true, but we see this as a positive point since most declarative languages do not allow such control.
+- Data layout: the data layout/node placement is changed by using action facts such as set-cpu or set-static. Set-cpu moves the node to another thread while set-static pins the node to a thread.
+- inplace update: it can be done by analyzing how facts are used in the rules.
+If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.
+
+-> reviewer B and E:
+
+- The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular:
+   - Regular/Regular: speedup of the unannotated (e.g.., regular) program on n-processors over the unannotated program on 1 processor.
    - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
@@
 
--> reviewer A and C:
-
-- The scalability results were not presented up to 24 threads due to an error in our presentation, that info will be in the final paper.
-(Just so you know, the graphs continue to look the same up through 24 threads.)
 Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems.
-That simply cannot be said for most of the other systems.
 The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
 We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.
-
-LocalWords:  Halide pragmas customizable Pingali pre GraphLab Ligra
-LocalWords:  Cilk OpenMP TBB Mea Culpa superlinear SBP scalability
-LocalWords:  unoptimized CLM's SSSP MiniMax scalable liveness
-LocalWords:  unannotated

### 2015-01-20 rebuttal2.txt
Update.

@@
-Coordination is a programming language design issue that requires novel approaches based on sound formal foundations.
+From the review scores, it looks like the paper may be reject.
+We note however that coordination is an highly important programming language design issue, specially when concerning declarative programming languages and parallelism.
+We disagree that the paper is a small increment
+
 The implementation challenges are important but not as important from a PL perspective.
@@
 
-CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
-Because the paper was written from this PL perspective, we unfairly neglected important related work.
-
 Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM.
@@
 Each reviewer asked for more (but different) information which would require us to remove something in the paper.
-We will happily take advice from the committee about what we should remove to make room for something else.
 
-Related Work:
+Talk about small increment?
 
-The reviews pointed out numerous systems that we did not cite.  Please note that CLM stands alone in
-- making coordination (both scheduling and partitioning) a first-class programming construct.
-- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
-- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
+==> Related Work:
 
-We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity.  We will do a better job of showing where the ideas come from.
+We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity. In reference to works cited by the reviewers:
 
-In reference to works cited by the reviewers:
 - Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
@@
 
-   We will expand the related work section, but stress the unique nature of CLM which results in clean concise coordination, e.g., we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).
-It is hard (or impossible) to write in all the above languages.
+We claim rather that CLM integrates coordination (scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
+- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
+- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
+- we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write in all the above languages.
 
-   Experiments:
+CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
+Because the paper was written from this PL perspective, we unfairly neglected important related work.
 
-   Mea Culpa on not explaining the superlinear speed-up.  We didn't implement the randomized scheduler needed for the best sequential implementation.  We will address this in the final version.
+==> Experiments: We explained the comparisons done against C++ and GraphLab and also the super linear speedup in the 1st response.
+Our system is competitive (2-3x) against GraphLab and performs reasonabily against C++ (2-10x).
+We did not include more datasets due to lack of space because we wanted to show a broad set of programs.
 
-   We did compare CLM to an optimized C++ implementation of 2 programs (HT and SBP): The CLM implementations are 1.8X-3X slower, but scale equally well.
-Our focus was on expressiveness and scalability, not raw performance.
-(CLM is currently interpreted by an unoptimized virtual machine.
-It could even be compiled.)  From an expressiveness point of view, the GraphLab scheduler is 500 lines of customized library code.
-CLM's version is less than 1 page!
-If requested, we can also include a third-party solution to SSSP.
-As well as including several different synthetic graphs to test the scalability over different data sets.
-
-We can include comparisons to sequential solutions:
-   - interpreted solutions (Python which is 2x-10x slower than CLM)
-   - compiled solutions in C (which is 2-10x faster depending on the program)
-
-Other Stuff (And addressing some of the smaller, but interesting points taken up by the reviewers):
-
-Please refer to 1st response for the other points.
-
--> reviewer D:
-
-- Small increment over Meld: we agree that this is true on the surface since the language is actually the same, but we do not agree that this is a point against since the language is able to describe normal computation and coordination using the same constructs.
-However, the runtime system actually looks quite different because now we have to deal with programmable scheduling and data partitioning.
-These two coordination mechanisms required significant changes in the implementation and compilation with the introduction of different queues for nodes and the ability to move and pin nodes to threads.
-- CLM target: our current target is shared memory multi-cores.
-However the CLM programming model fits and is easily ported to clusters of computers.
-It’s just a matter of implementing a suitable runtime system for those architectures.
-- Contribution of CLM: Galois and many others are not declarative programming languages.
-CLM solves a problem of programming language design by seamlessly combining computation and coordination into a single declarative programming model.
-Galois is a runtime system and does not solve this particular design problem.
-- Technical challenges: there is a PL design language: how to design a declarative language that allows the programmer to dynamically coordinate execution while keeping the language declarative; we show that linear logic allows us to do this easily.
-There is an implementation challenge where we had to add support for all the coordination directives.
-- Proofs: for most programs they do not complicate the proofs. However, for programs such as SBP, you have to take coordination into account because it changes how the program is computed.
-
-Reviewer E:
-
-- Programming efficiency: true, but we see this as a positive point since most declarative languages do not allow such control.
-- Data layout: the data layout/node placement is changed by using action facts such as set-cpu or set-static. Set-cpu moves the node to another thread while set-static pins the node to a thread.
-- inplace update: it can be done by analyzing how facts are used in the rules.
-If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.
-
--> reviewer B and E:
-
-- The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular:
-   - Regular/Regular: speedup of the unannotated (e.g.., regular) program on n-processors over the unannotated program on 1 processor.
-   - Coordinated/Regular: speedup of the coordinated program on       n-processors over the unannotated (e.g., regular program) on 1       processor.
-   - Coordinated/Coordinated: speedup of the coordinated program on       n-processors over the coordinated program on 1 processor.
-
-Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems.
-The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
-We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.
+// include other_stuff.txt

### 2015-01-20 rebuttal2.txt
update

@@
-From the review scores, it looks like the paper may be reject.
-We note however that coordination is an highly important programming language design issue, specially when concerning declarative programming languages and parallelism.
-We disagree that the paper is a small increment
+We note however that coordination is a highly important programming language design issue, specially when concerning declarative programming languages and parallelism.
+Our effort is not a small increment because it integrates dynamic coordination into a relatively general programming language, allowing the programmer to improve execution and partitioning if so desired.
+Our solution to this PL design challenge fits the scope of PLDI fairly well.
+We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.
 
@@
 While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue.
-Our solution to this PL design challenge fits the scope of PLDI fairly well.
-We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.
 
@@
 Talk about small increment?
+CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
+Because the paper was written from this PL perspective, we unfairly neglected important related work.
 
@@
 
-We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity. In reference to works cited by the reviewers:
-
+We do not claim novelty in the mechanisms per se, but rather in how they are integrated into a declarative language - as a first class entity.
+In reference to works cited by the reviewers:
 - Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
@@
 
-We claim rather that CLM integrates coordination (scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
+CLM integrates coordination (both scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
 - supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
 - is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
-- we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write in all the above languages.
-
-CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
-Because the paper was written from this PL perspective, we unfairly neglected important related work.
+- being able to specify complex coordination programs such as SBP in less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in all the above languages.
 

### 2015-01-20 rebuttal2.txt
update.

--- 2015-01-20 rebuttal2.txt
@@
-Coordination is a programming language design issue that requires novel approaches based on sound formal foundations.
-The implementation challenges are important but not as important from a PL perspective.
-While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue.
-Our solution to this PL design challenge fits the scope of PLDI fairly well.
-We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.
+Coordination is a highly important programming language design issue, specially
+when concerning declarative programming languages and parallelism. Our effort
+is not a small increment because it makes dynamic coordination a first class
+entity of a language and semantically equivalent to computation, allowing
+the programmer to improve execution if so desired. This is a novel design
+that fits well under the scope of PLDI. Unfortunately there is some mismatch between
+this important contribution and the reviewer’s perspective on the material.
 
-CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
-Because the paper was written from this PL perspective, we unfairly neglected important related work.
+We wrote the paper from a programming language perspective and thus some
+systems were unfarily neglected from the related work.
+However, those systems are not programming languages on their own right and thus
+do not solve the programming language design issue. While we think they
+solve many of the implementation issues, the programming language
+aspect remains unsolved. CLM is a attempt to solve this issue.
 
-Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM.
-This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program.
-Each reviewer asked for more (but different) information which would require us to remove something in the paper.
-We will happily take advice from the committee about what we should remove to make room for something else.
-
-Related Work:
-
-The reviews pointed out numerous systems that we did not cite.  Please note that CLM stands alone in
-- making coordination (both scheduling and partitioning) a first-class programming construct.
-- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
-- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
-
-We do not claim novelty in the mechanisms, but rather in how they are integrated into a declarative language - as a first class entity.  We will do a better job of showing where the ideas come from.
-
-In reference to works cited by the reviewers:
-- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
-- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
-- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
+We compare the contributions of those systems:
+- Halide supports distinct coordination, but only as "pragmas" in the domain of
+image processing, i.e., regular data-structures.
+- Galois supports a rich set of schedulers through customizable containers
+(see: Nguyen&Pingali), but it requires extra-language mechanisms to do things
+such as determining the location of data.
+- Elixir's uses a pre-defined set of schedulers and limits coordination to
+   scheduling.
 - Grace supports scheduling through priorities. (It is very much like GraphLab)
 - Ligra, also oriented towards graph processing, only handles partitioning.
-   - Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.
+- Cilk, OpenMP and TBB all provide low-level imperative approaches for
+coordination.
 
-   We will expand the related work section, but stress the unique nature of CLM which results in clean concise coordination, e.g., we can write the Splash scheduler used for Splash BP less than 60 lines (versus 500 in GraphLab).
-It is hard (or impossible) to write in all the above languages.
+CLM differs by integrating both scheduling and partitioning into a
+declarative language as a first class entity, in particular CLM stands alone
+in:
+- supporting data-driven dynamic (at run-time) coordination, particularly for
+   irregular data structures.
+- actually being a declarative programming language and not a runtime system
+   (Galois, Elixir and others), solving both the implementation and design issues.
+- being able to specify complex coordination programs such as SBP in less than
+   60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in
+   all the above languages.
 
-   Experiments:
-
-   Mea Culpa on not explaining the superlinear speed-up.  We didn't implement the randomized scheduler needed for the best sequential implementation.  We will address this in the final version.
-
-   We did compare CLM to an optimized C++ implementation of 2 programs (HT and SBP): The CLM implementations are 1.8X-3X slower, but scale equally well.
-Our focus was on expressiveness and scalability, not raw performance.
-(CLM is currently interpreted by an unoptimized virtual machine.
-It could even be compiled.)  From an expressiveness point of view, the GraphLab scheduler is 500 lines of customized library code.
-CLM's version is less than 1 page!
-If requested, we can also include a third-party solution to SSSP.
-As well as including several different synthetic graphs to test the scalability over different data sets.
-
-We can include comparisons to sequential solutions:
-   - interpreted solutions (Python which is 2x-10x slower than CLM)
-   - compiled solutions in C (which is 2-10x faster depending on the program)
+To show the expressive power of making coordination a first class programming
+construct we included five examples to give readers a flavor of the power of
+CLM. This approach highlights the conciseness and expressiveness of CLM as well
+as the readability of a high-level declarative approach to fine-tune the
+performance of a program. Each reviewer asked for more (but different)
+information which would require us to remove something in the paper.
+In relation to the experiments, we explained the comparisons done against C++
+and GraphLab and also the super linear speedup in the 1st response. Our system
+is competitive (2-3x) against GraphLab and performs reasonabily against C++
+(2-10x).
 
@@
 
-Please refer to 1st response for the other points.
+-> reviewer A, B and C: please refer to 1st response.
 
@@
 If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.
-
--> reviewer B and E:
-
 - The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular:
@@
 
-Finally, we note that CLM is noteworthy since it is not restricted to shared memory systems and can be easily extended to distributed systems.
-The use of linear logic, a sound logical framework, has allowed us to easily prove properties about programs using coordination.
-We thank the reviewers for the time and expertise they have invested in these reviews.  We will of course avail ourselves of all the comments on presentation, grammar, etc.  Again, many thanks.
+We thank the reviewers for the time and expertise they have invested in these reviews.
+We will happily take advice from the committee how we should improve our paper.

### 2015-01-20 rebuttal2.txt
Update.

--- 2015-01-20 rebuttal2.txt
@@
-We note however that coordination is a highly important programming language design issue, specially when concerning declarative programming languages and parallelism.
-Our effort is not a small increment because it integrates dynamic coordination into a relatively general programming language, allowing the programmer to improve execution and partitioning if so desired.
-Our solution to this PL design challenge fits the scope of PLDI fairly well.
-We think there is some mismatch between this important contribution and the reviewer’s perspective on the material.
+Coordination is a highly important programming language design issue, specially
+when concerning declarative programming languages and parallelism. Our effort
+is not a small increment because it makes dynamic coordination a first class
+entity of a language and semantically equivalent to computation, allowing
+the programmer to write arbitrary coordination algorithms. This is a novel design
+that fits well under the scope of PLDI. Unfortunately there has been some mismatch between
+this important contribution and the reviewer's perspective on the material.
 
-The implementation challenges are important but not as important from a PL perspective.
-While systems such as Galois and Elixir solve of those challenges, they are not programming languages on their own right and thus do not solve the programming language design issue.
+We wrote the paper from a programming language perspective and thus some
+systems were unfarily neglected from the related work.
+However, those systems are not programming languages on their own right and thus
+do not solve the programming language design issue. While we think they
+solve many important implementation issues, the programming language
+aspect remains unsolved. CLM is an elegant approach to this challenge.
 
-Another purpose of the paper is to show the expressive power of making coordination a first class programming construct and thus we chose to include five examples to give readers a flavor of the power of CLM.
-This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of a program.
-Each reviewer asked for more (but different) information which would require us to remove something in the paper.
-
-Talk about small increment?
-CLM makes scheduling and partitioning an integral part of a declarative programming language, indistinguishable and semantically equivalent to regular computation, allowing the programmer to write arbitrarily complex scheduling policies that are dynamic.
-Because the paper was written from this PL perspective, we unfairly neglected important related work.
-
-==> Related Work:
-
-We do not claim novelty in the mechanisms per se, but rather in how they are integrated into a declarative language - as a first class entity.
-In reference to works cited by the reviewers:
-- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
-- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
-- Elixir's uses a pre-defined set of schedulers and limits coordination to scheduling.
+We compare the contributions of those other systems:
+- Halide supports distinct coordination, but only as "pragmas" in the domain of
+image processing, i.e., regular data-structures.
+- Galois supports a rich set of schedulers through customizable containers
+(see: Nguyen&Pingali), but it requires extra-language mechanisms to do things
+such as determining the location of data.
+- Elixir's uses a pre-defined set of schedulers and limits coordination to
+   scheduling.
 - Grace supports scheduling through priorities. (It is very much like GraphLab)
 - Ligra, also oriented towards graph processing, only handles partitioning.
-- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.
+- Cilk, OpenMP and TBB all provide low-level imperative approaches for
+coordination.
 
-CLM integrates coordination (both scheduling and partitioning) into a declarative language as a first class entity, in particular CLM stands alone in:
-- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
-- is actually a declarative programming language and not a runtime system (Galois, Elixir and others), solving both the implementation and design issues.
-- being able to specify complex coordination programs such as SBP in less than 60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in all the above languages.
+CLM distinguishes itself by integrating both scheduling and partitioning into a
+declarative programming language as a first class entity. In particular CLM
+stands alone in:
+- supporting data-driven dynamic (at run-time) coordination, particularly for
+   irregular data structures.
+- actually being a declarative programming language and not a runtime system,
+   solving both the implementation and design issues.
+- being able to specify complex coordination programs such as SBP in less than
+   60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in
+   all the above languages.
 
-==> Experiments: We explained the comparisons done against C++ and GraphLab and also the super linear speedup in the 1st response.
-Our system is competitive (2-3x) against GraphLab and performs reasonabily against C++ (2-10x).
-We did not include more datasets due to lack of space because we wanted to show a broad set of programs.
+Each reviewer asked for more (but different) information which would require us
+to remove something in the paper.  We could have included more experiments and
+data sets but instead we decided to include five programs that give readers a
+flavor of the power of CLM. This approach highlights the conciseness and
+expressiveness of CLM as well as the readability of a high-level declarative
+approach to fine-tune the performance of programs. We remind our reviewers that
+our system is competitive (2-3x) against GraphLab and performs reasonabily well
+against C++ (2-10x slower).
 
-// include other_stuff.txt
+% include other_stuff.txt

### 2015-01-20 rebuttal2.txt
Update.

@@
 is not a small increment because it makes dynamic coordination a first class
-entity of a language and semantically equivalent to computation, allowing
-the programmer to write arbitrary coordination algorithms. This is a novel design
-that fits well under the scope of PLDI. Unfortunately there has been some mismatch between
-this important contribution and the reviewer's perspective on the material.
+entity of a declarative language and semantically equivalent to computation.
+Like the reviewer B said, this orthogonal design for writing coordinated
+algorithms will be of interest to many in the PLDI community.  Unfortunately,
+there has been some misunderstanding between this important
+contribution and the other reviewer's perspective on the material.
 

### 2015-01-20 rebuttal2.txt
Update rebuttal.

@@
-Coordination is a highly important programming language design issue, specially
-when concerning declarative programming languages and parallelism. Our effort
-is not a small increment because it makes dynamic coordination a first class
-entity of a declarative language and semantically equivalent to computation.
-Like the reviewer B said, this orthogonal design for writing coordinated
-algorithms will be of interest to many in the PLDI community.  Unfortunately,
-there has been some misunderstanding between this important
-contribution and the other reviewer's perspective on the material.
+Coordination is a highly important programming language design issue,
+specially when concerning declarative programming languages and
+parallelism. Our contribution makes dynamic coordination a first class
+entity of a declarative language and semantically equivalent to
+computation. Like reviewer B mentioned, this orthogonal design for
+writing coordinated algorithms will be of interest to many in the PLDI
+community. Unfortunately, there has been some misunderstanding between
+this important contribution and the other reviewer's perspective on
+the material.
 
-We wrote the paper from a programming language perspective and thus some
-systems were unfarily neglected from the related work.
-However, those systems are not programming languages on their own right and thus
-do not solve the programming language design issue. While we think they
+Each reviewer asked for more (but different) information which would
+require us to remove something in the paper. We could have included
+more experiments and data sets but instead we decided to include five
+programs that give readers a flavor of the power of CLM. This approach
+highlights the conciseness and expressiveness of CLM as well as the
+readability of a high-level declarative approach to fine-tune the
+performance of programs. Our results show that CLM is competitive
+(2-3x) against GraphLab and performs reasonabily well against C++
+(2-10x slower).
+
+We wrote the paper from a programming language perspective and thus
+some systems were unfarily neglected from the related work. However,
+those systems are not programming languages on their own right and
+thus do not solve the programming language design issue. While they
 solve many important implementation issues, the programming language
@@
 
-We compare the contributions of those other systems:
-- Halide supports distinct coordination, but only as "pragmas" in the domain of
-image processing, i.e., regular data-structures.
-- Galois supports a rich set of schedulers through customizable containers
-(see: Nguyen&Pingali), but it requires extra-language mechanisms to do things
-such as determining the location of data.
-- Elixir's uses a pre-defined set of schedulers and limits coordination to
-   scheduling.
-- Grace supports scheduling through priorities. (It is very much like GraphLab)
-- Ligra, also oriented towards graph processing, only handles partitioning.
+We briefly compare the contributions of those other systems:
+- Halide supports distinct coordination, but only as "pragmas" in the
+  domain of image processing, i.e., regular data-structures.
+- Galois supports a rich set of schedulers through customizable
+  containers (see: Nguyen&Pingali), but it requires extra-language
+  mechanisms to do things such as determining the location of data.
+- Elixir's uses a pre-defined set of schedulers and limits
+  coordination to scheduling.
+- Grace supports scheduling through priorities (it is very much like
+  GraphLab).
+- Ligra, also oriented towards graph processing, only handles
+  partitioning.
 - Cilk, OpenMP and TBB all provide low-level imperative approaches for
@@
 
-CLM distinguishes itself by integrating both scheduling and partitioning into a
-declarative programming language as a first class entity. In particular CLM
-stands alone in:
-- supporting data-driven dynamic (at run-time) coordination, particularly for
-   irregular data structures.
-- actually being a declarative programming language and not a runtime system,
-   solving both the implementation and design issues.
-- being able to specify complex coordination programs such as SBP in less than
-   60 lines (versus 500 in GraphLab).  It is hard (or impossible) to write it in
-   all the above languages.
-
-Each reviewer asked for more (but different) information which would require us
-to remove something in the paper.  We could have included more experiments and
-data sets but instead we decided to include five programs that give readers a
-flavor of the power of CLM. This approach highlights the conciseness and
-expressiveness of CLM as well as the readability of a high-level declarative
-approach to fine-tune the performance of programs. We remind our reviewers that
-our system is competitive (2-3x) against GraphLab and performs reasonabily well
-against C++ (2-10x slower).
+CLM distinguishes itself by integrating both scheduling and
+partitioning into a declarative programming language as a first class
+entity. In particular CLM stands alone in:
+- supporting data-driven dynamic (at run-time) coordination,
+  particularly for irregular data structures.
+- actually being a declarative programming language and not a runtime
+  system, solving both the implementation and design issues.
+- being able to specify complex coordination programs such as SBP in
+  less than 60 lines (versus 500 in GraphLab). It is hard (or
+  impossible) to write it in all the above languages.
 

### 2015-01-21 rebuttal2.txt
edits

@@
-Coordination is a highly important programming language design issue,
-specially when concerning declarative programming languages and
-parallelism. Our contribution makes dynamic coordination a first class
-entity of a declarative language and semantically equivalent to
-computation. Like reviewer B mentioned, this orthogonal design for
-writing coordinated algorithms will be of interest to many in the PLDI
-community. Unfortunately, there has been some misunderstanding between
-this important contribution and the other reviewer's perspective on
-the material.
+Coordination is an important programming language design issue, especially concerning declarative languages and parallelism.
+Our contribution makes dynamic coordination a first class entity of a declarative language and semantically equivalent to computation.
+Like reviewer B mentioned, this orthogonal design for writing coordinated algorithms will be of interest to many in the PLDI community.
+Unfortunately, there has been some misunderstanding between this important contribution and the other reviewer's perspective on the material.
 
-Each reviewer asked for more (but different) information which would
-require us to remove something in the paper. We could have included
-more experiments and data sets but instead we decided to include five
-programs that give readers a flavor of the power of CLM. This approach
-highlights the conciseness and expressiveness of CLM as well as the
-readability of a high-level declarative approach to fine-tune the
-performance of programs. Our results show that CLM is competitive
-(2-3x) against GraphLab and performs reasonabily well against C++
-(2-10x slower).
+Each reviewer asked for more (but different) information which would require us to remove something in the paper.
+We could have included more experiments and data sets but instead we decided to include five programs that give readers a flavor of the power of CLM.
+This approach highlights the conciseness and expressiveness of CLM as well as the readability of a high-level declarative approach to fine-tune the performance of programs.
+Our results show that CLM is competitive (2-3x) against GraphLab and performs reasonably well against C++ (2-10x slower).
+[See our first rebuttal for more detail]
 
-We wrote the paper from a programming language perspective and thus
-some systems were unfarily neglected from the related work. However,
-those systems are not programming languages on their own right and
-thus do not solve the programming language design issue. While they
-solve many important implementation issues, the programming language
-aspect remains unsolved. CLM is an elegant approach to this challenge.
+We wrote the paper from a language perspective and thus some systems were unfairly neglected.
+We will include them in the final paper.
+While those systems solve many important implementation issues, the programming language aspect remains unsolved.
+In Brief (See first response for more detail):
+- Halide supports distinct coordination, but only as "pragmas" in the domain of image processing, i.e., regular data-structures.
+- Galois supports a rich set of schedulers through customizable containers (see: Nguyen&Pingali), but it requires extra-language mechanisms to do things such as determining the location of data.
+- Elixir's uses a predefined set of schedulers and limits coordination to scheduling.
+- Grace supports scheduling through priorities (it is very much like GraphLab).
+- Ligra, also oriented towards graph processing, only handles partitioning.
+- Cilk, OpenMP and TBB all provide low-level imperative approaches for coordination.
 
-We briefly compare the contributions of those other systems:
-- Halide supports distinct coordination, but only as "pragmas" in the
-  domain of image processing, i.e., regular data-structures.
-- Galois supports a rich set of schedulers through customizable
-  containers (see: Nguyen&Pingali), but it requires extra-language
-  mechanisms to do things such as determining the location of data.
-- Elixir's uses a pre-defined set of schedulers and limits
-  coordination to scheduling.
-- Grace supports scheduling through priorities (it is very much like
-  GraphLab).
-- Ligra, also oriented towards graph processing, only handles
-  partitioning.
-- Cilk, OpenMP and TBB all provide low-level imperative approaches for
-  coordination.
+CLM distinguishes itself by integrating both scheduling and partitioning into a declarative language as a first class entity. In particular CLM stands alone in:
+- supporting data-driven dynamic (at run-time) coordination, particularly for irregular data structures.
+- actually being a declarative programming language and not a runtime system, solving both the implementation and design issues.
+- being able to specify complex coordination programs such as SBP in less than 60 lines (versus 500 in GraphLab). It is hard (or impossible) to write it in all the above languages.
 
-CLM distinguishes itself by integrating both scheduling and
-partitioning into a declarative programming language as a first class
-entity. In particular CLM stands alone in:
-- supporting data-driven dynamic (at run-time) coordination,
-  particularly for irregular data structures.
-- actually being a declarative programming language and not a runtime
-  system, solving both the implementation and design issues.
-- being able to specify complex coordination programs such as SBP in
-  less than 60 lines (versus 500 in GraphLab). It is hard (or
-  impossible) to write it in all the above languages.
+REVIEWER D:
 
-% include other_stuff.txt
+- Small increment over Meld: It is the beauty of our approach that the integration of coordination appears to be a small increment.
+While it appears to be a small increment on the surface the ability to specify coordination in a first class manner is a significant contribution requiring major changes to both the runtime and compiler.
+
+- CLM target: our current target is shared memory multicores. However, the CLM programming model fits and is easily ported to clusters of computers.
+
+- Contribution of CLM: Galois and many others are not declarative languages.
+CLM solves the language design issue by seamlessly combining computation and coordination into a single declarative model.
+Galois is a runtime system and does not solve this particular design problem.
+
+- Technical challenges: The PL design issue is: how to design a declarative language that allows the programmer to dynamically   coordinate execution while keeping the language declarative; we show   that linear logic allows us to do this easily.
+The implementation   challenge is in supporting the coordination directives.
+
+- Proofs: for most programs they do not complicate the   proofs. However, for programs such as SBP, you have to take coordination into account because it changes how the program is   computed.
+
+REVIEWER E:
+
+- Programming efficiency: No more or less then any other language, but unlike other declarative languages the programmer does get control if she wants it.
+
+- Data layout: the data layout/node placement is changed by using action facts such as set-cpu or set-static. Set-cpu moves the node to another thread while set-static pins the node to a thread.
+
+- inplace update: it can be done by analyzing how facts are used in the rules.
+If one fact fires a rule that updates the argument of another fact, then we could compile code that would update directly the argument of such fact instead of performing standard rule derivation.
+
+- The difference between Regular/Regular, Coordinated/Coordinated, Coordinated/Regular: This is confusing and will be clearer in the final paper.
+
+   - Regular/Regular: speedup of the unannotated (e.g.., regular) program on n-processors over the unannotated program on 1 processor.
+
+   - Coordinated/Regular: speedup of the coordinated program on n-processors over the unannotated (e.g., regular program) on 1 processor.
+
+   - Coordinated/Coordinated: speedup of the coordinated program on n-processors over the coordinated program on 1 processor.
+
+REVIEWER A, B and C: please refer to 1st response.
+
+We thank the reviewers for the time and expertise they have invested
+in these reviews.  We will happily take advice from the committee on
+ways to improve our paper.
+
+LocalWords:  PLDI CLM GraphLab Halide pragmas customizable Pingali
+LocalWords:  pre Ligra Cilk OpenMP TBB runtime SBP multi cpu inplace
+LocalWords:  unannotated
